### 2018-06-18 latex/manuscript.tex
Turned manuscript into tex file

\RequirePackage[l2tabu, orthodox]{nag}
\documentclass[paper=a4, 12pt]{scrartcl}
\usepackage[utf8]{inputenc}
\usepackage[colorlinks=true,
	linkcolor=black,
	citecolor=black,
	filecolor=black,
	urlcolor=black]{hyperref}
\usepackage{textcomp}
\usepackage[left=1cm,right=5cm,top=2cm,bottom=2cm,nohead,nofoot]{geometry}
\usepackage{todonotes}

\setlength{\marginparwidth}{4cm}

\begin{document}

\noindent
{\huge\sffamily\bfseries NamPRT and NNMT – key drivers of NAD-dependent signalling \par}

\vspace{2cm}

\noindent
Mathias Bockwoldt$^{1}$, Marc Niere$^{2}$, Toni I. Gossmann$^{3}$, Mathias Ziegler$^{2}$ and Ines Heiland$^{1,\S}$

\vspace{1cm}

\noindent
$^{1}$Department of Arctic and Marine Biology, UiT Arctic University of Norway, Naturfagbygget, Dramsveien 201, 9037 Tromsø, Norway

\noindent
$^{2}$Department of Molecular Biology, University of Bergen, Thormøhlensgt. 55
5020 Bergen, Norway

\noindent
$^{3}$Department of Animal and Plant Sciences, Western Bank, University of Sheffield
Sheffield, S10 2TN, United Kingdom

\noindent
§ Corresponding author: ines.heiland@uit.no

\section{Summary}

\section{Keywords}

\section{Highlights}

\section{Introduction}

NAD metabolism represents one of the most critical links that connect cellular signal transduction and energy metabolism. According to Kyoto Encyclopedia of Genes and Genomes KEGG (\url{http://www.genome.jp/kegg/}), more than 20\% of all biochemical reactions use NAD(P), mostly in redox reactions in which NAD(P)+ and NAD(P)H are reversibly interconverted. In addition to redox reactions, various signalling processes consume NAD+ by cleaving the molecule to nicotinamide (Nam) and ADP-ribose \cite{Verdin2015}. NAD+-dependent signalling reactions include poly- and mono-ADP-ribosylation, NAD-dependent protein deacylation by sirtuins as well as the synthesis of calcium mobilizing molecules such as cyclic ADP-ribose \cite{Opitz2015}. These NAD+-dependent signalling processes participate in the regulation of virtually all cellular activities. The enzymes involved in these processes are sensitive to the available NAD+ concentration \cite{Ruggieri2015}, which in turn is dependent on the NAD+/NADH redox ratio. Therefore, NAD+-dependent signalling can act as a transmitter of changes in the cellular energy homeostasis, for example, to regulate gene expression or metabolic activity \cite{Koch-Nolte2009}.

The significance of NAD+-dependent signalling for NAD homeostasis has long been underestimated. However, it has now been established that inhibition of NAD biosynthesis in mammalian cells leads to a rapid decline in NAD concentration suggesting that NAD+-dependent signalling consumes substantial amounts of NAD (Ref?), that is why we later refer to them as NAD-consuming reactions. As it has been shown that NAD concentrations change in a circadian manner \cite{Nakahata2009; Ramsey2009}, the cellular NAD-pool is turned over at least once per day\todo[author=Mathias]{Either the NAD-pool is turned over at least once a day or the concentration is maintained, but both … could be possible but sounds strange.}. To maintain the NAD concentration at physiological levels, NAD biosynthesis needs to proceed at an equally rapid rate. Imbalances in NAD-homeostasis have been linked to various, in particular, ageing-related diseases such as diabetes, neurodegenerative disorders and cancer \cite{Chiarugi2012; Verdin2015}\todo[author=Ines]{NAD turnover}. Various recent studies have demonstrated impressive health benefits of dietary supplementation with intermediates of NAD biosynthesis including NMN and nicotinamide riboside, NR \cite{Belenky2007}\todo[author=Ines]{That is not very recent. Better suggestions?}. Apparently, the exploitation of NAD biosynthetic routes, in addition to the use of nicotinamide as precursor (Fig. 1), results in increased NAD concentrations that stimulate NAD+-dependent signalling processes, in particular, protein deacetylation by sirtuins. \todo[author=Ines]{Shorten introduction}

In mammals, NAD biosynthesis predominantly relies on nicotinamide \cite{Bogan2008}\todo[author=Ines]{Not the perfect reference maybe}, a form of vitamin B3, as precursor. In the first step, Nam is converted to the mononucleotide, NMN, by Nam phosphoribosyltransferase (NamPRT) using phosphoribosylpyrophosphate (PRPP) as co-substrate. The key role of this enzyme is related to the nature of NAD+-dependent signalling reactions, namely, that in all these reactions Nam is cleaved off and potentially lost. If the released Nam were not recycled into NAD biosynthesis, humans would require a much higher daily vitamin B3 intake than the 15 mg that are the current daily recommendation (ref) \todo[author=Ines]{I have not been able to find a reference for that so far.}. The nearly complete recycling of Nam is achieved by an extraordinary high affinity of NamPRT to Nam, the Km being in the low nanomolar range \cite{Burgos2008}. Despite the importance of its salvage, Nam can also be marked for excretion by methylation \todo[author=Mathias]{It’s not too surprising. If there was no way of removing parts, we would accumulate more and more, as we still take up 15 mg every day.}. Indeed, the presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012} is among the most enigmatic and counterintuitive features of NAD metabolism. Why is there one enzyme (NamPRT) seemingly optimized to recycle even the faintest amounts of Nam back into NAD synthesis, while at the same time there is another one (NNMT) that seems to have no metabolic function other than to remove Nam from NAD metabolism\todo[author=Ines]{This paragraphs to the start including a short statement about the purpose of the analysis.}? This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deaminated Nam to nicotinic acid (NA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (Fig. 1). Previous phylogenetic analysis show that the two Nam recycling pathways are both ancestral showing a scattered distribution in bacteria \cite{Gazzaniga2009}. In contrast, NNMT has so far only been found in animals \cite{Gossmann2012}. The analysis of eukaryotic pathway evolution was previously limited by the low number of available genome sequences. A comprehensive evolutionary analysis is, however, needed to fully understand the phylogeny of NAD metabolism, in particular, Nam metabolizing enzymes and their relationship to signalling pathways.

This analysis may also help to unravel, how biosynthesis and signalling are coupled and influence each other, as this is not well understood so far. It is widely assumed that a major driving force of NAD+-consuming signalling reactions is the cellular NAD concentration. Most enzymes using NAD in signalling reactions that have been characterized in detail and it was shown that most sirtuins and PARPs are subject to inhibition by nicotinamide \cite{Borra2004; Ko2012}. To promote NAD+-dependent signalling reactions, the balance between elevated NAD levels and accumulating nicotinamide concentrations needs to be kept such that the inhibitory effect of Nam does not override the kinetic stimulation by high NAD levels \todo[author=Mathias]{Well… To promote [signalling] reactions, Nam simply has to be turned into NAD as quickly as possible. Besides that it sounds, form the description, as a self-regulating process.}. Given the complex nature of the NAD metabolome, it is not trivial to predict the optimal conditions for efficient NAD dependent signalling. However, it is obvious that Nam-converting enzymes must play a key role in this regard.

In the present study, we have conducted a comprehensive phylogenetic analysis of the genes that encode Nam-converting enzymes. The most surprising result was that with the emergence of Deuterostomes, there has been a strong tendency towards the co-existence of NamPRT and NNMT. Moreover, this selection for the co-existence of NamPRT and NNMT was accompanied by a marked increase in the number of genes encoding NAD+-consuming signalling reactions. To explain this observation, we built a kinetic model of NAD metabolism based on available kinetic parameters. This mathematical model provided a powerful tool to analyse the relationship between NAD biosynthetic and signalling fluxes when different subsets of Nam-converting enzymes are present. The model demonstrated that NNMT has a critical role to maintain high NAD+-consuming signalling fluxes by preventing accumulation of inhibitory Nam. At lower Nam concentrations, the impact of NNMT is minimal, because of the very high affinity of NamPRT. Taken together, our analyses suggest that the co-existence of NamPRT and NNMT has been a prerequisite to enable the evolutionary development of versatile NAD+-dependent signalling mechanisms present in vertebrates.

\section{Results}

\subsection{Phylogenetic analysis of NAD biosynthesis and consumption}

As shown in Fig. 1, NAD can be synthesized using several routes from three main precursors: tryptophan, nicotinamide (Nam) and nicotinic acid (NA). Nam and NA are together known as vitamin B3 or niacin. Alternatively, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead \cite{Bogan2008}. As in humans only 1\% of the tryptophan taken up with our diet is converted into NAD, vitamin B3 and to a lower extend NR are essential components of our diet, with Nam being the major NAD precursor in humans. Nam is also the product of NAD consuming signalling reactions such as sirtuins (NAD-dependent histone deacetylases) or PARPs (poly ADP ribosylases).

For the recycling of Nam, two different pathways exist. The pathway found in yeast and plants is using a four-step pathway starting with the deamination of Nam to nicotinic acid by Nam deamidase (NADA). The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates. The recycling pathway found in mammals directly converts Nam into the corresponding mononucleotide (NMN) a reaction catalysed by NamPRT and driven by a non-stoichiometric ATP-hydrolysis. A similar reaction catalysed by an evolutionary related enzyme NAPRT, converts NA into the NA mononucleotide in the Preiss-Handler pathway. NMN and NAMN are converted into dinucleotides by the Nam/NA adenylytransferases (NMNATs). The recycling pathway via NA finally requires an amination step catalysed by NADsynthase, driven by the conversion of ATP to AMP producing pyrophosphate. Even though the latter pathway seems to be very inefficient, it is the pathway preferentially used by most bacteria, fungi and plants (see Figure 2A), whereas most metazoans recycle Nam using NamPRT.

Analysing the phylogeny of the NAD recycling enzymes in Metazoa in more detail reveals that not only does NamPRT replace NADA, but in most organisms, especially in Deuterostomia, NamPRT is found together with the Nam methyltransferase NNMT (Figure 2B). NNMT methylates Nam to methyl-Nam that is in mammals excreted with the urine, thus removing Nam from recycling. NNMT seems to have arisen de novo in the common ancestor of Ecdysozoa and Lophotrochozoa, as we could not find any gene with considerable similarity in fungi or plants. Nematodes are the only organisms where we find NNMT together with NADA without NamPRT being present. In Deuterostomia the only large class that does only have NamPRT and seems to have lost NNMT again are Sauropsida and here especially birds. The reason why a lot of birds do not encode NNMT remains unclear, as the appearance is quite scattered (not shown)\todo[author=Mathias]{Maybe as supplementary figure?}. It might be related to the excretion system, as the product of NNMT methyl-nicotinamide is in mammals excreted with the urine. There are some species where we could not find NamPRT or NADA but NNMT, we assume that this is due to incomplete genomes in the database\todo[author=Ines]{Not sure we should discuss Branchisotoma and Tunicata here}.

In addition to the phylogenetic distribution of the two Nam salvage enzymes NADA and NamPRT, we looked at phylogenetic diversity of enzymes catalysing NAD-dependent signalling reactions. To do so we used the previously established classification into 10 different families \cite{Gossmann2012} (For details see materials and methods and supplementary information). The numbers in Figure 2B denote the average number of NAD-dependent signalling enzyme families we found in each taxa. With the exception of Cnidaria and Lophotrochozoa, we find in Protostomia on average 3 to 4 families, whereas most Deuterostomia have on average more than 8 families with an increasing diversification of enzymes within at least some of these families \cite{Gossmann2014}.

Taken together, we found that the presence of NamPRT and NNMT coincides with an increased diversification of NAD-dependent signalling. That is surprising as intuitively, one would expect that a decrease in precursor concentration caused by the precursor removal through NNMT, should cause a decrease of NAD availability and consequently less active NAD dependent signalling.

\subsection{Dynamics of NAD biosynthesis and consumption}

So why does the diversity of NAD-dependent signalling increase? And why does NADA disappear in Deuterostomia although it is the predominant pathway in bacteria, plants and fungi? Given the complexity of the NAD metabolic network, this question is difficult to be comprehensively addressed experimentally. Thus, to answer these questions we built a dynamic model of NAD metabolism using existing kinetic data from the literature (details see materials and methods and supplementary material).

To be able to compare metabolic features of evolutionary quite different systems in our simulations, and as we have limited information about expression levels of enzymes or changes of kinetic constants during evolution, we initially used the kinetic constants found for yeast or human enzymes for all systems analysed and used equal amounts of enzymes for all reactions. Wherever possible we did not only include substrate affinities but also known product inhibition or inhibition by downstream metabolites. As we assume that cell growth is, besides NAD-consuming reactions, a major driving force for NAD biosynthesis, we analysed different growth rates (cell division rates) by simulating different dilution rates for all metabolites. We assume furthermore that Nam availability is different for different organisms and thus in addition analysed this simulating different Nam import rates.

The pathway using NADA and recycling Nam via NA is superior \todo[author=Mathias]{What is meant with superior? NAD consumption flux is higher or lower with NNMT, depending on the condition. Free NAD concentration is higher without NNMT, but is that superior?} both in terms of steady state NAD concentration and NAD consumption flux (representing the activity of NAD-dependent signalling) in the absence of NNMT (Figure 3). In the presence of NNMT, the picture is slightly different (Figure 3C-D), even though NADA is still superior to NamPRT under most conditions \todo[author=Ines]{Check if not under all shown}\todo[author=Mathias]{Reply: From fig 4C, it looks like the flux is decreased at very high cell division rates while the Nam import rate is low.}. Only if Nam availability is very low, NamPRT is performing better because of its high substrate affinity. However, the disadvantage of NamPRT at equal amounts of enzyme, can be compensated by an increased expression of NamPRT. Although, to reach similar NAD concentrations in our model, NamPRT expression has to be tenfold higher than the expression of NAPRT (Figure S1). The NADA expression required is very low due to its high turnover. This might provide an explanation why we find NADA and with that the pathway via NA predominantly in bacteria, yeast, and plants, organisms that show high cell division rates. Under these conditions, protein expression costs are assumed to have a high impact on metabolic performance \todo[author=Mathias]{source} and thus pathways where low enzyme expression suffices, might be favourable, even though the pathway via NA is energetically less efficient.

NADA seems to be able to maintain higher NAD concentrations than NamPRT. This figure changes if we simulate two organisms or cells that are in direct competition for Nam. Under these conditions, the organism expressing NamPRT has a competitive advantage above NADA-expressing organisms, but this only holds if the competing organisms expresses NamPRT together with NNMT. (new Figure) This observation might explain, why NADA disappeared in Metazoa together with the appearance of NNMT. It might also explain, why many bacteria associated with mammals harbour NamPRT instead of NADA\todo[author=Mathias]{Needs some figure or source}.

Although these simulations already provide an idea of why we find NamPRT predominantly in combination with NNMT, we still do not know why this development coincides with an increased diversification of NAD-consuming enzymes. When we simulate the presence or absence of NNMT in the presence of NamPRT, we see that the impact of NNMT on NAD concentration is relatively small (Figure 4D), but we see that that NNMT increases the NAD consumption flux under most conditions in the presence of NamPRT (Figure 4C). NAD consumption flux can be further increased by increasing the expression NamPRT, which also compensates the slight reduction in NAD concentration in the presence of NNMT (see Figure S2).

These findings can be explained when looking in more detail into the kinetic parameters of NamPRT and NAD-consuming enzymes such as Sirt1p. The ability to maintain high a NAD concentration in the presence of NNMT and at low Nam availability, is due to the very high affinity of NamPRT for its substrate, having a half saturation constant (Km) in the low nanomolar range. The increase of NAD consumption flux is caused by the fact that most NAD-consuming enzymes are inhibited by their product Nam, which is also the reason why the presence of NNMT enables higher NAD consumption fluxes.

As the substrate affinity of NamPRT for Nam is extremely high (with a Km in the low nanomolar range) and as this might not have been the case throughout evolution, we analysed the effect of the NamPRT Km on NAD steady state concentration and NAD consumption flux, leaving all other kinetic parameters constant. In the absence of NNMT (Figure 5A-B) the Km has very little effect on steady state NAD concentration and NAD consumption flux. Without NNMT, NAD concentration and consumption flux are both considerably affected by cell division rates, at least if the enzyme expression is kept constant. This is of course an artificial scenario, as one would assume organisms to regulate enzyme expression to achieve similar levels of metabolite concentrations instead. In the absence of NNMT there appears furthermore to be a trade-off between achievable steady state NAD concentration and NAD consumption flux.

In the presence of NNMT, NAD consumption flux and NAD concentration increases with decreasing Km values (Figure 5C-D). And we note, that both the NAD steady state concentration and the consumption flux are relatively stable over a wide range of cell division rates even at constant expression levels of the involved enzymes, suggesting that NNMT might have an important role to maintain homeostasis of NAD metabolism.

When we compare NAD consumption and NAD concentration with and without NNMT with two different substrate affinities of NamPRT, we see that at a low affinity (Km of 100 nM, which is in the range of the Km of NAPRT for its substrate, or the Km of NADA for Nam), NAD consumption flux is only higher with NNMT at low cell division rates, whereas at high division rates, higher NAD consumption fluxes are achieved without NNMT (Figure 5 E-F). This might explain why we do not find NNMT in organism that tend to have high growth rates\todo[author=Ines]{I am not sure if we should go even further suggesting that decreasing substrate affinity by using a competitive inhibitor such as FK866 will influence the consumption rate in fast growing cells much more than the one in slow growing cells, being potentially relevant for cancer therapy. – Discussion}. We furthermore find, that the competitive advantage of NamPRT of NADA is only present at sufficiently high affinity of NamPRT.

Our analysis also suggests that NNMT might have exerted an evolutionary driving force on the substrate affinity of NamPRT, explaining the extreme values found for the human enzyme\todo[author=Ines]{Move to discussion?}.

The pathway dynamics are of course not solely dependent on one enzyme. Thus, what is the impact of the substrate affinity of NNMT that is competing with NamPRT for the same substrate? In Figure 6A we see that the substrate affinity values found in the human enzymes (indicated by black asterisks) are actually optimal with respect to both achievable steady state NAD concentration and consumption fluxes. Thus, a further increase of the affinity of NamPRT for Nam would not provide any advantage.

\subsection{Sequence variance acquired in metazoans enhances substrate affinity}

To see whether we can find sequence variations in the protein sequence of NamPRT that indicate evolutionary changes of NamPRT upon the appearance of NNMT, we created a multiple sequence alignment of selected eukaryotic NamPRT sequences. As shown in Figure 7A and Supplementary Figure 3\todo[author=Ines]{To be created}, Deuterostomia that have only NamPRT and NNMT (indicated by the number 6 in parenthesis \todo[author=Ines]{Maybe rather indicate this by colour and the kingdom through abbrev?}) have an insert of ten amino acids corresponding to positions 43 to 52 of the human enzyme. Looking at the crystal structure of human NamPRT this sequence insertion corresponds to a region that has not been structurally resolved in any of the currently available crystal structures (e.g. \cite{Wang2006} structure visualisation Figure 7B) and its function is unclear. The unresolved loop structure overlaps with a predicted weak nuclear localisation signal (figure 7A), that is not present without the insertion. The loop is in addition connected to one of the beta-sheets involved in substrate binding, potentially affecting the affinity or turnover of the enzyme.

The observed evolutionary change in the primary sequence of NamPRT could therefore have had different effects, that we wanted to test experimentally. We first investigated whether the deletion of the amino acids 43 to 52 has an effect on the localisation of the human enzyme, expressing a …. mutant lacking this sequence in HeLa cells. We could, however, not detect any changes in subcellular localisation (Figure 7C) compared to the wildtype fusion construct. We therefore conclude that the partial nuclear localisation of NamPRT is not affected by the deletion. We then tested the enzymatic activity of the NamPRT mutant recombinantly expressed in E. coli. As shown in Figure 7D the enzyme still forms a dimer, thus seems to be folded correctly. The enzymatic activity is, however, much lower compared to the wildtype enzyme. Using higher concentrations of Nam, it appears that the mutant is not saturated at 100 \textmu M, pointing to a decreased substrate affinity of the mutant enzyme.

\subsection{Comments and parts of old version not included yet}

Chordata: NADA disappears. These analyses also support the view that Tunicata and Branchiostoma are not part of the phylum chordata\todo[author=Mathias]{Is this view in any way argued upon by anyone??}.

In the evolutionary context, an additional question arises: Why do only a few organisms, mostly less complex animals, possess the gene for NADA in addition to NamPRT and NNMT. When including NADA into the simulations, we see that the effect of NADA on NAD consumption is very limited in the presence of NNMT (old Figure 4F) even at high expression levels (see old Suppl. Figure S1C and D) and could be compensated by increased expression of NamPRT (not shown).

Until this point we have neglected compartmentalisation of the pathway. We do however know from previous studies that in early vertebrate development a compartmentalisation of the pathway has occurred reflected by a gene triplication of NMNAT and the occurrence of compartment-specific domains called ISTIDs (Lau, 2010). Looking at the evolutionary timepoint \todo[author=Toni]{This time point can only be reconstructed, but is unknown. I think a figure with the likely reconstruction would help to follow the argumentation} of appearance of NNMT and the gene triplication of NMNATs, we see that NNMT occurs prior to the gene triplication and the first occurrence of ISTIDs and we have confirmed this in our own analysis (not shown).

\section{Discussion}

\subsection{Evolutionary and physiological role of NNMT}

The dynamic interactions among metabolism, signal transduction, and gene regulation is still not very well understood. New approaches are required to disentangle the underlying network, helping us to understand how alterations affect human physiology. We here combine phylogenetic analysis and mathematical modelling of a central metabolic pathway supported by experimental analysis to reveal the likely evolutionary development of the NAD biosynthesis and consumption pathway and the physiological role of the methyltransferase NNMT.

We show that the vitamin B3-degrading enzyme NNMT plays a vital role for the diversification of NAD-dependent signalling reactions and potentially also for NAD homeostasis. Due to the fact that NamPRT has a very high affinity for Nam whereas NNMT has a rather low affinity for it, NNMT is able to remove excess Nam that would lead to inhibition of NAD-dependent signalling reactions while maintaining NAD concentrations and even increasing NAD consumption fluxes.

The affinities measured for the human enzymes seem to be optimal and further increase in NAD consumption rates might only enabled by compartmentalisation of NAD biosynthesis.

\subsection{General applicability}

It is of course impossible to extend our conclusions to other metabolic processes without analysing them in more detail, but we have noted that the metabolism of other vitamins, such as pyridoxal (vitamin B6) metabolism, also contains vitamin degrading enzymes \todo[author=Mathias]{source}, that might thus have similar roles. Degrading enzymes do in general not receive the same attention as biosynthetic enzymes, reflected by the observation that NNMT has only recently been analysed in more detail, besides the much earlier recognition of NNMT as potential marker for some types of cancer (e.g. Ref?) and routine clinical measurements of urine methyl-Nam concentrations in the context of different diseases (e.g. Ref?). Our analysis show, however, that degrading enzymes do not solely modify substrates for better excretion, but can play a vital role both in human physiology as well as in the evolutionary development of biological processes.

\subsection{Limitations of our analysis and outlook}

Among others, we neglected the potential effects of the methyl donor S-adenosyl-methionin (SAM) and its precursor methionine in our analysis, although it is most likely contributing yet another regulatory level for Nam availability and thus an additional interaction point between gene regulation and metabolism. It has been shown earlier that in turn, NNMT expression influences protein methylation dependent on methionine availability \cite{Ulanovskaya2013}. One of the challenges in the analysis of this aspect is the availability of in vivo concentration measurements for SAM and the large amount of reactions using it as substrate. The same holds for the analysis of enzyme acetylation versus deacetylation by sirtuins that would enable prediction about enzyme activation or histone state if available.

\section{Experimental Procedure}

\subsection{Dynamic modelling}

Kinetic parameters (substrate affinity (Km) and turnover rates (kcat), substrate and product inhibitions) were retrieved from the enzyme database BRENDA and additionally evaluated by checking the original literature especially with respect to measurement conditions. Parameter values from mammals were used if available. For enzymes not present in mammals, values from yeast were used. The full list of kinetic parameters including reference to original literature can be found in Supplementary table 1. For NMNAT, the previously developed rate law for substrate competition was used \cite{Schauble2013}. Despite these modifications, Henri-Michaelis-Menten kinetics were used for all reactions except the import and efflux of Nam, which were simulated using constant flux and mass action kinetics, respectively. All simulations were performed using the steady state calculation and parameter scan options provided by COPASI 4.22 \cite{Hoops2006}. The model will be available at the Biomodels database accession number \todo[inline]{xxx}. Related figures were generated using gnuplot\todo[author=Mathias]{Version?}.

\subsection{Phylogenetic Analysis}

NADA, NamPRT, and NNMT enzymes or enzyme candidates were identified using Blastp with known enzymes against the NCBI non-redundant protein sequence database (nr). A list of functionally verified enzymes used as templates is given in supplementary table 2. This table also includes the length cut-offs for identified enzymes. The e-value cut-off was 1e-30 for all enzymes. Blastp parameters were set to yield maximum 20,000 target sequences, using the BLOSUM62 matrix with a word size of 6 and gap opening and extension costs of 11 and 1, respectively. Low-complexity filtering was disabled. Obvious sequence contaminations were removed by manual inspection of the results. The taxonomy IDs of the species for each enzyme was derived from the accession2taxonomy database provided by NCBI. Scripts for creating, analysing, and visualising the phylogenetic tree were written in Python, using the ETE3 toolkit (Huerta-Cepas, 2010).

\subsection{Generation of eukaryotic expression vectors encoding C-terminally FLAG-tagged NamPT proteins}

The open reading frame (ORF) encoding human NamPRT was inserted into pFLAG-CMV-5a (Merck - Sigma Aldrich) via EcoRI/BamHI sites. Using a PCR-based approach, this vector provided the basis for the generation of a plasmid encoding a NamPRT deletion mutant lacking amino acid residues 42-51 (NamPRT-$\Delta$42-51). The sequences of the inserted ORFs were verified by DNA sequence analysis.

\subsection{Transient transfection, immunocytochemstry and confocal laser scanning microscopy}

HeLa S3 cells cultivated in Ham’s F12 medium supplemented with 10\% (v/v) FCS, 2 mM L-glutamine, and penicillin/streptomycin, were seeded on cover slips in a 24 well plate. After one day, cells were transfected using Effectene transfection reagent (Qiagen) according to the manufacturer’s recommendations. Cells were fixed with 4\% paraformaldehyde in PBS 24 hours post transfection, permeabilised (0.5\% (v/v) Triton X-100 in PBS) and blocked for one hour with complete culture medium. After overnight incubation with primary FLAG-antibody (mouse M2, Sigma-Aldrich) diluted 1:2500 in complete medium, cells were washed and incubated for one hour with secondary AlexaFluor 594-conjugated goat anti mouse antibody (ThermoFisher, Invitrogen) diluted 1:1000 in complete culture medium. Nuclei were stained with DAPI and the cells washed. The cover slips were mounted on microscope slides using ProLong Gold (ThermoFisher, Invitrogen). Confocal laser scan imaging of cells was performed using a Leica TCS SP8 STED 3x microscope equipped with a 100x oil immersion objective (numerical aperture 1.4).

\section{Figure Legends}

\subsection{Figure1}

Schematic overview of NAD biosynthesis and consumption. NAD can be synthesized using several routes from altogether three main precursors: tryptophan (Trp), nicotinamide (Nam), and nicotinic acid (NA). Nam and NA are together known as vitamin B3 or niacin. To a lesser extend, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead. Only 1\% of the tryptophan taken up with our diet is converted into NAD, thus, vitamin B3 and, to a lower extend, NR are essential components of our diet, with Nam being the major NAD precursor in humans\todo[author=Mathias]{Whole sentence not necessarily part of the legend}. Nam is furthermore the product of NAD-consuming signalling reactions such as sirtuins (NAD-dependent histone deacetylases) or PARPs (poly-ADP-ribosylases).

For the recycling of Nam, two different pathways exist. The pathway found in yeast and many bacteria is using a four-step pathway starting with the deamination of Nam to nicotinic acid (NA) by Nam deamidase (NADA). The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates. The recycling pathway found in vertebrates directly converts Nam into the corresponding mononucleotide (NMN) a reaction catalysed by NamPRT and driven by a non-stoichiometric ATP-hydrolysis. A similar reaction catalysed by an evolutionary related enzyme NAPRT, converts NA into the NA mononucleotide (NAMN). NMN and NAMN are converted into dinucleotides by the Nam/NA adenylytransferases (NMNATs). The recycling pathway via NA finally requires an amination step catalysed by NADsynthase (NADS), driven by the conversion of ATP to AMP producing pyrophosphate.

\subsection{Figure 2}

Evolutionary distribution of NADA, NNMT and NamPRT and their relation to the number of NAD consumers. A) Distribution of NADA, NNMT and NamPRT in selected major taxa. NADA is dominant in Bacteria, Fungi, and Plants (Viridiplantae), whereas NamPRT together with NNMT is dominant in Deuterostomia. Numbers at the pie charts show, how many species of the taxon possess the respective enzyme combination indicated by the colour explained in the lower right of the figure. Below the taxon name, the number of species in that taxon is given.

B) Common tree of selected taxa within the Metazoa, including 334 species. The pie charts indicate the distribution of species within the respective taxon that have the enzyme combination indicated by the colour, explained in the lower right. The size of the pie charts is proportional to the logarithm of the number of species analysed in the particular taxon. The numbers below the taxon names indicate the average number of NAD-consuming enzyme families found in all sub-taxa. The branch length is arbitrary.

\subsection{Figure 3}

NAD consumption flux and NAD concentration in simulations of organisms with NADA(?) and with and without NNMT at different Nam import and cell division rates.

\subsection{Figure 4}

Impact of NNMT on NAD consumption and NAD concentration in simulations of organisms with NADA or NamPRT.

\subsection{Figure 5}

NAD consumption flux and NAD concentration in simulations of organisms with NamPRT and with and without NNMT at different KMs for NamPRT and cell division rates.

\subsection{Figure 6}

The substrate affinity of NNMT and NamPRT have opposite effects on NAD consumption (A) and concentration (B), as would be expected. The affinities previously measured for human enzymes (indicated by a black asterisk) appear to be close to optimal, as further improvements would have little or no effect on NAD consumption or concentration.

\subsection{Figure 7}

Deuterostomia that encode NNMT show a sequence insertion in the N-terminal region of NamPRT. A) Multiple sequence alignment of NamPRT of selected deuterostomes show a sequence insertion in organisms that encode NamPRT and NNMT. (Species names are given and the number in parenthesis indicates the enzyme combination encoded in these species; 3 – NADA and NamPRT, 6 – NamPRT and NNMT, 7 – NADA, NamPRT and NNMT. \todo[author=Mathias]{Note to myself: replace the numbers by something better and rewrite this part.}) The nuclear localisation signal (NLS\todo[author=Mathias]{Note to myself: Write “predicted NLS” in figure}) indicated, was predicted using using cNLS-mapper (\url{http://nls-mapper.iab.keio.ac.jp/cgi-bin/NLS_Mapper_form.cgi} \cite{Kosugi2009}).

B) The inserted region is not resolved in crystal structures of human NamPRT and thus appears to be a flexible loop structure at the surface of the NamPRT dimer, coloured in red. The visualisation is based on a structure prediction of SWISS-MODEL \cite{Arnold2006; Biasini2014} of the sequence of the human NamPRT (P43490) using the model 2H3D as template \cite{Wang2006}.

C) The localisation of wildtype NamPRT varies between pure cytosolic and mixed cytosolic-nuclear localisation. This is not changed in the mutant missing the insert that corresponds to the unresolved loop structure.

\end{document}

### 2018-06-20 latex/discussion.tex
splitted document in several files to speed up editing

--- 2018-06-18 latex/manuscript.tex
@@
-\RequirePackage[l2tabu, orthodox]{nag}
-\documentclass[paper=a4, 12pt]{scrartcl}
-\usepackage[utf8]{inputenc}
-\usepackage[colorlinks=true,
-	linkcolor=black,
-	citecolor=black,
-	filecolor=black,
-	urlcolor=black]{hyperref}
-\usepackage{textcomp}
-\usepackage[left=1cm,right=5cm,top=2cm,bottom=2cm,nohead,nofoot]{geometry}
-\usepackage{todonotes}
-
-\setlength{\marginparwidth}{4cm}
-
-\begin{document}
-
-\noindent
-{\huge\sffamily\bfseries NamPRT and NNMT – key drivers of NAD-dependent signalling \par}
-
-\vspace{2cm}
-
-\noindent
-Mathias Bockwoldt$^{1}$, Marc Niere$^{2}$, Toni I. Gossmann$^{3}$, Mathias Ziegler$^{2}$ and Ines Heiland$^{1,\S}$
-
-\vspace{1cm}
-
-\noindent
-$^{1}$Department of Arctic and Marine Biology, UiT Arctic University of Norway, Naturfagbygget, Dramsveien 201, 9037 Tromsø, Norway
-
-\noindent
-$^{2}$Department of Molecular Biology, University of Bergen, Thormøhlensgt. 55
-5020 Bergen, Norway
-
-\noindent
-$^{3}$Department of Animal and Plant Sciences, Western Bank, University of Sheffield
-Sheffield, S10 2TN, United Kingdom
-
-\noindent
-§ Corresponding author: ines.heiland@uit.no
-
-\section{Summary}
-
-\section{Keywords}
-
-\section{Highlights}
-
-\section{Introduction}
-
-NAD metabolism represents one of the most critical links that connect cellular signal transduction and energy metabolism.
-According to Kyoto Encyclopedia of Genes and Genomes KEGG (\url{http://www.genome.jp/kegg/}), more than 20\% of all biochemical reactions use NAD(P), mostly in redox reactions in which NAD(P)+ and NAD(P)H are reversibly interconverted.
-In addition to redox reactions, various signalling processes consume NAD+ by cleaving the molecule to nicotinamide (Nam) and ADP-ribose \cite{Verdin2015}.
-NAD+-dependent signalling reactions include poly- and mono-ADP-ribosylation, NAD-dependent protein deacylation by sirtuins as well as the synthesis of calcium mobilizing molecules such as cyclic ADP-ribose \cite{Opitz2015}.
-These NAD+-dependent signalling processes participate in the regulation of virtually all cellular activities.
-The enzymes involved in these processes are sensitive to the available NAD+ concentration \cite{Ruggieri2015}, which in turn is dependent on the NAD+/NADH redox ratio.
-Therefore, NAD+-dependent signalling can act as a transmitter of changes in the cellular energy homeostasis, for example, to regulate gene expression or metabolic activity \cite{Koch-Nolte2009}.
-
-The significance of NAD+-dependent signalling for NAD homeostasis has long been underestimated.
-However, it has now been established that inhibition of NAD biosynthesis in mammalian cells leads to a rapid decline in NAD concentration suggesting that NAD+-dependent signalling consumes substantial amounts of NAD (Ref?), that is why we later refer to them as NAD-consuming reactions.
-As it has been shown that NAD concentrations change in a circadian manner \cite{Nakahata2009; Ramsey2009}, the cellular NAD-pool is turned over at least once per day\todo[author=Mathias]{Either the NAD-pool is turned over at least once a day or the concentration is maintained, but both … could be possible but sounds strange.}.
-To maintain the NAD concentration at physiological levels, NAD biosynthesis needs to proceed at an equally rapid rate.
-Imbalances in NAD-homeostasis have been linked to various, in particular, ageing-related diseases such as diabetes, neurodegenerative disorders and cancer \cite{Chiarugi2012; Verdin2015}\todo[author=Ines]{NAD turnover}.
-Various recent studies have demonstrated impressive health benefits of dietary supplementation with intermediates of NAD biosynthesis including NMN and nicotinamide riboside, NR \cite{Belenky2007}\todo[author=Ines]{That is not very recent.
-Better suggestions?}.
-Apparently, the exploitation of NAD biosynthetic routes, in addition to the use of nicotinamide as precursor (Fig.
-1), results in increased NAD concentrations that stimulate NAD+-dependent signalling processes, in particular, protein deacetylation by sirtuins.
-\todo[author=Ines]{Shorten introduction}
-
-In mammals, NAD biosynthesis predominantly relies on nicotinamide \cite{Bogan2008}\todo[author=Ines]{Not the perfect reference maybe}, a form of vitamin B3, as precursor.
-In the first step, Nam is converted to the mononucleotide, NMN, by Nam phosphoribosyltransferase (NamPRT) using phosphoribosylpyrophosphate (PRPP) as co-substrate.
-The key role of this enzyme is related to the nature of NAD+-dependent signalling reactions, namely, that in all these reactions Nam is cleaved off and potentially lost.
-If the released Nam were not recycled into NAD biosynthesis, humans would require a much higher daily vitamin B3 intake than the 15 mg that are the current daily recommendation (ref) \todo[author=Ines]{I have not been able to find a reference for that so far.}.
-The nearly complete recycling of Nam is achieved by an extraordinary high affinity of NamPRT to Nam, the Km being in the low nanomolar range \cite{Burgos2008}.
-Despite the importance of its salvage, Nam can also be marked for excretion by methylation \todo[author=Mathias]{It’s not too surprising.
-If there was no way of removing parts, we would accumulate more and more, as we still take up 15 mg every day.}.
-Indeed, the presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012} is among the most enigmatic and counterintuitive features of NAD metabolism.
-Why is there one enzyme (NamPRT) seemingly optimized to recycle even the faintest amounts of Nam back into NAD synthesis, while at the same time there is another one (NNMT) that seems to have no metabolic function other than to remove Nam from NAD metabolism\todo[author=Ines]{This paragraphs to the start including a short statement about the purpose of the analysis.}?
-This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deaminated Nam to nicotinic acid (NA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (Fig.
-1).
-Previous phylogenetic analysis show that the two Nam recycling pathways are both ancestral showing a scattered distribution in bacteria \cite{Gazzaniga2009}.
-In contrast, NNMT has so far only been found in animals \cite{Gossmann2012}.
-The analysis of eukaryotic pathway evolution was previously limited by the low number of available genome sequences.
-A comprehensive evolutionary analysis is, however, needed to fully understand the phylogeny of NAD metabolism, in particular, Nam metabolizing enzymes and their relationship to signalling pathways.
-
-This analysis may also help to unravel, how biosynthesis and signalling are coupled and influence each other, as this is not well understood so far.
-It is widely assumed that a major driving force of NAD+-consuming signalling reactions is the cellular NAD concentration.
-Most enzymes using NAD in signalling reactions that have been characterized in detail and it was shown that most sirtuins and PARPs are subject to inhibition by nicotinamide \cite{Borra2004; Ko2012}.
-To promote NAD+-dependent signalling reactions, the balance between elevated NAD levels and accumulating nicotinamide concentrations needs to be kept such that the inhibitory effect of Nam does not override the kinetic stimulation by high NAD levels \todo[author=Mathias]{Well… To promote [signalling] reactions, Nam simply has to be turned into NAD as quickly as possible.
-Besides that it sounds, form the description, as a self-regulating process.}.
-Given the complex nature of the NAD metabolome, it is not trivial to predict the optimal conditions for efficient NAD dependent signalling.
-However, it is obvious that Nam-converting enzymes must play a key role in this regard.
-
-In the present study, we have conducted a comprehensive phylogenetic analysis of the genes that encode Nam-converting enzymes.
-The most surprising result was that with the emergence of Deuterostomes, there has been a strong tendency towards the co-existence of NamPRT and NNMT.
-Moreover, this selection for the co-existence of NamPRT and NNMT was accompanied by a marked increase in the number of genes encoding NAD+-consuming signalling reactions.
-To explain this observation, we built a kinetic model of NAD metabolism based on available kinetic parameters.
-This mathematical model provided a powerful tool to analyse the relationship between NAD biosynthetic and signalling fluxes when different subsets of Nam-converting enzymes are present.
-The model demonstrated that NNMT has a critical role to maintain high NAD+-consuming signalling fluxes by preventing accumulation of inhibitory Nam.
-At lower Nam concentrations, the impact of NNMT is minimal, because of the very high affinity of NamPRT.
-Taken together, our analyses suggest that the co-existence of NamPRT and NNMT has been a prerequisite to enable the evolutionary development of versatile NAD+-dependent signalling mechanisms present in vertebrates.
-
-\section{Results}
-
-\subsection{Phylogenetic analysis of NAD biosynthesis and consumption}
-
-As shown in Fig.
-1, NAD can be synthesized using several routes from three main precursors: tryptophan, nicotinamide (Nam) and nicotinic acid (NA).
-Nam and NA are together known as vitamin B3 or niacin.
-Alternatively, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead \cite{Bogan2008}.
-As in humans only 1\% of the tryptophan taken up with our diet is converted into NAD, vitamin B3 and to a lower extend NR are essential components of our diet, with Nam being the major NAD precursor in humans.
-Nam is also the product of NAD consuming signalling reactions such as sirtuins (NAD-dependent histone deacetylases) or PARPs (poly ADP ribosylases).
-
-For the recycling of Nam, two different pathways exist.
-The pathway found in yeast and plants is using a four-step pathway starting with the deamination of Nam to nicotinic acid by Nam deamidase (NADA).
-The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates.
-The recycling pathway found in mammals directly converts Nam into the corresponding mononucleotide (NMN) a reaction catalysed by NamPRT and driven by a non-stoichiometric ATP-hydrolysis.
-A similar reaction catalysed by an evolutionary related enzyme NAPRT, converts NA into the NA mononucleotide in the Preiss-Handler pathway.
-NMN and NAMN are converted into dinucleotides by the Nam/NA adenylytransferases (NMNATs).
-The recycling pathway via NA finally requires an amination step catalysed by NADsynthase, driven by the conversion of ATP to AMP producing pyrophosphate.
-Even though the latter pathway seems to be very inefficient, it is the pathway preferentially used by most bacteria, fungi and plants (see Figure 2A), whereas most metazoans recycle Nam using NamPRT.
-
-Analysing the phylogeny of the NAD recycling enzymes in Metazoa in more detail reveals that not only does NamPRT replace NADA, but in most organisms, especially in Deuterostomia, NamPRT is found together with the Nam methyltransferase NNMT (Figure 2B).
-NNMT methylates Nam to methyl-Nam that is in mammals excreted with the urine, thus removing Nam from recycling.
-NNMT seems to have arisen de novo in the common ancestor of Ecdysozoa and Lophotrochozoa, as we could not find any gene with considerable similarity in fungi or plants.
-Nematodes are the only organisms where we find NNMT together with NADA without NamPRT being present.
-In Deuterostomia the only large class that does only have NamPRT and seems to have lost NNMT again are Sauropsida and here especially birds.
-The reason why a lot of birds do not encode NNMT remains unclear, as the appearance is quite scattered (not shown)\todo[author=Mathias]{Maybe as supplementary figure?}.
-It might be related to the excretion system, as the product of NNMT methyl-nicotinamide is in mammals excreted with the urine.
-There are some species where we could not find NamPRT or NADA but NNMT, we assume that this is due to incomplete genomes in the database\todo[author=Ines]{Not sure we should discuss Branchisotoma and Tunicata here}.
-
-In addition to the phylogenetic distribution of the two Nam salvage enzymes NADA and NamPRT, we looked at phylogenetic diversity of enzymes catalysing NAD-dependent signalling reactions.
-To do so we used the previously established classification into 10 different families \cite{Gossmann2012} (For details see materials and methods and supplementary information).
-The numbers in Figure 2B denote the average number of NAD-dependent signalling enzyme families we found in each taxa.
-With the exception of Cnidaria and Lophotrochozoa, we find in Protostomia on average 3 to 4 families, whereas most Deuterostomia have on average more than 8 families with an increasing diversification of enzymes within at least some of these families \cite{Gossmann2014}.
-
-Taken together, we found that the presence of NamPRT and NNMT coincides with an increased diversification of NAD-dependent signalling.
-That is surprising as intuitively, one would expect that a decrease in precursor concentration caused by the precursor removal through NNMT, should cause a decrease of NAD availability and consequently less active NAD dependent signalling.
-
-\subsection{Dynamics of NAD biosynthesis and consumption}
-
-So why does the diversity of NAD-dependent signalling increase?
-And why does NADA disappear in Deuterostomia although it is the predominant pathway in bacteria, plants and fungi?
-Given the complexity of the NAD metabolic network, this question is difficult to be comprehensively addressed experimentally.
-Thus, to answer these questions we built a dynamic model of NAD metabolism using existing kinetic data from the literature (details see materials and methods and supplementary material).
-
-To be able to compare metabolic features of evolutionary quite different systems in our simulations, and as we have limited information about expression levels of enzymes or changes of kinetic constants during evolution, we initially used the kinetic constants found for yeast or human enzymes for all systems analysed and used equal amounts of enzymes for all reactions.
-Wherever possible we did not only include substrate affinities but also known product inhibition or inhibition by downstream metabolites.
-As we assume that cell growth is, besides NAD-consuming reactions, a major driving force for NAD biosynthesis, we analysed different growth rates (cell division rates) by simulating different dilution rates for all metabolites.
-We assume furthermore that Nam availability is different for different organisms and thus in addition analysed this simulating different Nam import rates.
-
-The pathway using NADA and recycling Nam via NA is superior \todo[author=Mathias]{What is meant with superior?
-NAD consumption flux is higher or lower with NNMT, depending on the condition.
-Free NAD concentration is higher without NNMT, but is that superior?} both in terms of steady state NAD concentration and NAD consumption flux (representing the activity of NAD-dependent signalling) in the absence of NNMT (Figure 3).
-In the presence of NNMT, the picture is slightly different (Figure 3C-D), even though NADA is still superior to NamPRT under most conditions \todo[author=Ines]{Check if not under all shown}\todo[author=Mathias]{Reply: From fig 4C, it looks like the flux is decreased at very high cell division rates while the Nam import rate is low.}.
-Only if Nam availability is very low, NamPRT is performing better because of its high substrate affinity.
-However, the disadvantage of NamPRT at equal amounts of enzyme, can be compensated by an increased expression of NamPRT.
-Although, to reach similar NAD concentrations in our model, NamPRT expression has to be tenfold higher than the expression of NAPRT (Figure S1).
-The NADA expression required is very low due to its high turnover.
-This might provide an explanation why we find NADA and with that the pathway via NA predominantly in bacteria, yeast, and plants, organisms that show high cell division rates.
-Under these conditions, protein expression costs are assumed to have a high impact on metabolic performance \todo[author=Mathias]{source} and thus pathways where low enzyme expression suffices, might be favourable, even though the pathway via NA is energetically less efficient.
-
-NADA seems to be able to maintain higher NAD concentrations than NamPRT.
-This figure changes if we simulate two organisms or cells that are in direct competition for Nam.
-Under these conditions, the organism expressing NamPRT has a competitive advantage above NADA-expressing organisms, but this only holds if the competing organisms expresses NamPRT together with NNMT.
-(new Figure) This observation might explain, why NADA disappeared in Metazoa together with the appearance of NNMT.
-It might also explain, why many bacteria associated with mammals harbour NamPRT instead of NADA\todo[author=Mathias]{Needs some figure or source}.
-
-Although these simulations already provide an idea of why we find NamPRT predominantly in combination with NNMT, we still do not know why this development coincides with an increased diversification of NAD-consuming enzymes.
-When we simulate the presence or absence of NNMT in the presence of NamPRT, we see that the impact of NNMT on NAD concentration is relatively small (Figure 4D), but we see that that NNMT increases the NAD consumption flux under most conditions in the presence of NamPRT (Figure 4C).
-NAD consumption flux can be further increased by increasing the expression NamPRT, which also compensates the slight reduction in NAD concentration in the presence of NNMT (see Figure S2).
-
-These findings can be explained when looking in more detail into the kinetic parameters of NamPRT and NAD-consuming enzymes such as Sirt1p.
-The ability to maintain high a NAD concentration in the presence of NNMT and at low Nam availability, is due to the very high affinity of NamPRT for its substrate, having a half saturation constant (Km) in the low nanomolar range.
-The increase of NAD consumption flux is caused by the fact that most NAD-consuming enzymes are inhibited by their product Nam, which is also the reason why the presence of NNMT enables higher NAD consumption fluxes.
-
-As the substrate affinity of NamPRT for Nam is extremely high (with a Km in the low nanomolar range) and as this might not have been the case throughout evolution, we analysed the effect of the NamPRT Km on NAD steady state concentration and NAD consumption flux, leaving all other kinetic parameters constant.
-In the absence of NNMT (Figure 5A-B) the Km has very little effect on steady state NAD concentration and NAD consumption flux.
-Without NNMT, NAD concentration and consumption flux are both considerably affected by cell division rates, at least if the enzyme expression is kept constant.
-This is of course an artificial scenario, as one would assume organisms to regulate enzyme expression to achieve similar levels of metabolite concentrations instead.
-In the absence of NNMT there appears furthermore to be a trade-off between achievable steady state NAD concentration and NAD consumption flux.
-
-In the presence of NNMT, NAD consumption flux and NAD concentration increases with decreasing Km values (Figure 5C-D).
-And we note, that both the NAD steady state concentration and the consumption flux are relatively stable over a wide range of cell division rates even at constant expression levels of the involved enzymes, suggesting that NNMT might have an important role to maintain homeostasis of NAD metabolism.
-
-When we compare NAD consumption and NAD concentration with and without NNMT with two different substrate affinities of NamPRT, we see that at a low affinity (Km of 100 nM, which is in the range of the Km of NAPRT for its substrate, or the Km of NADA for Nam), NAD consumption flux is only higher with NNMT at low cell division rates, whereas at high division rates, higher NAD consumption fluxes are achieved without NNMT (Figure 5 E-F).
-This might explain why we do not find NNMT in organism that tend to have high growth rates\todo[author=Ines]{I am not sure if we should go even further suggesting that decreasing substrate affinity by using a competitive inhibitor such as FK866 will influence the consumption rate in fast growing cells much more than the one in slow growing cells, being potentially relevant for cancer therapy.
-– Discussion}.
-We furthermore find, that the competitive advantage of NamPRT of NADA is only present at sufficiently high affinity of NamPRT.
-
-Our analysis also suggests that NNMT might have exerted an evolutionary driving force on the substrate affinity of NamPRT, explaining the extreme values found for the human enzyme\todo[author=Ines]{Move to discussion?}.
-
-The pathway dynamics are of course not solely dependent on one enzyme.
-Thus, what is the impact of the substrate affinity of NNMT that is competing with NamPRT for the same substrate?
-In Figure 6A we see that the substrate affinity values found in the human enzymes (indicated by black asterisks) are actually optimal with respect to both achievable steady state NAD concentration and consumption fluxes.
-Thus, a further increase of the affinity of NamPRT for Nam would not provide any advantage.
-
-\subsection{Sequence variance acquired in metazoans enhances substrate affinity}
-
-To see whether we can find sequence variations in the protein sequence of NamPRT that indicate evolutionary changes of NamPRT upon the appearance of NNMT, we created a multiple sequence alignment of selected eukaryotic NamPRT sequences.
-As shown in Figure 7A and Supplementary Figure 3\todo[author=Ines]{To be created}, Deuterostomia that have only NamPRT and NNMT (indicated by the number 6 in parenthesis \todo[author=Ines]{Maybe rather indicate this by colour and the kingdom through abbrev?}) have an insert of ten amino acids corresponding to positions 43 to 52 of the human enzyme.
-Looking at the crystal structure of human NamPRT this sequence insertion corresponds to a region that has not been structurally resolved in any of the currently available crystal structures (e.g.
-\cite{Wang2006} structure visualisation Figure 7B) and its function is unclear.
-The unresolved loop structure overlaps with a predicted weak nuclear localisation signal (figure 7A), that is not present without the insertion.
-The loop is in addition connected to one of the beta-sheets involved in substrate binding, potentially affecting the affinity or turnover of the enzyme.
-
-The observed evolutionary change in the primary sequence of NamPRT could therefore have had different effects, that we wanted to test experimentally.
-We first investigated whether the deletion of the amino acids 43 to 52 has an effect on the localisation of the human enzyme, expressing a ….
-mutant lacking this sequence in HeLa cells.
-We could, however, not detect any changes in subcellular localisation (Figure 7C) compared to the wildtype fusion construct.
-We therefore conclude that the partial nuclear localisation of NamPRT is not affected by the deletion.
-We then tested the enzymatic activity of the NamPRT mutant recombinantly expressed in E. coli.
-As shown in Figure 7D the enzyme still forms a dimer, thus seems to be folded correctly.
-The enzymatic activity is, however, much lower compared to the wildtype enzyme.
-Using higher concentrations of Nam, it appears that the mutant is not saturated at 100 \textmu M, pointing to a decreased substrate affinity of the mutant enzyme.
-
-\subsection{Comments and parts of old version not included yet}
-
-Chordata: NADA disappears.
-These analyses also support the view that Tunicata and Branchiostoma are not part of the phylum chordata\todo[author=Mathias]{Is this view in any way argued upon by anyone??}.
-
-In the evolutionary context, an additional question arises: Why do only a few organisms, mostly less complex animals, possess the gene for NADA in addition to NamPRT and NNMT.
-When including NADA into the simulations, we see that the effect of NADA on NAD consumption is very limited in the presence of NNMT (old Figure 4F) even at high expression levels (see old Suppl.
-Figure S1C and D) and could be compensated by increased expression of NamPRT (not shown).
-
-Until this point we have neglected compartmentalisation of the pathway.
-We do however know from previous studies that in early vertebrate development a compartmentalisation of the pathway has occurred reflected by a gene triplication of NMNAT and the occurrence of compartment-specific domains called ISTIDs (Lau, 2010).
-Looking at the evolutionary timepoint \todo[author=Toni]{This time point can only be reconstructed, but is unknown.
-I think a figure with the likely reconstruction would help to follow the argumentation} of appearance of NNMT and the gene triplication of NMNATs, we see that NNMT occurs prior to the gene triplication and the first occurrence of ISTIDs and we have confirmed this in our own analysis (not shown).
+%%% TeX-master: "manuscript"
 
@@
 Confocal laser scan imaging of cells was performed using a Leica TCS SP8 STED 3x microscope equipped with a 100x oil immersion objective (numerical aperture 1.4).
-
-\section{Figure Legends}
-
-\subsection{Figure1}
-
-Schematic overview of NAD biosynthesis and consumption.
-NAD can be synthesized using several routes from altogether three main precursors: tryptophan (Trp), nicotinamide (Nam), and nicotinic acid (NA).
-Nam and NA are together known as vitamin B3 or niacin.
-To a lesser extend, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead.
-Only 1\% of the tryptophan taken up with our diet is converted into NAD, thus, vitamin B3 and, to a lower extend, NR are essential components of our diet, with Nam being the major NAD precursor in humans\todo[author=Mathias]{Whole sentence not necessarily part of the legend}.
-Nam is furthermore the product of NAD-consuming signalling reactions such as sirtuins (NAD-dependent histone deacetylases) or PARPs (poly-ADP-ribosylases).
-
-For the recycling of Nam, two different pathways exist.
-The pathway found in yeast and many bacteria is using a four-step pathway starting with the deamination of Nam to nicotinic acid (NA) by Nam deamidase (NADA).
-The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates.
-The recycling pathway found in vertebrates directly converts Nam into the corresponding mononucleotide (NMN) a reaction catalysed by NamPRT and driven by a non-stoichiometric ATP-hydrolysis.
-A similar reaction catalysed by an evolutionary related enzyme NAPRT, converts NA into the NA mononucleotide (NAMN).
-NMN and NAMN are converted into dinucleotides by the Nam/NA adenylytransferases (NMNATs).
-The recycling pathway via NA finally requires an amination step catalysed by NADsynthase (NADS), driven by the conversion of ATP to AMP producing pyrophosphate.
-
-\subsection{Figure 2}
-
-Evolutionary distribution of NADA, NNMT and NamPRT and their relation to the number of NAD consumers.
-A) Distribution of NADA, NNMT and NamPRT in selected major taxa.
-NADA is dominant in Bacteria, Fungi, and Plants (Viridiplantae), whereas NamPRT together with NNMT is dominant in Deuterostomia.
-Numbers at the pie charts show, how many species of the taxon possess the respective enzyme combination indicated by the colour explained in the lower right of the figure.
-Below the taxon name, the number of species in that taxon is given.
-
-B) Common tree of selected taxa within the Metazoa, including 334 species.
-The pie charts indicate the distribution of species within the respective taxon that have the enzyme combination indicated by the colour, explained in the lower right.
-The size of the pie charts is proportional to the logarithm of the number of species analysed in the particular taxon.
-The numbers below the taxon names indicate the average number of NAD-consuming enzyme families found in all sub-taxa.
-The branch length is arbitrary.
-
-\subsection{Figure 3}
-
-NAD consumption flux and NAD concentration in simulations of organisms with NADA(?) and with and without NNMT at different Nam import and cell division rates.
-
-\subsection{Figure 4}
-
-Impact of NNMT on NAD consumption and NAD concentration in simulations of organisms with NADA or NamPRT.
-
-\subsection{Figure 5}
-
-NAD consumption flux and NAD concentration in simulations of organisms with NamPRT and with and without NNMT at different KMs for NamPRT and cell division rates.
-
-\subsection{Figure 6}
-
-The substrate affinity of NNMT and NamPRT have opposite effects on NAD consumption (A) and concentration (B), as would be expected.
-The affinities previously measured for human enzymes (indicated by a black asterisk) appear to be close to optimal, as further improvements would have little or no effect on NAD consumption or concentration.
-
-\subsection{Figure 7}
-
-Deuterostomia that encode NNMT show a sequence insertion in the N-terminal region of NamPRT.
-A) Multiple sequence alignment of NamPRT of selected deuterostomes show a sequence insertion in organisms that encode NamPRT and NNMT.
-(Species names are given and the number in parenthesis indicates the enzyme combination encoded in these species; 3 – NADA and NamPRT, 6 – NamPRT and NNMT, 7 – NADA, NamPRT and NNMT.
-\todo[author=Mathias]{Note to myself: replace the numbers by something better and rewrite this part.}) The nuclear localisation signal (NLS\todo[author=Mathias]{Note to myself: Write “predicted NLS” in figure}) indicated, was predicted using using cNLS-mapper (\url{http://nls-mapper.iab.keio.ac.jp/cgi-bin/NLS_Mapper_form.cgi} \cite{Kosugi2009}).
-
-B) The inserted region is not resolved in crystal structures of human NamPRT and thus appears to be a flexible loop structure at the surface of the NamPRT dimer, coloured in red.
-The visualisation is based on a structure prediction of SWISS-MODEL \cite{Arnold2006; Biasini2014} of the sequence of the human NamPRT (P43490) using the model 2H3D as template \cite{Wang2006}.
-
-C) The localisation of wildtype NamPRT varies between pure cytosolic and mixed cytosolic-nuclear localisation.
-This is not changed in the mutant missing the insert that corresponds to the unresolved loop structure.
-
-\end{document}

### 2018-06-20 latex/figurelegends.tex
splitted document in several files to speed up editing

--- 2018-06-18 latex/manuscript.tex
@@
-\RequirePackage[l2tabu, orthodox]{nag}
-\documentclass[paper=a4, 12pt]{scrartcl}
-\usepackage[utf8]{inputenc}
-\usepackage[colorlinks=true,
-	linkcolor=black,
-	citecolor=black,
-	filecolor=black,
-	urlcolor=black]{hyperref}
-\usepackage{textcomp}
-\usepackage[left=1cm,right=5cm,top=2cm,bottom=2cm,nohead,nofoot]{geometry}
-\usepackage{todonotes}
-
-\setlength{\marginparwidth}{4cm}
-
-\begin{document}
-
-\noindent
-{\huge\sffamily\bfseries NamPRT and NNMT – key drivers of NAD-dependent signalling \par}
-
-\vspace{2cm}
-
-\noindent
-Mathias Bockwoldt$^{1}$, Marc Niere$^{2}$, Toni I. Gossmann$^{3}$, Mathias Ziegler$^{2}$ and Ines Heiland$^{1,\S}$
-
-\vspace{1cm}
-
-\noindent
-$^{1}$Department of Arctic and Marine Biology, UiT Arctic University of Norway, Naturfagbygget, Dramsveien 201, 9037 Tromsø, Norway
-
-\noindent
-$^{2}$Department of Molecular Biology, University of Bergen, Thormøhlensgt. 55
-5020 Bergen, Norway
-
-\noindent
-$^{3}$Department of Animal and Plant Sciences, Western Bank, University of Sheffield
-Sheffield, S10 2TN, United Kingdom
-
-\noindent
-§ Corresponding author: ines.heiland@uit.no
-
-\section{Summary}
-
-\section{Keywords}
-
-\section{Highlights}
-
-\section{Introduction}
-
-NAD metabolism represents one of the most critical links that connect cellular signal transduction and energy metabolism.
-According to Kyoto Encyclopedia of Genes and Genomes KEGG (\url{http://www.genome.jp/kegg/}), more than 20\% of all biochemical reactions use NAD(P), mostly in redox reactions in which NAD(P)+ and NAD(P)H are reversibly interconverted.
-In addition to redox reactions, various signalling processes consume NAD+ by cleaving the molecule to nicotinamide (Nam) and ADP-ribose \cite{Verdin2015}.
-NAD+-dependent signalling reactions include poly- and mono-ADP-ribosylation, NAD-dependent protein deacylation by sirtuins as well as the synthesis of calcium mobilizing molecules such as cyclic ADP-ribose \cite{Opitz2015}.
-These NAD+-dependent signalling processes participate in the regulation of virtually all cellular activities.
-The enzymes involved in these processes are sensitive to the available NAD+ concentration \cite{Ruggieri2015}, which in turn is dependent on the NAD+/NADH redox ratio.
-Therefore, NAD+-dependent signalling can act as a transmitter of changes in the cellular energy homeostasis, for example, to regulate gene expression or metabolic activity \cite{Koch-Nolte2009}.
-
-The significance of NAD+-dependent signalling for NAD homeostasis has long been underestimated.
-However, it has now been established that inhibition of NAD biosynthesis in mammalian cells leads to a rapid decline in NAD concentration suggesting that NAD+-dependent signalling consumes substantial amounts of NAD (Ref?), that is why we later refer to them as NAD-consuming reactions.
-As it has been shown that NAD concentrations change in a circadian manner \cite{Nakahata2009; Ramsey2009}, the cellular NAD-pool is turned over at least once per day\todo[author=Mathias]{Either the NAD-pool is turned over at least once a day or the concentration is maintained, but both … could be possible but sounds strange.}.
-To maintain the NAD concentration at physiological levels, NAD biosynthesis needs to proceed at an equally rapid rate.
-Imbalances in NAD-homeostasis have been linked to various, in particular, ageing-related diseases such as diabetes, neurodegenerative disorders and cancer \cite{Chiarugi2012; Verdin2015}\todo[author=Ines]{NAD turnover}.
-Various recent studies have demonstrated impressive health benefits of dietary supplementation with intermediates of NAD biosynthesis including NMN and nicotinamide riboside, NR \cite{Belenky2007}\todo[author=Ines]{That is not very recent.
-Better suggestions?}.
-Apparently, the exploitation of NAD biosynthetic routes, in addition to the use of nicotinamide as precursor (Fig.
-1), results in increased NAD concentrations that stimulate NAD+-dependent signalling processes, in particular, protein deacetylation by sirtuins.
-\todo[author=Ines]{Shorten introduction}
-
-In mammals, NAD biosynthesis predominantly relies on nicotinamide \cite{Bogan2008}\todo[author=Ines]{Not the perfect reference maybe}, a form of vitamin B3, as precursor.
-In the first step, Nam is converted to the mononucleotide, NMN, by Nam phosphoribosyltransferase (NamPRT) using phosphoribosylpyrophosphate (PRPP) as co-substrate.
-The key role of this enzyme is related to the nature of NAD+-dependent signalling reactions, namely, that in all these reactions Nam is cleaved off and potentially lost.
-If the released Nam were not recycled into NAD biosynthesis, humans would require a much higher daily vitamin B3 intake than the 15 mg that are the current daily recommendation (ref) \todo[author=Ines]{I have not been able to find a reference for that so far.}.
-The nearly complete recycling of Nam is achieved by an extraordinary high affinity of NamPRT to Nam, the Km being in the low nanomolar range \cite{Burgos2008}.
-Despite the importance of its salvage, Nam can also be marked for excretion by methylation \todo[author=Mathias]{It’s not too surprising.
-If there was no way of removing parts, we would accumulate more and more, as we still take up 15 mg every day.}.
-Indeed, the presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012} is among the most enigmatic and counterintuitive features of NAD metabolism.
-Why is there one enzyme (NamPRT) seemingly optimized to recycle even the faintest amounts of Nam back into NAD synthesis, while at the same time there is another one (NNMT) that seems to have no metabolic function other than to remove Nam from NAD metabolism\todo[author=Ines]{This paragraphs to the start including a short statement about the purpose of the analysis.}?
-This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deaminated Nam to nicotinic acid (NA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (Fig.
-1).
-Previous phylogenetic analysis show that the two Nam recycling pathways are both ancestral showing a scattered distribution in bacteria \cite{Gazzaniga2009}.
-In contrast, NNMT has so far only been found in animals \cite{Gossmann2012}.
-The analysis of eukaryotic pathway evolution was previously limited by the low number of available genome sequences.
-A comprehensive evolutionary analysis is, however, needed to fully understand the phylogeny of NAD metabolism, in particular, Nam metabolizing enzymes and their relationship to signalling pathways.
-
-This analysis may also help to unravel, how biosynthesis and signalling are coupled and influence each other, as this is not well understood so far.
-It is widely assumed that a major driving force of NAD+-consuming signalling reactions is the cellular NAD concentration.
-Most enzymes using NAD in signalling reactions that have been characterized in detail and it was shown that most sirtuins and PARPs are subject to inhibition by nicotinamide \cite{Borra2004; Ko2012}.
-To promote NAD+-dependent signalling reactions, the balance between elevated NAD levels and accumulating nicotinamide concentrations needs to be kept such that the inhibitory effect of Nam does not override the kinetic stimulation by high NAD levels \todo[author=Mathias]{Well… To promote [signalling] reactions, Nam simply has to be turned into NAD as quickly as possible.
-Besides that it sounds, form the description, as a self-regulating process.}.
-Given the complex nature of the NAD metabolome, it is not trivial to predict the optimal conditions for efficient NAD dependent signalling.
-However, it is obvious that Nam-converting enzymes must play a key role in this regard.
-
-In the present study, we have conducted a comprehensive phylogenetic analysis of the genes that encode Nam-converting enzymes.
-The most surprising result was that with the emergence of Deuterostomes, there has been a strong tendency towards the co-existence of NamPRT and NNMT.
-Moreover, this selection for the co-existence of NamPRT and NNMT was accompanied by a marked increase in the number of genes encoding NAD+-consuming signalling reactions.
-To explain this observation, we built a kinetic model of NAD metabolism based on available kinetic parameters.
-This mathematical model provided a powerful tool to analyse the relationship between NAD biosynthetic and signalling fluxes when different subsets of Nam-converting enzymes are present.
-The model demonstrated that NNMT has a critical role to maintain high NAD+-consuming signalling fluxes by preventing accumulation of inhibitory Nam.
-At lower Nam concentrations, the impact of NNMT is minimal, because of the very high affinity of NamPRT.
-Taken together, our analyses suggest that the co-existence of NamPRT and NNMT has been a prerequisite to enable the evolutionary development of versatile NAD+-dependent signalling mechanisms present in vertebrates.
-
-\section{Results}
-
-\subsection{Phylogenetic analysis of NAD biosynthesis and consumption}
-
-As shown in Fig.
-1, NAD can be synthesized using several routes from three main precursors: tryptophan, nicotinamide (Nam) and nicotinic acid (NA).
-Nam and NA are together known as vitamin B3 or niacin.
-Alternatively, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead \cite{Bogan2008}.
-As in humans only 1\% of the tryptophan taken up with our diet is converted into NAD, vitamin B3 and to a lower extend NR are essential components of our diet, with Nam being the major NAD precursor in humans.
-Nam is also the product of NAD consuming signalling reactions such as sirtuins (NAD-dependent histone deacetylases) or PARPs (poly ADP ribosylases).
-
-For the recycling of Nam, two different pathways exist.
-The pathway found in yeast and plants is using a four-step pathway starting with the deamination of Nam to nicotinic acid by Nam deamidase (NADA).
-The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates.
-The recycling pathway found in mammals directly converts Nam into the corresponding mononucleotide (NMN) a reaction catalysed by NamPRT and driven by a non-stoichiometric ATP-hydrolysis.
-A similar reaction catalysed by an evolutionary related enzyme NAPRT, converts NA into the NA mononucleotide in the Preiss-Handler pathway.
-NMN and NAMN are converted into dinucleotides by the Nam/NA adenylytransferases (NMNATs).
-The recycling pathway via NA finally requires an amination step catalysed by NADsynthase, driven by the conversion of ATP to AMP producing pyrophosphate.
-Even though the latter pathway seems to be very inefficient, it is the pathway preferentially used by most bacteria, fungi and plants (see Figure 2A), whereas most metazoans recycle Nam using NamPRT.
-
-Analysing the phylogeny of the NAD recycling enzymes in Metazoa in more detail reveals that not only does NamPRT replace NADA, but in most organisms, especially in Deuterostomia, NamPRT is found together with the Nam methyltransferase NNMT (Figure 2B).
-NNMT methylates Nam to methyl-Nam that is in mammals excreted with the urine, thus removing Nam from recycling.
-NNMT seems to have arisen de novo in the common ancestor of Ecdysozoa and Lophotrochozoa, as we could not find any gene with considerable similarity in fungi or plants.
-Nematodes are the only organisms where we find NNMT together with NADA without NamPRT being present.
-In Deuterostomia the only large class that does only have NamPRT and seems to have lost NNMT again are Sauropsida and here especially birds.
-The reason why a lot of birds do not encode NNMT remains unclear, as the appearance is quite scattered (not shown)\todo[author=Mathias]{Maybe as supplementary figure?}.
-It might be related to the excretion system, as the product of NNMT methyl-nicotinamide is in mammals excreted with the urine.
-There are some species where we could not find NamPRT or NADA but NNMT, we assume that this is due to incomplete genomes in the database\todo[author=Ines]{Not sure we should discuss Branchisotoma and Tunicata here}.
-
-In addition to the phylogenetic distribution of the two Nam salvage enzymes NADA and NamPRT, we looked at phylogenetic diversity of enzymes catalysing NAD-dependent signalling reactions.
-To do so we used the previously established classification into 10 different families \cite{Gossmann2012} (For details see materials and methods and supplementary information).
-The numbers in Figure 2B denote the average number of NAD-dependent signalling enzyme families we found in each taxa.
-With the exception of Cnidaria and Lophotrochozoa, we find in Protostomia on average 3 to 4 families, whereas most Deuterostomia have on average more than 8 families with an increasing diversification of enzymes within at least some of these families \cite{Gossmann2014}.
-
-Taken together, we found that the presence of NamPRT and NNMT coincides with an increased diversification of NAD-dependent signalling.
-That is surprising as intuitively, one would expect that a decrease in precursor concentration caused by the precursor removal through NNMT, should cause a decrease of NAD availability and consequently less active NAD dependent signalling.
-
-\subsection{Dynamics of NAD biosynthesis and consumption}
-
-So why does the diversity of NAD-dependent signalling increase?
-And why does NADA disappear in Deuterostomia although it is the predominant pathway in bacteria, plants and fungi?
-Given the complexity of the NAD metabolic network, this question is difficult to be comprehensively addressed experimentally.
-Thus, to answer these questions we built a dynamic model of NAD metabolism using existing kinetic data from the literature (details see materials and methods and supplementary material).
-
-To be able to compare metabolic features of evolutionary quite different systems in our simulations, and as we have limited information about expression levels of enzymes or changes of kinetic constants during evolution, we initially used the kinetic constants found for yeast or human enzymes for all systems analysed and used equal amounts of enzymes for all reactions.
-Wherever possible we did not only include substrate affinities but also known product inhibition or inhibition by downstream metabolites.
-As we assume that cell growth is, besides NAD-consuming reactions, a major driving force for NAD biosynthesis, we analysed different growth rates (cell division rates) by simulating different dilution rates for all metabolites.
-We assume furthermore that Nam availability is different for different organisms and thus in addition analysed this simulating different Nam import rates.
-
-The pathway using NADA and recycling Nam via NA is superior \todo[author=Mathias]{What is meant with superior?
-NAD consumption flux is higher or lower with NNMT, depending on the condition.
-Free NAD concentration is higher without NNMT, but is that superior?} both in terms of steady state NAD concentration and NAD consumption flux (representing the activity of NAD-dependent signalling) in the absence of NNMT (Figure 3).
-In the presence of NNMT, the picture is slightly different (Figure 3C-D), even though NADA is still superior to NamPRT under most conditions \todo[author=Ines]{Check if not under all shown}\todo[author=Mathias]{Reply: From fig 4C, it looks like the flux is decreased at very high cell division rates while the Nam import rate is low.}.
-Only if Nam availability is very low, NamPRT is performing better because of its high substrate affinity.
-However, the disadvantage of NamPRT at equal amounts of enzyme, can be compensated by an increased expression of NamPRT.
-Although, to reach similar NAD concentrations in our model, NamPRT expression has to be tenfold higher than the expression of NAPRT (Figure S1).
-The NADA expression required is very low due to its high turnover.
-This might provide an explanation why we find NADA and with that the pathway via NA predominantly in bacteria, yeast, and plants, organisms that show high cell division rates.
-Under these conditions, protein expression costs are assumed to have a high impact on metabolic performance \todo[author=Mathias]{source} and thus pathways where low enzyme expression suffices, might be favourable, even though the pathway via NA is energetically less efficient.
-
-NADA seems to be able to maintain higher NAD concentrations than NamPRT.
-This figure changes if we simulate two organisms or cells that are in direct competition for Nam.
-Under these conditions, the organism expressing NamPRT has a competitive advantage above NADA-expressing organisms, but this only holds if the competing organisms expresses NamPRT together with NNMT.
-(new Figure) This observation might explain, why NADA disappeared in Metazoa together with the appearance of NNMT.
-It might also explain, why many bacteria associated with mammals harbour NamPRT instead of NADA\todo[author=Mathias]{Needs some figure or source}.
-
-Although these simulations already provide an idea of why we find NamPRT predominantly in combination with NNMT, we still do not know why this development coincides with an increased diversification of NAD-consuming enzymes.
-When we simulate the presence or absence of NNMT in the presence of NamPRT, we see that the impact of NNMT on NAD concentration is relatively small (Figure 4D), but we see that that NNMT increases the NAD consumption flux under most conditions in the presence of NamPRT (Figure 4C).
-NAD consumption flux can be further increased by increasing the expression NamPRT, which also compensates the slight reduction in NAD concentration in the presence of NNMT (see Figure S2).
-
-These findings can be explained when looking in more detail into the kinetic parameters of NamPRT and NAD-consuming enzymes such as Sirt1p.
-The ability to maintain high a NAD concentration in the presence of NNMT and at low Nam availability, is due to the very high affinity of NamPRT for its substrate, having a half saturation constant (Km) in the low nanomolar range.
-The increase of NAD consumption flux is caused by the fact that most NAD-consuming enzymes are inhibited by their product Nam, which is also the reason why the presence of NNMT enables higher NAD consumption fluxes.
-
-As the substrate affinity of NamPRT for Nam is extremely high (with a Km in the low nanomolar range) and as this might not have been the case throughout evolution, we analysed the effect of the NamPRT Km on NAD steady state concentration and NAD consumption flux, leaving all other kinetic parameters constant.
-In the absence of NNMT (Figure 5A-B) the Km has very little effect on steady state NAD concentration and NAD consumption flux.
-Without NNMT, NAD concentration and consumption flux are both considerably affected by cell division rates, at least if the enzyme expression is kept constant.
-This is of course an artificial scenario, as one would assume organisms to regulate enzyme expression to achieve similar levels of metabolite concentrations instead.
-In the absence of NNMT there appears furthermore to be a trade-off between achievable steady state NAD concentration and NAD consumption flux.
-
-In the presence of NNMT, NAD consumption flux and NAD concentration increases with decreasing Km values (Figure 5C-D).
-And we note, that both the NAD steady state concentration and the consumption flux are relatively stable over a wide range of cell division rates even at constant expression levels of the involved enzymes, suggesting that NNMT might have an important role to maintain homeostasis of NAD metabolism.
-
-When we compare NAD consumption and NAD concentration with and without NNMT with two different substrate affinities of NamPRT, we see that at a low affinity (Km of 100 nM, which is in the range of the Km of NAPRT for its substrate, or the Km of NADA for Nam), NAD consumption flux is only higher with NNMT at low cell division rates, whereas at high division rates, higher NAD consumption fluxes are achieved without NNMT (Figure 5 E-F).
-This might explain why we do not find NNMT in organism that tend to have high growth rates\todo[author=Ines]{I am not sure if we should go even further suggesting that decreasing substrate affinity by using a competitive inhibitor such as FK866 will influence the consumption rate in fast growing cells much more than the one in slow growing cells, being potentially relevant for cancer therapy.
-– Discussion}.
-We furthermore find, that the competitive advantage of NamPRT of NADA is only present at sufficiently high affinity of NamPRT.
-
-Our analysis also suggests that NNMT might have exerted an evolutionary driving force on the substrate affinity of NamPRT, explaining the extreme values found for the human enzyme\todo[author=Ines]{Move to discussion?}.
-
-The pathway dynamics are of course not solely dependent on one enzyme.
-Thus, what is the impact of the substrate affinity of NNMT that is competing with NamPRT for the same substrate?
-In Figure 6A we see that the substrate affinity values found in the human enzymes (indicated by black asterisks) are actually optimal with respect to both achievable steady state NAD concentration and consumption fluxes.
-Thus, a further increase of the affinity of NamPRT for Nam would not provide any advantage.
-
-\subsection{Sequence variance acquired in metazoans enhances substrate affinity}
-
-To see whether we can find sequence variations in the protein sequence of NamPRT that indicate evolutionary changes of NamPRT upon the appearance of NNMT, we created a multiple sequence alignment of selected eukaryotic NamPRT sequences.
-As shown in Figure 7A and Supplementary Figure 3\todo[author=Ines]{To be created}, Deuterostomia that have only NamPRT and NNMT (indicated by the number 6 in parenthesis \todo[author=Ines]{Maybe rather indicate this by colour and the kingdom through abbrev?}) have an insert of ten amino acids corresponding to positions 43 to 52 of the human enzyme.
-Looking at the crystal structure of human NamPRT this sequence insertion corresponds to a region that has not been structurally resolved in any of the currently available crystal structures (e.g.
-\cite{Wang2006} structure visualisation Figure 7B) and its function is unclear.
-The unresolved loop structure overlaps with a predicted weak nuclear localisation signal (figure 7A), that is not present without the insertion.
-The loop is in addition connected to one of the beta-sheets involved in substrate binding, potentially affecting the affinity or turnover of the enzyme.
-
-The observed evolutionary change in the primary sequence of NamPRT could therefore have had different effects, that we wanted to test experimentally.
-We first investigated whether the deletion of the amino acids 43 to 52 has an effect on the localisation of the human enzyme, expressing a ….
-mutant lacking this sequence in HeLa cells.
-We could, however, not detect any changes in subcellular localisation (Figure 7C) compared to the wildtype fusion construct.
-We therefore conclude that the partial nuclear localisation of NamPRT is not affected by the deletion.
-We then tested the enzymatic activity of the NamPRT mutant recombinantly expressed in E. coli.
-As shown in Figure 7D the enzyme still forms a dimer, thus seems to be folded correctly.
-The enzymatic activity is, however, much lower compared to the wildtype enzyme.
-Using higher concentrations of Nam, it appears that the mutant is not saturated at 100 \textmu M, pointing to a decreased substrate affinity of the mutant enzyme.
-
-\subsection{Comments and parts of old version not included yet}
-
-Chordata: NADA disappears.
-These analyses also support the view that Tunicata and Branchiostoma are not part of the phylum chordata\todo[author=Mathias]{Is this view in any way argued upon by anyone??}.
-
-In the evolutionary context, an additional question arises: Why do only a few organisms, mostly less complex animals, possess the gene for NADA in addition to NamPRT and NNMT.
-When including NADA into the simulations, we see that the effect of NADA on NAD consumption is very limited in the presence of NNMT (old Figure 4F) even at high expression levels (see old Suppl.
-Figure S1C and D) and could be compensated by increased expression of NamPRT (not shown).
-
-Until this point we have neglected compartmentalisation of the pathway.
-We do however know from previous studies that in early vertebrate development a compartmentalisation of the pathway has occurred reflected by a gene triplication of NMNAT and the occurrence of compartment-specific domains called ISTIDs (Lau, 2010).
-Looking at the evolutionary timepoint \todo[author=Toni]{This time point can only be reconstructed, but is unknown.
-I think a figure with the likely reconstruction would help to follow the argumentation} of appearance of NNMT and the gene triplication of NMNATs, we see that NNMT occurs prior to the gene triplication and the first occurrence of ISTIDs and we have confirmed this in our own analysis (not shown).
-
-\section{Discussion}
-
-\subsection{Evolutionary and physiological role of NNMT}
-
-The dynamic interactions among metabolism, signal transduction, and gene regulation is still not very well understood.
-New approaches are required to disentangle the underlying network, helping us to understand how alterations affect human physiology.
-We here combine phylogenetic analysis and mathematical modelling of a central metabolic pathway supported by experimental analysis to reveal the likely evolutionary development of the NAD biosynthesis and consumption pathway and the physiological role of the methyltransferase NNMT.
-
-We show that the vitamin B3-degrading enzyme NNMT plays a vital role for the diversification of NAD-dependent signalling reactions and potentially also for NAD homeostasis.
-Due to the fact that NamPRT has a very high affinity for Nam whereas NNMT has a rather low affinity for it, NNMT is able to remove excess Nam that would lead to inhibition of NAD-dependent signalling reactions while maintaining NAD concentrations and even increasing NAD consumption fluxes.
-
-The affinities measured for the human enzymes seem to be optimal and further increase in NAD consumption rates might only enabled by compartmentalisation of NAD biosynthesis.
-
-\subsection{General applicability}
-
-It is of course impossible to extend our conclusions to other metabolic processes without analysing them in more detail, but we have noted that the metabolism of other vitamins, such as pyridoxal (vitamin B6) metabolism, also contains vitamin degrading enzymes \todo[author=Mathias]{source}, that might thus have similar roles.
-Degrading enzymes do in general not receive the same attention as biosynthetic enzymes, reflected by the observation that NNMT has only recently been analysed in more detail, besides the much earlier recognition of NNMT as potential marker for some types of cancer (e.g.
-Ref?) and routine clinical measurements of urine methyl-Nam concentrations in the context of different diseases (e.g.
-Ref?).
-Our analysis show, however, that degrading enzymes do not solely modify substrates for better excretion, but can play a vital role both in human physiology as well as in the evolutionary development of biological processes.
-
-\subsection{Limitations of our analysis and outlook}
-
-Among others, we neglected the potential effects of the methyl donor S-adenosyl-methionin (SAM) and its precursor methionine in our analysis, although it is most likely contributing yet another regulatory level for Nam availability and thus an additional interaction point between gene regulation and metabolism.
-It has been shown earlier that in turn, NNMT expression influences protein methylation dependent on methionine availability \cite{Ulanovskaya2013}.
-One of the challenges in the analysis of this aspect is the availability of in vivo concentration measurements for SAM and the large amount of reactions using it as substrate.
-The same holds for the analysis of enzyme acetylation versus deacetylation by sirtuins that would enable prediction about enzyme activation or histone state if available.
-
-\section{Experimental Procedure}
-
-\subsection{Dynamic modelling}
-
-Kinetic parameters (substrate affinity (Km) and turnover rates (kcat), substrate and product inhibitions) were retrieved from the enzyme database BRENDA and additionally evaluated by checking the original literature especially with respect to measurement conditions.
-Parameter values from mammals were used if available.
-For enzymes not present in mammals, values from yeast were used.
-The full list of kinetic parameters including reference to original literature can be found in Supplementary table 1.
-For NMNAT, the previously developed rate law for substrate competition was used \cite{Schauble2013}.
-Despite these modifications, Henri-Michaelis-Menten kinetics were used for all reactions except the import and efflux of Nam, which were simulated using constant flux and mass action kinetics, respectively.
-All simulations were performed using the steady state calculation and parameter scan options provided by COPASI 4.22 \cite{Hoops2006}.
-The model will be available at the Biomodels database accession number \todo[inline]{xxx}.
-Related figures were generated using gnuplot\todo[author=Mathias]{Version?}.
-
-\subsection{Phylogenetic Analysis}
-
-NADA, NamPRT, and NNMT enzymes or enzyme candidates were identified using Blastp with known enzymes against the NCBI non-redundant protein sequence database (nr).
-A list of functionally verified enzymes used as templates is given in supplementary table 2.
-This table also includes the length cut-offs for identified enzymes.
-The e-value cut-off was 1e-30 for all enzymes.
-Blastp parameters were set to yield maximum 20,000 target sequences, using the BLOSUM62 matrix with a word size of 6 and gap opening and extension costs of 11 and 1, respectively.
-Low-complexity filtering was disabled.
-Obvious sequence contaminations were removed by manual inspection of the results.
-The taxonomy IDs of the species for each enzyme was derived from the accession2taxonomy database provided by NCBI.
-Scripts for creating, analysing, and visualising the phylogenetic tree were written in Python, using the ETE3 toolkit (Huerta-Cepas, 2010).
-
-\subsection{Generation of eukaryotic expression vectors encoding C-terminally FLAG-tagged NamPT proteins}
-
-The open reading frame (ORF) encoding human NamPRT was inserted into pFLAG-CMV-5a (Merck - Sigma Aldrich) via EcoRI/BamHI sites.
-Using a PCR-based approach, this vector provided the basis for the generation of a plasmid encoding a NamPRT deletion mutant lacking amino acid residues 42-51 (NamPRT-$\Delta$42-51).
-The sequences of the inserted ORFs were verified by DNA sequence analysis.
-
-\subsection{Transient transfection, immunocytochemstry and confocal laser scanning microscopy}
-
-HeLa S3 cells cultivated in Ham’s F12 medium supplemented with 10\% (v/v) FCS, 2 mM L-glutamine, and penicillin/streptomycin, were seeded on cover slips in a 24 well plate.
-After one day, cells were transfected using Effectene transfection reagent (Qiagen) according to the manufacturer’s recommendations.
-Cells were fixed with 4\% paraformaldehyde in PBS 24 hours post transfection, permeabilised (0.5\% (v/v) Triton X-100 in PBS) and blocked for one hour with complete culture medium.
-After overnight incubation with primary FLAG-antibody (mouse M2, Sigma-Aldrich) diluted 1:2500 in complete medium, cells were washed and incubated for one hour with secondary AlexaFluor 594-conjugated goat anti mouse antibody (ThermoFisher, Invitrogen) diluted 1:1000 in complete culture medium.
-Nuclei were stained with DAPI and the cells washed.
-The cover slips were mounted on microscope slides using ProLong Gold (ThermoFisher, Invitrogen).
-Confocal laser scan imaging of cells was performed using a Leica TCS SP8 STED 3x microscope equipped with a 100x oil immersion objective (numerical aperture 1.4).
-
 \section{Figure Legends}
@@
 NAD consumption flux and NAD concentration in simulations of organisms with NamPRT and with and without NNMT at different KMs for NamPRT and cell division rates.
+%%% TeX-master: "manuscript"
 
@@
 This is not changed in the mutant missing the insert that corresponds to the unresolved loop structure.
-
-\end{document}

### 2018-06-20 latex/introduction.tex
splitted document in several files to speed up editing

%%% TeX-master: "manuscript"

\section{Introduction}

NAD metabolism represents one of the most critical links that connect cellular signal transduction and energy metabolism.  Even though best known as cofactor for various redox-reactions , NAD is involved in a number of signalling processes that consume NAD+ by cleaving the molecule to nicotinamide (Nam) and ADP-ribose \cite{Verdin2015}. These NAD+-dependent signalling reactions include but are not limited to poly- and mono-ADP-ribosylation, NAD+-dependent protein deacylation by sirtuins as well as the synthesis of calcium mobilizing molecules such as cyclic ADP-ribose \cite{Opitz2015}. These NAD+-dependent signalling processes participate in the regulation of virtually all cellular activities. The enzymes involved in these processes are sensitive to the available NAD+ concentration \cite{Ruggieri2015}, which in turn is dependent on the NAD+/NADH redox ratio. Therefore, NAD+-dependent signalling can act as a transmitter of changes in the cellular energy homeostasis, for example, to regulate gene expression or metabolic activity \cite{Koch-Nolte2009}.

The significance of NAD+-dependent signalling for NAD homeostasis has long been underestimated. However, it has now been established that substances affecting NAD biosynthesis  lead to a rapid decline  of the NAD concentration \cite{Vacor2018}\todo[author=Ines]{Insert bibtex reference} suggesting that NAD+-dependent signalling consumes substantial amounts of NAD, that is why we later refer to them also as NAD-consuming reactions. The resulting NAD turnover differs in a cell type specific manner and can lead to an NAD-halflife as short as 2 hours  \cite{Rabinowitz2018}\todo[author=Ines]. To maintain the NAD concentration at physiological levels, NAD biosynthesis needs to act at an equally rapid rate. Imbalances in NAD-homeostasis have been linked to various, in particular, ageing-related diseases such as diabetes, neurodegenerative disorders and cancer \cite{Chiarugi2012; Verdin2015}\todo[author=Ines]{{Insert bibtex reference, potentially adjust ref}}. Severalrecent studies have demonstrated impressive health benefits of dietary supplementation with intermediates of NAD biosynthesis including NMN and nicotinamide riboside, NR \cite{ImaiCellmetab2018}\todo[author=Ines]{Insert bibtex reference}. Apparently, the exploitation of NAD biosynthetic routes, in addition to the use of nicotinamide as precursor (Fig. 1), results in increased NAD concentrations that stimulate NAD+-dependent signalling processes, in particular, protein deacetylation by sirtuins\cite{?}\todo[author=Mathias B.]{Please add references}.

In mammals, NAD biosynthesis predominantly relies on nicotinamide \cite{?}\todo[author=Mathias Z.]{Please add references}, a form of vitamin B3, as precursor. In the first step, Nam is converted to the mononucleotide, NMN, by Nam phosphoribosyltransferase (NamPRT) using phosphoribosylpyrophosphate (PRPP) as co-substrate. The key role of this enzyme is related to the nature of NAD+-dependent signalling reactions, namely, that in all these reactions Nam is cleaved off and potentially lost. If the released Nam were not recycled into NAD biosynthesis, humans would require a much higher daily vitamin B3 intake than the 15 mg that are the current daily recommendation  \cite{?}(\todo[author=Mathias Z.]{I have not been able to find a reference for that so far.}. The nearly complete recycling of Nam is achieved by an extraordinary high affinity of NamPRT to Nam, the Km being in the low nanomolar range \cite{Burgos2008}. Despite the importance of its salvage, Nam can also be marked for excretion by methylation. Indeed, the presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012} is among the most enigmatic and counterintuitive features of NAD metabolism. Why is there one enzyme (NamPRT) seemingly optimized to recycle even the faintest amounts of Nam back into NAD synthesis, while at the same time there is another one (NNMT) that seems to have no metabolic function other than to remove Nam from NAD metabolism, This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deaminated Nam to nicotinic acid (NA) using the nicotinamide deamidase (NADA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (Fig. 1).

We here present a phylogenetic analys of NAD-pathway in eukaryotes that  in contrast to our previous analysis ,that has been lmited by the low number of eularyotic genomes available at the time, now comprehensively analysis the evolution of the pathway. In the present study, we furthermore combine the phylogenetic analysis with a mathematical modelling approach. This has proven to be  a powerful tool to analyse the relationship between NAD biosynthetic and signalling fluxes when different subsets of Nam-converting enzymes are present. that enable us to explain the counterintuitive pathway evolution, such as that in Deuterostomes, there has been a strong tendency towards the co-existence of NamPRT and NNMT, while the pathway dominant in bacteria is lost in vertebrates. Moreover, this selection for the co-existence of NamPRT and NNMT was accompanied by a marked increase in the number of genes encoding NAD+-consuming signalling reactions.  The model demonstrated that NNMT has a critical role to maintain high NAD+-consuming signalling fluxes by preventing accumulation of inhibitory Nam. We furthermore show that the presence of high affinity NamPRT together with NNMT makes the NADA dependent pathway obsolete. We  furthermore identify a sequence insertion only found in Deuterostomes that encode NNMT but not NADA and experimentally verify that this insertion is indeed changing the affinity of NamPRT.
Taken together, our analyses suggest that the co-existence of NamPRT and NNMT has been a prerequisite to enable the evolutionary development of versatile NAD+-dependent signalling mechanisms present in vertebrates, and that NNMT likely exerted an evolutionary pressure on the NamPRT affinity development.

### 2018-06-20 latex/manuscript.tex
splitted document in several files to speed up editing

@@
 \noindent
-Mathias Bockwoldt$^{1}$, Marc Niere$^{2}$, Toni I. Gossmann$^{3}$, Mathias Ziegler$^{2}$ and Ines Heiland$^{1,\S}$
+Mathias Bockwoldt$^{1}$, Dorothee Houry$^{2}$,  Marc Niere$^{2}$, Toni I. Gossmann$^{3}$, Mathias Ziegler$^{2}$ and Ines Heiland$^{1,\S}$
 
@@
 
+NAD is best known as cofactor in redox reactions, but it is also substrate of NAD-dependent signalling reactions that consume NAD and release nicotinamide (Nam).
+In eukaryotes, two different Nam salvage pathways exist.
+While in lower organisms the initial deamidation of Nam is prevalent, in animals the direct conversion of Nam to the mononucleotide by Nam phosphoribosyltransferase (NamPRT) dominates and eventually remains as the single Nam recycling route in vertebrates.
+Strikingly, loss of the deamidation pathway in early vertebrates is accompanied by the occurrence of a new enzyme that marks Nam for excretion by methylation – nicotinamide N-methyltransferase (NNMT).
+The physiological relevance of NNMT is still enigmatic.
+Why is there an enzyme that removes Nam from recycling, seemingly not having any other physiological role?
+And why is the occurrence of NNMT accompanied by a diversification of NAD-consuming enzymes?
+We have used mathematical modelling approaches to resolve these counterintuitive observations.
+Our results indicate that NNMT is required to enable high NAD consumption fluxes necessitated by the increasing diversification of NAD-dependent signaling pathways.
+This kinetic regulation requires a high substrate affinity of the key enzyme for Nam salvage, NamPRT.
+Indeed, the affinity of NamPRT to Nam has previously been measured to be in the nanomolar range.
+Mathematical modelling supports the hypothesis that NNMT exerted an evolutionary pressure on NamPRT enforcing the development of its unusually high substrate affinity.
+Using multiple sequence alignments, we identified a sequence insertion, first occurring in vertebrates, that parallels an - experimentally verified - increase in the substrate affinity of the enzyme.
+Further simulations show that the deamidation pathway became obsolete owing to the high substrate affinity of NamPRT.
+Collectively, our results illustrate a close evolutionary relationship between NAD biosynthesis and the diversification NAD-dependent signaling pathways, potentially driven by the concomitant occurrence of a regulator of Nam salvage, NNMT.
+
 \section{Keywords}
 
-\section{Highlights}
-
-\section{Introduction}
-
-NAD metabolism represents one of the most critical links that connect cellular signal transduction and energy metabolism.
-According to Kyoto Encyclopedia of Genes and Genomes KEGG (\url{http://www.genome.jp/kegg/}), more than 20\% of all biochemical reactions use NAD(P), mostly in redox reactions in which NAD(P)+ and NAD(P)H are reversibly interconverted.
-In addition to redox reactions, various signalling processes consume NAD+ by cleaving the molecule to nicotinamide (Nam) and ADP-ribose \cite{Verdin2015}.
-NAD+-dependent signalling reactions include poly- and mono-ADP-ribosylation, NAD-dependent protein deacylation by sirtuins as well as the synthesis of calcium mobilizing molecules such as cyclic ADP-ribose \cite{Opitz2015}.
-These NAD+-dependent signalling processes participate in the regulation of virtually all cellular activities.
-The enzymes involved in these processes are sensitive to the available NAD+ concentration \cite{Ruggieri2015}, which in turn is dependent on the NAD+/NADH redox ratio.
-Therefore, NAD+-dependent signalling can act as a transmitter of changes in the cellular energy homeostasis, for example, to regulate gene expression or metabolic activity \cite{Koch-Nolte2009}.
-
-The significance of NAD+-dependent signalling for NAD homeostasis has long been underestimated.
-However, it has now been established that inhibition of NAD biosynthesis in mammalian cells leads to a rapid decline in NAD concentration suggesting that NAD+-dependent signalling consumes substantial amounts of NAD (Ref?), that is why we later refer to them as NAD-consuming reactions.
-As it has been shown that NAD concentrations change in a circadian manner \cite{Nakahata2009; Ramsey2009}, the cellular NAD-pool is turned over at least once per day\todo[author=Mathias]{Either the NAD-pool is turned over at least once a day or the concentration is maintained, but both … could be possible but sounds strange.}.
-To maintain the NAD concentration at physiological levels, NAD biosynthesis needs to proceed at an equally rapid rate.
-Imbalances in NAD-homeostasis have been linked to various, in particular, ageing-related diseases such as diabetes, neurodegenerative disorders and cancer \cite{Chiarugi2012; Verdin2015}\todo[author=Ines]{NAD turnover}.
-Various recent studies have demonstrated impressive health benefits of dietary supplementation with intermediates of NAD biosynthesis including NMN and nicotinamide riboside, NR \cite{Belenky2007}\todo[author=Ines]{That is not very recent.
-Better suggestions?}.
-Apparently, the exploitation of NAD biosynthetic routes, in addition to the use of nicotinamide as precursor (Fig.
-1), results in increased NAD concentrations that stimulate NAD+-dependent signalling processes, in particular, protein deacetylation by sirtuins.
-\todo[author=Ines]{Shorten introduction}
-
-In mammals, NAD biosynthesis predominantly relies on nicotinamide \cite{Bogan2008}\todo[author=Ines]{Not the perfect reference maybe}, a form of vitamin B3, as precursor.
-In the first step, Nam is converted to the mononucleotide, NMN, by Nam phosphoribosyltransferase (NamPRT) using phosphoribosylpyrophosphate (PRPP) as co-substrate.
-The key role of this enzyme is related to the nature of NAD+-dependent signalling reactions, namely, that in all these reactions Nam is cleaved off and potentially lost.
-If the released Nam were not recycled into NAD biosynthesis, humans would require a much higher daily vitamin B3 intake than the 15 mg that are the current daily recommendation (ref) \todo[author=Ines]{I have not been able to find a reference for that so far.}.
-The nearly complete recycling of Nam is achieved by an extraordinary high affinity of NamPRT to Nam, the Km being in the low nanomolar range \cite{Burgos2008}.
-Despite the importance of its salvage, Nam can also be marked for excretion by methylation \todo[author=Mathias]{It’s not too surprising.
-If there was no way of removing parts, we would accumulate more and more, as we still take up 15 mg every day.}.
-Indeed, the presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012} is among the most enigmatic and counterintuitive features of NAD metabolism.
-Why is there one enzyme (NamPRT) seemingly optimized to recycle even the faintest amounts of Nam back into NAD synthesis, while at the same time there is another one (NNMT) that seems to have no metabolic function other than to remove Nam from NAD metabolism\todo[author=Ines]{This paragraphs to the start including a short statement about the purpose of the analysis.}?
-This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deaminated Nam to nicotinic acid (NA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (Fig.
-1).
-Previous phylogenetic analysis show that the two Nam recycling pathways are both ancestral showing a scattered distribution in bacteria \cite{Gazzaniga2009}.
-In contrast, NNMT has so far only been found in animals \cite{Gossmann2012}.
-The analysis of eukaryotic pathway evolution was previously limited by the low number of available genome sequences.
-A comprehensive evolutionary analysis is, however, needed to fully understand the phylogeny of NAD metabolism, in particular, Nam metabolizing enzymes and their relationship to signalling pathways.
-
-This analysis may also help to unravel, how biosynthesis and signalling are coupled and influence each other, as this is not well understood so far.
-It is widely assumed that a major driving force of NAD+-consuming signalling reactions is the cellular NAD concentration.
-Most enzymes using NAD in signalling reactions that have been characterized in detail and it was shown that most sirtuins and PARPs are subject to inhibition by nicotinamide \cite{Borra2004; Ko2012}.
-To promote NAD+-dependent signalling reactions, the balance between elevated NAD levels and accumulating nicotinamide concentrations needs to be kept such that the inhibitory effect of Nam does not override the kinetic stimulation by high NAD levels \todo[author=Mathias]{Well… To promote [signalling] reactions, Nam simply has to be turned into NAD as quickly as possible.
-Besides that it sounds, form the description, as a self-regulating process.}.
-Given the complex nature of the NAD metabolome, it is not trivial to predict the optimal conditions for efficient NAD dependent signalling.
-However, it is obvious that Nam-converting enzymes must play a key role in this regard.
-
-In the present study, we have conducted a comprehensive phylogenetic analysis of the genes that encode Nam-converting enzymes.
-The most surprising result was that with the emergence of Deuterostomes, there has been a strong tendency towards the co-existence of NamPRT and NNMT.
-Moreover, this selection for the co-existence of NamPRT and NNMT was accompanied by a marked increase in the number of genes encoding NAD+-consuming signalling reactions.
-To explain this observation, we built a kinetic model of NAD metabolism based on available kinetic parameters.
-This mathematical model provided a powerful tool to analyse the relationship between NAD biosynthetic and signalling fluxes when different subsets of Nam-converting enzymes are present.
-The model demonstrated that NNMT has a critical role to maintain high NAD+-consuming signalling fluxes by preventing accumulation of inhibitory Nam.
-At lower Nam concentrations, the impact of NNMT is minimal, because of the very high affinity of NamPRT.
-Taken together, our analyses suggest that the co-existence of NamPRT and NNMT has been a prerequisite to enable the evolutionary development of versatile NAD+-dependent signalling mechanisms present in vertebrates.
-
-\section{Results}
-
-\subsection{Phylogenetic analysis of NAD biosynthesis and consumption}
-
-As shown in Fig.
-1, NAD can be synthesized using several routes from three main precursors: tryptophan, nicotinamide (Nam) and nicotinic acid (NA).
-Nam and NA are together known as vitamin B3 or niacin.
-Alternatively, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead \cite{Bogan2008}.
-As in humans only 1\% of the tryptophan taken up with our diet is converted into NAD, vitamin B3 and to a lower extend NR are essential components of our diet, with Nam being the major NAD precursor in humans.
-Nam is also the product of NAD consuming signalling reactions such as sirtuins (NAD-dependent histone deacetylases) or PARPs (poly ADP ribosylases).
-
-For the recycling of Nam, two different pathways exist.
-The pathway found in yeast and plants is using a four-step pathway starting with the deamination of Nam to nicotinic acid by Nam deamidase (NADA).
-The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates.
-The recycling pathway found in mammals directly converts Nam into the corresponding mononucleotide (NMN) a reaction catalysed by NamPRT and driven by a non-stoichiometric ATP-hydrolysis.
-A similar reaction catalysed by an evolutionary related enzyme NAPRT, converts NA into the NA mononucleotide in the Preiss-Handler pathway.
-NMN and NAMN are converted into dinucleotides by the Nam/NA adenylytransferases (NMNATs).
-The recycling pathway via NA finally requires an amination step catalysed by NADsynthase, driven by the conversion of ATP to AMP producing pyrophosphate.
-Even though the latter pathway seems to be very inefficient, it is the pathway preferentially used by most bacteria, fungi and plants (see Figure 2A), whereas most metazoans recycle Nam using NamPRT.
-
-Analysing the phylogeny of the NAD recycling enzymes in Metazoa in more detail reveals that not only does NamPRT replace NADA, but in most organisms, especially in Deuterostomia, NamPRT is found together with the Nam methyltransferase NNMT (Figure 2B).
-NNMT methylates Nam to methyl-Nam that is in mammals excreted with the urine, thus removing Nam from recycling.
-NNMT seems to have arisen de novo in the common ancestor of Ecdysozoa and Lophotrochozoa, as we could not find any gene with considerable similarity in fungi or plants.
-Nematodes are the only organisms where we find NNMT together with NADA without NamPRT being present.
-In Deuterostomia the only large class that does only have NamPRT and seems to have lost NNMT again are Sauropsida and here especially birds.
-The reason why a lot of birds do not encode NNMT remains unclear, as the appearance is quite scattered (not shown)\todo[author=Mathias]{Maybe as supplementary figure?}.
-It might be related to the excretion system, as the product of NNMT methyl-nicotinamide is in mammals excreted with the urine.
-There are some species where we could not find NamPRT or NADA but NNMT, we assume that this is due to incomplete genomes in the database\todo[author=Ines]{Not sure we should discuss Branchisotoma and Tunicata here}.
-
-In addition to the phylogenetic distribution of the two Nam salvage enzymes NADA and NamPRT, we looked at phylogenetic diversity of enzymes catalysing NAD-dependent signalling reactions.
-To do so we used the previously established classification into 10 different families \cite{Gossmann2012} (For details see materials and methods and supplementary information).
-The numbers in Figure 2B denote the average number of NAD-dependent signalling enzyme families we found in each taxa.
-With the exception of Cnidaria and Lophotrochozoa, we find in Protostomia on average 3 to 4 families, whereas most Deuterostomia have on average more than 8 families with an increasing diversification of enzymes within at least some of these families \cite{Gossmann2014}.
-
-Taken together, we found that the presence of NamPRT and NNMT coincides with an increased diversification of NAD-dependent signalling.
-That is surprising as intuitively, one would expect that a decrease in precursor concentration caused by the precursor removal through NNMT, should cause a decrease of NAD availability and consequently less active NAD dependent signalling.
-
-\subsection{Dynamics of NAD biosynthesis and consumption}
-
-So why does the diversity of NAD-dependent signalling increase?
-And why does NADA disappear in Deuterostomia although it is the predominant pathway in bacteria, plants and fungi?
-Given the complexity of the NAD metabolic network, this question is difficult to be comprehensively addressed experimentally.
-Thus, to answer these questions we built a dynamic model of NAD metabolism using existing kinetic data from the literature (details see materials and methods and supplementary material).
-
-To be able to compare metabolic features of evolutionary quite different systems in our simulations, and as we have limited information about expression levels of enzymes or changes of kinetic constants during evolution, we initially used the kinetic constants found for yeast or human enzymes for all systems analysed and used equal amounts of enzymes for all reactions.
-Wherever possible we did not only include substrate affinities but also known product inhibition or inhibition by downstream metabolites.
-As we assume that cell growth is, besides NAD-consuming reactions, a major driving force for NAD biosynthesis, we analysed different growth rates (cell division rates) by simulating different dilution rates for all metabolites.
-We assume furthermore that Nam availability is different for different organisms and thus in addition analysed this simulating different Nam import rates.
-
-The pathway using NADA and recycling Nam via NA is superior \todo[author=Mathias]{What is meant with superior?
-NAD consumption flux is higher or lower with NNMT, depending on the condition.
-Free NAD concentration is higher without NNMT, but is that superior?} both in terms of steady state NAD concentration and NAD consumption flux (representing the activity of NAD-dependent signalling) in the absence of NNMT (Figure 3).
-In the presence of NNMT, the picture is slightly different (Figure 3C-D), even though NADA is still superior to NamPRT under most conditions \todo[author=Ines]{Check if not under all shown}\todo[author=Mathias]{Reply: From fig 4C, it looks like the flux is decreased at very high cell division rates while the Nam import rate is low.}.
-Only if Nam availability is very low, NamPRT is performing better because of its high substrate affinity.
-However, the disadvantage of NamPRT at equal amounts of enzyme, can be compensated by an increased expression of NamPRT.
-Although, to reach similar NAD concentrations in our model, NamPRT expression has to be tenfold higher than the expression of NAPRT (Figure S1).
-The NADA expression required is very low due to its high turnover.
-This might provide an explanation why we find NADA and with that the pathway via NA predominantly in bacteria, yeast, and plants, organisms that show high cell division rates.
-Under these conditions, protein expression costs are assumed to have a high impact on metabolic performance \todo[author=Mathias]{source} and thus pathways where low enzyme expression suffices, might be favourable, even though the pathway via NA is energetically less efficient.
-
-NADA seems to be able to maintain higher NAD concentrations than NamPRT.
-This figure changes if we simulate two organisms or cells that are in direct competition for Nam.
-Under these conditions, the organism expressing NamPRT has a competitive advantage above NADA-expressing organisms, but this only holds if the competing organisms expresses NamPRT together with NNMT.
-(new Figure) This observation might explain, why NADA disappeared in Metazoa together with the appearance of NNMT.
-It might also explain, why many bacteria associated with mammals harbour NamPRT instead of NADA\todo[author=Mathias]{Needs some figure or source}.
-
-Although these simulations already provide an idea of why we find NamPRT predominantly in combination with NNMT, we still do not know why this development coincides with an increased diversification of NAD-consuming enzymes.
-When we simulate the presence or absence of NNMT in the presence of NamPRT, we see that the impact of NNMT on NAD concentration is relatively small (Figure 4D), but we see that that NNMT increases the NAD consumption flux under most conditions in the presence of NamPRT (Figure 4C).
-NAD consumption flux can be further increased by increasing the expression NamPRT, which also compensates the slight reduction in NAD concentration in the presence of NNMT (see Figure S2).
-
-These findings can be explained when looking in more detail into the kinetic parameters of NamPRT and NAD-consuming enzymes such as Sirt1p.
-The ability to maintain high a NAD concentration in the presence of NNMT and at low Nam availability, is due to the very high affinity of NamPRT for its substrate, having a half saturation constant (Km) in the low nanomolar range.
-The increase of NAD consumption flux is caused by the fact that most NAD-consuming enzymes are inhibited by their product Nam, which is also the reason why the presence of NNMT enables higher NAD consumption fluxes.
-
-As the substrate affinity of NamPRT for Nam is extremely high (with a Km in the low nanomolar range) and as this might not have been the case throughout evolution, we analysed the effect of the NamPRT Km on NAD steady state concentration and NAD consumption flux, leaving all other kinetic parameters constant.
-In the absence of NNMT (Figure 5A-B) the Km has very little effect on steady state NAD concentration and NAD consumption flux.
-Without NNMT, NAD concentration and consumption flux are both considerably affected by cell division rates, at least if the enzyme expression is kept constant.
-This is of course an artificial scenario, as one would assume organisms to regulate enzyme expression to achieve similar levels of metabolite concentrations instead.
-In the absence of NNMT there appears furthermore to be a trade-off between achievable steady state NAD concentration and NAD consumption flux.
-
-In the presence of NNMT, NAD consumption flux and NAD concentration increases with decreasing Km values (Figure 5C-D).
-And we note, that both the NAD steady state concentration and the consumption flux are relatively stable over a wide range of cell division rates even at constant expression levels of the involved enzymes, suggesting that NNMT might have an important role to maintain homeostasis of NAD metabolism.
-
-When we compare NAD consumption and NAD concentration with and without NNMT with two different substrate affinities of NamPRT, we see that at a low affinity (Km of 100 nM, which is in the range of the Km of NAPRT for its substrate, or the Km of NADA for Nam), NAD consumption flux is only higher with NNMT at low cell division rates, whereas at high division rates, higher NAD consumption fluxes are achieved without NNMT (Figure 5 E-F).
-This might explain why we do not find NNMT in organism that tend to have high growth rates\todo[author=Ines]{I am not sure if we should go even further suggesting that decreasing substrate affinity by using a competitive inhibitor such as FK866 will influence the consumption rate in fast growing cells much more than the one in slow growing cells, being potentially relevant for cancer therapy.
-– Discussion}.
-We furthermore find, that the competitive advantage of NamPRT of NADA is only present at sufficiently high affinity of NamPRT.
-
-Our analysis also suggests that NNMT might have exerted an evolutionary driving force on the substrate affinity of NamPRT, explaining the extreme values found for the human enzyme\todo[author=Ines]{Move to discussion?}.
-
-The pathway dynamics are of course not solely dependent on one enzyme.
-Thus, what is the impact of the substrate affinity of NNMT that is competing with NamPRT for the same substrate?
-In Figure 6A we see that the substrate affinity values found in the human enzymes (indicated by black asterisks) are actually optimal with respect to both achievable steady state NAD concentration and consumption fluxes.
-Thus, a further increase of the affinity of NamPRT for Nam would not provide any advantage.
-
-\subsection{Sequence variance acquired in metazoans enhances substrate affinity}
-
-To see whether we can find sequence variations in the protein sequence of NamPRT that indicate evolutionary changes of NamPRT upon the appearance of NNMT, we created a multiple sequence alignment of selected eukaryotic NamPRT sequences.
-As shown in Figure 7A and Supplementary Figure 3\todo[author=Ines]{To be created}, Deuterostomia that have only NamPRT and NNMT (indicated by the number 6 in parenthesis \todo[author=Ines]{Maybe rather indicate this by colour and the kingdom through abbrev?}) have an insert of ten amino acids corresponding to positions 43 to 52 of the human enzyme.
-Looking at the crystal structure of human NamPRT this sequence insertion corresponds to a region that has not been structurally resolved in any of the currently available crystal structures (e.g.
-\cite{Wang2006} structure visualisation Figure 7B) and its function is unclear.
-The unresolved loop structure overlaps with a predicted weak nuclear localisation signal (figure 7A), that is not present without the insertion.
-The loop is in addition connected to one of the beta-sheets involved in substrate binding, potentially affecting the affinity or turnover of the enzyme.
-
-The observed evolutionary change in the primary sequence of NamPRT could therefore have had different effects, that we wanted to test experimentally.
-We first investigated whether the deletion of the amino acids 43 to 52 has an effect on the localisation of the human enzyme, expressing a ….
-mutant lacking this sequence in HeLa cells.
-We could, however, not detect any changes in subcellular localisation (Figure 7C) compared to the wildtype fusion construct.
-We therefore conclude that the partial nuclear localisation of NamPRT is not affected by the deletion.
-We then tested the enzymatic activity of the NamPRT mutant recombinantly expressed in E. coli.
-As shown in Figure 7D the enzyme still forms a dimer, thus seems to be folded correctly.
-The enzymatic activity is, however, much lower compared to the wildtype enzyme.
-Using higher concentrations of Nam, it appears that the mutant is not saturated at 100 \textmu M, pointing to a decreased substrate affinity of the mutant enzyme.
+\input{introduction}
+\input{results}
+\input{discussion}
+\input{figurelegends}
 
@@
 
-\section{Discussion}
-
-\subsection{Evolutionary and physiological role of NNMT}
-
-The dynamic interactions among metabolism, signal transduction, and gene regulation is still not very well understood.
-New approaches are required to disentangle the underlying network, helping us to understand how alterations affect human physiology.
-We here combine phylogenetic analysis and mathematical modelling of a central metabolic pathway supported by experimental analysis to reveal the likely evolutionary development of the NAD biosynthesis and consumption pathway and the physiological role of the methyltransferase NNMT.
-
-We show that the vitamin B3-degrading enzyme NNMT plays a vital role for the diversification of NAD-dependent signalling reactions and potentially also for NAD homeostasis.
-Due to the fact that NamPRT has a very high affinity for Nam whereas NNMT has a rather low affinity for it, NNMT is able to remove excess Nam that would lead to inhibition of NAD-dependent signalling reactions while maintaining NAD concentrations and even increasing NAD consumption fluxes.
-
-The affinities measured for the human enzymes seem to be optimal and further increase in NAD consumption rates might only enabled by compartmentalisation of NAD biosynthesis.
-
-\subsection{General applicability}
-
-It is of course impossible to extend our conclusions to other metabolic processes without analysing them in more detail, but we have noted that the metabolism of other vitamins, such as pyridoxal (vitamin B6) metabolism, also contains vitamin degrading enzymes \todo[author=Mathias]{source}, that might thus have similar roles.
-Degrading enzymes do in general not receive the same attention as biosynthetic enzymes, reflected by the observation that NNMT has only recently been analysed in more detail, besides the much earlier recognition of NNMT as potential marker for some types of cancer (e.g.
-Ref?) and routine clinical measurements of urine methyl-Nam concentrations in the context of different diseases (e.g.
-Ref?).
-Our analysis show, however, that degrading enzymes do not solely modify substrates for better excretion, but can play a vital role both in human physiology as well as in the evolutionary development of biological processes.
-
-\subsection{Limitations of our analysis and outlook}
-
-Among others, we neglected the potential effects of the methyl donor S-adenosyl-methionin (SAM) and its precursor methionine in our analysis, although it is most likely contributing yet another regulatory level for Nam availability and thus an additional interaction point between gene regulation and metabolism.
-It has been shown earlier that in turn, NNMT expression influences protein methylation dependent on methionine availability \cite{Ulanovskaya2013}.
-One of the challenges in the analysis of this aspect is the availability of in vivo concentration measurements for SAM and the large amount of reactions using it as substrate.
-The same holds for the analysis of enzyme acetylation versus deacetylation by sirtuins that would enable prediction about enzyme activation or histone state if available.
-
-\section{Experimental Procedure}
-
-\subsection{Dynamic modelling}
-
-Kinetic parameters (substrate affinity (Km) and turnover rates (kcat), substrate and product inhibitions) were retrieved from the enzyme database BRENDA and additionally evaluated by checking the original literature especially with respect to measurement conditions.
-Parameter values from mammals were used if available.
-For enzymes not present in mammals, values from yeast were used.
-The full list of kinetic parameters including reference to original literature can be found in Supplementary table 1.
-For NMNAT, the previously developed rate law for substrate competition was used \cite{Schauble2013}.
-Despite these modifications, Henri-Michaelis-Menten kinetics were used for all reactions except the import and efflux of Nam, which were simulated using constant flux and mass action kinetics, respectively.
-All simulations were performed using the steady state calculation and parameter scan options provided by COPASI 4.22 \cite{Hoops2006}.
-The model will be available at the Biomodels database accession number \todo[inline]{xxx}.
-Related figures were generated using gnuplot\todo[author=Mathias]{Version?}.
-
-\subsection{Phylogenetic Analysis}
-
-NADA, NamPRT, and NNMT enzymes or enzyme candidates were identified using Blastp with known enzymes against the NCBI non-redundant protein sequence database (nr).
-A list of functionally verified enzymes used as templates is given in supplementary table 2.
-This table also includes the length cut-offs for identified enzymes.
-The e-value cut-off was 1e-30 for all enzymes.
-Blastp parameters were set to yield maximum 20,000 target sequences, using the BLOSUM62 matrix with a word size of 6 and gap opening and extension costs of 11 and 1, respectively.
-Low-complexity filtering was disabled.
-Obvious sequence contaminations were removed by manual inspection of the results.
-The taxonomy IDs of the species for each enzyme was derived from the accession2taxonomy database provided by NCBI.
-Scripts for creating, analysing, and visualising the phylogenetic tree were written in Python, using the ETE3 toolkit (Huerta-Cepas, 2010).
-
-\subsection{Generation of eukaryotic expression vectors encoding C-terminally FLAG-tagged NamPT proteins}
-
-The open reading frame (ORF) encoding human NamPRT was inserted into pFLAG-CMV-5a (Merck - Sigma Aldrich) via EcoRI/BamHI sites.
-Using a PCR-based approach, this vector provided the basis for the generation of a plasmid encoding a NamPRT deletion mutant lacking amino acid residues 42-51 (NamPRT-$\Delta$42-51).
-The sequences of the inserted ORFs were verified by DNA sequence analysis.
-
-\subsection{Transient transfection, immunocytochemstry and confocal laser scanning microscopy}
-
-HeLa S3 cells cultivated in Ham’s F12 medium supplemented with 10\% (v/v) FCS, 2 mM L-glutamine, and penicillin/streptomycin, were seeded on cover slips in a 24 well plate.
-After one day, cells were transfected using Effectene transfection reagent (Qiagen) according to the manufacturer’s recommendations.
-Cells were fixed with 4\% paraformaldehyde in PBS 24 hours post transfection, permeabilised (0.5\% (v/v) Triton X-100 in PBS) and blocked for one hour with complete culture medium.
-After overnight incubation with primary FLAG-antibody (mouse M2, Sigma-Aldrich) diluted 1:2500 in complete medium, cells were washed and incubated for one hour with secondary AlexaFluor 594-conjugated goat anti mouse antibody (ThermoFisher, Invitrogen) diluted 1:1000 in complete culture medium.
-Nuclei were stained with DAPI and the cells washed.
-The cover slips were mounted on microscope slides using ProLong Gold (ThermoFisher, Invitrogen).
-Confocal laser scan imaging of cells was performed using a Leica TCS SP8 STED 3x microscope equipped with a 100x oil immersion objective (numerical aperture 1.4).
-
-\section{Figure Legends}
-
-\subsection{Figure1}
-
-Schematic overview of NAD biosynthesis and consumption.
-NAD can be synthesized using several routes from altogether three main precursors: tryptophan (Trp), nicotinamide (Nam), and nicotinic acid (NA).
-Nam and NA are together known as vitamin B3 or niacin.
-To a lesser extend, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead.
-Only 1\% of the tryptophan taken up with our diet is converted into NAD, thus, vitamin B3 and, to a lower extend, NR are essential components of our diet, with Nam being the major NAD precursor in humans\todo[author=Mathias]{Whole sentence not necessarily part of the legend}.
-Nam is furthermore the product of NAD-consuming signalling reactions such as sirtuins (NAD-dependent histone deacetylases) or PARPs (poly-ADP-ribosylases).
-
-For the recycling of Nam, two different pathways exist.
-The pathway found in yeast and many bacteria is using a four-step pathway starting with the deamination of Nam to nicotinic acid (NA) by Nam deamidase (NADA).
-The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates.
-The recycling pathway found in vertebrates directly converts Nam into the corresponding mononucleotide (NMN) a reaction catalysed by NamPRT and driven by a non-stoichiometric ATP-hydrolysis.
-A similar reaction catalysed by an evolutionary related enzyme NAPRT, converts NA into the NA mononucleotide (NAMN).
-NMN and NAMN are converted into dinucleotides by the Nam/NA adenylytransferases (NMNATs).
-The recycling pathway via NA finally requires an amination step catalysed by NADsynthase (NADS), driven by the conversion of ATP to AMP producing pyrophosphate.
-
-\subsection{Figure 2}
-
-Evolutionary distribution of NADA, NNMT and NamPRT and their relation to the number of NAD consumers.
-A) Distribution of NADA, NNMT and NamPRT in selected major taxa.
-NADA is dominant in Bacteria, Fungi, and Plants (Viridiplantae), whereas NamPRT together with NNMT is dominant in Deuterostomia.
-Numbers at the pie charts show, how many species of the taxon possess the respective enzyme combination indicated by the colour explained in the lower right of the figure.
-Below the taxon name, the number of species in that taxon is given.
-
-B) Common tree of selected taxa within the Metazoa, including 334 species.
-The pie charts indicate the distribution of species within the respective taxon that have the enzyme combination indicated by the colour, explained in the lower right.
-The size of the pie charts is proportional to the logarithm of the number of species analysed in the particular taxon.
-The numbers below the taxon names indicate the average number of NAD-consuming enzyme families found in all sub-taxa.
-The branch length is arbitrary.
-
-\subsection{Figure 3}
-
-NAD consumption flux and NAD concentration in simulations of organisms with NADA(?) and with and without NNMT at different Nam import and cell division rates.
-
-\subsection{Figure 4}
-
-Impact of NNMT on NAD consumption and NAD concentration in simulations of organisms with NADA or NamPRT.
-
-\subsection{Figure 5}
-
-NAD consumption flux and NAD concentration in simulations of organisms with NamPRT and with and without NNMT at different KMs for NamPRT and cell division rates.
-
-\subsection{Figure 6}
-
-The substrate affinity of NNMT and NamPRT have opposite effects on NAD consumption (A) and concentration (B), as would be expected.
-The affinities previously measured for human enzymes (indicated by a black asterisk) appear to be close to optimal, as further improvements would have little or no effect on NAD consumption or concentration.
-
-\subsection{Figure 7}
-
-Deuterostomia that encode NNMT show a sequence insertion in the N-terminal region of NamPRT.
-A) Multiple sequence alignment of NamPRT of selected deuterostomes show a sequence insertion in organisms that encode NamPRT and NNMT.
-(Species names are given and the number in parenthesis indicates the enzyme combination encoded in these species; 3 – NADA and NamPRT, 6 – NamPRT and NNMT, 7 – NADA, NamPRT and NNMT.
-\todo[author=Mathias]{Note to myself: replace the numbers by something better and rewrite this part.}) The nuclear localisation signal (NLS\todo[author=Mathias]{Note to myself: Write “predicted NLS” in figure}) indicated, was predicted using using cNLS-mapper (\url{http://nls-mapper.iab.keio.ac.jp/cgi-bin/NLS_Mapper_form.cgi} \cite{Kosugi2009}).
-
-B) The inserted region is not resolved in crystal structures of human NamPRT and thus appears to be a flexible loop structure at the surface of the NamPRT dimer, coloured in red.
-The visualisation is based on a structure prediction of SWISS-MODEL \cite{Arnold2006; Biasini2014} of the sequence of the human NamPRT (P43490) using the model 2H3D as template \cite{Wang2006}.
-
-C) The localisation of wildtype NamPRT varies between pure cytosolic and mixed cytosolic-nuclear localisation.
-This is not changed in the mutant missing the insert that corresponds to the unresolved loop structure.
-
 \end{document}

### 2018-06-20 latex/results.tex
splitted document in several files to speed up editing

\section{Results}

\subsection{Phylogenetic analysis of NAD biosynthesis and consumption}

As shown in Fig. 1, NAD can be synthesized using several routes from three main precursors: tryptophan, nicotinamide (Nam) and nicotinic acid (NA). Nam and NA are together known as vitamin B3 or niacin. Alternatively, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead \cite{Imai2018}. Du to the hight turnover of NAD observed and the fact that only 1\% of the tryptophan taken up with our diet is converted into NAD, vitamin B3 (Nam and NA) and to a lower extend NR are the major precursor of NAD-biosynthesis.

For the recycling of Nam, two different pathways exist. The pathway found in yeast and plants is using a four-step pathway starting with the deamination of Nam to nicotinic acid by NADA. The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates. The recycling pathway found in mammals directly converts Nam into the corresponding mononucleotide (NMN) a reaction catalysed by NamPRT and driven by a non-stoichiometric ATP-hydrolysis. A similar reaction catalysed by an evolutionary related enzyme NAPRT, converts NA into the NA mononucleotide in the Preiss-Handler pathway. NMN and NAMN are converted into dinucleotides by the Nam/NA adenylytransferases (NMNATs). The recycling pathway via NA finally requires an amination step catalysed by NADsynthase, driven by the conversion of ATP to AMP producing pyrophosphate. Even though the latter pathway seems to be very inefficient, it is the pathway preferentially used by most bacteria, fungi and plants (see Figure 2A), whereas most metazoans recycle Nam using NamPRT.

Analysing the phylogeny of the NAD recycling enzymes in Metazoa in more detail reveals that not only does NamPRT replace NADA, but in most organisms, especially in Deuterostomia, NamPRT is found together with the Nam methyltransferase NNMT (Figure 2B). NNMT methylates Nam to methyl-Nam that is in mammals excreted with the urine, thus removing Nam from recycling. NNMT seems to have arisen de novo in the common ancestor of Ecdysozoa and Lophotrochozo.  We were not be able to find any gene with considerable similarity in fungi or plants, even though in fungi a gene named NNMT can be found in databases. These genes do, however, show very limited sequence similarities to the NNMT of nematodes or deuterostomes and in the yeast protein has later on be found to be a lysin-protein-methyltransferase  \cite{?}(\todo[author=Mathias B.]{Please look up the reference for this.}. Nematodes are the only organisms where we find NNMT together with NADA without NamPRT being present. In Deuterostomia the only large class that does only have NamPRT and seems to have lost NNMT again are Sauropsida and here especially birds. The reason why a lot of birds do not encode NNMT remains unclear, as the appearance is quite scattered (not shown)\todo[author=Mathias]{Maybe as supplementary figure?}. It might be related to the excretion system, as the product of NNMT methyl-nicotinamide is in mammals excreted with the urine. There are some species where we could not find NamPRT or NADA but NNMT, we assume that this is due to incomplete genomes in the database.

In addition to the phylogenetic distribution of the two Nam salvage enzymes NADA and NamPRT, we looked at phylogenetic diversity of enzymes catalysing NAD-dependent signalling reactions. To do so we used the previously established classification into 10 different families \cite{Gossmann2012} (For details see materials and methods and supplementary information). The numbers in Figure 2B denote the average number of NAD-dependent signalling enzyme families found in each taxa. With the exception of Cnidaria and Lophotrochozoa, we find average 3 to 4 families in Protostomia, whereas most Deuterostomia have on average more than 8 families with an increasing diversification of enzymes within some of these families \cite{Gossmann2014}.

Taken together, we found that the presence of NamPRT and NNMT coincides with an increased diversification of NAD-dependent signalling, whilst NADA is lost in vertebrates. That is  counterintuitive, as one would expect that a decrease in precursor concentration caused by the precursor removal through NNMT, should cause a decrease of NAD availability and consequently less active NAD dependent signalling.

\subsection{Dynamics of NAD biosynthesis and consumption}

So why does the diversity of NAD-dependent signalling increase? And why does NADA disappear in Deuterostomia although it is the predominant pathway in bacteria, plants and fungi? Given the complexity of the NAD metabolic network, this question is difficult to be comprehensively addressed experimentally. Thus, to answer these questions we built a dynamic model of NAD metabolism using existing kinetic data from the literature (details see materials and methods and supplementary material).

To be able to compare metabolic features of evolutionary quite different systems in our simulations, and as we have limited information about expression levels of enzymes or changes of kinetic constants during evolution, we initially used the kinetic constants found for yeast or human enzymes for all systems analysed and used equal amounts of enzymes for all reactions. Wherever possible we did not only include substrate affinities but also known product inhibition or inhibition by downstream metabolites. As we assume that cell growth is, besides NAD-consuming reactions, a major driving force for NAD biosynthesis, we analysed different growth rates (cell division rates) by simulating different dilution rates for all metabolites. We assume furthermore that Nam availability is different for different organisms and thus in addition analysed this simulating different Nam import rates.

The pathway using NADA and recycling Nam via NA is superior \todo[author=Mathias]{What is meant with superior? NAD consumption flux is higher or lower with NNMT, depending on the condition. Free NAD concentration is higher without NNMT, but is that superior?} both in terms of steady state NAD concentration and NAD consumption flux (representing the activity of NAD-dependent signalling) in the absence of NNMT (Figure 3). In the presence of NNMT, the picture is slightly different (Figure 3C-D), even though NADA is still superior to NamPRT under most conditions \todo[author=Ines]{Check if not under all shown}\todo[author=Mathias]{Reply: From fig 4C, it looks like the flux is decreased at very high cell division rates while the Nam import rate is low.}. Only if Nam availability is very low, NamPRT is performing better because of its high substrate affinity. However, the disadvantage of NamPRT at equal amounts of enzyme, can be compensated by an increased expression of NamPRT. Although, to reach similar NAD concentrations in our model, NamPRT expression has to be tenfold higher than the expression of NAPRT (Figure S1). The NADA expression required is very low due to its high turnover. This might provide an explanation why we find NADA and with that the pathway via NA predominantly in bacteria, yeast, and plants, organisms that show high cell division rates. Under these conditions, protein expression costs are assumed to have a high impact on metabolic performance \todo[author=Mathias]{source} and thus pathways where low enzyme expression suffices, might be favourable, even though the pathway via NA is energetically less efficient.

NADA seems to be able to maintain higher NAD concentrations than NamPRT. This figure changes if we simulate two organisms or cells that are in direct competition for Nam. Under these conditions, the organism expressing NamPRT has a competitive advantage above NADA-expressing organisms, but this only holds if the competing organisms expresses NamPRT together with NNMT. (new Figure) This observation might explain, why NADA disappeared in Metazoa together with the appearance of NNMT. It might also explain, why many bacteria associated with mammals harbour NamPRT instead of NADA\todo[author=Mathias]{Needs some figure or source}.

Although these simulations already provide an idea of why we find NamPRT predominantly in combination with NNMT, we still do not know why this development coincides with an increased diversification of NAD-consuming enzymes. When we simulate the presence or absence of NNMT in the presence of NamPRT, we see that the impact of NNMT on NAD concentration is relatively small (Figure 4D), but we see that that NNMT increases the NAD consumption flux under most conditions in the presence of NamPRT (Figure 4C). NAD consumption flux can be further increased by increasing the expression NamPRT, which also compensates the slight reduction in NAD concentration in the presence of NNMT (see Figure S2).

These findings can be explained when looking in more detail into the kinetic parameters of NamPRT and NAD-consuming enzymes such as Sirt1p. The ability to maintain high a NAD concentration in the presence of NNMT and at low Nam availability, is due to the very high affinity of NamPRT for its substrate, having a half saturation constant (Km) in the low nanomolar range. The increase of NAD consumption flux is caused by the fact that most NAD-consuming enzymes are inhibited by their product Nam, which is also the reason why the presence of NNMT enables higher NAD consumption fluxes.

As the substrate affinity of NamPRT for Nam is extremely high (with a Km in the low nanomolar range) and as this might not have been the case throughout evolution, we analysed the effect of the NamPRT Km on NAD steady state concentration and NAD consumption flux, leaving all other kinetic parameters constant. In the absence of NNMT (Figure 5A-B) the Km has very little effect on steady state NAD concentration and NAD consumption flux. Without NNMT, NAD concentration and consumption flux are both considerably affected by cell division rates, at least if the enzyme expression is kept constant. This is of course an artificial scenario, as one would assume organisms to regulate enzyme expression to achieve similar levels of metabolite concentrations instead. In the absence of NNMT there appears furthermore to be a trade-off between achievable steady state NAD concentration and NAD consumption flux.

In the presence of NNMT, NAD consumption flux and NAD concentration increases with decreasing Km values (Figure 5C-D). And we note, that both the NAD steady state concentration and the consumption flux are relatively stable over a wide range of cell division rates even at constant expression levels of the involved enzymes, suggesting that NNMT might have an important role to maintain homeostasis of NAD metabolism.

When we compare NAD consumption and NAD concentration with and without NNMT with two different substrate affinities of NamPRT, we see that at a low affinity (Km of 100 nM, which is in the range of the Km of NAPRT for its substrate, or the Km of NADA for Nam), NAD consumption flux is only higher with NNMT at low cell division rates, whereas at high division rates, higher NAD consumption fluxes are achieved without NNMT (Figure 5 E-F). This might explain why we do not find NNMT in organism that tend to have high growth rates\todo[author=Ines]{I am not sure if we should go even further suggesting that decreasing substrate affinity by using a competitive inhibitor such as FK866 will influence the consumption rate in fast growing cells much more than the one in slow growing cells, being potentially relevant for cancer therapy. – Discussion}. We furthermore find, that the competitive advantage of NamPRT of NADA is only present at sufficiently high affinity of NamPRT.

Our analysis also suggests that NNMT might have exerted an evolutionary driving force on the substrate affinity of NamPRT, explaining the extreme values found for the human enzyme\todo[author=Ines]{Move to discussion?}.

The pathway dynamics are of course not solely dependent on one enzyme. Thus, what is the impact of the substrate affinity of NNMT that is competing with NamPRT for the same substrate? In Figure 6A we see that the substrate affinity values found in the human enzymes (indicated by black asterisks) are actually optimal with respect to both achievable steady state NAD concentration and consumption fluxes. Thus, a further increase of the affinity of NamPRT for Nam would not provide any advantage.

\subsection{Sequence variance acquired in metazoans enhances substrate affinity}

To see whether we can find sequence variations in the protein sequence of NamPRT that indicate evolutionary changes of NamPRT upon the appearance of NNMT, we created a multiple sequence alignment of selected eukaryotic NamPRT sequences. As shown in Figure 7A and Supplementary Figure 3\todo[author=Ines]{To be created}, Deuterostomia that have only NamPRT and NNMT (indicated by the number 6 in parenthesis \todo[author=Ines]{Maybe rather indicate this by colour and the kingdom through abbrev?}) have an insert of ten amino acids corresponding to positions 43 to 52 of the human enzyme. Looking at the crystal structure of human NamPRT this sequence insertion corresponds to a region that has not been structurally resolved in any of the currently available crystal structures (e.g. \cite{Wang2006} structure visualisation Figure 7B) and its function is unclear. The unresolved loop structure overlaps with a predicted weak nuclear localisation signal (figure 7A), that is not present without the insertion. The loop is in addition connected to one of the beta-sheets involved in substrate binding, potentially affecting the affinity or turnover of the enzyme.

The observed evolutionary change in the primary sequence of NamPRT could therefore have had different effects, that we wanted to test experimentally. We first investigated whether the deletion of the amino acids 43 to 52 has an effect on the localisation of the human enzyme, expressing a …. mutant lacking this sequence in HeLa cells. We could, however, not detect any changes in subcellular localisation (Figure 7C) compared to the wildtype fusion construct. We therefore conclude that the partial nuclear localisation of NamPRT is not affected by the deletion. We then tested the enzymatic activity of the NamPRT mutant recombinantly expressed in E. coli. As shown in Figure 7D the enzyme still forms a dimer, thus seems to be folded correctly. The enzymatic activity is, however, much lower compared to the wildtype enzyme. Using higher concentrations of Nam, it appears that the mutant is not saturated at 100 \textmu M, pointing to a decreased substrate affinity of the mutant enzyme.

### 2018-06-20 latex/introduction.tex
Added bibtex library

@@
 However, it has now been established that substances affecting NAD biosynthesis  lead to a rapid decline  of the NAD concentration \cite{Vacor2018}\todo[author=Ines]{Insert bibtex reference} suggesting that NAD+-dependent signalling consumes substantial amounts of NAD, that is why we later refer to them also as NAD-consuming reactions.
-The resulting NAD turnover differs in a cell type specific manner and can lead to an NAD-halflife as short as 2 hours  \cite{Rabinowitz2018}\todo[author=Ines].
+The resulting NAD turnover differs in a cell type specific manner and can lead to an NAD-halflife as short as 2 hours  \cite{Rabinowitz2018}\todo[author=Ines]{Insert bibtex reference}.
 To maintain the NAD concentration at physiological levels, NAD biosynthesis needs to act at an equally rapid rate.
-Imbalances in NAD-homeostasis have been linked to various, in particular, ageing-related diseases such as diabetes, neurodegenerative disorders and cancer \cite{Chiarugi2012; Verdin2015}\todo[author=Ines]{{Insert bibtex reference, potentially adjust ref}}.
+Imbalances in NAD-homeostasis have been linked to various, in particular, ageing-related diseases such as diabetes, neurodegenerative disorders and cancer \cite{Chiarugi2012,Verdin2015}\todo[author=Ines]{Potentially adjust ref}.
 Severalrecent studies have demonstrated impressive health benefits of dietary supplementation with intermediates of NAD biosynthesis including NMN and nicotinamide riboside, NR \cite{ImaiCellmetab2018}\todo[author=Ines]{Insert bibtex reference}.
 Apparently, the exploitation of NAD biosynthetic routes, in addition to the use of nicotinamide as precursor (Fig.
-1), results in increased NAD concentrations that stimulate NAD+-dependent signalling processes, in particular, protein deacetylation by sirtuins\cite{?}\todo[author=Mathias B.]{Please add references}.
+1), results in increased NAD concentrations that stimulate NAD+-dependent signalling processes, in particular, protein deacetylation by sirtuins \todo[author=Mathias B.]{Please add references}.
 
-In mammals, NAD biosynthesis predominantly relies on nicotinamide \cite{?}\todo[author=Mathias Z.]{Please add references}, a form of vitamin B3, as precursor.
+In mammals, NAD biosynthesis predominantly relies on nicotinamide \todo[author=Mathias Z.]{Please add references}, a form of vitamin B3, as precursor.
 In the first step, Nam is converted to the mononucleotide, NMN, by Nam phosphoribosyltransferase (NamPRT) using phosphoribosylpyrophosphate (PRPP) as co-substrate.
 The key role of this enzyme is related to the nature of NAD+-dependent signalling reactions, namely, that in all these reactions Nam is cleaved off and potentially lost.
-If the released Nam were not recycled into NAD biosynthesis, humans would require a much higher daily vitamin B3 intake than the 15 mg that are the current daily recommendation  \cite{?}(\todo[author=Mathias Z.]{I have not been able to find a reference for that so far.}.
+If the released Nam were not recycled into NAD biosynthesis, humans would require a much higher daily vitamin B3 intake than the 15 mg that are the current daily recommendation \todo[author=Mathias Z.]{I have not been able to find a reference for that so far.}.
 The nearly complete recycling of Nam is achieved by an extraordinary high affinity of NamPRT to Nam, the Km being in the low nanomolar range \cite{Burgos2008}.

### 2018-06-20 latex/results.tex
Added bibtex library

@@
+%%% TeX-master: "manuscript"
+
 \section{Results}
@@
 Nam and NA are together known as vitamin B3 or niacin.
-Alternatively, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead \cite{Imai2018}.
+Alternatively, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead \cite{Imai2018}\todo[author=Mathias]{Insert bibtex reference}.
 Du to the hight turnover of NAD observed and the fact that only 1\% of the tryptophan taken up with our diet is converted into NAD, vitamin B3 (Nam and NA) and to a lower extend NR are the major precursor of NAD-biosynthesis.
@@
 We were not be able to find any gene with considerable similarity in fungi or plants, even though in fungi a gene named NNMT can be found in databases.
-These genes do, however, show very limited sequence similarities to the NNMT of nematodes or deuterostomes and in the yeast protein has later on be found to be a lysin-protein-methyltransferase  \cite{?}(\todo[author=Mathias B.]{Please look up the reference for this.}.
+These genes do, however, show very limited sequence similarities to the NNMT of nematodes or deuterostomes and in the yeast protein has later on be found to be a lysin-protein-methyltransferase (\todo[author=Mathias B.]{Please look up the reference for this.}.
 Nematodes are the only organisms where we find NNMT together with NADA without NamPRT being present.

### 2018-06-20 latex/introduction.tex
Added paragraph about yeast NNMT and minor formatting

@@
 The significance of NAD+-dependent signalling for NAD homeostasis has long been underestimated.
-However, it has now been established that substances affecting NAD biosynthesis  lead to a rapid decline  of the NAD concentration \cite{Vacor2018}\todo[author=Ines]{Insert bibtex reference} suggesting that NAD+-dependent signalling consumes substantial amounts of NAD, that is why we later refer to them also as NAD-consuming reactions.
+However, it has now been established that substances affecting NAD biosynthesis lead to a rapid decline of the NAD concentration \cite{Vacor2018}\todo[author=Ines]{Insert bibtex reference} suggesting that NAD+-dependent signalling consumes substantial amounts of NAD, which is why we later refer to them also as NAD-consuming reactions.
 The resulting NAD turnover differs in a cell type specific manner and can lead to an NAD-halflife as short as 2 hours \cite{Rabinowitz2018}\todo[author=Ines]{Insert bibtex reference}.
@@
 Indeed, the presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012} is among the most enigmatic and counterintuitive features of NAD metabolism.
-Why is there one enzyme (NamPRT) seemingly optimized to recycle even the faintest amounts of Nam back into NAD synthesis, while at the same time there is another one (NNMT) that seems to have no metabolic function other than to remove Nam from NAD metabolism, This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deaminated Nam to nicotinic acid (NA) using the nicotinamide deamidase (NADA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (Fig.
+Why is there one enzyme (NamPRT) seemingly optimized to recycle even the faintest amounts of Nam back into NAD synthesis, while at the same time there is another enzyme (NNMT) that seems to have no metabolic function other than to remove Nam from NAD metabolism, This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deaminated Nam to nicotinic acid (NA) using the nicotinamide deamidase (NADA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (Fig.
 1).
 
-We here present a phylogenetic analys of NAD-pathway in eukaryotes that  in contrast to our previous analysis ,that has been lmited by the low number of eularyotic genomes available at the time, now comprehensively analysis the evolution of the pathway.
+We here present a phylogenetic analys of NAD-pathway in eukaryotes that in contrast to our previous analysis that has been lmited by the low number of eularyotic genomes available at the time, now comprehensively analysis the evolution of the pathway.
 In the present study, we furthermore combine the phylogenetic analysis with a mathematical modelling approach.
@@
 The model demonstrated that NNMT has a critical role to maintain high NAD+-consuming signalling fluxes by preventing accumulation of inhibitory Nam.
-We furthermore show that the presence of high affinity NamPRT together with NNMT makes the NADA dependent pathway obsolete.
+We show that the presence of high affinity NamPRT together with NNMT makes the NADA dependent pathway obsolete.
 We furthermore identify a sequence insertion only found in Deuterostomes that encode NNMT but not NADA and experimentally verify that this insertion is indeed changing the affinity of NamPRT.
+
 Taken together, our analyses suggest that the co-existence of NamPRT and NNMT has been a prerequisite to enable the evolutionary development of versatile NAD+-dependent signalling mechanisms present in vertebrates, and that NNMT likely exerted an evolutionary pressure on the NamPRT affinity development.

### 2018-06-20 latex/introduction.tex
Added some more references

@@
 Apparently, the exploitation of NAD biosynthetic routes, in addition to the use of nicotinamide as precursor (Fig.
-1), results in increased NAD concentrations that stimulate NAD+-dependent signalling processes, in particular, protein deacetylation by sirtuins \todo[author=Mathias B.]{Please add references}.
+1), results in increased NAD concentrations that stimulate NAD+-dependent signalling processes, in particular, protein deacetylation by sirtuins \cite{North2004}.
 
@@
 The key role of this enzyme is related to the nature of NAD+-dependent signalling reactions, namely, that in all these reactions Nam is cleaved off and potentially lost.
-If the released Nam were not recycled into NAD biosynthesis, humans would require a much higher daily vitamin B3 intake than the 15 mg that are the current daily recommendation \todo[author=Mathias Z.]{I have not been able to find a reference for that so far.}.
+If the released Nam were not recycled into NAD biosynthesis, humans would require a much higher daily vitamin B3 intake than the 16 mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}.
 The nearly complete recycling of Nam is achieved by an extraordinary high affinity of NamPRT to Nam, the Km being in the low nanomolar range \cite{Burgos2008}.

### 2018-06-20 latex/results.tex
Added some more references

@@
 We were not be able to find any gene with considerable similarity in fungi or plants, even though in fungi a gene named NNMT can be found in databases.
-These genes do, however, show very limited sequence similarities to the NNMT of nematodes or deuterostomes and in the yeast protein has later on be found to be a lysin-protein-methyltransferase (\todo[author=Mathias B.]{Please look up the reference for this.}.
+These genes do, however, show very limited sequence similarities to the NNMT of nematodes or deuterostomes and in the yeast protein has later on be found to be a lysin-protein-methyltransferase \cite{Wlodarski2011}.
 Nematodes are the only organisms where we find NNMT together with NADA without NamPRT being present.

### 2018-06-20 latex/introduction.tex
Updated some references

@@
 The significance of NAD+-dependent signalling for NAD homeostasis has long been underestimated.
-However, it has now been established that substances affecting NAD biosynthesis lead to a rapid decline of the NAD concentration \cite{Vacor2018}\todo[author=Ines]{Insert bibtex reference} suggesting that NAD+-dependent signalling consumes substantial amounts of NAD, which is why we later refer to them also as NAD-consuming reactions.
-The resulting NAD turnover differs in a cell type specific manner and can lead to an NAD-halflife as short as 2 hours \cite{Rabinowitz2018}\todo[author=Ines]{Insert bibtex reference}.
+However, it has now been established that substances affecting NAD biosynthesis lead to a rapid decline of the NAD concentration \cite{Buonvicino2018} suggesting that NAD+-dependent signalling consumes substantial amounts of NAD, which is why we later refer to them also as NAD-consuming reactions.
+The resulting NAD turnover differs in a cell type specific manner and can lead to an NAD-halflife as short as 2 hours \cite{Liu2018}.
 To maintain the NAD concentration at physiological levels, NAD biosynthesis needs to act at an equally rapid rate.
 Imbalances in NAD-homeostasis have been linked to various, in particular, ageing-related diseases such as diabetes, neurodegenerative disorders and cancer \cite{Chiarugi2012,Verdin2015}\todo[author=Ines]{Potentially adjust ref}.
-Severalrecent studies have demonstrated impressive health benefits of dietary supplementation with intermediates of NAD biosynthesis including NMN and nicotinamide riboside, NR \cite{ImaiCellmetab2018}\todo[author=Ines]{Insert bibtex reference}.
+Several recent studies have demonstrated impressive health benefits of dietary supplementation with intermediates of NAD biosynthesis including NMN and nicotinamide riboside, NR \cite{Yoshino2018}.
 Apparently, the exploitation of NAD biosynthetic routes, in addition to the use of nicotinamide as precursor (Fig.

### 2018-06-20 latex/results.tex
Updated some references

@@
 Nam and NA are together known as vitamin B3 or niacin.
-Alternatively, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead \cite{Imai2018}\todo[author=Mathias]{Insert bibtex reference}.
+Alternatively, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead \cite{Yoshino2018}.
 Du to the hight turnover of NAD observed and the fact that only 1\% of the tryptophan taken up with our diet is converted into NAD, vitamin B3 (Nam and NA) and to a lower extend NR are the major precursor of NAD-biosynthesis.

### 2018-06-21 latex/results.tex
Edited results

@@
 Nam and NA are together known as vitamin B3 or niacin.
-Alternatively, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead \cite{Yoshino2018}.
+Alternatively, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead \cite{Imai2018}.
 Du to the hight turnover of NAD observed and the fact that only 1\% of the tryptophan taken up with our diet is converted into NAD, vitamin B3 (Nam and NA) and to a lower extend NR are the major precursor of NAD-biosynthesis.
@@
 We were not be able to find any gene with considerable similarity in fungi or plants, even though in fungi a gene named NNMT can be found in databases.
-These genes do, however, show very limited sequence similarities to the NNMT of nematodes or deuterostomes and in the yeast protein has later on be found to be a lysin-protein-methyltransferase \cite{Wlodarski2011}.
+These genes do, however, show very limited sequence similarities to the NNMT of nematodes or deuterostomes and in the yeast protein has later on be found to be a lysin-protein-methyltransferase  \cite{?}(\todo[author=Mathias B.]{Please look up the reference for this.}.
 Nematodes are the only organisms where we find NNMT together with NADA without NamPRT being present.
@@
 Given the complexity of the NAD metabolic network, this question is difficult to be comprehensively addressed experimentally.
-Thus, to answer these questions we built a dynamic model of NAD metabolism using existing kinetic data from the literature (details see materials and methods and supplementary material).
+Thus, to answer these questions we built a dynamic model of NAD metabolism using existing kinetic data from the literature (for details see materials and methods and supplementary material).
 
@@
 As we assume that cell growth is, besides NAD-consuming reactions, a major driving force for NAD biosynthesis, we analysed different growth rates (cell division rates) by simulating different dilution rates for all metabolites.
-We assume furthermore that Nam availability is different for different organisms and thus in addition analysed this simulating different Nam import rates.
 
-The pathway using NADA and recycling Nam via NA is superior \todo[author=Mathias]{What is meant with superior?
-NAD consumption flux is higher or lower with NNMT, depending on the condition.
-Free NAD concentration is higher without NNMT, but is that superior?} both in terms of steady state NAD concentration and NAD consumption flux (representing the activity of NAD-dependent signalling) in the absence of NNMT (Figure 3).
-In the presence of NNMT, the picture is slightly different (Figure 3C-D), even though NADA is still superior to NamPRT under most conditions \todo[author=Ines]{Check if not under all shown}\todo[author=Mathias]{Reply: From fig 4C, it looks like the flux is decreased at very high cell division rates while the Nam import rate is low.}.
-Only if Nam availability is very low, NamPRT is performing better because of its high substrate affinity.
-However, the disadvantage of NamPRT at equal amounts of enzyme, can be compensated by an increased expression of NamPRT.
-Although, to reach similar NAD concentrations in our model, NamPRT expression has to be tenfold higher than the expression of NAPRT (Figure S1).
-The NADA expression required is very low due to its high turnover.
-This might provide an explanation why we find NADA and with that the pathway via NA predominantly in bacteria, yeast, and plants, organisms that show high cell division rates.
-Under these conditions, protein expression costs are assumed to have a high impact on metabolic performance \todo[author=Mathias]{source} and thus pathways where low enzyme expression suffices, might be favourable, even though the pathway via NA is energetically less efficient.
-
-NADA seems to be able to maintain higher NAD concentrations than NamPRT.
-This figure changes if we simulate two organisms or cells that are in direct competition for Nam.
-Under these conditions, the organism expressing NamPRT has a competitive advantage above NADA-expressing organisms, but this only holds if the competing organisms expresses NamPRT together with NNMT.
-(new Figure) This observation might explain, why NADA disappeared in Metazoa together with the appearance of NNMT.
-It might also explain, why many bacteria associated with mammals harbour NamPRT instead of NADA\todo[author=Mathias]{Needs some figure or source}.
-
-Although these simulations already provide an idea of why we find NamPRT predominantly in combination with NNMT, we still do not know why this development coincides with an increased diversification of NAD-consuming enzymes.
-When we simulate the presence or absence of NNMT in the presence of NamPRT, we see that the impact of NNMT on NAD concentration is relatively small (Figure 4D), but we see that that NNMT increases the NAD consumption flux under most conditions in the presence of NamPRT (Figure 4C).
-NAD consumption flux can be further increased by increasing the expression NamPRT, which also compensates the slight reduction in NAD concentration in the presence of NNMT (see Figure S2).
+First we addressed the question of the predominant coexitence of NamPRT and NNMT in vertebrates that coincides with an increase in the number of NAD-consuming enzyme families.
+We thus used our dynamic model of the NAD biosynthesis and consmption pathway to simulate the effect of the  presence or absence of NNMT.
+We see that the presence of NNMT enables higher NAD consumption rates  (Figure3A) while, as exptected, reducing the steady state concentration of NAD (Figure 3B).
+The decline in NAD-concnetrtion can be compensated by a higher expression of NamPRT (not shown) further increasing NAD consumption flux.
 
 These findings can be explained when looking in more detail into the kinetic parameters of NamPRT and NAD-consuming enzymes such as Sirt1p.
-The ability to maintain high a NAD concentration in the presence of NNMT and at low Nam availability, is due to the very high affinity of NamPRT for its substrate, having a half saturation constant (Km) in the low nanomolar range.
-The increase of NAD consumption flux is caused by the fact that most NAD-consuming enzymes are inhibited by their product Nam, which is also the reason why the presence of NNMT enables higher NAD consumption fluxes.
+The increase of NAD consumption flux is caused by the fact that most NAD-consuming enzymes are inhibited by their product Nam, explaing why the presence of NNMT enables higher NAD consumption fluxes.
+At the same time, the high substrate affinity of NamPRT maintains a sufficiently high NAD-concetration.
 
-As the substrate affinity of NamPRT for Nam is extremely high (with a Km in the low nanomolar range) and as this might not have been the case throughout evolution, we analysed the effect of the NamPRT Km on NAD steady state concentration and NAD consumption flux, leaving all other kinetic parameters constant.
-In the absence of NNMT (Figure 5A-B) the Km has very little effect on steady state NAD concentration and NAD consumption flux.
-Without NNMT, NAD concentration and consumption flux are both considerably affected by cell division rates, at least if the enzyme expression is kept constant.
-This is of course an artificial scenario, as one would assume organisms to regulate enzyme expression to achieve similar levels of metabolite concentrations instead.
-In the absence of NNMT there appears furthermore to be a trade-off between achievable steady state NAD concentration and NAD consumption flux.
+As kinetic parameters of NamPRT are only available for the human enzyme \cite{Burgos2008} , we went on to theoretically analysed the potential effect of the NamPRT affinity (Km) on NAD steady state concentration and NAD consumption flux, leaving all other kinetic parameters constant.
+In the absence of NNMT (Figure 3C-D) a change in Km has very little effect on steady state NAD concentration and NAD consumption flux.
+In the presence of NNMT, however, NAD consumption flux and NAD concentration increases with decreasing Km values (Figure 5E-F).
 
-In the presence of NNMT, NAD consumption flux and NAD concentration increases with decreasing Km values (Figure 5C-D).
-And we note, that both the NAD steady state concentration and the consumption flux are relatively stable over a wide range of cell division rates even at constant expression levels of the involved enzymes, suggesting that NNMT might have an important role to maintain homeostasis of NAD metabolism.
+It is intersting to note that, without NNMT, NAD concentration and consumption flux are both considerably affected by cell division rates, at least if the enzyme expression is kept constant.
+Even though, this is of course an artificial scenario, as one would assume organisms to regulate enzyme expression to achieve similar levels of metabolite concentrations instead.
+However, there seems to be an apparent trade off between maintainable NAD concentration and consumption flux.
 
-When we compare NAD consumption and NAD concentration with and without NNMT with two different substrate affinities of NamPRT, we see that at a low affinity (Km of 100 nM, which is in the range of the Km of NAPRT for its substrate, or the Km of NADA for Nam), NAD consumption flux is only higher with NNMT at low cell division rates, whereas at high division rates, higher NAD consumption fluxes are achieved without NNMT (Figure 5 E-F).
-This might explain why we do not find NNMT in organism that tend to have high growth rates\todo[author=Ines]{I am not sure if we should go even further suggesting that decreasing substrate affinity by using a competitive inhibitor such as FK866 will influence the consumption rate in fast growing cells much more than the one in slow growing cells, being potentially relevant for cancer therapy.
-– Discussion}.
-We furthermore find, that the competitive advantage of NamPRT of NADA is only present at sufficiently high affinity of NamPRT.
-
-Our analysis also suggests that NNMT might have exerted an evolutionary driving force on the substrate affinity of NamPRT, explaining the extreme values found for the human enzyme\todo[author=Ines]{Move to discussion?}.
+When we compare NAD consumption and NAD concentration with and without NNMT with two different substrate affinities of NamPRT, we see that at a low affinity (Km of 1 mikroM, which is in the range of the Km of NADA for Nam, NAD consumption flux is only higher with NNMT at low cell division rates.
+This effect becomes stronger if we further lower the affinity (not shown) (Figure 3G-H).
 
@@
 Thus, what is the impact of the substrate affinity of NNMT that is competing with NamPRT for the same substrate?
-In Figure 6A we see that the substrate affinity values found in the human enzymes (indicated by black asterisks) are actually optimal with respect to both achievable steady state NAD concentration and consumption fluxes.
+In Figure 4A we see that the substrate affinity values found in the human enzymes (indicated by black asterisks) are actually optimal with respect to both achievable steady state NAD concentration and consumption fluxes.
 Thus, a further increase of the affinity of NamPRT for Nam would not provide any advantage.
@@
 
-To see whether we can find sequence variations in the protein sequence of NamPRT that indicate evolutionary changes of NamPRT upon the appearance of NNMT, we created a multiple sequence alignment of selected eukaryotic NamPRT sequences.
-As shown in Figure 7A and Supplementary Figure 3\todo[author=Ines]{To be created}, Deuterostomia that have only NamPRT and NNMT (indicated by the number 6 in parenthesis \todo[author=Ines]{Maybe rather indicate this by colour and the kingdom through abbrev?}) have an insert of ten amino acids corresponding to positions 43 to 52 of the human enzyme.
-Looking at the crystal structure of human NamPRT this sequence insertion corresponds to a region that has not been structurally resolved in any of the currently available crystal structures (e.g.
-\cite{Wang2006} structure visualisation Figure 7B) and its function is unclear.
-The unresolved loop structure overlaps with a predicted weak nuclear localisation signal (figure 7A), that is not present without the insertion.
-The loop is in addition connected to one of the beta-sheets involved in substrate binding, potentially affecting the affinity or turnover of the enzyme.
+As our simulations suggest that NNMT might have excerted an evolitionary pressure on the development of NamPRT.
+We performed multiple sequence analysis to see if we can find sequence variations in the protein sequence of NamPRT that indicate evolutionary changes of NamPRT upon the appearance of NNMT.
+The multiple sequence alignment of selected sequences is shown in Figure 5A and Supplementary Figure ?\todo[author=Mathias B.]{Full MSA to be prepared as suppl figure.} We recognized that Deuterostomia that have only NamPRT and NNMT (indicated by the number 6 in parenthesis \todo[author=Mathias B.]{Maybe rather indicate this by colour and the kingdom through abbrev?}) have an insert of ten amino acids corresponding to positions 43 to 52 of the human enzyme, that overlaps with a predicted weak nuclear localisation signal, that is lost when removing this sequence stretch.
+Furthermore, the inserted sequence corresponce to an structurallly unresolved loop structure in all avaibable crystal structure of human NamPR (e.g.
+\cite{Wang2006} structure visualisation Figure 5B).
+The loop is connected to one of the beta-sheets involved in substrate binding.
 
-The observed evolutionary change in the primary sequence of NamPRT could therefore have had different effects, that we wanted to test experimentally.
-We first investigated whether the deletion of the amino acids 43 to 52 has an effect on the localisation of the human enzyme, expressing a ….
-mutant lacking this sequence in HeLa cells.
-We could, however, not detect any changes in subcellular localisation (Figure 7C) compared to the wildtype fusion construct.
-We therefore conclude that the partial nuclear localisation of NamPRT is not affected by the deletion.
-We then tested the enzymatic activity of the NamPRT mutant recombinantly expressed in E.
+From these observations we derived two hypotheses regarding the potential function of the loop for NamPRT function.
+Firstly, we analysed whether the deletion of the amino acids 43 to 52 has an effect on the localisation of the human enzyme.
+But both wildtype and mutant protein show a mixed localisation in the cytosol and nucleus (Figure 5C).
+We therefore conclude that the partial nuclear localisation of NamPRT is not affected by the sequence deletion.
+The second hypothesis was that the sequence insertion influences NamPRT affinity and activity, which is also what our mathematical simulations would predict.
+We thus recombinantly expressed  both wildttype and mutant  NamPRT  in E.
 coli.
-As shown in Figure 7D the enzyme still forms a dimer, thus seems to be folded correctly.
-The enzymatic activity is, however, much lower compared to the wildtype enzyme.
-Using higher concentrations of Nam, it appears that the mutant is not saturated at 100 \textmu M, pointing to a decreased substrate affinity of the mutant enzyme.
+Upon purifictaion we meassured the activity using NMR detecttion of NMN produced in the presence and absence of ATP.
+It appears that the enzymatic activity of the mutant enzyme is much lower than that of the wildtype enzyme both with and without ATP.
+We furthermore see that the activity of the mutant enzyme can be increase with higher concentration of Nam and PRPP whereas that is not the case for the wildtype enzyme.
+Pointing to a decrease substrate affinity of the mutant enzyme.
+Supporting the second hypothesis and thus supporting the predictions derived from our mathematical modelling approach.
+
+\subsection{Loss of NADA in vertebrates}
+
+But why is NADA lost in vertebrates?
+Our simulations show that in direct competition NADA is outcompeted by the combined presence of NamPRT and NNMT but not by NamPRT alone.
+This effect deisappears of NamPRT affinity is lowered.
+....

### 2018-06-21 latex/figurelegends.tex

@@
+%%% TeX-master: "manuscript"
+
 \section{Figure Legends}
@@
 B) The inserted region is not resolved in crystal structures of human NamPRT and thus appears to be a flexible loop structure at the surface of the NamPRT dimer, coloured in red.
-The visualisation is based on a structure prediction of SWISS-MODEL \cite{Arnold2006; Biasini2014} of the sequence of the human NamPRT (P43490) using the model 2H3D as template \cite{Wang2006}.
+The visualisation is based on a structure prediction of SWISS-MODEL \cite{Arnold2006,Biasini2014} of the sequence of the human NamPRT (P43490) using the model 2H3D as template \cite{Wang2006}.
 

### 2018-06-21 latex/manuscript.tex

@@
 \documentclass[paper=a4, 12pt]{scrartcl}
+
 \usepackage[utf8]{inputenc}
@@
 \noindent
-Mathias Bockwoldt$^{1}$, Dorothee Houry$^{2}$,  Marc Niere$^{2}$, Toni I. Gossmann$^{3}$, Mathias Ziegler$^{2}$ and Ines Heiland$^{1,\S}$
+Mathias Bockwoldt$^{1}$, Dorothée Houry$^{2}$,  Marc Niere$^{2}$, Toni I. Gossmann$^{3}$, Mathias Ziegler$^{2}$ and Ines Heiland$^{1,\S}$
 
@@
 While in lower organisms the initial deamidation of Nam is prevalent, in animals the direct conversion of Nam to the mononucleotide by Nam phosphoribosyltransferase (NamPRT) dominates and eventually remains as the single Nam recycling route in vertebrates.
+
 Strikingly, loss of the deamidation pathway in early vertebrates is accompanied by the occurrence of a new enzyme that marks Nam for excretion by methylation – nicotinamide N-methyltransferase (NNMT).
@@
 And why is the occurrence of NNMT accompanied by a diversification of NAD-consuming enzymes?
+
 We have used mathematical modelling approaches to resolve these counterintuitive observations.
@@
 Mathematical modelling supports the hypothesis that NNMT exerted an evolutionary pressure on NamPRT enforcing the development of its unusually high substrate affinity.
-Using multiple sequence alignments, we identified a sequence insertion, first occurring in vertebrates, that parallels an - experimentally verified - increase in the substrate affinity of the enzyme.
+Using multiple sequence alignments, we identified a sequence insertion, first occurring in vertebrates, that parallels an – experimentally verified – increase in the substrate affinity of the enzyme.
 Further simulations show that the deamidation pathway became obsolete owing to the high substrate affinity of NamPRT.
@@
 \input{introduction}
+
 \input{results}
+
 \input{discussion}
+
 \input{figurelegends}
 
-\subsection{Comments and parts of old version not included yet}
+\section{Comments and parts of old version not included yet}
 
@@
 
+\bibliographystyle{plain}
+\bibliography{library}
+
 \end{document}

### 2018-06-21 latex/results.tex

@@
 Nam and NA are together known as vitamin B3 or niacin.
-Alternatively, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead \cite{Imai2018}.
+Alternatively, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead \cite{Yoshino2018}.
 Du to the hight turnover of NAD observed and the fact that only 1\% of the tryptophan taken up with our diet is converted into NAD, vitamin B3 (Nam and NA) and to a lower extend NR are the major precursor of NAD-biosynthesis.
@@
 We were not be able to find any gene with considerable similarity in fungi or plants, even though in fungi a gene named NNMT can be found in databases.
-These genes do, however, show very limited sequence similarities to the NNMT of nematodes or deuterostomes and in the yeast protein has later on be found to be a lysin-protein-methyltransferase  \cite{?}(\todo[author=Mathias B.]{Please look up the reference for this.}.
+These genes do, however, show very limited sequence similarities to the NNMT of nematodes or deuterostomes and in the yeast protein has later on be found to be a lysin-protein-methyltransferase \cite{Wlodarski2011}.
 Nematodes are the only organisms where we find NNMT together with NADA without NamPRT being present.

### 2018-06-21 figures/NamPRT_activity/make_figure.py
Added figure with NamPRT activity

@@ -1,0 +1,80 @@
+import sys
+import os
+import numpy as np
+import matplotlib as mpl
+import matplotlib.pyplot as plt
+from matplotlib.ticker import FormatStrFormatter
+import seaborn as sns
+
+fs = 9
+img_width = 4.5 # 3.42 or 4.5 or 7.0 for PNAS
+img_height = 3.0
+
+plt.rc('font', size=fs)        # controls default text sizes
+plt.rc('axes', titlesize=fs)   # fontsize of the axes title
+plt.rc('axes', labelsize=fs)   # fontsize of the x and y labels
+plt.rc('xtick', labelsize=fs)  # fontsize of the tick labels
+plt.rc('ytick', labelsize=fs)  # fontsize of the tick labels
+plt.rc('legend', fontsize=fs)  # legend fontsize
+plt.rc('figure', titlesize=fs) # fontsize of the figure title
+
+#color = sns.color_palette("colorblind", n_colors=2)
+
+majorlabel = ('wildtype', 'mutant')
+minorlabel = ('Nam 1\nPRPP 1\nATP 0', '1\n0.1\n0', '0.1\n1\n0', '0.1\n0.1\n0', '0.1\n0.1\n1')
+
+wt = []
+mu = []
+wtstd = []
+mustd = []
+
+with open('activity.csv') as f:
+	next(f)
+	next(f)
+	for l in minorlabel:
+		wtbuf = []
+		mubuf = []
+		for _ in range(3):
+			d = next(f).rstrip().replace(',', '.').split('\t')
+			wtbuf.append(float(d[5]))
+			mubuf.append(float(d[6]))
+		wt.append(np.mean(wtbuf))
+		mu.append(np.mean(mubuf))
+		wtstd.append(np.std(wtbuf))
+		mustd.append(np.std(mubuf))
+
+sns.set_style("whitegrid")
+
+fig = plt.figure(figsize = (img_width, img_height))
+ax1 = fig.add_subplot(1,1,1) # numrows, numcols, fignum
+
+ind = np.arange(5)  # the x locations for the groups
+width = 0.33       # the width of the bars
+offset = width/2
+
+wt_bar = ax1.bar(ind+offset+offset, wt, width, yerr=wtstd, color='black', align='center', error_kw={'ecolor':'black'})
[…]

### 2018-06-21 latex/figurelegends.tex
Indicated enzymes in figure by colour instead of number as before. Adjusted text accordingly.

@@
 A) Multiple sequence alignment of NamPRT of selected deuterostomes show a sequence insertion in organisms that encode NamPRT and NNMT.
-(Species names are given and the number in parenthesis indicates the enzyme combination encoded in these species; 3 – NADA and NamPRT, 6 – NamPRT and NNMT, 7 – NADA, NamPRT and NNMT.
-\todo[author=Mathias]{Note to myself: replace the numbers by something better and rewrite this part.}) The nuclear localisation signal (NLS\todo[author=Mathias]{Note to myself: Write “predicted NLS” in figure}) indicated, was predicted using using cNLS-mapper (\url{http://nls-mapper.iab.keio.ac.jp/cgi-bin/NLS_Mapper_form.cgi} \cite{Kosugi2009}).
+Coloured circles indicate the enzymes present in the species; blue: NamPRT and NNMT; black: NamPRT, NADA, and NNMT; yellow: NamPRT and NADA.
 
@@
 The visualisation is based on a structure prediction of SWISS-MODEL \cite{Arnold2006,Biasini2014} of the sequence of the human NamPRT (P43490) using the model 2H3D as template \cite{Wang2006}.
-
-C) The localisation of wildtype NamPRT varies between pure cytosolic and mixed cytosolic-nuclear localisation.
-This is not changed in the mutant missing the insert that corresponds to the unresolved loop structure.

### 2018-06-21 latex/results.tex
Indicated enzymes in figure by colour instead of number as before. Adjusted text accordingly.

@@
 We performed multiple sequence analysis to see if we can find sequence variations in the protein sequence of NamPRT that indicate evolutionary changes of NamPRT upon the appearance of NNMT.
-The multiple sequence alignment of selected sequences is shown in Figure 5A and Supplementary Figure ?\todo[author=Mathias B.]{Full MSA to be prepared as suppl figure.} We recognized that Deuterostomia that have only NamPRT and NNMT (indicated by the number 6 in parenthesis \todo[author=Mathias B.]{Maybe rather indicate this by colour and the kingdom through abbrev?}) have an insert of ten amino acids corresponding to positions 43 to 52 of the human enzyme, that overlaps with a predicted weak nuclear localisation signal, that is lost when removing this sequence stretch.
+The multiple sequence alignment of selected sequences is shown in Figure 5A and Supplementary Figure ?\todo[author=Mathias B.]{Full MSA to be prepared as suppl figure.} We recognized that Deuterostomia that have only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 43 to 52 of the human enzyme, that overlaps with a predicted weak nuclear localisation signal, that is lost when removing this sequence stretch.
 Furthermore, the inserted sequence corresponce to an structurallly unresolved loop structure in all avaibable crystal structure of human NamPR (e.g.

### 2018-06-21 simulations/symlink.sh
added script for symlink conversion

@@ -1,0 +1,11 @@
+#!/bin/bash
+# converts all symlinks from the current directory and below into real links
+for f in $(find . -type l) ; do
+    ft=$(dirname ${f})/$(readlink ${f})
+    if [ -f ${ft} ] ; then
+        mv ${f} ${f}.bak
+        ln ${ft} ${f}
+    else
+        echo not a file: ${ft}
+    fi
+done

### 2018-06-21 latex/discussion.tex
Split methods and discussion; Updated affiliations; added some references

@@
 
+In contrast to our results, it was proposed that NNMT is present in yeast (\textit{Saccharomyces cerevisiae} \cite{Anderson2003}.
+The enzyme was later further characterised as methyltransferase of the transcription elongation factors TEF1 and TEF2 but not Nam \cite{Wlodarski2011}.
+
 \subsection{General applicability}
 
-It is of course impossible to extend our conclusions to other metabolic processes without analysing them in more detail, but we have noted that the metabolism of other vitamins, such as pyridoxal (vitamin B6) metabolism, also contains vitamin degrading enzymes \todo[author=Mathias]{source}, that might thus have similar roles.
+It is of course impossible to extend our conclusions to other metabolic processes without analysing them in more detail, but we have noted that the metabolism of other vitamins, such as pyridoxal (vitamin B6) metabolism, also contains vitamin degrading enzymes \todo{source}, that might thus have similar roles.
 Degrading enzymes do in general not receive the same attention as biosynthetic enzymes, reflected by the observation that NNMT has only recently been analysed in more detail, besides the much earlier recognition of NNMT as potential marker for some types of cancer (e.g.
-Ref?) and routine clinical measurements of urine methyl-Nam concentrations in the context of different diseases (e.g.
-Ref?).
+\cite{Okamura1998}) and routine clinical measurements of urine methyl-Nam concentrations in the context of different diseases (e.g.
+\cite{Pumpo2001,Delaney2005}).
 Our analysis show, however, that degrading enzymes do not solely modify substrates for better excretion, but can play a vital role both in human physiology as well as in the evolutionary development of biological processes.
@@
 The same holds for the analysis of enzyme acetylation versus deacetylation by sirtuins that would enable prediction about enzyme activation or histone state if available.
-
-\section{Experimental Procedure}
-
-\subsection{Dynamic modelling}
-
-Kinetic parameters (substrate affinity (Km) and turnover rates (kcat), substrate and product inhibitions) were retrieved from the enzyme database BRENDA and additionally evaluated by checking the original literature especially with respect to measurement conditions.
-Parameter values from mammals were used if available.
-For enzymes not present in mammals, values from yeast were used.
-The full list of kinetic parameters including reference to original literature can be found in Supplementary table 1.
-For NMNAT, the previously developed rate law for substrate competition was used \cite{Schauble2013}.
-Despite these modifications, Henri-Michaelis-Menten kinetics were used for all reactions except the import and efflux of Nam, which were simulated using constant flux and mass action kinetics, respectively.
-All simulations were performed using the steady state calculation and parameter scan options provided by COPASI 4.22 \cite{Hoops2006}.
-The model will be available at the Biomodels database accession number \todo[inline]{xxx}.
-Related figures were generated using gnuplot\todo[author=Mathias]{Version?}.
-
-\subsection{Phylogenetic Analysis}
-
-NADA, NamPRT, and NNMT enzymes or enzyme candidates were identified using Blastp with known enzymes against the NCBI non-redundant protein sequence database (nr).
-A list of functionally verified enzymes used as templates is given in supplementary table 2.
-This table also includes the length cut-offs for identified enzymes.
-The e-value cut-off was 1e-30 for all enzymes.
-Blastp parameters were set to yield maximum 20,000 target sequences, using the BLOSUM62 matrix with a word size of 6 and gap opening and extension costs of 11 and 1, respectively.
-Low-complexity filtering was disabled.
-Obvious sequence contaminations were removed by manual inspection of the results.
-The taxonomy IDs of the species for each enzyme was derived from the accession2taxonomy database provided by NCBI.
-Scripts for creating, analysing, and visualising the phylogenetic tree were written in Python, using the ETE3 toolkit (Huerta-Cepas, 2010).
-
-\subsection{Generation of eukaryotic expression vectors encoding C-terminally FLAG-tagged NamPT proteins}
-
-The open reading frame (ORF) encoding human NamPRT was inserted into pFLAG-CMV-5a (Merck - Sigma Aldrich) via EcoRI/BamHI sites.
-Using a PCR-based approach, this vector provided the basis for the generation of a plasmid encoding a NamPRT deletion mutant lacking amino acid residues 42-51 (NamPRT-$\Delta$42-51).
-The sequences of the inserted ORFs were verified by DNA sequence analysis.
-
-\subsection{Transient transfection, immunocytochemstry and confocal laser scanning microscopy}
-
-HeLa S3 cells cultivated in Ham’s F12 medium supplemented with 10\% (v/v) FCS, 2 mM L-glutamine, and penicillin/streptomycin, were seeded on cover slips in a 24 well plate.
-After one day, cells were transfected using Effectene transfection reagent (Qiagen) according to the manufacturer’s recommendations.
-Cells were fixed with 4\% paraformaldehyde in PBS 24 hours post transfection, permeabilised (0.5\% (v/v) Triton X-100 in PBS) and blocked for one hour with complete culture medium.
-After overnight incubation with primary FLAG-antibody (mouse M2, Sigma-Aldrich) diluted 1:2500 in complete medium, cells were washed and incubated for one hour with secondary AlexaFluor 594-conjugated goat anti mouse antibody (ThermoFisher, Invitrogen) diluted 1:1000 in complete culture medium.
-Nuclei were stained with DAPI and the cells washed.
-The cover slips were mounted on microscope slides using ProLong Gold (ThermoFisher, Invitrogen).
-Confocal laser scan imaging of cells was performed using a Leica TCS SP8 STED 3x microscope equipped with a 100x oil immersion objective (numerical aperture 1.4).

### 2018-06-21 latex/figurelegends.tex
Split methods and discussion; Updated affiliations; added some references

@@
 
-\subsection{Figure1}
+\subsection{Figure 1}
 
@@
 To a lesser extend, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead.
-Only 1\% of the tryptophan taken up with our diet is converted into NAD, thus, vitamin B3 and, to a lower extend, NR are essential components of our diet, with Nam being the major NAD precursor in humans\todo[author=Mathias]{Whole sentence not necessarily part of the legend}.
+Only 1\% of the tryptophan taken up with our diet is converted into NAD, thus, vitamin B3 and, to a lower extend, NR are essential components of our diet, with Nam being the major NAD precursor in humans\todo{Whole sentence (starting with "Only") not necessarily part of the legend.
+What do you think, Ines?}.
 Nam is furthermore the product of NAD-consuming signalling reactions such as sirtuins (NAD-dependent histone deacetylases) or PARPs (poly-ADP-ribosylases).

### 2018-06-21 latex/introduction.tex
Split methods and discussion; Updated affiliations; added some references

@@
 Despite the importance of its salvage, Nam can also be marked for excretion by methylation.
-Indeed, the presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012} is among the most enigmatic and counterintuitive features of NAD metabolism.
+Indeed, the presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012FEBS} is among the most enigmatic and counterintuitive features of NAD metabolism.
 Why is there one enzyme (NamPRT) seemingly optimized to recycle even the faintest amounts of Nam back into NAD synthesis, while at the same time there is another enzyme (NNMT) that seems to have no metabolic function other than to remove Nam from NAD metabolism, This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deaminated Nam to nicotinic acid (NA) using the nicotinamide deamidase (NADA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (Fig.

### 2018-06-21 latex/methods.tex
Split methods and discussion; Updated affiliations; added some references

%%% TeX-master: "manuscript"

\section{Experimental Procedure}

\subsection{Dynamic modelling}

Kinetic parameters (substrate affinity (Km) and turnover rates (kcat), substrate and product inhibitions) were retrieved from the enzyme database BRENDA and additionally evaluated by checking the original literature especially with respect to measurement conditions. Parameter values from mammals were used if available. For enzymes not present in mammals, values from yeast were used. The full list of kinetic parameters including reference to original literature can be found in Supplementary table 1. For NMNAT, the previously developed rate law for substrate competition was used \cite{Schauble2013}. Despite these modifications, Henri-Michaelis-Menten kinetics were used for all reactions except the import and efflux of Nam, which were simulated using constant flux and mass action kinetics, respectively. All simulations were performed using the steady state calculation and parameter scan options provided by COPASI 4.22 \cite{Hoops2006}. The model will be available at the Biomodels database accession number xxx\todo[author=Ines]{Insert number}. Related figures were generated using gnuplot\todo[author=Ines]{Version?}.

\subsection{Phylogenetic Analysis}

NADA, NamPRT, and NNMT enzymes or enzyme candidates were identified using Blastp with known enzymes against the NCBI non-redundant protein sequence database (nr). A list of functionally verified enzymes used as templates is given in supplementary table 2. This table also includes the length cut-offs for identified enzymes. The e-value cut-off was 1e-30 for all enzymes. Blastp parameters were set to yield maximum 20,000 target sequences, using the BLOSUM62 matrix with a word size of 6 and gap opening and extension costs of 11 and 1, respectively. Low-complexity filtering was disabled. Obvious sequence contaminations were removed by manual inspection of the results. The taxonomy IDs of the species for each enzyme was derived from the accession2taxonomy database provided by NCBI. Scripts for creating, analysing, and visualising the phylogenetic tree were written in Python, using the ETE3 toolkit (Huerta-Cepas, 2010).

\subsection{Generation of eukaryotic expression vectors encoding C-terminally FLAG-tagged NamPT proteins}

The open reading frame (ORF) encoding human NamPRT was inserted into pFLAG-CMV-5a (Merck - Sigma Aldrich) via EcoRI/BamHI sites. Using a PCR-based approach, this vector provided the basis for the generation of a plasmid encoding a NamPRT deletion mutant lacking amino acid residues 42-51 (NamPRT-$\Delta$42-51). The sequences of the inserted ORFs were verified by DNA sequence analysis.

\subsection{Transient transfection, immunocytochemstry and confocal laser scanning microscopy}

HeLa S3 cells cultivated in Ham’s F12 medium supplemented with 10\% (v/v) FCS, 2 mM L-glutamine, and penicillin/streptomycin, were seeded on cover slips in a 24 well plate. After one day, cells were transfected using Effectene transfection reagent (Qiagen) according to the manufacturer’s recommendations. Cells were fixed with 4\% paraformaldehyde in PBS 24 hours post transfection, permeabilised (0.5\% (v/v) Triton X-100 in PBS) and blocked for one hour with complete culture medium. After overnight incubation with primary FLAG-antibody (mouse M2, Sigma-Aldrich) diluted 1:2500 in complete medium, cells were washed and incubated for one hour with secondary AlexaFluor 594-conjugated goat anti mouse antibody (ThermoFisher, Invitrogen) diluted 1:1000 in complete culture medium. Nuclei were stained with DAPI and the cells washed. The cover slips were mounted on microscope slides using ProLong Gold (ThermoFisher, Invitrogen). Confocal laser scan imaging of cells was performed using a Leica TCS SP8 STED 3x microscope equipped with a 100x oil immersion objective (numerical aperture 1.4).

### 2018-06-21 latex/results.tex
Split methods and discussion; Updated affiliations; added some references

@@
 In addition to the phylogenetic distribution of the two Nam salvage enzymes NADA and NamPRT, we looked at phylogenetic diversity of enzymes catalysing NAD-dependent signalling reactions.
-To do so we used the previously established classification into 10 different families \cite{Gossmann2012} (For details see materials and methods and supplementary information).
+To do so we used the previously established classification into 10 different families \cite{Gossmann2012FEBS} (For details see materials and methods and supplementary information).
 The numbers in Figure 2B denote the average number of NAD-dependent signalling enzyme families found in each taxa.
-With the exception of Cnidaria and Lophotrochozoa, we find average 3 to 4 families in Protostomia, whereas most Deuterostomia have on average more than 8 families with an increasing diversification of enzymes within some of these families \cite{Gossmann2014}.
+With the exception of Cnidaria and Lophotrochozoa, we find average 3 to 4 families in Protostomia, whereas most Deuterostomia have on average more than 8 families with an increasing diversification of enzymes within some of these families \cite{Gossmann2014DNAR}.
 

### 2018-06-22 figures/NamPRT_activity/make_figure.py
Split NamPRT activity into two panels

@@ -21,54 +21,81 @@ plt.rc('figure', titlesize=fs) # fontsize of the figure title
 #color = sns.color_palette("colorblind", n_colors=2)
 
 majorlabel = ('wildtype', 'mutant')
-minorlabel = ('Nam 1\nPRPP 1\nATP 0', '1\n0.1\n0', '0.1\n1\n0', '0.1\n0.1\n0', '0.1\n0.1\n1')
+minorlabel1 = ('Nam 1\nPRPP 1\nATP 0', '1\n0.1\n0', '0.1\n1\n0')
+minorlabel2 = ('0.1\n0.1\n0', '0.1\n0.1\n1')
 
-wt = []
-mu = []
-wtstd = []
-mustd = []
+wt1 = []
+mu1 = []
+wtstd1 = []
+mustd1 = []
+
+wt2 = []
+mu2 = []
+wtstd2 = []
+mustd2 = []
 
 with open('activity.csv') as f:
 	next(f)
 	next(f)
-	for l in minorlabel:
+	for _ in minorlabel1:
 		wtbuf = []
 		mubuf = []
 		for _ in range(3):
 			d = next(f).rstrip().replace(',', '.').split('\t')
 			wtbuf.append(float(d[5]))
 			mubuf.append(float(d[6]))
-		wt.append(np.mean(wtbuf))
-		mu.append(np.mean(mubuf))
-		wtstd.append(np.std(wtbuf))
-		mustd.append(np.std(mubuf))
+		wt1.append(np.mean(wtbuf))
+		mu1.append(np.mean(mubuf))
+		wtstd1.append(np.std(wtbuf))
+		mustd1.append(np.std(mubuf))
+
+	rel = None
+	for _ in minorlabel2:
+		wtbuf = []
+		mubuf = []
+		for _ in range(3):
+			d = next(f).rstrip().replace(',', '.').split('\t')
+			wtbuf.append(float(d[5]))
+			mubuf.append(float(d[6]))
+		if rel is None:
+			rel = np.mean(wtbuf)
+		wt2.append(np.mean(wtbuf)/rel)
+		mu2.append(np.mean(mubuf)/rel)
+		wtstd2.append(np.std(wtbuf)/rel)
+		mustd2.append(np.std(mubuf)/rel)
 
 sns.set_style("whitegrid")
 
 fig = plt.figure(figsize = (img_width, img_height))
[…]

### 2018-06-22 figures/Unresolved_loop_large/get_attributes.py
Added supplementary figure alignment of NamPRT with focus on loop

@@ -1,0 +1,29 @@
+
+from taxfinder import TaxFinder
+
+spec = {}
+for line in open('renamed.fasta'):
+	if line.startswith('>'):
+		spec[line.strip()[1:].replace('-', ' ')] = 0
+
+TF = TaxFinder()
+
+NADA = 1
+NamPRT = 2
+NNMT = 4
+
+# aD -> NADA
+# aL -> NNMT
+# aM -> NamPRT
+attrs = {'aD': NADA, 'aL': NNMT, 'aM': NamPRT}
+
+for line in open('attributes.txt'):
+	lline = line.split()
+	name = TF.getNameFromID(lline[0])
+	if name in spec:
+		att = lline[1]
+		for a in attrs:
+			if a in att:
+				spec[name] += attrs[a]
+
+print(spec)

### 2018-06-22 figures/Unresolved_loop_large/make_fasta_for_alignment.py

@@ -1,0 +1,32 @@
+
+seqs = {}
+with open('fastas_euks.fa') as f:
+	for line in f:
+		if line.startswith('>'):
+			#acc = line.split()[0][5:]
+			spec = line.split('[')[1][:-2].replace(' ', '-')
+			seq = next(f).rstrip()
+			if not 400 < len(seq) < 550:
+				continue
+
+			if spec not in seqs or len(seqs[spec]) < len(seq):
+				seqs[spec] = seq
+
+print(len(seqs), 'seqs found')
+
+order = []
+
+for line in open('commontree.txt'):
+	for char in r'|\+-':
+		line = line.replace(char, '')
+	line = line.strip().replace(' ', '-')
+	if line in seqs:
+		order.append(line)
+
+for spec in seqs:
+	if spec not in order:
+		print(spec, 'not found in order')
+
+with open('renamed.fasta', 'w') as out:
+	for spec in order:
+		out.write('>{}\n{}\n\n'.format(spec, seqs[spec]))

### 2018-06-22 figures/Unresolved_loop_large/make_html.py

@@ -1,0 +1,94 @@
+#!/usr/bin/env python3
+
+import sys
+from taxfinder import TaxFinder
+
+template = '''<!DOCTYPE html>
+<html>
+<head>
+<meta charset="utf-8" />
+<title>MSA</title>
+<style type="text/css">
+body {{font-family: monospace;}}
+.bg {{width: 25px;}}
+.white {{background-color: #ffffff;}}
+.red {{background-color: #aa0000;}}
+.green {{background-color: #00aa00;}}
+.lightblue {{background-color: #55aaff;}}
+.yellow {{background-color: #dddd00;}}
+.purple {{background-color: #aa00aa;}}
+.blue {{background-color: #3333ff;}}
+.black {{background-color: #000000; color: white;}}
+table {{border-collapse: collapse;}}
+td, tr {{margin: 0px; padding: 0px;}}
+</style>
+</head>
+<body>
+<table>
+{}
+</table>
+</body>
+</html>'''
+
+line_template = '''<tr>
+<td class="bg {}">&nbsp;</td>
+<td><nobr>&nbsp;&nbsp;{}&nbsp;&nbsp;</nobr></td>
+<td><nobr>{}</nobr></td>
+</tr>'''
+
+fasta = {}
+spec = None
+seq = []
+attributes = {}
+for line in open('alignmentmuscle.fasta'):
+	line = line.rstrip()
+	if line.startswith('>'):
+		if spec:
+			fasta[spec] = ''.join(seq)[232:305]
+		spec = line[1:].replace('-', ' ')
+		attributes[spec] = 0
+		seq = []
+	else:
+		seq.append(line)
+
+
+TF = TaxFinder()
+
+NADA = 1
+NamPRT = 2
+NNMT = 4
+attrs = {'aD': NADA, 'aL': NNMT, 'aM': NamPRT}
+
+colors = {0: 'white',
+NADA: 'red',
+NamPRT: 'green',
+NNMT: 'lightblue',
+NADA+NamPRT: 'yellow',
+NADA+NNMT: 'purple',
+NamPRT+NNMT: 'blue',
+NADA+NamPRT+NNMT: 'black'}
+
+for line in open('attributes.txt'):
+	lline = line.split()
+	name = TF.getNameFromID(lline[0])
+	if name in attributes:
+		for a in attrs:
+			if a in lline[1]:
+				attributes[name] += attrs[a]
+
+order = []
+for line in open('commontree.txt'):
+	for char in r'|\+-':
+		line = line.replace(char, '')
[…]

### 2018-06-22 latex/results.tex
Linked new unresolved loop supp figure

@@
 We performed multiple sequence analysis to see if we can find sequence variations in the protein sequence of NamPRT that indicate evolutionary changes of NamPRT upon the appearance of NNMT.
-The multiple sequence alignment of selected sequences is shown in Figure 5A and Supplementary Figure ?\todo[author=Mathias B.]{Full MSA to be prepared as suppl figure.} We recognized that Deuterostomia that have only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 43 to 52 of the human enzyme, that overlaps with a predicted weak nuclear localisation signal, that is lost when removing this sequence stretch.
+The multiple sequence alignment of selected sequences is shown in Figure 5A and Supplementary Figure ?\todo[author=Ines]{Decide figure number} We recognized that Deuterostomia that have only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 43 to 52 of the human enzyme, that overlaps with a predicted weak nuclear localisation signal, that is lost when removing this sequence stretch.
 Furthermore, the inserted sequence corresponce to an structurallly unresolved loop structure in all avaibable crystal structure of human NamPR (e.g.

### 2018-06-24 simulations/symlink.sh
symlinkscript modified

@@ -1,7 +1,8 @@
 #!/bin/bash
 # converts all symlinks from the current directory and below into real links
 for f in $(find . -type l) ; do
-    ft=$(dirname ${f})/$(readlink ${f})
+    ft=$(readlink ${f})
+    echo ${ft} | grep -q '^/' || ft=$(dirname ${f})/${ft}
     if [ -f ${ft} ] ; then
         mv ${f} ${f}.bak
         ln ${ft} ${f}

### 2018-06-24 simulations/symlink.sh
symlinkscript modified

@@ -1,8 +1,7 @@
 #!/bin/bash
 # converts all symlinks from the current directory and below into real links
 for f in $(find . -type l) ; do
-    ft=$(readlink ${f})
-    echo ${ft} | grep -q '^/' || ft=$(dirname ${f})/${ft}
+    ft=$(dirname ${f})/$(readlink ${f})
     if [ -f ${ft} ] ; then
         mv ${f} ${f}.bak
         ln ${ft} ${f}

### 2018-06-24 figures/NamPRT_activity/make_figure.py
Modified NamPRT activity plot

@@ -4,7 +4,7 @@ import numpy as np
 import matplotlib as mpl
 import matplotlib.pyplot as plt
 from matplotlib.ticker import FormatStrFormatter
-import seaborn as sns
+import scipy.stats as stats
 
 fs = 9
 img_width = 4.5 # 3.42 or 4.5 or 7.0 for PNAS
@@ -20,7 +20,7 @@ plt.rc('figure', titlesize=fs) # fontsize of the figure title
 
 #color = sns.color_palette("colorblind", n_colors=2)
 
-majorlabel = ('wildtype', 'mutant')
+majorlabel = ('wt NamPRT', '$\Delta$42-51 NamPRT')
 minorlabel1 = ('Nam 1\nPRPP 1\nATP 0', '1\n0.1\n0', '0.1\n1\n0')
 minorlabel2 = ('0.1\n0.1\n0', '0.1\n0.1\n1')
 
@@ -28,11 +28,15 @@ wt1 = []
 mu1 = []
 wtstd1 = []
 mustd1 = []
+wtall1 = []
+muall1 = []
 
 wt2 = []
 mu2 = []
 wtstd2 = []
 mustd2 = []
+wtall2 = []
+muall2 = []
 
 with open('activity.csv') as f:
 	next(f)
@@ -48,6 +52,8 @@ with open('activity.csv') as f:
 		mu1.append(np.mean(mubuf))
 		wtstd1.append(np.std(wtbuf))
 		mustd1.append(np.std(mubuf))
+		wtall1.append(tuple(wtbuf))
+		muall1.append(tuple(mubuf))
 
 	rel = None
 	for _ in minorlabel2:
@@ -63,8 +69,10 @@ with open('activity.csv') as f:
 		mu2.append(np.mean(mubuf)/rel)
 		wtstd2.append(np.std(wtbuf)/rel)
 		mustd2.append(np.std(mubuf)/rel)
+		wtall2.append(tuple(wtbuf))
+		muall2.append(tuple(mubuf))
 
-sns.set_style("whitegrid")
+wtstd2[0] = 0 # disable error bar for first element in second subplot
 
 fig = plt.figure(figsize = (img_width, img_height))
 ax1 = fig.add_subplot(1,2,1) # numrows, numcols, fignum
@@ -82,22 +90,41 @@ offset = width/2
[…]

### 2018-06-24 figures/NamPRT_activity/make_figure.py
Changed NamPRT activity figure

@@ -41,7 +41,9 @@ muall2 = []
 with open('activity.csv') as f:
 	next(f)
 	next(f)
-	for _ in minorlabel1:
+	for i in range(len(minorlabel1)):
+		if i == 1:
+			continue
 		wtbuf = []
 		mubuf = []
 		for _ in range(3):
@@ -72,13 +74,13 @@ with open('activity.csv') as f:
 		wtall2.append(tuple(wtbuf))
 		muall2.append(tuple(mubuf))
 
-wtstd2[0] = 0 # disable error bar for first element in second subplot
+minorlabel1 = [minorlabel1[0], minorlabel1[2]]
 
 fig = plt.figure(figsize = (img_width, img_height))
 ax1 = fig.add_subplot(1,2,1) # numrows, numcols, fignum
 ax2 = fig.add_subplot(1,2,2) # numrows, numcols, fignum
 
-ind1 = np.arange(3)  # the x locations for the groups
+ind1 = np.arange(2)  # the x locations for the groups
 ind2 = np.arange(2)  # the x locations for the groups
 width = 0.33       # the width of the bars
 offset = width/2
@@ -94,7 +96,7 @@ ax1.legend((wt_bar1[0], mu_bar1[0]), majorlabel, loc='upper right', frameon=Fals
 ax2.legend((wt_bar2[0], mu_bar2[0]), majorlabel, loc='upper left', frameon=False)
 
 ax1.xaxis.grid(False)
-ax1.set_xlim(0, 3)
+ax1.set_xlim(0, 2)
 ax1.set_ylim(0, 11)
 ax1.set_xticks(ind1 + width + 0.165)
 ax1.set_xticklabels(minorlabel1)
@@ -115,7 +117,7 @@ ax2.yaxis.set_ticks_position('left')
 plt.tight_layout()
 
 
-print('T-test pvalues: Nam,PRPP,ATP')
+'''print('T-test pvalues: Nam,PRPP,ATP')
 print('mu 1 vs. mu 3', stats.ttest_ind(muall1[0], muall1[2]).pvalue)
 print('mu 1 vs. mu 2', stats.ttest_ind(muall1[0], muall1[1]).pvalue)
 print('mu 2 vs. mu 3', stats.ttest_ind(muall1[1], muall1[2]).pvalue)
[…]

### 2018-06-25 figures/NamPRT_activity/make_figure.py
Calculated kcat and updated figures

@@ -18,10 +18,8 @@ plt.rc('ytick', labelsize=fs)  # fontsize of the tick labels
 plt.rc('legend', fontsize=fs)  # legend fontsize
 plt.rc('figure', titlesize=fs) # fontsize of the figure title
 
-#color = sns.color_palette("colorblind", n_colors=2)
-
 majorlabel = ('wt NamPRT', '$\Delta$42-51 NamPRT')
-minorlabel1 = ('Nam 1\nPRPP 1\nATP 0', '1\n0.1\n0', '0.1\n1\n0')
+minorlabel1 = ('Nam 1\nPRPP 1\nATP 0', '0.1\n1\n0')
 minorlabel2 = ('0.1\n0.1\n0', '0.1\n0.1\n1')
 
 wt1 = []
@@ -41,15 +39,13 @@ muall2 = []
 with open('activity.csv') as f:
 	next(f)
 	next(f)
-	for i in range(len(minorlabel1)):
-		if i == 1:
-			continue
+	for _ in minorlabel1:
 		wtbuf = []
 		mubuf = []
 		for _ in range(3):
 			d = next(f).rstrip().replace(',', '.').split('\t')
-			wtbuf.append(float(d[5]))
-			mubuf.append(float(d[6]))
+			wtbuf.append(float(d[7]))
+			mubuf.append(float(d[8]))
 		wt1.append(np.mean(wtbuf))
 		mu1.append(np.mean(mubuf))
 		wtstd1.append(np.std(wtbuf))
@@ -63,8 +59,8 @@ with open('activity.csv') as f:
 		mubuf = []
 		for _ in range(3):
 			d = next(f).rstrip().replace(',', '.').split('\t')
-			wtbuf.append(float(d[5]))
-			mubuf.append(float(d[6]))
+			wtbuf.append(float(d[7]))
+			mubuf.append(float(d[8]))
 		if rel is None:
 			rel = np.mean(wtbuf)
 		wt2.append(np.mean(wtbuf)/rel)
@@ -73,8 +69,6 @@ with open('activity.csv') as f:
 		mustd2.append(np.std(mubuf)/rel)
 		wtall2.append(tuple(wtbuf))
 		muall2.append(tuple(mubuf))
-
-minorlabel1 = [minorlabel1[0], minorlabel1[2]]
 
[…]

### 2018-06-25 latex/figurelegends.tex
adjusted Figure order and partially revised Figure legends

@@
 
-Evolutionary distribution of NADA, NNMT and NamPRT and their relation to the number of NAD consumers.
-A) Distribution of NADA, NNMT and NamPRT in selected major taxa.
+\textbf{Evolutionary distribution of NADA, NNMT and NamPRT and their relation to the number of NAD consumers} {A) Distribution of NADA, NNMT and NamPRT in selected major taxa.
 NADA is dominant in Bacteria, Fungi, and Plants (Viridiplantae), whereas NamPRT together with NNMT is dominant in Deuterostomia.
@@
 Below the taxon name, the number of species in that taxon is given.
-
 B) Common tree of selected taxa within the Metazoa, including 334 species.
@@
 
-NAD consumption flux and NAD concentration in simulations of organisms with NADA(?) and with and without NNMT at different Nam import and cell division rates.
+\textbf{NNMT enables high NAD-consumption flux} We used a dynamic model of NAD-biosynthesis and consumption (details see Materials and Methods)  to simulate NAD consumption flux (A) and NAD concentration (B) in the presence of NamPRT and with or without NNMT at different cell division rates.
+NMMT enables higher steady state NAD-consumption flux despite reduced NAD-concentrations.
 
 \subsection{Figure 4}
-
-Impact of NNMT on NAD consumption and NAD concentration in simulations of organisms with NADA or NamPRT.
+\textbf{Role of NamPRT affinity for Nam} Using the dynamic model of NAD-biosynthesis and consumption we simulated the affect of different Michaelis Menten constants ($K_M$)  of NamPRT for Nam on steady state NAD consumption flux and NAD concentration at different cell devision rates.
+In the absence of NNMT (A and B) The $K_M$ of NamPRT has little influence on NAD consumption and concentration, but both are changing with cell devision rates.
+In the presence of NNMT (C and D), decreasing $K_M$ of NamPRT enables incresing NAD consumption flux and NAD concentration.
+NNMT furthermore makes both relatively independent of cell devision rates.
+Comparing the situation with and without NNMT (E and F) at different NamPRT $K_M$ reveals that at low $K_M$ and high cell devision rates NNMT no longer enables higher NAD consumption rates compared to NamPRT alone
 
 \subsection{Figure 5}
-
-NAD consumption flux and NAD concentration in simulations of organisms with NamPRT and with and without NNMT at different KMs for NamPRT and cell division rates.
-%%% TeX-master: "manuscript"
-
-\subsection{Figure 6}
 
@@
 
-\subsection{Figure 7}
+\subsection{Figure 6}
 
@@
 The visualisation is based on a structure prediction of SWISS-MODEL \cite{Arnold2006,Biasini2014} of the sequence of the human NamPRT (P43490) using the model 2H3D as template \cite{Wang2006}.
+
+\subsection{Figure 7}

### 2018-06-25 latex/figurelegends.tex
first draft figure legends

@@
 
-The substrate affinity of NNMT and NamPRT have opposite effects on NAD consumption (A) and concentration (B), as would be expected.
-The affinities previously measured for human enzymes (indicated by a black asterisk) appear to be close to optimal, as further improvements would have little or no effect on NAD consumption or concentration.
+\textbf{The substrate affinities of human NNMT and NamPRT are optimal}.
+Both  NAD consumption (A) and NAD concentration (B) are increasing with decreasing $K_M$ of NamPRT but decreeasing $K_M$ of NNMT.
+The affinities reported for human enzymes (indicated by a black asterisk) appear to be close to optimal, as further improvements would have little or no effect on NAD consumption or concentration.
 
@@
 
-Deuterostomia that encode NNMT show a sequence insertion in the N-terminal region of NamPRT.
-A) Multiple sequence alignment of NamPRT of selected deuterostomes show a sequence insertion in organisms that encode NamPRT and NNMT.
+\textbf{The function of the structurally unresolved loop structure of NamPRT} Most deuterostomia that encode NNMT show a sequence insertion in the N-terminal region of NamPRT that has been revealed by multiple sequence alignment of NamPRT from different species (A).
 Coloured circles indicate the enzymes present in the species; blue: NamPRT and NNMT; black: NamPRT, NADA, and NNMT; yellow: NamPRT and NADA.
-
-B) The inserted region is not resolved in crystal structures of human NamPRT and thus appears to be a flexible loop structure at the surface of the NamPRT dimer, coloured in red.
+The inserted region is not resolved in crystal structures of human NamPRT and thus appears to be a flexible loop structure at the surface of the NamPRT dimer, coloured in red  (B) .
 The visualisation is based on a structure prediction of SWISS-MODEL \cite{Arnold2006,Biasini2014} of the sequence of the human NamPRT (P43490) using the model 2H3D as template \cite{Wang2006}.
+The localisation of the FLAG-tagged mutant protein is not changed (E) compared to NamPRT-wildtype.
+Both show a heterogenous nuclear cytosolic localisation in immunofluerescence images in ....
+cells (C).
+But, in vitro measurements using recombinant protein show  that a mutant of NamPRT missing this insertion has lower activity (D)  and is not activated by ATP (E).
 
 \subsection{Figure 7}
+\textbf{NNMT provides a competitive advantage and makes NADA obsolete} To simulate competition for common resources we created a two compartment model, where one compartment contained NADA but no NamPRT and one compartment contained NamPRT either with or withour NNMT but no NADA.
+In the absence of NNMT (A-B) the compartment containing NADA has slightly lower NAD-consumption rates (A) but much higher NAD concentrations (B)  In the presence of NNMT, however, both NAD consumption (C) and NAD concentration (D) are lower in the NADA compartment, but this effect is only observed at low NamPRT $K_M$.

### 2018-06-25 latex/results.tex
updated results

@@
 We performed multiple sequence analysis to see if we can find sequence variations in the protein sequence of NamPRT that indicate evolutionary changes of NamPRT upon the appearance of NNMT.
-The multiple sequence alignment of selected sequences is shown in Figure 5A and Supplementary Figure ?\todo[author=Ines]{Decide figure number} We recognized that Deuterostomia that have only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 43 to 52 of the human enzyme, that overlaps with a predicted weak nuclear localisation signal, that is lost when removing this sequence stretch.
-Furthermore, the inserted sequence corresponce to an structurallly unresolved loop structure in all avaibable crystal structure of human NamPR (e.g.
+The multiple sequence alignment of selected sequences is shown in Figure 5A.
+A more comprehensive multiple sequence alignment containing a larger number of species can be found in Supplementary Figure 1.
+We recognized that most Deuterostomia that have only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 43 to 52 of the human enzyme.
+This insert overlaps with a predicted weak nuclear localisation signal, that is lost when removing this sequence stretch.
+The inserted sequence corresponce to a structurallly unresolved loop structure in all avaibable crystal structure of human NamPR (e.g.
 \cite{Wang2006} structure visualisation Figure 5B).
@@
 From these observations we derived two hypotheses regarding the potential function of the loop for NamPRT function.
-Firstly, we analysed whether the deletion of the amino acids 43 to 52 has an effect on the localisation of the human enzyme.
-But both wildtype and mutant protein show a mixed localisation in the cytosol and nucleus (Figure 5C).
+Firstly, we analysed whether the deletion of the amino acids 42 to 51 has an effect on the localisation of the human enzyme.
+But both wildtype and mutant protein show a mixed localisation in the cytosol and nucleus (Figure 6C).
 We therefore conclude that the partial nuclear localisation of NamPRT is not affected by the sequence deletion.
@@
 Upon purifictaion we meassured the activity using NMR detecttion of NMN produced in the presence and absence of ATP.
-It appears that the enzymatic activity of the mutant enzyme is much lower than that of the wildtype enzyme both with and without ATP.
-We furthermore see that the activity of the mutant enzyme can be increase with higher concentration of Nam and PRPP whereas that is not the case for the wildtype enzyme.
-Pointing to a decrease substrate affinity of the mutant enzyme.
-Supporting the second hypothesis and thus supporting the predictions derived from our mathematical modelling approach.
+It appears that the enzymatic activity of the mutant enzyme is much lower than that of the wildtype enzyme (Figure 6D).
+We furthermore see that in contrast to NamPRT wildtype, the mutant does not show an increased activity in the presence of ATP (Figure 6E).
 
@@
 
-But why is NADA lost in vertebrates?
-Our simulations show that in direct competition NADA is outcompeted by the combined presence of NamPRT and NNMT but not by NamPRT alone.
-This effect deisappears of NamPRT affinity is lowered.
-....
+But why is NADA lost in vertebrates while NADA and NamPRT are co-existing existing in bacteria?
+To analyse this question we built a two compartment model that has a shared Nam source (details see Materials an Methods and Supplementary Material).
+One compartment contains NADA while the other contains either NamPRT alone or together with NNMT.
+Without NNMT the compartment containing NADA has slightly lower NAD consumption rates (Figure 7A), but is able to maintain much higher steady state NAD concentrations at low cell division rates  (Figure 7B).
+At high cell division rates, however, steady state concentration in both compartments are similar.
+The latter might explain why in bacteria that have relatively high growth rates both systems coexists.
+If in contrast, NNMT is present in the NamPRT compartment, this compartment outcompetes the NADA containing compartment, both with respect to NAD-consumption rates and steady state NAD concentrations (Figure 7 C and D).
+This effect is dependent on a high affinity of NamPRT, as at low affinity the NADA containing system again maintains higher NAD-concentrations, the NAD consumption flux still being higher  in the NamPRT and NNMT containing compartment.
+Taken together this might explain why upon the development of a sufficiently high affinity of NamPRT for its substrate, which seems to have been induced by the appearance of NNMT, NADA is lost.

### 2018-06-25 latex/discussion.tex
removed old text and subsectionf of discussion ... needs to be rewritten

@@
 
-\subsection{Evolutionary and physiological role of NNMT}
-
 The dynamic interactions among metabolism, signal transduction, and gene regulation is still not very well understood.
 New approaches are required to disentangle the underlying network, helping us to understand how alterations affect human physiology.
-We here combine phylogenetic analysis and mathematical modelling of a central metabolic pathway supported by experimental analysis to reveal the likely evolutionary development of the NAD biosynthesis and consumption pathway and the physiological role of the methyltransferase NNMT.
+We here combined phylogenetic analysis and mathematical modelling of a central metabolic pathway supported by experimental analysis to reveal the likely evolutionary development of the NAD biosynthesis and consumption pathway and the physiological role of the methyltransferase NNMT.
 
@@
 The affinities measured for the human enzymes seem to be optimal and further increase in NAD consumption rates might only enabled by compartmentalisation of NAD biosynthesis.
-
-In contrast to our results, it was proposed that NNMT is present in yeast (\textit{Saccharomyces cerevisiae} \cite{Anderson2003}.
-The enzyme was later further characterised as methyltransferase of the transcription elongation factors TEF1 and TEF2 but not Nam \cite{Wlodarski2011}.
-
-\subsection{General applicability}
 
@@
 
-\subsection{Limitations of our analysis and outlook}
-
 Among others, we neglected the potential effects of the methyl donor S-adenosyl-methionin (SAM) and its precursor methionine in our analysis, although it is most likely contributing yet another regulatory level for Nam availability and thus an additional interaction point between gene regulation and metabolism.
@@
 The same holds for the analysis of enzyme acetylation versus deacetylation by sirtuins that would enable prediction about enzyme activation or histone state if available.
+
+\section{Conclusions}
+
+Beside the evolutionary changes analysed here we have detected further evolutionary events in the development of NAD biosynthesis.
+We noted for example that the unresolved loop insertion in NamPRT co-appears tigether with the first gene duplication of the NMNAT and thus the development of NMNAT2 \cite{Lau2010}.
+At the point of the second gene duplication of NMNAT and thus the development of NNMT3 coincides with a positive selection event in NNMT (overview Figure 8) \todo{I think we should leave this for a later manuscript}.

### 2018-06-25 latex/manuscript.tex
Decreased font size of todo notes. Updated pdf
Split methods and discussion; Updated affiliations; added some references
removed old text and subsectionf of discussion ... needs to be rewritten

@@
 \usepackage[left=1cm,right=5cm,top=2cm,bottom=2cm,nohead,nofoot]{geometry}
-\usepackage{todonotes}
+\usepackage[textsize=scriptsize]{todonotes}
 
@@
 \noindent
-$^{1}$Department of Arctic and Marine Biology, UiT Arctic University of Norway, Naturfagbygget, Dramsveien 201, 9037 Tromsø, Norway
+$^{1}$Department of Arctic and Marine Biology, UiT The Arctic University of Norway, Biologibygget, Framstredet 39, 9017 Tromsø, Norway
 
@@
 
+\input{methods}
+
 \input{figurelegends}
-
-\section{Comments and parts of old version not included yet}
-
-Chordata: NADA disappears.
-These analyses also support the view that Tunicata and Branchiostoma are not part of the phylum chordata\todo[author=Mathias]{Is this view in any way argued upon by anyone??}.
-
-In the evolutionary context, an additional question arises: Why do only a few organisms, mostly less complex animals, possess the gene for NADA in addition to NamPRT and NNMT.
-When including NADA into the simulations, we see that the effect of NADA on NAD consumption is very limited in the presence of NNMT (old Figure 4F) even at high expression levels (see old Suppl.
-Figure S1C and D) and could be compensated by increased expression of NamPRT (not shown).
-
-Until this point we have neglected compartmentalisation of the pathway.
-We do however know from previous studies that in early vertebrate development a compartmentalisation of the pathway has occurred reflected by a gene triplication of NMNAT and the occurrence of compartment-specific domains called ISTIDs (Lau, 2010).
-Looking at the evolutionary timepoint \todo[author=Toni]{This time point can only be reconstructed, but is unknown.
-I think a figure with the likely reconstruction would help to follow the argumentation} of appearance of NNMT and the gene triplication of NMNATs, we see that NNMT occurs prior to the gene triplication and the first occurrence of ISTIDs and we have confirmed this in our own analysis (not shown).
 

### 2018-06-26 latex/introduction.tex
edited introduction

--- 2018-06-20 latex/introduction.tex
@@
 Even though best known as cofactor for various redox-reactions , NAD is involved in a number of signalling processes that consume NAD+ by cleaving the molecule to nicotinamide (Nam) and ADP-ribose \cite{Verdin2015}.
-These NAD+-dependent signalling reactions include but are not limited to poly- and mono-ADP-ribosylation, NAD+-dependent protein deacylation by sirtuins as well as the synthesis of calcium mobilizing molecules such as cyclic ADP-ribose \cite{Opitz2015}.
+These NAD+-dependent signalling reactions include but are not limited to poly- and mono-ADP-ribosylation, NAD+-dependent protein deacylation by sirtuins as well as the synthesis of calcium mobilizing molecules such as cyclic ADP-ribose \cite{?}.
 These NAD+-dependent signalling processes participate in the regulation of virtually all cellular activities.
 The enzymes involved in these processes are sensitive to the available NAD+ concentration \cite{Ruggieri2015}, which in turn is dependent on the NAD+/NADH redox ratio.
-Therefore, NAD+-dependent signalling can act as a transmitter of changes in the cellular energy homeostasis, for example, to regulate gene expression or metabolic activity \cite{Koch-Nolte2009}.
+Therefore, NAD+-dependent signalling can act as a transmitter of changes in the cellular energy homeostasis, for example, to regulate gene expression or metabolic activity \cite{Koch-Nolte2009}.\todo[author=Mathias B.]{Please add more references to the first paragraph.
+I think there are better references then my own review.}
 
@@
 To maintain the NAD concentration at physiological levels, NAD biosynthesis needs to act at an equally rapid rate.
-Imbalances in NAD-homeostasis have been linked to various, in particular, ageing-related diseases such as diabetes, neurodegenerative disorders and cancer \cite{Chiarugi2012,Verdin2015}\todo[author=Ines]{Potentially adjust ref}.
-Several recent studies have demonstrated impressive health benefits of dietary supplementation with intermediates of NAD biosynthesis including NMN and nicotinamide riboside, NR \cite{Yoshino2018}.
+Imbalances in NAD-homeostasis have been linked to various, in particular, ageing-related diseases such as diabetes, neurodegenerative disorders and cancer \cite{Chiarugi2012,Verdin2015}.
+Several recent studies have demonstrated impressive health benefits of dietary supplementation with intermediates of NAD biosynthesis including NMN and nicotinamide riboside, (NR) \cite{Yoshino2018}.
 Apparently, the exploitation of NAD biosynthetic routes, in addition to the use of nicotinamide as precursor (Fig.
@@
 
-In mammals, NAD biosynthesis predominantly relies on nicotinamide \todo[author=Mathias Z.]{Please add references}, a form of vitamin B3, as precursor.
-In the first step, Nam is converted to the mononucleotide, NMN, by Nam phosphoribosyltransferase (NamPRT) using phosphoribosylpyrophosphate (PRPP) as co-substrate.
-The key role of this enzyme is related to the nature of NAD+-dependent signalling reactions, namely, that in all these reactions Nam is cleaved off and potentially lost.
-If the released Nam were not recycled into NAD biosynthesis, humans would require a much higher daily vitamin B3 intake than the 16 mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}.
+Due to the constant release of Nam through NAD-consuming signalling reactions, the NAD salvage pathway using Nam as precursor is the most important NAD-synthesis pathway.
+If Nam would not be constantly recycled into NAD, humans would require a much higher daily vitamin B3 intake than the 16 mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}.
+In the first step of the salvage pathway, Nam is converted to the mononucleotide, NMN, by Nam phosphoribosyltransferase (NamPRT) using phosphoribosylpyrophosphate (PRPP) as co-substrate.
 The nearly complete recycling of Nam is achieved by an extraordinary high affinity of NamPRT to Nam, the Km being in the low nanomolar range \cite{Burgos2008}.
 Despite the importance of its salvage, Nam can also be marked for excretion by methylation.
-Indeed, the presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012} is among the most enigmatic and counterintuitive features of NAD metabolism.
+The presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012FEBS} is among the most enigmatic and counterintuitive features of NAD metabolism.
 Why is there one enzyme (NamPRT) seemingly optimized to recycle even the faintest amounts of Nam back into NAD synthesis, while at the same time there is another enzyme (NNMT) that seems to have no metabolic function other than to remove Nam from NAD metabolism, This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deaminated Nam to nicotinic acid (NA) using the nicotinamide deamidase (NADA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (Fig.
@@
 
-We here present a phylogenetic analys of NAD-pathway in eukaryotes that in contrast to our previous analysis that has been lmited by the low number of eularyotic genomes available at the time, now comprehensively analysis the evolution of the pathway.
-In the present study, we furthermore combine the phylogenetic analysis with a mathematical modelling approach.
-This has proven to be a powerful tool to analyse the relationship between NAD biosynthetic and signalling fluxes when different subsets of Nam-converting enzymes are present.
-that enable us to explain the counterintuitive pathway evolution, such as that in Deuterostomes, there has been a strong tendency towards the co-existence of NamPRT and NNMT, while the pathway dominant in bacteria is lost in vertebrates.
-Moreover, this selection for the co-existence of NamPRT and NNMT was accompanied by a marked increase in the number of genes encoding NAD+-consuming signalling reactions.
-The model demonstrated that NNMT has a critical role to maintain high NAD+-consuming signalling fluxes by preventing accumulation of inhibitory Nam.
-We show that the presence of high affinity NamPRT together with NNMT makes the NADA dependent pathway obsolete.
-We furthermore identify a sequence insertion only found in Deuterostomes that encode NNMT but not NADA and experimentally verify that this insertion is indeed changing the affinity of NamPRT.
+We here present a phylogenetic analys of NAD-pathway in eukaryotes that.
+In contrast to our previous analysis, which has been lmited by the low number of eularyotic genomes available at the time, we are now able to provide a comprehensively analysis of the eukaryotich pathway evolution.
+The results show there has seemingly been a selection for the co-existence of NamPRT and NNMT in Deuterostomia, while the pathway dominant in bacteria is lost.
+This was accompanied by a marked increase in the number of  NAD+-consuming signalling enzymes.
+To explain these counterintuitive results we built a mathematical model of the pathway and demonstratet that NNMT has a critical role to maintain high NAD+-consuming signalling fluxes by preventing accumulation of inhibitory Nam.
+Our model furthermore predicts that NNMT likely exerted an evolutionary pressure on the NamPRT affinity development.
+Simulating the resource competition we furthermore show that the presence of high affinity NamPRT together with NNMT makes the NADA dependent pathway obsolete.
+Based on multiple sequence alignments,  we identified a sequence insertion in NamPRT in Deuterostomes  that is affecting the affinity of NamPRT,  undermining the predictions derived from our mathematical modelling approach
 
-Taken together, our analyses suggest that the co-existence of NamPRT and NNMT has been a prerequisite to enable the evolutionary development of versatile NAD+-dependent signalling mechanisms present in vertebrates, and that NNMT likely exerted an evolutionary pressure on the NamPRT affinity development.
+Taken together, our analyses suggest that the co-existence of NamPRT and NNMT has been a prerequisite to enable the evolutionary development of versatile NAD+-dependent signalling mechanisms present in vertebrates.

### 2018-06-26 latex/manuscript.tex
edited introduction

@@
 
-Strikingly, loss of the deamidation pathway in early vertebrates is accompanied by the occurrence of a new enzyme that marks Nam for excretion by methylation – nicotinamide N-methyltransferase (NNMT).
+Strikingly, loss of the deamidation pathway in early vertebrates is preceded by the occurrence of a new enzyme that marks Nam for excretion by methylation – nicotinamide N-methyltransferase (NNMT).
 The physiological relevance of NNMT is still enigmatic.

### 2018-06-26 supplements/KineticsTable/Table_S1.tex
Added symlinks to all figures and tables

\documentclass[a4paper,10pt]{article} \usepackage[utf8x]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{booktabs,colortbl,tabularx}
\usepackage{longtable}
\usepackage{float}
\usepackage{footmisc}
\usepackage{breqn}
\restylefloat{table}

\usepackage[pdftex,bookmarks=false,
pdftex,
%linkcolor=black,
linkcolor=blue,
citecolor=blue,
colorlinks=true,                %%% hyper-references for pdflatex
urlcolor=blue]{hyperref}

\begin{document}

\section*{Supplementary Table 1}
\addcontentsline{toc}{section}{Supplementary Table 1}
\begin{longtable}{p{3cm}p{2cm}p{3cm}cc}
\multicolumn{5}{l}{\textbf{Overview of kinetic constants used for the
construction of the model.}} \cr\toprule
  \cr\textbf{Enzyme} &
  \textbf{EC\linebreak number}&
  \textbf{Kinetic\linebreak parameter} &
  \textbf{References} &
  \textbf{Rate Law} \\
\cr\hline
\endfirsthead
\multicolumn{5}{c}{Table 1 -- Continued}
\cr\hline
\cr\textbf{Enzyme} &
 \textbf{EC\linebreak number}&
  \textbf{Kinetic\linebreak parameter} &
  \textbf{Reference} &
  \textbf{Rate Law} \\
\cr\hline
\endhead
NADA & \href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC3/5/1/19.html}{3.5.1.19} &$K_M$:9.6$\mu$M  & \cite{Smith2012} & Product inhibition\\
& & $K_{iP}$:120$\mu$M  & & 
    \\
    & & $k_{cat}$:0.65$s^{-1}$  & & 
    \\
    \hline
NADS & \href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC6/3/5/1.html}{6.3.5.1}
&$K_M$:190$\mu$M  & \cite{Yi1972} & HMM\\
    & & $k_{cat}$:21$s^{-1}$  & & 
    \\  \hline
NMNAT &
\href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC2/7/7/1.html}{2.7.7.1}/\href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC2/7/7/18.html}{2.7.7.18}
&$K_{M_{NaMN}}$:67.7$\mu$M  & \cite{Sorci2007}\footnote{Values for NMNAT1 used}
& Substrate Competition\\
    & & $k_{{cat}_{NaMN}}$:42.9$s^{-1}$  & & \\
    & & $K_{M_{NMN}}$:22.3$\mu$M & & \\
    & & $k_{{cat}_{NMN}}$:53.8$s^{-1}$  & & \\
	& & $K_{M_{NMN}}$:59$\mu$M  & & \\
    & & $k_{{cat}_{NAD}}$:129.1$s^{-1}$  &\cite{Berger2005}\footnote{Keq used
    for calculation of turnover rate of reverse reaction} &
    \\
    & & $K_{M_{NaAD}}$:502$\mu$M  & & \\
    & & $k_{{cat}_{NaAD}}$:103.8$s^{-1}$  &\cite{Berger2005}\footnote{Equilibrium constant used for calculation of turnover rate of reverse reaction} &
    \\ \hline
NMNT & \href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC2/1/1/1.html}{2.1.1.1}
&$K_M$:400$\mu$M  & \cite{Aksoy1994} & Product inhibition\\
& & $K_{iP}$:60$\mu$M  & & 
    \\
    & & $k_{cat}$:8.1$s^{-1}$  & \cite{Alston1988}& 
    \\  \hline
NamPRT & \href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC6/3/5/1.html}{6.3.5.1}
&$K_M$:5nM  & \cite{Burgos2008} & HMM\\
    & & $k_{cat}$:0.0077$s^{-1}$  & & 
    \\  \hline
    NAPRT &
       \href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC2/4/2/11.html}{2.4.2.11}
    &$K_M$:23$\mu$M  & \cite{Hanna1983} & HMM\\
    & & $k_{cat}$:3.3$s^{-1}$  & & 
    \\  \hline
    SIRT1 &
    \href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC3/5/1/index.html}{3.5.1.-}
    &$K_M$:29$\mu$M  & \cite{Borra2004} & Product inhibition\\
    & & $K_{iP}$:60$\mu$M  & & 
    \\
    & & $k_{cat}$:0.67$s^{-1}$  & & 
    \\  \hline
    NT5 &
    \href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC3/1/3/5.html}{3.1.3.5}
    &$K_M$:100$\mu$M  & \cite{} & HMM \\
    & & $k_{cat}$:1$s^{-1}$ 
    & & \\  \hline   
  \bottomrule
  \label{tab:kinetic}
\end{longtable}

The total enzyme concentration was set to 10 for all enzymes if not mentioned
otherwise. Concentration of potentiantial cosubstrate was assumed to be constant
and considered to be not limitng the reaction and thus represented by the
maximal velocities given.

\subsection*{Kinetic Rate Laws}

\subsubsection*{Product Inhibition}
\begin{equation}
v=\cfrac{E_T\cdot k_{cat}\cdot S}{K_M + S + \cfrac{K_M\cdot P}{K_{iP}} }
\end{equation}

\subsubsection*{Henry-Michaelis Menten for irreversible reactions (HMM)}
\begin{equation}
v=\frac{E_T\cdot k_{cat}\cdot S}{K_M + S}
\end{equation}

\subsubsection*{Substrate Competition at NMNAT}
\begin{equation}
v=E_T \cdot \cfrac{\cfrac{k_{{cat}_{A}}\cdot A \cdot
B}{K_{{M}_A}}-\cfrac{k_{{cat}_{P}}\cdot P \cdot
Q}{K_{{M}_P}}}{1+\cfrac{A}{K_{{M}_A}}+\cfrac{B}{K_{{M}_B}}+\cfrac{P}{K_{{M}_P}}+\cfrac{Q}{K_{{M}_Q}}}
\end{equation}

\bibliographystyle{plos2009}
\bibliography{TableS1}
\end{document}

### 2018-06-26 latex/results.tex
edited Figure 3 and results section

@@
 
-As shown in Fig.
-1, NAD can be synthesized using several routes from three main precursors: tryptophan, nicotinamide (Nam) and nicotinic acid (NA).
+NAD can be synthesized using several routes from three main precursors: tryptophan, nicotinamide (Nam) and nicotinic acid (NA) (Overview see Fig 1.).
 Nam and NA are together known as vitamin B3 or niacin.
 Alternatively, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead \cite{Yoshino2018}.
-Du to the hight turnover of NAD observed and the fact that only 1\% of the tryptophan taken up with our diet is converted into NAD, vitamin B3 (Nam and NA) and to a lower extend NR are the major precursor of NAD-biosynthesis.
+Du to the hight turnover of NAD observed and the fact that only 1\% of the tryptophan taken up with our diet is converted into NAD, vitamin B3 (Nam and NA) and to a lower extend NR are the major precursor of human NAD-biosynthesis.
 
-For the recycling of Nam, two different pathways exist.
-The pathway found in yeast and plants is using a four-step pathway starting with the deamination of Nam to nicotinic acid by NADA.
+Looking acroos all known species,   two different pathways exist that recylce NAM.
+The major pathway found in yeast and plants is using a four-step pathway starting with the deamination of Nam to nicotinic acid by NADA.
 The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates.
@@
 NNMT seems to have arisen de novo in the common ancestor of Ecdysozoa and Lophotrochozo.
-We were not be able to find any gene with considerable similarity in fungi or plants, even though in fungi a gene named NNMT can be found in databases.
-These genes do, however, show very limited sequence similarities to the NNMT of nematodes or deuterostomes and in the yeast protein has later on be found to be a lysin-protein-methyltransferase \cite{Wlodarski2011}.
-Nematodes are the only organisms where we find NNMT together with NADA without NamPRT being present.
+We were not be able to find any gene with considerable similarity in fungi or plants, even though in fungi a genes named NNMT can be found in databases.
+These genes do, however, show very limited sequence similarities to the NNMT of nematodes or deuterostomes and the yeast protein has later on shown to be a lysin-protein-methyltransferase \cite{Wlodarski2011}.
+Nematodes are the only organisms that encode NNMT together with NADA without NamPRT being present.
 In Deuterostomia the only large class that does only have NamPRT and seems to have lost NNMT again are Sauropsida and here especially birds.
@@
 In addition to the phylogenetic distribution of the two Nam salvage enzymes NADA and NamPRT, we looked at phylogenetic diversity of enzymes catalysing NAD-dependent signalling reactions.
-To do so we used the previously established classification into 10 different families \cite{Gossmann2012FEBS} (For details see materials and methods and supplementary information).
+To do so we used the previously established classification into 10 different enzyme families \cite{Gossmann2012FEBS}.
+The detailed list of templates used for the phylogenetic analysis can be found in  Supplementary Table 1?
+\todo[author=Mathias]{Inlcude number and order and name Supplementary Tables accordingly}.
 The numbers in Figure 2B denote the average number of NAD-dependent signalling enzyme families found in each taxa.
@@
 Taken together, we found that the presence of NamPRT and NNMT coincides with an increased diversification of NAD-dependent signalling, whilst NADA is lost in vertebrates.
-That is  counterintuitive, as one would expect that a decrease in precursor concentration caused by the precursor removal through NNMT, should cause a decrease of NAD availability and consequently less active NAD dependent signalling.
+This seems  counterintuitive, as one would expect that a decrease in precursor concentration caused by the precursor removal through NNMT, should cause a decrease of NAD availability and consequently less active NAD dependent signalling.
 
@@
 And why does NADA disappear in Deuterostomia although it is the predominant pathway in bacteria, plants and fungi?
-Given the complexity of the NAD metabolic network, this question is difficult to be comprehensively addressed experimentally.
-Thus, to answer these questions we built a dynamic model of NAD metabolism using existing kinetic data from the literature (for details see materials and methods and supplementary material).
+Given the complexity of the NAD metabolic network, this questions are difficult to be comprehensively addressed experimentally.
+We therefore built a dynamic model of NAD metabolism using existing kinetic data from the literature (for details see materials and methods and supplementary table 2).
 
@@
 First we addressed the question of the predominant coexitence of NamPRT and NNMT in vertebrates that coincides with an increase in the number of NAD-consuming enzyme families.
-We thus used our dynamic model of the NAD biosynthesis and consmption pathway to simulate the effect of the  presence or absence of NNMT.
+We thus used our dynamic model to simulate steady state concentrations and NAD-consumption rates simulating NAD biosynthesis via  NamPRT in the  presence or absence of NNMT.
 We see that the presence of NNMT enables higher NAD consumption rates  (Figure3A) while, as exptected, reducing the steady state concentration of NAD (Figure 3B).
-The decline in NAD-concnetrtion can be compensated by a higher expression of NamPRT (not shown) further increasing NAD consumption flux.
+The decline in NAD-concentration can be compensated by a higher expression of NamPRT further increasing NAD consumption flux.
 

### 2018-06-26 latex/figurelegends.tex

--- 2018-06-21 latex/figurelegends.tex
@@
 
-Schematic overview of NAD biosynthesis and consumption.
-NAD can be synthesized using several routes from altogether three main precursors: tryptophan (Trp), nicotinamide (Nam), and nicotinic acid (NA).
-Nam and NA are together known as vitamin B3 or niacin.
-To a lesser extend, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead.
-Only 1\% of the tryptophan taken up with our diet is converted into NAD, thus, vitamin B3 and, to a lower extend, NR are essential components of our diet, with Nam being the major NAD precursor in humans\todo{Whole sentence (starting with "Only") not necessarily part of the legend.
-What do you think, Ines?}.
-Nam is furthermore the product of NAD-consuming signalling reactions such as sirtuins (NAD-dependent histone deacetylases) or PARPs (poly-ADP-ribosylases).
-
+\textbf{Schematic overview of NAD biosynthesis and consumption.} NAD can be synthesized from tryptophan (Trp), nicotinamide (Nam), nicotinic acid (NA), and, to a lesser extend, nicotinamide ribose (NR).
+Nam is the main precursor and also the product of NAD-consuming signalling reactions by enzymes such as sirtuins (NAD-dependent histone deacetylases) or PARPs (poly-ADP-ribosylases).
 For the recycling of Nam, two different pathways exist.
-The pathway found in yeast and many bacteria is using a four-step pathway starting with the deamination of Nam to nicotinic acid (NA) by Nam deamidase (NADA).
+The pathway found in yeast and many bacteria starts with the deamination of Nam by Nam deamidase (NADA).
 The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates.
-The recycling pathway found in vertebrates directly converts Nam into the corresponding mononucleotide (NMN) a reaction catalysed by NamPRT and driven by a non-stoichiometric ATP-hydrolysis.
-A similar reaction catalysed by an evolutionary related enzyme NAPRT, converts NA into the NA mononucleotide (NAMN).
-NMN and NAMN are converted into dinucleotides by the Nam/NA adenylytransferases (NMNATs).
-The recycling pathway via NA finally requires an amination step catalysed by NADsynthase (NADS), driven by the conversion of ATP to AMP producing pyrophosphate.
+The pathway found in vertebrates directly converts Nam into the corresponding mononucleotide (NMN) by the Nam phosphoribosyltransferase (NamPRT).
+A third enzyme can consume Nam, the Nam N-methyltransferase (NNMT).
+For more details and abbreviations, see text.
 
@@
 
-Evolutionary distribution of NADA, NNMT and NamPRT and their relation to the number of NAD consumers.
-A) Distribution of NADA, NNMT and NamPRT in selected major taxa.
-NADA is dominant in Bacteria, Fungi, and Plants (Viridiplantae), whereas NamPRT together with NNMT is dominant in Deuterostomia.
+\textbf{Evolutionary distribution of NADA, NNMT and NamPRT and their relation to the number of NAD consumers.} A) Distribution of NADA, NNMT and NamPRT in selected major taxa.
+NADA is dominant in Bacteria, Fungi, and Plants (Viridiplantae), whereas NamPRT together with NNMT is dominant in Metazoa.
 Numbers at the pie charts show, how many species of the taxon possess the respective enzyme combination indicated by the colour explained in the lower right of the figure.
 Below the taxon name, the number of species in that taxon is given.
-
 B) Common tree of selected taxa within the Metazoa, including 334 species.
-The pie charts indicate the distribution of species within the respective taxon that have the enzyme combination indicated by the colour, explained in the lower right.
+The pie charts indicate the distribution of species within the respective taxon that have the enzyme combination indicated by the colour explained in the lower right.
 The size of the pie charts is proportional to the logarithm of the number of species analysed in the particular taxon.
@@
 
-NAD consumption flux and NAD concentration in simulations of organisms with NADA(?) and with and without NNMT at different Nam import and cell division rates.
+\textbf{NNMT enables high NAD-consumption flux.} We used a dynamic model of NAD biosynthesis and consumption (details see Materials and Methods) to simulate NAD consumption flux (A) and NAD concentration (B) in the presence of NamPRT and with or without NNMT at different cell division rates.
+NMMT enables higher steady-state NAD consumption flux despite reduced NAD concentrations.
 
@@
 
-Impact of NNMT on NAD consumption and NAD concentration in simulations of organisms with NADA or NamPRT.
+\textbf{Role of NamPRT affinity for Nam.} Using the dynamic model of NAD biosynthesis and consumption we simulated the affect of different Michaelis Menten constants ($K_M$) of NamPRT for Nam on steady-state NAD consumption flux and NAD concentration at different cell devision rates.
+In the absence of NNMT (A and B), the $K_M$ of NamPRT has little influence on NAD consumption and concentration, but both are changing with cell devision rates.
+In the presence of NNMT (C and D), decreasing $K_M$ of NamPRT enables increasing NAD consumption flux and NAD concentration.
+NNMT furthermore makes both, consumption flux and concentration, relatively independent of cell division rates.
+Comparing the situation with and without NNMT (E and F) at different NamPRT $K_M$ reveals that at low $K_M$ and high cell devision rates NNMT no longer enables higher NAD consumption rates compared to NamPRT alone.
 
@@
 
-NAD consumption flux and NAD concentration in simulations of organisms with NamPRT and with and without NNMT at different KMs for NamPRT and cell division rates.
-%%% TeX-master: "manuscript"
+\textbf{The substrate affinities of human NNMT and NamPRT are optimal.} Both NAD consumption flux (A) and NAD concentration (B) are increasing with decreasing $K_M$ of NamPRT but decreeasing $K_M$ of NNMT.
+The affinities reported for human enzymes (indicated by a black asterisk) appear to be close to optimal, as further improvements would have little or no effect on NAD consumption or concentration.
 
@@
 
-The substrate affinity of NNMT and NamPRT have opposite effects on NAD consumption (A) and concentration (B), as would be expected.
-The affinities previously measured for human enzymes (indicated by a black asterisk) appear to be close to optimal, as further improvements would have little or no effect on NAD consumption or concentration.
+\textbf{The function of the structurally unresolved loop structure of NamPRT.} Most Deuterostomia that encode NNMT show a sequence insertion in the N-terminal region of NamPRT that has been revealed by multiple sequence alignment of NamPRT from different species (A).
+Coloured circles indicate the enzymes present in the species besides NamPRT; blue: NNMT; black: NADA and NNMT; yellow: NADA.
+B) The visualisation of human NamPRT is based on a structure prediction of SWISS-MODEL \cite{Arnold2006,Biasini2014} of the sequence of the human NamPRT using the model 2H3D as template \cite{Wang2006}.
+The inserted region is not resolved in crystal structures of NamPRT and thus appears to be a flexible loop structure at the surface of the NamPRT dimer, coloured in red.
+C) The localisation of the FLAG-tagged mutant protein lacking the unresolved loop is not changed compared to wildtype human NamPRT.
+Both show a heterogenous nuclear cytosolic localisation in immunofluerescence images in HeLa S3 cells.
+But \textit{in vitro} measurements using recombinant protein show that the mutant NamPRT has lower activity than the wildtype enzyme (D) and is not activated by ATP (E).
 
@@
 
-Deuterostomia that encode NNMT show a sequence insertion in the N-terminal region of NamPRT.
-A) Multiple sequence alignment of NamPRT of selected deuterostomes show a sequence insertion in organisms that encode NamPRT and NNMT.
-Coloured circles indicate the enzymes present in the species; blue: NamPRT and NNMT; black: NamPRT, NADA, and NNMT; yellow: NamPRT and NADA.
-
-B) The inserted region is not resolved in crystal structures of human NamPRT and thus appears to be a flexible loop structure at the surface of the NamPRT dimer, coloured in red.
-The visualisation is based on a structure prediction of SWISS-MODEL \cite{Arnold2006,Biasini2014} of the sequence of the human NamPRT (P43490) using the model 2H3D as template \cite{Wang2006}.
+\textbf{NNMT provides a competitive advantage and makes NADA obsolete.} To simulate competition for common resources, we created a two compartment model where one compartment contained NADA but no NamPRT and the other compartment contained NamPRT either with or without NNMT but no NADA.
+In the absence of NNMT (A and B) the compartment containing NADA has slightly lower NAD consumption rates (A) but much higher NAD concentrations (B).
+In the presence of NNMT, however, both NAD consumption (C) and NAD concentration (D) are lower in the NADA compartment, but this effect is only observed at low NamPRT $K_M$.

### 2018-06-26 latex/results.tex
revised results section

@@
 
-NAD can be synthesized using several routes from three main precursors: tryptophan, nicotinamide (Nam) and nicotinic acid (NA) (Overview see Fig 1.).
+NAD can be synthesized using several routes from three main precursors: tryptophan, nicotinamide (Nam) and nicotinic acid (NA) (Overview see Figure 1.).
 Nam and NA are together known as vitamin B3 or niacin.
@@
 We see that the presence of NNMT enables higher NAD consumption rates  (Figure3A) while, as exptected, reducing the steady state concentration of NAD (Figure 3B).
-The decline in NAD-concentration can be compensated by a higher expression of NamPRT further increasing NAD consumption flux.
+The decline in NAD-concentration can be compensated by a higher expression of NamPRT further increasing NAD consumption flux (Figure 3A and B dashed lines).
 
-These findings can be explained when looking in more detail into the kinetic parameters of NamPRT and NAD-consuming enzymes such as Sirt1p.
+These findings can be explained when looking in more detail at the kinetic parameters of NamPRT and NAD-consuming enzymes such as Sirt1p.
 The increase of NAD consumption flux is caused by the fact that most NAD-consuming enzymes are inhibited by their product Nam, explaing why the presence of NNMT enables higher NAD consumption fluxes.
@@
 
-As kinetic parameters of NamPRT are only available for the human enzyme \cite{Burgos2008} , we went on to theoretically analysed the potential effect of the NamPRT affinity (Km) on NAD steady state concentration and NAD consumption flux, leaving all other kinetic parameters constant.
-In the absence of NNMT (Figure 3C-D) a change in Km has very little effect on steady state NAD concentration and NAD consumption flux.
-In the presence of NNMT, however, NAD consumption flux and NAD concentration increases with decreasing Km values (Figure 5E-F).
+As kinetic parameters of NamPRT are only available for the human enzyme \cite{Burgos2008}, we went on to theoretically analysed the potential effect of the NamPRT affinity ($K_M$) on NAD steady state concentration and NAD consumption flux.
+In the absence of NNMT (Figure 4A-B) a change in $K_M$ has very little effect on steady state NAD concentration and NAD consumption flux.
+In the presence of NNMT, however, NAD consumption flux and NAD concentration increases with decreasing $K_M$ values (Figure 4C-D).
 
 It is intersting to note that, without NNMT, NAD concentration and consumption flux are both considerably affected by cell division rates, at least if the enzyme expression is kept constant.
-Even though, this is of course an artificial scenario, as one would assume organisms to regulate enzyme expression to achieve similar levels of metabolite concentrations instead.
-However, there seems to be an apparent trade off between maintainable NAD concentration and consumption flux.
+This is of course an artificial scenario, as one would assume organisms to regulate enzyme expression to achieve similar levels of metabolite concentrations instead.
+However, there seems to be an apparent trade off between maintainable NAD concentration and consumption flux in the absecne of NNMT.
+In contrast,  NAD consumption rates increase with NAD concentration in the presence of NNMT and both are relatively independent of cell devision rates.
 
-When we compare NAD consumption and NAD concentration with and without NNMT with two different substrate affinities of NamPRT, we see that at a low affinity (Km of 1 mikroM, which is in the range of the Km of NADA for Nam, NAD consumption flux is only higher with NNMT at low cell division rates.
-This effect becomes stronger if we further lower the affinity (not shown) (Figure 3G-H).
+When we compare NAD consumption and NAD concentration with and without NNMT with two different substrate affinities of NamPRT, we see that at low affinity ($K_M$ of 1 $\mu$M, which is in the range of the $K_M$  of NADA for Nam), NAD consumption flux is only higher with NNMT at low cell division rates.
+This effect becomes stronger with decreasing affinity (Figure 4E-F).
 
@@
 Thus, what is the impact of the substrate affinity of NNMT that is competing with NamPRT for the same substrate?
-In Figure 4A we see that the substrate affinity values found in the human enzymes (indicated by black asterisks) are actually optimal with respect to both achievable steady state NAD concentration and consumption fluxes.
+In Figure 5 we see that the substrate affinity values found in the human enzymes (indicated by black asterisks) are actually optimal with respect to both achievable steady state NAD concentration and consumption fluxes.
 Thus, a further increase of the affinity of NamPRT for Nam would not provide any advantage.
@@
 
-As our simulations suggest that NNMT might have excerted an evolitionary pressure on the development of NamPRT.
+As our simulations suggest that NNMT might have excerted an evolutionary pressure on the development of NamPRT.
 We performed multiple sequence analysis to see if we can find sequence variations in the protein sequence of NamPRT that indicate evolutionary changes of NamPRT upon the appearance of NNMT.
-The multiple sequence alignment of selected sequences is shown in Figure 5A.
-A more comprehensive multiple sequence alignment containing a larger number of species can be found in Supplementary Figure 1.
+The multiple sequence alignment of selected sequences is shown in Figure 5A and a more comprehensive multiple sequence alignment containing a larger number of species can be found in Supplementary Figure 1.
 We recognized that most Deuterostomia that have only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 43 to 52 of the human enzyme.
 This insert overlaps with a predicted weak nuclear localisation signal, that is lost when removing this sequence stretch.
-The inserted sequence corresponce to a structurallly unresolved loop structure in all avaibable crystal structure of human NamPR (e.g.
+The inserted sequence corresponce to a structurallly unresolved loop in all avaibable crystal structure of human NamPR (e.g.
 \cite{Wang2006} structure visualisation Figure 5B).
@@
 
-From these observations we derived two hypotheses regarding the potential function of the loop for NamPRT function.
+From these observations we derived two possible hypotheses regarding the role of the loop for NamPRT function.
 Firstly, we analysed whether the deletion of the amino acids 42 to 51 has an effect on the localisation of the human enzyme.
-But both wildtype and mutant protein show a mixed localisation in the cytosol and nucleus (Figure 6C).
+We thus performed immunoflourescence imaging with FLAG-tagged proteins, but both wildtype and mutant protein showed a mixed cytosolic nuclear localisation (Figure 6C).
 We therefore conclude that the partial nuclear localisation of NamPRT is not affected by the sequence deletion.
-The second hypothesis was that the sequence insertion influences NamPRT affinity and activity, which is also what our mathematical simulations would predict.
+The second hypothesis was that the sequence insertion influences NamPRT affinity, which is what our mathematical simulations would predict.
 We thus recombinantly expressed  both wildttype and mutant  NamPRT  in E.
@@
 Upon purifictaion we meassured the activity using NMR detecttion of NMN produced in the presence and absence of ATP.
-It appears that the enzymatic activity of the mutant enzyme is much lower than that of the wildtype enzyme (Figure 6D).
-We furthermore see that in contrast to NamPRT wildtype, the mutant does not show an increased activity in the presence of ATP (Figure 6E).
+It appears that the enzymatic activity of the mutant enzyme is much lower than that of the wildtype enzyme (Figure 6D) and that an increase of the substrate concentration has a much stronger effect on the activity of the mutant compared to the wildtype enzyme, indicating a lower affinity.
+We furthermore showed  that in contrast to NamPRT wildtype, the mutant does not exhibit an increased activity in the presence of ATP (Figure 6E).
 
@@
 
-But why is NADA lost in vertebrates while NADA and NamPRT are co-existing existing in bacteria?
+The remaining question is now, why NADA is lost in vertebrates, while NADA and NamPRT are co-existing in bacteria?
 To analyse this question we built a two compartment model that has a shared Nam source (details see Materials an Methods and Supplementary Material).
-One compartment contains NADA while the other contains either NamPRT alone or together with NNMT.
+One compartment contains NADA while the other contains either NamPRT alone at equal expression levels, or together with NNMT.
 Without NNMT the compartment containing NADA has slightly lower NAD consumption rates (Figure 7A), but is able to maintain much higher steady state NAD concentrations at low cell division rates  (Figure 7B).
-At high cell division rates, however, steady state concentration in both compartments are similar.
-The latter might explain why in bacteria that have relatively high growth rates both systems coexists.
-If in contrast, NNMT is present in the NamPRT compartment, this compartment outcompetes the NADA containing compartment, both with respect to NAD-consumption rates and steady state NAD concentrations (Figure 7 C and D).
-This effect is dependent on a high affinity of NamPRT, as at low affinity the NADA containing system again maintains higher NAD-concentrations, the NAD consumption flux still being higher  in the NamPRT and NNMT containing compartment.
+At high cell division rates, steady state concentrations in both compartments are similar.
+The latter might explain why in bacteria that have relatively high growth rates both systems co-exist.
+In the presence of NNMT the NamPRT compartment outcompetes the NADA containing compartment, both with respect to NAD consumption rates and steady state NAD concentrations (Figure 7 C and D).
+This effect is dependent on a high affinity of NamPRT, as at low affinity (high $K_M$ ) the NADA containing system is again able to maintain higher NAD-concentrations.
+The NAD consumption flux is still  higher  in the NamPRT and NNMT containing compartment.
 Taken together this might explain why upon the development of a sufficiently high affinity of NamPRT for its substrate, which seems to have been induced by the appearance of NNMT, NADA is lost.

### 2018-06-26 latex/supplementaries.tex
Added supplementary legends and methods by Dorothée

%%% TeX-master: "manuscript"

\section{Supplementary Legends}

\subsection{Supplementary figure S1}

\textbf{The structurally unresolved loop structure of NamPRT.} Sequence alignment of NamPRT of different species cropped to the region around the unresolved loop structure. Coloured rectangles indicate the enzymes present in the species besides NamPRT; blue: NNMT; black: NADA and NNMT; yellow: NADA; green: NamPRT only. Major clades are indicated for better orientation. Number of amino acid indicated at the top refer to the human protein.

\subsection{Supplementary figure S2}

\todo{Figure must be made!}\textbf{Distribution of NNMT in birds.} ...

\subsection{Supplementary table S1}

\textbf{Overview of kinetic constants used for the construction of the model.}

\subsection{Supplementary table S2}

\textbf{Query proteins used for Blast searches.}

### 2018-06-26 latex/methods.tex

@@
 The e-value cut-off was 1e-30 for all enzymes.
-Blastp parameters were set to yield maximum 20,000 target sequences, using the BLOSUM62 matrix with a word size of 6 and gap opening and extension costs of 11 and 1, respectively.
+Blastp parameters were set to yield maximum 20\,000 target sequences, using the BLOSUM62 matrix with a word size of 6 and gap opening and extension costs of 11 and 1, respectively.
 Low-complexity filtering was disabled.
@@
 
-\subsection{Generation of eukaryotic expression vectors encoding C-terminally FLAG-tagged NamPT proteins}
+\subsection{Generation of eukaryotic expression vectors encoding C-terminally FLAG-tagged NamPRT proteins}
 
@@
 
-\subsection{Transient transfection, immunocytochemstry and confocal laser scanning microscopy}
+\subsection{Transient transfection, immunocytochemstry, and confocal laser scanning microscopy}
 
-HeLa S3 cells cultivated in Ham’s F12 medium supplemented with 10\% (v/v) FCS, 2 mM L-glutamine, and penicillin/streptomycin, were seeded on cover slips in a 24 well plate.
+HeLa S3 cells cultivated in Ham’s F12 medium supplemented with 10\% (v/v) FCS, 2\,mM L-glutamine, and penicillin/streptomycin, were seeded on cover slips in a 24 well plate.
 After one day, cells were transfected using Effectene transfection reagent (Qiagen) according to the manufacturer’s recommendations.
@@
 Confocal laser scan imaging of cells was performed using a Leica TCS SP8 STED 3x microscope equipped with a 100x oil immersion objective (numerical aperture 1.4).
+
+\subsection{Generation of bacterial expression vectors encoding mutated or wildtype NamPRT proteins}
+
+BL21-codonPlus (DE3) RIL cells were transformed with pQE-30 NamPRT WT/pREP4 or pQE-30 NamPRT $\Delta$42-51/pREP4.
+Bacterial cells were grown at 37\,°C in 250\,rpm shaking incubator in 1\,L of LB medium containing 100\,µg/mL ampicillin, 50\,µg/mL kanamycin, and 32\,µg/mL chloramphenicol.
+Protein expression was induced with 0.2\,mM isopropyl-$\beta$-D-thiogalactoside (IPTG) at an OD600 of 0.4 to 0.6.
+Induction was done overnight at 18\,\textdegree C in 250\,rpm shaking incubator.
+
+\subsection{Purification of NamPRT}
+
+The cells were harvested by centrifugation and resuspended in lysis buffer (20\,mM Tris-HCl pH 8.0, 500\,mM NaCl, 4\,mM dithiltheritol (DTT), 1\,mg/mL lysozyme, 1X Complete EDTA-free protease inhibitor cocktail (Roche)).
+After sonification, the lysate was centrifuged for 30\,min at 13000\,g, and the clear lysate was incubated with 2\,mL of Nickel-NTA resin (Qiagen).
+Non-specific protein binding was removed with washing buffer (20\,mM Tris-HCl pH 8.0, 500\,mM NaCl, 1\,mM DTT, 20\,mM imidazole).
+The protein was eluted with 2.5\,mL of elution buffer (20 mM\,Tris-HCl pH 8.0, 500\,mM NaCl, 300\,mM imidazole).
+
+The eluted protein was immediately subjected to size exclusion chromatography (SEC) on an ÄKTA pure system (GE Healthcare) and loaded onto a HiLoad 16/60 Superdex 200\,pg column (GE Healthcare), run at a flow rate of 1\,mL/min with SEC buffer (20\,mM Tris-HCl pH 8.0, 500\,mM NaCl).
+Fractions corresponding to the size of recombinant protein were pooled and used for enzymatic assay.
+The purity and size of the protein were assessed by SDS-PAGE.
+
+\subsection{Enzymatic Assay}
+
+2\,µM of enzyme were incubated with 5-phospho-D-ribose 1-diphosphate (PRPP, 0.1\,mM or 1\,mM) and nicotinamide (Nam, 0.1\,mM or 1\,mM) in reaction buffer (20\,mM Tris-HCl pH 8.0, 500\,mM NaCl, 2\,mM MgCl2, and 0.03\% BSA), in absence or presence of 1\,mM of adenosine triphosphate (ATP).
+The 1.2\,mL reaction was incubated for 10 minutes at 30\,°C and the enzymatic activity stopped with 0.1\,mM of FK866, the samples were frozen in liquid nitrogen.
+
+\subsection{Sample preparation and NMR spectroscopy}
+
+The samples were dried with an Eppendorf Vacufuge Concentrator, and then resuspended with 200\,µl of NMR buffer containing 5\% deuterated H2O (D2O) and 1\,mM 4,4-dimethyl-4-silapentane-1-sulfonate (DSS).
+
+1D 1H NMR spectra were acquired on a 850\,MHz Ascend Bruker spectrometer equipped with 5\,mm TCI triple-resonance CryoProbe and a pulse field gradients along the z-axis.
+The experiments were acquired with zgesgppe pulse sequence, allowing water suppression using excitation sculpting with gradients and perfect echo.
+The temperature was kept constant at 300\,K and the acquisition parameters were the following: number of scans = 2000, relaxation delay = 1\,s, acquisition time = 1.6\,s, data points = 65K, and spectral width = 14\,ppm.
+
+The spectra phase and baseline were automatically and manually corrected using TopSpin 3.5 software (Bruker Biospin).
+Quantification of nicotinamide mononucleotide (NMN) was done by the integration of the peak at 9.57\,ppm and DSS used as an internal standard.

### 2018-06-26 latex/introduction.tex

@@
 %%% TeX-master: "manuscript"
-
+% !TeX spellcheck=en_GB
 \section{Introduction}
@@
 The presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012FEBS} is among the most enigmatic and counterintuitive features of NAD metabolism.
-Why is there one enzyme (NamPRT) seemingly optimized to recycle even the faintest amounts of Nam back into NAD synthesis, while at the same time there is another enzyme (NNMT) that seems to have no metabolic function other than to remove Nam from NAD metabolism, This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deaminated Nam to nicotinic acid (NA) using the nicotinamide deamidase (NADA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (Fig.
+Why is there one enzyme (NamPRT) seemingly optimized to recycle even the faintest amounts of Nam back into NAD synthesis, while at the same time there is another enzyme (NNMT) that seems to have no metabolic function other than to remove Nam from NAD metabolism, This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deamidated Nam to nicotinic acid (NA) using the nicotinamide deamidase (NADA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (Fig.
 1).
 
-We here present a phylogenetic analys of NAD-pathway in eukaryotes that.
-In contrast to our previous analysis, which has been lmited by the low number of eularyotic genomes available at the time, we are now able to provide a comprehensively analysis of the eukaryotich pathway evolution.
+We here present a phylogenetic analysis of NAD-pathway in eukaryotes that.
+In contrast to our previous analysis, which has been limited by the low number of eukaryotic genomes available at the time, we are now able to provide a comprehensively analysis of the eukaryotic pathway evolution.
 The results show there has seemingly been a selection for the co-existence of NamPRT and NNMT in Deuterostomia, while the pathway dominant in bacteria is lost.
 This was accompanied by a marked increase in the number of  NAD+-consuming signalling enzymes.
-To explain these counterintuitive results we built a mathematical model of the pathway and demonstratet that NNMT has a critical role to maintain high NAD+-consuming signalling fluxes by preventing accumulation of inhibitory Nam.
+To explain these counterintuitive results we built a mathematical model of the pathway and demonstrated that NNMT has a critical role to maintain high NAD+-consuming signalling fluxes by preventing accumulation of inhibitory Nam.
 Our model furthermore predicts that NNMT likely exerted an evolutionary pressure on the NamPRT affinity development.

### 2018-06-26 latex/results.tex

@@
 %%% TeX-master: "manuscript"
-
 \section{Results}
@@
 NMN and NAMN are converted into dinucleotides by the Nam/NA adenylytransferases (NMNATs).
-The recycling pathway via NA finally requires an amination step catalysed by NADsynthase, driven by the conversion of ATP to AMP producing pyrophosphate.
+The recycling pathway via NA finally requires an amidation step catalysed by NADsynthase, driven by the conversion of ATP to AMP producing pyrophosphate.
 Even though the latter pathway seems to be very inefficient, it is the pathway preferentially used by most bacteria, fungi and plants (see Figure 2A), whereas most metazoans recycle Nam using NamPRT.
@@
 To do so we used the previously established classification into 10 different enzyme families \cite{Gossmann2012FEBS}.
-The detailed list of templates used for the phylogenetic analysis can be found in  Supplementary Table 1?
-\todo[author=Mathias]{Inlcude number and order and name Supplementary Tables accordingly}.
+The detailed list of templates used for the phylogenetic analysis can be found in  Supplementary Table 1 \todo[author=Mathias]{Inlcude number and order and name Supplementary Tables accordingly}.
 The numbers in Figure 2B denote the average number of NAD-dependent signalling enzyme families found in each taxa.
@@
 
-First we addressed the question of the predominant coexitence of NamPRT and NNMT in vertebrates that coincides with an increase in the number of NAD-consuming enzyme families.
+First we addressed the question of the predominant co-existence of NamPRT and NNMT in vertebrates that coincides with an increase in the number of NAD-consuming enzyme families.
 We thus used our dynamic model to simulate steady state concentrations and NAD-consumption rates simulating NAD biosynthesis via  NamPRT in the  presence or absence of NNMT.
-We see that the presence of NNMT enables higher NAD consumption rates  (Figure3A) while, as exptected, reducing the steady state concentration of NAD (Figure 3B).
+We see that the presence of NNMT enables higher NAD consumption rates  (Figure3A) while, as expected, reducing the steady state concentration of NAD (Figure 3B).
 The decline in NAD-concentration can be compensated by a higher expression of NamPRT further increasing NAD consumption flux (Figure 3A and B dashed lines).
@@
 These findings can be explained when looking in more detail at the kinetic parameters of NamPRT and NAD-consuming enzymes such as Sirt1p.
-The increase of NAD consumption flux is caused by the fact that most NAD-consuming enzymes are inhibited by their product Nam, explaing why the presence of NNMT enables higher NAD consumption fluxes.
-At the same time, the high substrate affinity of NamPRT maintains a sufficiently high NAD-concetration.
+The increase of NAD consumption flux is caused by the fact that most NAD-consuming enzymes are inhibited by their product Nam, explaining why the presence of NNMT enables higher NAD consumption fluxes.
+At the same time, the high substrate affinity of NamPRT maintains a sufficiently high NAD-concentration.
 
@@
 
-It is intersting to note that, without NNMT, NAD concentration and consumption flux are both considerably affected by cell division rates, at least if the enzyme expression is kept constant.
+It is interesting to note that, without NNMT, NAD concentration and consumption flux are both considerably affected by cell division rates, at least if the enzyme expression is kept constant.
 This is of course an artificial scenario, as one would assume organisms to regulate enzyme expression to achieve similar levels of metabolite concentrations instead.
-However, there seems to be an apparent trade off between maintainable NAD concentration and consumption flux in the absecne of NNMT.
-In contrast,  NAD consumption rates increase with NAD concentration in the presence of NNMT and both are relatively independent of cell devision rates.
+However, there seems to be an apparent trade off between maintainable NAD concentration and consumption flux in the absence of NNMT.
+In contrast,  NAD consumption rates increase with NAD concentration in the presence of NNMT and both are relatively independent of cell division rates.
 
@@
 Firstly, we analysed whether the deletion of the amino acids 42 to 51 has an effect on the localisation of the human enzyme.
-We thus performed immunoflourescence imaging with FLAG-tagged proteins, but both wildtype and mutant protein showed a mixed cytosolic nuclear localisation (Figure 6C).
+We thus performed immunofluorescence imaging with FLAG-tagged proteins, but both wildtype and mutant protein showed a mixed cytosolic nuclear localisation (Figure 6C).
 We therefore conclude that the partial nuclear localisation of NamPRT is not affected by the sequence deletion.
 The second hypothesis was that the sequence insertion influences NamPRT affinity, which is what our mathematical simulations would predict.
-We thus recombinantly expressed  both wildttype and mutant  NamPRT  in E.
+We thus recombinantly expressed  both wildtype and mutant  NamPRT  in E.
 coli.
-Upon purifictaion we meassured the activity using NMR detecttion of NMN produced in the presence and absence of ATP.
+Upon purification we measured the activity using NMR detection of NMN produced in the presence and absence of ATP.
 It appears that the enzymatic activity of the mutant enzyme is much lower than that of the wildtype enzyme (Figure 6D) and that an increase of the substrate concentration has a much stronger effect on the activity of the mutant compared to the wildtype enzyme, indicating a lower affinity.
@@
 The latter might explain why in bacteria that have relatively high growth rates both systems co-exist.
-In the presence of NNMT the NamPRT compartment outcompetes the NADA containing compartment, both with respect to NAD consumption rates and steady state NAD concentrations (Figure 7 C and D).
+In the presence of NNMT the NamPRT compartment out-competes the NADA containing compartment, both with respect to NAD consumption rates and steady state NAD concentrations (Figure 7 C and D).
 This effect is dependent on a high affinity of NamPRT, as at low affinity (high $K_M$ ) the NADA containing system is again able to maintain higher NAD-concentrations.

### 2018-06-26 latex/figurelegends.tex

@@
 %%% TeX-master: "manuscript"
+
+\newpage
 
@@
 
-\textbf{Schematic overview of NAD biosynthesis and consumption.} NAD can be synthesized from tryptophan (Trp), nicotinamide (Nam), nicotinic acid (NA), and, to a lesser extend, nicotinamide ribose (NR).
+\begin{figure}[ht]
+  \centering
+  \includegraphics{fig_1_pathway_overview}
+  \caption{\textbf{Schematic overview of NAD biosynthesis and consumption.} NAD can be synthesized from tryptophan (Trp), nicotinamide (Nam), nicotinic acid (NA), and, to a lesser extend, nicotinamide ribose (NR).
 Nam is the main precursor and also the product of NAD-consuming signalling reactions by enzymes such as sirtuins (NAD-dependent histone deacetylases) or PARPs (poly-ADP-ribosylases).
@@
 A third enzyme can consume Nam, the Nam N-methyltransferase (NNMT).
-For more details and abbreviations, see text.
+For more details and abbreviations, see text.}
+  \label{fig:pathway_overview}
+\end{figure}
+
+\newpage
 
@@
 
-\textbf{Evolutionary distribution of NADA, NNMT and NamPRT and their relation to the number of NAD consumers.} A) Distribution of NADA, NNMT and NamPRT in selected major taxa.
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.75\textwidth]{fig_2_phylo_distribution}
+  \caption{\textbf{Evolutionary distribution of NADA, NNMT and NamPRT and their relation to the number of NAD consumers.} A) Distribution of NADA, NNMT and NamPRT in selected major taxa.
 NADA is dominant in Bacteria, Fungi, and Plants (Viridiplantae), whereas NamPRT together with NNMT is dominant in Metazoa.
@@
 The numbers below the taxon names indicate the average number of NAD-consuming enzyme families found in all sub-taxa.
-The branch length is arbitrary.
+The branch length is arbitrary.}
+  \label{fig:phylo_distribution}
+\end{figure}
+
+\newpage
 
@@
 
-\textbf{NNMT enables high NAD-consumption flux.} We used a dynamic model of NAD biosynthesis and consumption (details see Materials and Methods) to simulate NAD consumption flux (A) and NAD concentration (B) in the presence of NamPRT and with or without NNMT at different cell division rates.
-NMMT enables higher steady-state NAD consumption flux despite reduced NAD concentrations.
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth]{fig_3_NNMT_NAD_flux}
+  \caption{\textbf{NNMT enables high NAD-consumption flux.} We used a dynamic model of NAD biosynthesis and consumption (details see Materials and Methods) to simulate NAD consumption flux (A) and NAD concentration (B) in the presence of NamPRT and with or without NNMT at different cell division rates.
+NMMT enables higher steady-state NAD consumption flux despite reduced NAD concentrations.}
+  \label{fig:NNMT_NAD_flux}
+\end{figure}
+
+\newpage
 
@@
 
-\textbf{Role of NamPRT affinity for Nam.} Using the dynamic model of NAD biosynthesis and consumption we simulated the affect of different Michaelis Menten constants ($K_M$) of NamPRT for Nam on steady-state NAD consumption flux and NAD concentration at different cell devision rates.
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.9\textwidth]{fig_4_NamPRT_affinity_Nam}
+  \caption{\textbf{Role of NamPRT affinity for Nam.} Using the dynamic model of NAD biosynthesis and consumption we simulated the affect of different Michaelis Menten constants ($K_M$) of NamPRT for Nam on steady-state NAD consumption flux and NAD concentration at different cell devision rates.
 In the absence of NNMT (A and B), the $K_M$ of NamPRT has little influence on NAD consumption and concentration, but both are changing with cell devision rates.
@@
 NNMT furthermore makes both, consumption flux and concentration, relatively independent of cell division rates.
-Comparing the situation with and without NNMT (E and F) at different NamPRT $K_M$ reveals that at low $K_M$ and high cell devision rates NNMT no longer enables higher NAD consumption rates compared to NamPRT alone.
+Comparing the situation with and without NNMT (E and F) at different NamPRT $K_M$ reveals that at low $K_M$ and high cell devision rates NNMT no longer enables higher NAD consumption rates compared to NamPRT alone.}
+  \label{fig:NamPRT_affinity_Nam}
+\end{figure}
+
+\newpage
 
@@
 
-\textbf{The substrate affinities of human NNMT and NamPRT are optimal.} Both NAD consumption flux (A) and NAD concentration (B) are increasing with decreasing $K_M$ of NamPRT but decreeasing $K_M$ of NNMT.
-The affinities reported for human enzymes (indicated by a black asterisk) appear to be close to optimal, as further improvements would have little or no effect on NAD consumption or concentration.
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth]{fig_5_optimal_substr_affinities}
+  \caption{\textbf{The substrate affinities of human NNMT and NamPRT are optimal.} Both NAD consumption flux (A) and NAD concentration (B) are increasing with decreasing $K_M$ of NamPRT but decreeasing $K_M$ of NNMT.
+The affinities reported for human enzymes (indicated by a black asterisk) appear to be close to optimal, as further improvements would have little or no effect on NAD consumption or concentration.}
+  \label{fig:optimal_substr_affinities}
+\end{figure}
+
+\newpage
 
@@
 
-\textbf{The function of the structurally unresolved loop structure of NamPRT.} Most Deuterostomia that encode NNMT show a sequence insertion in the N-terminal region of NamPRT that has been revealed by multiple sequence alignment of NamPRT from different species (A).
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.75\textwidth]{fig_6_unresolved_loop}
+  \caption{\textbf{The function of the structurally unresolved loop structure of NamPRT.} Most Deuterostomia that encode NNMT show a sequence insertion in the N-terminal region of NamPRT that has been revealed by multiple sequence alignment of NamPRT from different species (A).
 Coloured circles indicate the enzymes present in the species besides NamPRT; blue: NNMT; black: NADA and NNMT; yellow: NADA.
+For a more comprehensive alignment, please see supplementary figure~S1.
 B) The visualisation of human NamPRT is based on a structure prediction of SWISS-MODEL \cite{Arnold2006,Biasini2014} of the sequence of the human NamPRT using the model 2H3D as template \cite{Wang2006}.
@@
 But \textit{in vitro} measurements using recombinant protein show that the mutant NamPRT has lower activity than the wildtype enzyme (D) and is not activated by ATP (E).
+Bars in D) and E) with different letters indicate significant difference as estimated with a T test assuming independent samples and significance at $p < 0.05$.}
+  \label{fig:unresolved_loop}
+\end{figure}
+
+\newpage
 
@@
 
-\textbf{NNMT provides a competitive advantage and makes NADA obsolete.} To simulate competition for common resources, we created a two compartment model where one compartment contained NADA but no NamPRT and the other compartment contained NamPRT either with or without NNMT but no NADA.
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth]{fig_7_NNMT_comp_advantage}
+  \caption{\textbf{NNMT provides a competitive advantage and makes NADA obsolete.} To simulate competition for common resources, we created a two compartment model where one compartment contained NADA but no NamPRT and the other compartment contained NamPRT either with or without NNMT but no NADA.
 In the absence of NNMT (A and B) the compartment containing NADA has slightly lower NAD consumption rates (A) but much higher NAD concentrations (B).
-In the presence of NNMT, however, both NAD consumption (C) and NAD concentration (D) are lower in the NADA compartment, but this effect is only observed at low NamPRT $K_M$.
+In the presence of NNMT, however, both NAD consumption (C) and NAD concentration (D) are lower in the NADA compartment, but this effect is only observed at low NamPRT $K_M$.}
+  \label{fig:NNMT_comp_advantage}
+\end{figure}
+
+\newpage

### 2018-06-26 latex/methods.tex

@@
 
-\subsection{Generation of eukaryotic expression vectors encoding C-terminally FLAG-tagged NamPRT proteins}
+\subsection{Generation of expression vectors encoding encoding wild-type and mutant human NamPRT}
 
-The open reading frame (ORF) encoding human NamPRT was inserted into pFLAG-CMV-5a (Merck - Sigma Aldrich) via EcoRI/BamHI sites.
-Using a PCR-based approach, this vector provided the basis for the generation of a plasmid encoding a NamPRT deletion mutant lacking amino acid residues 42-51 (NamPRT-$\Delta$42-51).
-The sequences of the inserted ORFs were verified by DNA sequence analysis.
+For eukaryotic expression with a C-terminal FLAG-epitope, the open reading frame (ORF) encoding human NamPT was inserted into pFLAG-CMV-5a (Merck - Sigma Aldrich) via EcoRI/BamHI sites.
+Using a PCR-based approach, this vector provided the basis for the generation of a plasmid encoding a NamPT deletion mutant lacking amino acid residues 42-51 (NamPT-$\Delta$42-51).
+For prokaryotic expression with an N-terminal 6xHis-tag, the wild-type and mutant ORFs were inserted into pQE-30 (Qiagen) via BamHI and PstI-sites.
+All cloned sequences were verified by DNA sequence analysis.
 
@@
 Confocal laser scan imaging of cells was performed using a Leica TCS SP8 STED 3x microscope equipped with a 100x oil immersion objective (numerical aperture 1.4).
-
-\subsection{Generation of bacterial expression vectors encoding mutated or wildtype NamPRT proteins}
-
-BL21-codonPlus (DE3) RIL cells were transformed with pQE-30 NamPRT WT/pREP4 or pQE-30 NamPRT $\Delta$42-51/pREP4.
-Bacterial cells were grown at 37\,°C in 250\,rpm shaking incubator in 1\,L of LB medium containing 100\,µg/mL ampicillin, 50\,µg/mL kanamycin, and 32\,µg/mL chloramphenicol.
-Protein expression was induced with 0.2\,mM isopropyl-$\beta$-D-thiogalactoside (IPTG) at an OD600 of 0.4 to 0.6.
-Induction was done overnight at 18\,\textdegree C in 250\,rpm shaking incubator.
 

### 2018-06-26 latex/discussion.tex
revised discussion

--- 2018-06-21 latex/discussion.tex
@@
 %%% TeX-master: "manuscript"
-
+% !TeX spellcheck=en_GB
 \section{Discussion}
-
-\subsection{Evolutionary and physiological role of NNMT}
 
@@
 New approaches are required to disentangle the underlying network, helping us to understand how alterations affect human physiology.
-We here combine phylogenetic analysis and mathematical modelling of a central metabolic pathway supported by experimental analysis to reveal the likely evolutionary development of the NAD biosynthesis and consumption pathway and the physiological role of the methyltransferase NNMT.
+We here combined phylogenetic analysis and mathematical modelling the NAD biosynthesis and consumption pathway to better understand the evolution of this pathway and  especially the role of the methyltransferase NNMT.
+Our analysis reveals that that NNMT plays a vital role for the diversification of NAD-dependent signalling reactions and potentially also for NAD homeostasis.
+We show that the presence of NNMT enables higher NAD consumption rates in comparison to NamPRT or NADA alone.
+It furthermore appears that NNMT makes both NAD concentration and NAD consumption relatively independent of other NAD requiring processes such as cell growth.
+We furthermore suggest that NNMT exerted an evolutionary pressure on NamPRT driving the development of the high affinity of NamPRT for its substrates found in the human enzyme \cite{Burgos2008}.
+These finding shed new light on the role of NNMT, which has  earlier  been recognized  as potential marker for some types of cancer (e.g.
+\cite{Okamura1998}), hinting that increased NNMT expression in cancer might serve to remove  NAM produced by increased NAD-dependent signalling.
+Further analysis is required to better understand which NAD-signalling reactions are affected.
+But given a better mechanistic understanding and as routine clinical measurements of urine methyl-Nam (MNam) concentrations are available already, MNam could potentially serve as biomarker in the context of different diseases (e.g\cite{Pumpo2001,Delaney2005}).
 
-We show that the vitamin B3-degrading enzyme NNMT plays a vital role for the diversification of NAD-dependent signalling reactions and potentially also for NAD homeostasis.
-Due to the fact that NamPRT has a very high affinity for Nam whereas NNMT has a rather low affinity for it, NNMT is able to remove excess Nam that would lead to inhibition of NAD-dependent signalling reactions while maintaining NAD concentrations and even increasing NAD consumption fluxes.
+In our analysis we furthermore noted that NNMT only provides an advantage as long as the NamPRT affinity for Nam is sufficiently high, this implies that treatment with competitive inhibitors that affect the apparent $K_M$ would have stronger effects on cells that express NNMT compared to cells that don´t express NNMT (see Figure 4E-F).
+This could be important for cancer treatments as the expression of NNMT is relatively low in normal tissues except liver, whereas several types of cancer overexpress NNMT.
 
-The affinities measured for the human enzymes seem to be optimal and further increase in NAD consumption rates might only enabled by compartmentalisation of NAD biosynthesis.
+We here neglected the potential effects of the methyl donor S-adenosyl-methionin (SAM) and its precursor methionine, although it is most likely contributing yet another regulatory level for Nam availability and thus an additional interaction point between gene regulation and metabolism.
+It has been shown earlier that NNMT expression influences protein methylation dependent on methionine availability \cite{Ulanovskaya2013}.
+One of the challenges in the analysis of this aspect is the availability of \textit{in vivo} concentration measurements for SAM and the large amount of reactions using it as substrate.
+A mathematical model of this pathway has  been published \cite{PMID14637052}  providing the basis for future studies.
 
-In contrast to our results, it was proposed that NNMT is present in yeast (\textit{Saccharomyces cerevisiae} \cite{Anderson2003}.
-The enzyme was later further characterised as methyltransferase of the transcription elongation factors TEF1 and TEF2 but not Nam \cite{Wlodarski2011}.
+Beside gaining new insights into the role of NNMT for NAD-metabolism, our analyses furthermore provided a suggestion why NADA disappears in vertebrates while it co-exists with NamPRT in bacteria.
+It seems that the combined enzymatic activity of NNMT and NamPRT is optimised to maintain stable NAD-concentration and high NAD-consumption flux, making NADA obsolete.
 
-\subsection{General applicability}
+\section{Conclusions}
 
-It is of course impossible to extend our conclusions to other metabolic processes without analysing them in more detail, but we have noted that the metabolism of other vitamins, such as pyridoxal (vitamin B6) metabolism, also contains vitamin degrading enzymes \todo{source}, that might thus have similar roles.
-Degrading enzymes do in general not receive the same attention as biosynthetic enzymes, reflected by the observation that NNMT has only recently been analysed in more detail, besides the much earlier recognition of NNMT as potential marker for some types of cancer (e.g.
-\cite{Okamura1998}) and routine clinical measurements of urine methyl-Nam concentrations in the context of different diseases (e.g.
-\cite{Pumpo2001,Delaney2005}).
-Our analysis show, however, that degrading enzymes do not solely modify substrates for better excretion, but can play a vital role both in human physiology as well as in the evolutionary development of biological processes.
-
-\subsection{Limitations of our analysis and outlook}
-
-Among others, we neglected the potential effects of the methyl donor S-adenosyl-methionin (SAM) and its precursor methionine in our analysis, although it is most likely contributing yet another regulatory level for Nam availability and thus an additional interaction point between gene regulation and metabolism.
-It has been shown earlier that in turn, NNMT expression influences protein methylation dependent on methionine availability \cite{Ulanovskaya2013}.
-One of the challenges in the analysis of this aspect is the availability of in vivo concentration measurements for SAM and the large amount of reactions using it as substrate.
-The same holds for the analysis of enzyme acetylation versus deacetylation by sirtuins that would enable prediction about enzyme activation or histone state if available.
+We here highlight the co-evolution of several enzymes of the NAD-pathway with the appearance of NNMT seemingly initiating and driving complex alterations of the pathway such as an increase and diversification of NAD dependent signalling, followed by an increase in NamPRT substrate affinity (schematic overview see Figure 8).
+This again appears to be accompanied by the loss of NADA in vertebrates and the first gene duplication of NMNATs \cite{Lau2010}.
+We also noted that the second gene duplication of NMNATs and thus the further compartmentalisation of NAD metabolism is co-occurring with a site specific positive selection event in NNMT (data not shown).
+To our knowledge this is the first study that comprehensively demonstrates a functional co-evolution of several enzymes of a pathway.

### 2018-06-26 latex/manuscript.tex
revised results section
Added figures to document
Fixed typos and added acknowledgements section

@@
+% !TeX spellcheck=en_GB
+
 \RequirePackage[l2tabu, orthodox]{nag}
@@
 
+\usepackage{graphicx}
+\usepackage[font=footnotesize]{caption}
 \usepackage[utf8]{inputenc}
@@
 
+\frenchspacing
+\graphicspath{ {../} }
+
 \begin{document}
@@
 
-\vspace{2cm}
+\vspace{15mm}
 
 \noindent
-Mathias Bockwoldt$^{1}$, Dorothée Houry$^{2}$,  Marc Niere$^{2}$, Toni I. Gossmann$^{3}$, Mathias Ziegler$^{2}$ and Ines Heiland$^{1,\S}$
+Mathias Bockwoldt\textsuperscript{1}, Dorothée Houry\textsuperscript{2}, Marc Niere\textsuperscript{2}, Toni I.
+Gossmann\textsuperscript{3}, Mathias Ziegler\textsuperscript{2}, and Ines Heiland\textsuperscript{1,§}
 
@@
 \noindent
-$^{1}$Department of Arctic and Marine Biology, UiT The Arctic University of Norway, Biologibygget, Framstredet 39, 9017 Tromsø, Norway
+\textsuperscript{1}Department of Arctic and Marine Biology, UiT The Arctic University of Norway, Biologibygget, Framstredet 39, 9017 Tromsø, Norway
 
 \noindent
-$^{2}$Department of Molecular Biology, University of Bergen, Thormøhlensgt. 55
-5020 Bergen, Norway
+\textsuperscript{2}Department of Molecular Biology, University of Bergen, Thormøhlensgt. 55, 5020 Bergen, Norway
 
 \noindent
-$^{3}$Department of Animal and Plant Sciences, Western Bank, University of Sheffield
-Sheffield, S10 2TN, United Kingdom
+\textsuperscript{3}Department of Animal and Plant Sciences, Western Bank, University of Sheffield, Sheffield, S10 2TN, United Kingdom
 
@@
 In eukaryotes, two different Nam salvage pathways exist.
-While in lower organisms the initial deamidation of Nam is prevalent, in animals the direct conversion of Nam to the mononucleotide by Nam phosphoribosyltransferase (NamPRT) dominates and eventually remains as the single Nam recycling route in vertebrates.
+While in lower organisms the initial deamidation of Nam is prevalent, the direct conversion of Nam to the mononucleotide by Nam phosphoribosyltransferase (NamPRT) dominates in animals and eventually remains as the single Nam recycling route in vertebrates.
 
@@
 We have used mathematical modelling approaches to resolve these counterintuitive observations.
-Our results indicate that NNMT is required to enable high NAD consumption fluxes necessitated by the increasing diversification of NAD-dependent signaling pathways.
+Our results indicate that NNMT is required to enable high NAD consumption fluxes necessitated by the increasing diversification of NAD-dependent signalling pathways.
 This kinetic regulation requires a high substrate affinity of the key enzyme for Nam salvage, NamPRT.
@@
 Mathematical modelling supports the hypothesis that NNMT exerted an evolutionary pressure on NamPRT enforcing the development of its unusually high substrate affinity.
-Using multiple sequence alignments, we identified a sequence insertion, first occurring in vertebrates, that parallels an – experimentally verified – increase in the substrate affinity of the enzyme.
+Using multiple sequence alignments, we identified a sequence insertion, first occurring in vertebrates, that parallels an experimentally verified increase in the substrate affinity of the enzyme.
 Further simulations show that the deamidation pathway became obsolete owing to the high substrate affinity of NamPRT.
-Collectively, our results illustrate a close evolutionary relationship between NAD biosynthesis and the diversification NAD-dependent signaling pathways, potentially driven by the concomitant occurrence of a regulator of Nam salvage, NNMT.
+Collectively, our results illustrate a close evolutionary relationship between NAD biosynthesis and the diversification NAD-dependent signalling pathways, potentially driven by the concomitant occurrence of a regulator of Nam salvage, NNMT.
 
@@
 
+\input{supplementaries}
+
+\section{Acknowledgements}
+
 \bibliographystyle{plain}

### 2018-06-26 latex/figurelegends.tex
First round of spell checking done

@@
 %%% TeX-master: "manuscript"
+% !TeX spellcheck=en_GB
 
@@
   \includegraphics{fig_1_pathway_overview}
-  \caption{\textbf{Schematic overview of NAD biosynthesis and consumption.} NAD can be synthesized from tryptophan (Trp), nicotinamide (Nam), nicotinic acid (NA), and, to a lesser extend, nicotinamide ribose (NR).
+  \caption{\textbf{Schematic overview of NAD biosynthesis and consumption.} NAD can be synthesised from tryptophan (Trp), nicotinamide (Nam), nicotinic acid (NA), and, to a lesser extend, nicotinamide ribose (NR).
 Nam is the main precursor and also the product of NAD-consuming signalling reactions by enzymes such as sirtuins (NAD-dependent histone deacetylases) or PARPs (poly-ADP-ribosylases).
@@
   \includegraphics[width=\textwidth]{fig_3_NNMT_NAD_flux}
-  \caption{\textbf{NNMT enables high NAD-consumption flux.} We used a dynamic model of NAD biosynthesis and consumption (details see Materials and Methods) to simulate NAD consumption flux (A) and NAD concentration (B) in the presence of NamPRT and with or without NNMT at different cell division rates.
-NMMT enables higher steady-state NAD consumption flux despite reduced NAD concentrations.}
+  \caption{\textbf{NNMT enables high NAD consumption flux.} We used a dynamic model of NAD biosynthesis and consumption (details see Materials and Methods) to simulate NAD consumption flux (A) and NAD concentration (B) in the presence of NamPRT and with or without NNMT at different cell division rates.
+NNMT enables higher steady state NAD consumption flux despite reduced NAD concentrations.}
   \label{fig:NNMT_NAD_flux}
@@
   \includegraphics[width=0.9\textwidth]{fig_4_NamPRT_affinity_Nam}
-  \caption{\textbf{Role of NamPRT affinity for Nam.} Using the dynamic model of NAD biosynthesis and consumption we simulated the affect of different Michaelis Menten constants ($K_M$) of NamPRT for Nam on steady-state NAD consumption flux and NAD concentration at different cell devision rates.
-In the absence of NNMT (A and B), the $K_M$ of NamPRT has little influence on NAD consumption and concentration, but both are changing with cell devision rates.
+  \caption{\textbf{Role of NamPRT affinity for Nam.} Using the dynamic model of NAD biosynthesis and consumption we simulated the affect of different Michaelis-Menten constants ($K_M$) of NamPRT for Nam on steady state NAD consumption flux and NAD concentration at different cell division rates.
+In the absence of NNMT (A and B), the $K_M$ of NamPRT has little influence on NAD consumption and concentration, but both are changing with cell division rates.
 In the presence of NNMT (C and D), decreasing $K_M$ of NamPRT enables increasing NAD consumption flux and NAD concentration.
 NNMT furthermore makes both, consumption flux and concentration, relatively independent of cell division rates.
-Comparing the situation with and without NNMT (E and F) at different NamPRT $K_M$ reveals that at low $K_M$ and high cell devision rates NNMT no longer enables higher NAD consumption rates compared to NamPRT alone.}
+Comparing the situation with and without NNMT (E and F) at different NamPRT $K_M$ reveals that at low $K_M$ and high cell division rates NNMT no longer enables higher NAD consumption rates compared to NamPRT alone.}
   \label{fig:NamPRT_affinity_Nam}
@@
   \includegraphics[width=\textwidth]{fig_5_optimal_substr_affinities}
-  \caption{\textbf{The substrate affinities of human NNMT and NamPRT are optimal.} Both NAD consumption flux (A) and NAD concentration (B) are increasing with decreasing $K_M$ of NamPRT but decreeasing $K_M$ of NNMT.
+  \caption{\textbf{The substrate affinities of human NNMT and NamPRT are optimal.} Both NAD consumption flux (A) and NAD concentration (B) are increasing with decreasing $K_M$ of NamPRT but decreasing $K_M$ of NNMT.
 The affinities reported for human enzymes (indicated by a black asterisk) appear to be close to optimal, as further improvements would have little or no effect on NAD consumption or concentration.}
@@
 C) The localisation of the FLAG-tagged mutant protein lacking the unresolved loop is not changed compared to wildtype human NamPRT.
-Both show a heterogenous nuclear cytosolic localisation in immunofluerescence images in HeLa S3 cells.
+Both show a heterogeneous nuclear cytosolic localisation in immunofluorescence images in HeLa S3 cells.
 But \textit{in vitro} measurements using recombinant protein show that the mutant NamPRT has lower activity than the wildtype enzyme (D) and is not activated by ATP (E).

### 2018-06-26 latex/introduction.tex
First round of spell checking done

@@
 % !TeX spellcheck=en_GB
+
 \section{Introduction}
@@
 NAD metabolism represents one of the most critical links that connect cellular signal transduction and energy metabolism.
-Even though best known as cofactor for various redox-reactions , NAD is involved in a number of signalling processes that consume NAD+ by cleaving the molecule to nicotinamide (Nam) and ADP-ribose \cite{Verdin2015}.
-These NAD+-dependent signalling reactions include but are not limited to poly- and mono-ADP-ribosylation, NAD+-dependent protein deacylation by sirtuins as well as the synthesis of calcium mobilizing molecules such as cyclic ADP-ribose \cite{?}.
-These NAD+-dependent signalling processes participate in the regulation of virtually all cellular activities.
-The enzymes involved in these processes are sensitive to the available NAD+ concentration \cite{Ruggieri2015}, which in turn is dependent on the NAD+/NADH redox ratio.
-Therefore, NAD+-dependent signalling can act as a transmitter of changes in the cellular energy homeostasis, for example, to regulate gene expression or metabolic activity \cite{Koch-Nolte2009}.\todo[author=Mathias B.]{Please add more references to the first paragraph.
+Even though best known as cofactor for various redox-reactions , NAD is involved in a number of signalling processes that consume NAD$^{+}$ by cleaving the molecule to nicotinamide (Nam) and ADP-ribose \cite{Verdin2015}.
+These NAD-dependent signalling reactions include but are not limited to poly- and mono-ADP-ribosylation, NAD-dependent protein deacylation by sirtuins as well as the synthesis of calcium-mobilizing molecules such as cyclic ADP-ribose \cite{?}.
+These NAD-dependent signalling processes participate in the regulation of virtually all cellular activities.
+The enzymes involved in these processes are sensitive to the available NAD concentration \cite{Ruggieri2015}, which in turn is dependent on the NAD$^{+}$/NADH redox ratio.
+Therefore, NAD-dependent signalling can act as a transmitter of changes in the cellular energy homeostasis, for example, to regulate gene expression or metabolic activity \cite{Koch-Nolte2009}.\todo[author=Mathias B.]{Please add more references to the first paragraph.
 I think there are better references then my own review.}
 
-The significance of NAD+-dependent signalling for NAD homeostasis has long been underestimated.
-However, it has now been established that substances affecting NAD biosynthesis lead to a rapid decline of the NAD concentration \cite{Buonvicino2018} suggesting that NAD+-dependent signalling consumes substantial amounts of NAD, which is why we later refer to them also as NAD-consuming reactions.
-The resulting NAD turnover differs in a cell type specific manner and can lead to an NAD-halflife as short as 2 hours \cite{Liu2018}.
+The significance of NAD-dependent signalling for NAD homeostasis has long been underestimated.
+However, it has now been established that substances affecting NAD biosynthesis lead to a rapid decline of the NAD concentration \cite{Buonvicino2018} suggesting that NAD-dependent signalling consumes substantial amounts of NAD, which is why we later refer to them also as NAD-consuming reactions.
+The resulting NAD turnover differs in a cell type specific manner and can lead to an NAD half-life as short as 2 hours \cite{Liu2018}.
 To maintain the NAD concentration at physiological levels, NAD biosynthesis needs to act at an equally rapid rate.
-Imbalances in NAD-homeostasis have been linked to various, in particular, ageing-related diseases such as diabetes, neurodegenerative disorders and cancer \cite{Chiarugi2012,Verdin2015}.
+Imbalances in NAD homeostasis have been linked to various, in particular, ageing-related diseases such as diabetes, neurodegenerative disorders and cancer \cite{Chiarugi2012,Verdin2015}.
 Several recent studies have demonstrated impressive health benefits of dietary supplementation with intermediates of NAD biosynthesis including NMN and nicotinamide riboside, (NR) \cite{Yoshino2018}.
 Apparently, the exploitation of NAD biosynthetic routes, in addition to the use of nicotinamide as precursor (Fig.
-1), results in increased NAD concentrations that stimulate NAD+-dependent signalling processes, in particular, protein deacetylation by sirtuins \cite{North2004}.
+1), results in increased NAD concentrations that stimulate NAD-dependent signalling processes, in particular, protein deacetylation by sirtuins \cite{North2004}.
 
-Due to the constant release of Nam through NAD-consuming signalling reactions, the NAD salvage pathway using Nam as precursor is the most important NAD-synthesis pathway.
+Due to the constant release of Nam through NAD-consuming signalling reactions, the NAD salvage pathway using Nam as precursor is the most important NAD synthesis pathway.
 If Nam would not be constantly recycled into NAD, humans would require a much higher daily vitamin B3 intake than the 16 mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}.
-In the first step of the salvage pathway, Nam is converted to the mononucleotide, NMN, by Nam phosphoribosyltransferase (NamPRT) using phosphoribosylpyrophosphate (PRPP) as co-substrate.
-The nearly complete recycling of Nam is achieved by an extraordinary high affinity of NamPRT to Nam, the Km being in the low nanomolar range \cite{Burgos2008}.
+In the first step of the salvage pathway, Nam is converted to the mononucleotide, NMN, by Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate.
+The nearly complete recycling of Nam is achieved by an extraordinary high affinity of NamPRT to Nam, the $K_{M}$ being in the low nanomolar range \cite{Burgos2008}.
 Despite the importance of its salvage, Nam can also be marked for excretion by methylation.
 The presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012FEBS} is among the most enigmatic and counterintuitive features of NAD metabolism.
-Why is there one enzyme (NamPRT) seemingly optimized to recycle even the faintest amounts of Nam back into NAD synthesis, while at the same time there is another enzyme (NNMT) that seems to have no metabolic function other than to remove Nam from NAD metabolism, This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deamidated Nam to nicotinic acid (NA) using the nicotinamide deamidase (NADA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (Fig.
+Why is there one enzyme (NamPRT) seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, while at the same time there is another enzyme (NNMT) that seems to have no metabolic function other than to remove Nam from NAD metabolism, This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deamidated Nam to nicotinic acid (NA) using the nicotinamide deamidase (NADA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (Fig.
 1).
 
-We here present a phylogenetic analysis of NAD-pathway in eukaryotes that.
+We here present a phylogenetic analysis of NAD pathway in eukaryotes that.
 In contrast to our previous analysis, which has been limited by the low number of eukaryotic genomes available at the time, we are now able to provide a comprehensively analysis of the eukaryotic pathway evolution.
 The results show there has seemingly been a selection for the co-existence of NamPRT and NNMT in Deuterostomia, while the pathway dominant in bacteria is lost.
-This was accompanied by a marked increase in the number of  NAD+-consuming signalling enzymes.
-To explain these counterintuitive results we built a mathematical model of the pathway and demonstrated that NNMT has a critical role to maintain high NAD+-consuming signalling fluxes by preventing accumulation of inhibitory Nam.
+This was accompanied by a marked increase in the number of NAD-consuming signalling enzymes.
+To explain these counterintuitive results we built a mathematical model of the pathway and demonstrated that NNMT has a critical role to maintain high NAD-consuming signalling fluxes by preventing accumulation of inhibitory Nam.
 Our model furthermore predicts that NNMT likely exerted an evolutionary pressure on the NamPRT affinity development.
@@
 
-Taken together, our analyses suggest that the co-existence of NamPRT and NNMT has been a prerequisite to enable the evolutionary development of versatile NAD+-dependent signalling mechanisms present in vertebrates.
+Taken together, our analyses suggest that the co-existence of NamPRT and NNMT has been a prerequisite to enable the evolutionary development of versatile NAD-dependent signalling mechanisms present in vertebrates.

### 2018-06-26 latex/methods.tex
First round of spell checking done

@@
 
-Kinetic parameters (substrate affinity (Km) and turnover rates (kcat), substrate and product inhibitions) were retrieved from the enzyme database BRENDA and additionally evaluated by checking the original literature especially with respect to measurement conditions.
+Kinetic parameters (substrate affinity ($K_{M}$) and turnover rates ($k_{cat}$), substrate and product inhibitions) were retrieved from the enzyme database BRENDA and additionally evaluated by checking the original literature especially with respect to measurement conditions.
 Parameter values from mammals were used if available.
@@
 
-For eukaryotic expression with a C-terminal FLAG-epitope, the open reading frame (ORF) encoding human NamPT was inserted into pFLAG-CMV-5a (Merck - Sigma Aldrich) via EcoRI/BamHI sites.
-Using a PCR-based approach, this vector provided the basis for the generation of a plasmid encoding a NamPT deletion mutant lacking amino acid residues 42-51 (NamPT-$\Delta$42-51).
+For eukaryotic expression with a C-terminal FLAG-epitope, the open reading frame (ORF) encoding human NamPRT was inserted into pFLAG-CMV-5a (Merck - Sigma Aldrich) via EcoRI/BamHI sites.
+Using a PCR-based approach, this vector provided the basis for the generation of a plasmid encoding a NamPRT deletion mutant lacking amino acid residues 42-51 (NamPRT-$\Delta$42-51).
 For prokaryotic expression with an N-terminal 6xHis-tag, the wild-type and mutant ORFs were inserted into pQE-30 (Qiagen) via BamHI and PstI-sites.
@@
 
-The cells were harvested by centrifugation and resuspended in lysis buffer (20\,mM Tris-HCl pH 8.0, 500\,mM NaCl, 4\,mM dithiltheritol (DTT), 1\,mg/mL lysozyme, 1X Complete EDTA-free protease inhibitor cocktail (Roche)).
+The cells were harvested by centrifugation and resuspended in lysis buffer (20\,mM Tris-HCl pH 8.0, 500\,mM NaCl, 4\,mM dithiothreitol (DTT), 1\,mg/mL lysozyme, 1X Complete EDTA-free protease inhibitor cocktail (Roche)).
 After sonification, the lysate was centrifuged for 30\,min at 13000\,g, and the clear lysate was incubated with 2\,mL of Nickel-NTA resin (Qiagen).

### 2018-06-26 latex/results.tex
First round of spell checking done

@@
 %%% TeX-master: "manuscript"
+% !TeX spellcheck=en_GB
+
 \section{Results}
@@
 
-NAD can be synthesized using several routes from three main precursors: tryptophan, nicotinamide (Nam) and nicotinic acid (NA) (Overview see Figure 1.).
+NAD can be synthesised using several routes from three main precursors: tryptophan, nicotinamide (Nam) and nicotinic acid (NA) (Overview see Figure 1.).
 Nam and NA are together known as vitamin B3 or niacin.
 Alternatively, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead \cite{Yoshino2018}.
-Du to the hight turnover of NAD observed and the fact that only 1\% of the tryptophan taken up with our diet is converted into NAD, vitamin B3 (Nam and NA) and to a lower extend NR are the major precursor of human NAD-biosynthesis.
+Due to the high turnover of NAD observed and the fact that only 1\% of the tryptophan taken up with our diet is converted into NAD, vitamin B3 (Nam and NA) and to a lower extend NR are the major precursor of human NAD biosynthesis.
 
-Looking acroos all known species,   two different pathways exist that recylce NAM.
+Looking across all known species, two different pathways exist that recycle Nam.
 The major pathway found in yeast and plants is using a four-step pathway starting with the deamination of Nam to nicotinic acid by NADA.
@@
 NMN and NAMN are converted into dinucleotides by the Nam/NA adenylytransferases (NMNATs).
-The recycling pathway via NA finally requires an amidation step catalysed by NADsynthase, driven by the conversion of ATP to AMP producing pyrophosphate.
+The recycling pathway via NA finally requires an amidation step catalysed by NAD synthase, driven by the conversion of ATP to AMP producing pyrophosphate.
 Even though the latter pathway seems to be very inefficient, it is the pathway preferentially used by most bacteria, fungi and plants (see Figure 2A), whereas most metazoans recycle Nam using NamPRT.
@@
 NNMT methylates Nam to methyl-Nam that is in mammals excreted with the urine, thus removing Nam from recycling.
-NNMT seems to have arisen de novo in the common ancestor of Ecdysozoa and Lophotrochozo.
+NNMT seems to have arisen \textit{de novo} in the common ancestor of Ecdysozoa and Lophotrochozoa.
 We were not be able to find any gene with considerable similarity in fungi or plants, even though in fungi a genes named NNMT can be found in databases.
@@
 In Deuterostomia the only large class that does only have NamPRT and seems to have lost NNMT again are Sauropsida and here especially birds.
-The reason why a lot of birds do not encode NNMT remains unclear, as the appearance is quite scattered (not shown)\todo[author=Mathias]{Maybe as supplementary figure?}.
+The reason why a lot of birds do not encode NNMT remains unclear, as the appearance is quite scattered (figure S2\todo[author=Mathias B.]{Has to be made!}).
 It might be related to the excretion system, as the product of NNMT methyl-nicotinamide is in mammals excreted with the urine.
@@
 To do so we used the previously established classification into 10 different enzyme families \cite{Gossmann2012FEBS}.
-The detailed list of templates used for the phylogenetic analysis can be found in  Supplementary Table 1 \todo[author=Mathias]{Inlcude number and order and name Supplementary Tables accordingly}.
+The detailed list of templates used for the phylogenetic analysis can be found in Supplementary Table 1 \todo[author=Mathias B.]{Include number and order and name Supplementary Tables accordingly}.
 The numbers in Figure 2B denote the average number of NAD-dependent signalling enzyme families found in each taxa.
@@
 First we addressed the question of the predominant co-existence of NamPRT and NNMT in vertebrates that coincides with an increase in the number of NAD-consuming enzyme families.
-We thus used our dynamic model to simulate steady state concentrations and NAD-consumption rates simulating NAD biosynthesis via  NamPRT in the  presence or absence of NNMT.
+We thus used our dynamic model to simulate steady state concentrations and NAD consumption rates simulating NAD biosynthesis via NamPRT in the presence or absence of NNMT.
 We see that the presence of NNMT enables higher NAD consumption rates (Figure3A) while, as expected, reducing the steady state concentration of NAD (Figure 3B).
-The decline in NAD-concentration can be compensated by a higher expression of NamPRT further increasing NAD consumption flux (Figure 3A and B dashed lines).
+The decline in NAD concentration can be compensated by a higher expression of NamPRT further increasing NAD consumption flux (Figure 3A and B dashed lines).
 
@@
 The increase of NAD consumption flux is caused by the fact that most NAD-consuming enzymes are inhibited by their product Nam, explaining why the presence of NNMT enables higher NAD consumption fluxes.
-At the same time, the high substrate affinity of NamPRT maintains a sufficiently high NAD-concentration.
+At the same time, the high substrate affinity of NamPRT maintains a sufficiently high NAD concentration.
 
-As kinetic parameters of NamPRT are only available for the human enzyme \cite{Burgos2008}, we went on to theoretically analysed the potential effect of the NamPRT affinity ($K_M$) on NAD steady state concentration and NAD consumption flux.
-In the absence of NNMT (Figure 4A-B) a change in $K_M$ has very little effect on steady state NAD concentration and NAD consumption flux.
-In the presence of NNMT, however, NAD consumption flux and NAD concentration increases with decreasing $K_M$ values (Figure 4C-D).
+As kinetic parameters of NamPRT are only available for the human enzyme \cite{Burgos2008}, we went on to theoretically analysed the potential effect of the NamPRT affinity ($K_{M}$) on NAD steady state concentration and NAD consumption flux.
+In the absence of NNMT (Figure 4A-B) a change in $K_{M}$ has very little effect on steady state NAD concentration and NAD consumption flux.
+In the presence of NNMT, however, NAD consumption flux and NAD concentration increases with decreasing $K_{M}$ values (Figure 4C-D).
 
@@
 
-When we compare NAD consumption and NAD concentration with and without NNMT with two different substrate affinities of NamPRT, we see that at low affinity ($K_M$ of 1 $\mu$M, which is in the range of the $K_M$  of NADA for Nam), NAD consumption flux is only higher with NNMT at low cell division rates.
+When we compare NAD consumption and NAD concentration with and without NNMT with two different substrate affinities of NamPRT, we see that at low affinity ($K_{M}$ of 1 $\mu$M, which is in the range of the $K_{M}$ of NADA for Nam), NAD consumption flux is only higher with NNMT at low cell division rates.
 This effect becomes stronger with decreasing affinity (Figure 4E-F).
@@
 
-As our simulations suggest that NNMT might have excerted an evolutionary pressure on the development of NamPRT.
-We performed multiple sequence analysis to see if we can find sequence variations in the protein sequence of NamPRT that indicate evolutionary changes of NamPRT upon the appearance of NNMT.
+As our simulations suggest that NNMT might have exerted an evolutionary pressure on the development of NamPRT, we created a multiple sequence alignment to see if we can find sequence variations in the protein sequence of NamPRT that indicate evolutionary changes of NamPRT upon the appearance of NNMT.
 The multiple sequence alignment of selected sequences is shown in Figure 5A and a more comprehensive multiple sequence alignment containing a larger number of species can be found in Supplementary Figure 1.
-We recognized that most Deuterostomia that have only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 43 to 52 of the human enzyme.
+We recognised that most Deuterostomia that have only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 43 to 52 of the human enzyme.
 This insert overlaps with a predicted weak nuclear localisation signal, that is lost when removing this sequence stretch.
-The inserted sequence corresponce to a structurallly unresolved loop in all avaibable crystal structure of human NamPR (e.g.
+The inserted sequence corresponds to a structurally unresolved loop in all available crystal structure of human NamPRT (e.g.
 \cite{Wang2006} structure visualisation Figure 5B).
@@
 The second hypothesis was that the sequence insertion influences NamPRT affinity, which is what our mathematical simulations would predict.
-We thus recombinantly expressed  both wildtype and mutant  NamPRT  in E.
-coli.
+We thus recombinantly expressed both wildtype and mutant NamPRT in \textit{E.~coli}.
 Upon purification we measured the activity using NMR detection of NMN produced in the presence and absence of ATP.
@@
 The latter might explain why in bacteria that have relatively high growth rates both systems co-exist.
+
 In the presence of NNMT the NamPRT compartment out-competes the NADA containing compartment, both with respect to NAD consumption rates and steady state NAD concentrations (Figure 7 C and D).
-This effect is dependent on a high affinity of NamPRT, as at low affinity (high $K_M$ ) the NADA containing system is again able to maintain higher NAD-concentrations.
+This effect is dependent on a high affinity of NamPRT, as at low affinity (high $K_{M}$) the NADA containing system is again able to maintain higher NAD concentrations.
 The NAD consumption flux is still higher in the NamPRT and NNMT containing compartment.

### 2018-06-26 latex/latex_to_word.py
Some minor changes to the converter

@@ -61,9 +61,9 @@ first_replacements = {
 hard_replacements = {
 	r'\newpage': '',
 	r'\S': '§',
-	r'\beta': 'β',
-	r'\mu': 'µ',
-	r'\Delta': 'Δ',
+	r'$\beta$': 'β',
+	r'$\mu$': 'µ',
+	r'$\Delta$': 'Δ',
 	r'\,': '&thinsp;'}
 
 for key in first_replacements:
@@ -177,6 +177,10 @@ for line in lines:
 			line = re.sub(r'\\url\{([^\}\{\\]+)\}', r'<a href="\1">\1</a>', line)
 		if r'\href' in line:
 			line = re.sub(r'\\href\{([^\}\{\\]+)\}\{([^\}\{\\]+)\}', r'<a href="\1">\2</a>', line)
+		if r'\textsuperscript' in line:
+			line = re.sub(r'\\textsuperscript\{([^\}]+)\}', r'<sup>\1</sup>', line)
+		if r'\textsubscript' in line:
+			line = re.sub(r'\\textsubscript\{([^\}]+)\}', r'<sub>\1</sub>', line)
 
 		counter += 1
 

### 2018-06-26 latex/discussion.tex

@@
 New approaches are required to disentangle the underlying network, helping us to understand how alterations affect human physiology.
-We here combined phylogenetic analysis and mathematical modelling the NAD biosynthesis and consumption pathway to better understand the evolution of this pathway and  especially the role of the methyltransferase NNMT.
-Our analysis reveals that that NNMT plays a vital role for the diversification of NAD-dependent signalling reactions and potentially also for NAD homeostasis.
+We here combined phylogenetic analysis and mathematical modelling of the NAD biosynthesis and consumption pathway to better understand the evolution of this pathway and  especially the role of the methyltransferase NNMT.
+Our analysis reveals that NNMT plays a vital role for the diversification of NAD-dependent signalling reactions and potentially also for NAD homeostasis.
 We show that the presence of NNMT enables higher NAD consumption rates in comparison to NamPRT or NADA alone.
 It furthermore appears that NNMT makes both NAD concentration and NAD consumption relatively independent of other NAD requiring processes such as cell growth.
-We furthermore suggest that NNMT exerted an evolutionary pressure on NamPRT driving the development of the high affinity of NamPRT for its substrates found in the human enzyme \cite{Burgos2008}.
-These finding shed new light on the role of NNMT, which has  earlier  been recognized  as potential marker for some types of cancer (e.g.
-\cite{Okamura1998}), hinting that increased NNMT expression in cancer might serve to remove  NAM produced by increased NAD-dependent signalling.
-Further analysis is required to better understand which NAD-signalling reactions are affected.
-But given a better mechanistic understanding and as routine clinical measurements of urine methyl-Nam (MNam) concentrations are available already, MNam could potentially serve as biomarker in the context of different diseases (e.g\cite{Pumpo2001,Delaney2005}).
+We furthermore suggest that NNMT exerted an evolutionary pressure on NamPRT, driving the development of the high affinity of NamPRT for its substrates found in the human enzyme \cite{Burgos2008}.
+These findings shed new light on the role of NNMT, which has earlier been recognized  as potential marker for some types of cancer (e.g.
+\cite{Okamura1998}).
+Increased NNMT expression in cancer might thus serve to remove  NAM produced by increased NAD-dependent signalling.
+Further analysis is required to better understand which NAD-dependent signalling reactions are affected.
 
@@
 
-We here neglected the potential effects of the methyl donor S-adenosyl-methionin (SAM) and its precursor methionine, although it is most likely contributing yet another regulatory level for Nam availability and thus an additional interaction point between gene regulation and metabolism.
+We here neglected the potential effects of the methyl donor S-adenosyl-methionin (SAM) and its precursor methionine.
+Changing SAM availibility is most likely contributing yet another regulatory level for Nam availability and thus an additional interaction point between gene regulation and metabolism.
 It has been shown earlier that NNMT expression influences protein methylation dependent on methionine availability \cite{Ulanovskaya2013}.
 One of the challenges in the analysis of this aspect is the availability of \textit{in vivo} concentration measurements for SAM and the large amount of reactions using it as substrate.
-A mathematical model of this pathway has  been published \cite{PMID14637052}  providing the basis for future studies.
+A simplified mathematical model of this pathway has  been published earlier \cite{PMID14637052}  providing the basis for future studies.
 
-Beside gaining new insights into the role of NNMT for NAD-metabolism, our analyses furthermore provided a suggestion why NADA disappears in vertebrates while it co-exists with NamPRT in bacteria.
-It seems that the combined enzymatic activity of NNMT and NamPRT is optimised to maintain stable NAD-concentration and high NAD-consumption flux, making NADA obsolete.
+Beside gaining new insights into the role of NNMT for NAD-metabolism, our analyses  provided a suggestion why NADA disappears in vertebrates, while it co-exists with NamPRT in bacteria.
+It seems that the combined enzymatic activity of NNMT and NamPRT has been optimised in vertebrates to maintain stable NAD-concentration and high NAD-consumption flux, making NADA obsolete.
 
@@
 This again appears to be accompanied by the loss of NADA in vertebrates and the first gene duplication of NMNATs \cite{Lau2010}.
-We also noted that the second gene duplication of NMNATs and thus the further compartmentalisation of NAD metabolism is co-occurring with a site specific positive selection event in NNMT (data not shown).
+We also noted that the second gene duplication of NMNATs and thus the further compartmentalisation of NAD metabolism, is co-occurring with a site specific positive selection event in NNMT (unpublished results).
 To our knowledge this is the first study that comprehensively demonstrates a functional co-evolution of several enzymes of a pathway.

### 2018-06-26 latex/manuscript.tex
added acknowledgment text and revised summary

@@
 In eukaryotes, two different Nam salvage pathways exist.
-While in lower organisms the initial deamidation of Nam is prevalent, the direct conversion of Nam to the mononucleotide by Nam phosphoribosyltransferase (NamPRT) dominates in animals and eventually remains as the single Nam recycling route in vertebrates.
-
+Phylogenetic analysis of the pathway show that in lower organisms the initial deamidation of Nam is prevalent, whilst the direct conversion of Nam to the mononucleotide by Nam phosphoribosyltransferase (NamPRT) dominates in animals and eventually remains as the single Nam recycling route in vertebrates.
 Strikingly, loss of the deamidation pathway in early vertebrates is preceded by the occurrence of a new enzyme that marks Nam for excretion by methylation – nicotinamide N-methyltransferase (NNMT).
@@
 Why is there an enzyme that removes Nam from recycling, seemingly not having any other physiological role?
-And why is the occurrence of NNMT accompanied by a diversification of NAD-consuming enzymes?
+Despite the occurance of this Nam degrading enzyme NNMT,  we observe  a diversification of NAD dependent signalling enzymes in vertebrates.
+This seems counterintuitive.
 
-We have used mathematical modelling approaches to resolve these counterintuitive observations.
+To better understand these evolutionary changes and the role of NNMT for NAD metabolism, we built a mathematical model of the pathway using published enzyme kinetics data.
 Our results indicate that NNMT is required to enable high NAD consumption fluxes necessitated by the increasing diversification of NAD-dependent signalling pathways.
@@
 Indeed, the affinity of NamPRT to Nam has previously been measured to be in the nanomolar range.
-Mathematical modelling supports the hypothesis that NNMT exerted an evolutionary pressure on NamPRT enforcing the development of its unusually high substrate affinity.
+Based on our mathematical modelling approach we furthermore suggest that NNMT exerted an evolutionary pressure on NamPRT, enforcing the development of its unusually high substrate affinity.
 Using multiple sequence alignments, we identified a sequence insertion, first occurring in vertebrates, that parallels an experimentally verified increase in the substrate affinity of the enzyme.
-Further simulations show that the deamidation pathway became obsolete owing to the high substrate affinity of NamPRT.
+Additional simulations show that the deamidation pathway became obsolete owing to the high substrate affinity of NamPRT.
 Collectively, our results illustrate a close evolutionary relationship between NAD biosynthesis and the diversification NAD-dependent signalling pathways, potentially driven by the concomitant occurrence of a regulator of Nam salvage, NNMT.
@@
 
-\section{Acknowledgements}
+\section*{Acknowledgements}
+
+We thank the Norwegian Research Council for funding (grant no.
+250395/F20).
+We furthermore thank for the  computation time provided  through UNINETT Sigma2 – the National Infrastructure for High Performance Computing and Data Storage in Norway.
 

### 2018-06-26 latex/introduction.tex
Added some references and fixed some sentences

@@
 NAD metabolism represents one of the most critical links that connect cellular signal transduction and energy metabolism.
-Even though best known as cofactor for various redox-reactions , NAD is involved in a number of signalling processes that consume NAD$^{+}$ by cleaving the molecule to nicotinamide (Nam) and ADP-ribose \cite{Verdin2015}.
-These NAD-dependent signalling reactions include but are not limited to poly- and mono-ADP-ribosylation, NAD-dependent protein deacylation by sirtuins as well as the synthesis of calcium-mobilizing molecules such as cyclic ADP-ribose \cite{?}.
+Even though best known as cofactor for various redox-reactions, NAD is involved in a number of signalling processes that consume NAD$^{+}$ by cleaving the molecule to nicotinamide (Nam) and ADP-ribose \cite{Verdin2015}.
+These NAD-dependent signalling reactions include but are not limited to poly- and mono-ADP-ribosylation \cite{Butepage2015,DeVos2012}, NAD-dependent protein deacylation by sirtuins \cite{Osborne2016}, and the synthesis of calcium-mobilizing molecules such as cyclic ADP-ribose \cite{Lee2012}.
 These NAD-dependent signalling processes participate in the regulation of virtually all cellular activities.
 The enzymes involved in these processes are sensitive to the available NAD concentration \cite{Ruggieri2015}, which in turn is dependent on the NAD$^{+}$/NADH redox ratio.
-Therefore, NAD-dependent signalling can act as a transmitter of changes in the cellular energy homeostasis, for example, to regulate gene expression or metabolic activity \cite{Koch-Nolte2009}.\todo[author=Mathias B.]{Please add more references to the first paragraph.
-I think there are better references then my own review.}
+Therefore, NAD-dependent signalling can act as a transmitter of changes in the cellular energy homeostasis, for example, to regulate gene expression or metabolic activity \cite{Koch-Nolte2009}.
 
@@
 However, it has now been established that substances affecting NAD biosynthesis lead to a rapid decline of the NAD concentration \cite{Buonvicino2018} suggesting that NAD-dependent signalling consumes substantial amounts of NAD, which is why we later refer to them also as NAD-consuming reactions.
-The resulting NAD turnover differs in a cell type specific manner and can lead to an NAD half-life as short as 2 hours \cite{Liu2018}.
+The resulting NAD turnover differs in a cell type specific manner and can lead to an NAD half-life as short as two hours \cite{Liu2018}.
 To maintain the NAD concentration at physiological levels, NAD biosynthesis needs to act at an equally rapid rate.
-Imbalances in NAD homeostasis have been linked to various, in particular, ageing-related diseases such as diabetes, neurodegenerative disorders and cancer \cite{Chiarugi2012,Verdin2015}.
-Several recent studies have demonstrated impressive health benefits of dietary supplementation with intermediates of NAD biosynthesis including NMN and nicotinamide riboside, (NR) \cite{Yoshino2018}.
-Apparently, the exploitation of NAD biosynthetic routes, in addition to the use of nicotinamide as precursor (Fig.
-1), results in increased NAD concentrations that stimulate NAD-dependent signalling processes, in particular, protein deacetylation by sirtuins \cite{North2004}.
+Imbalances in NAD homeostasis have been linked to various diseases, in particular related to ageing, such as diabetes, neurodegenerative disorders, and cancer \cite{Chiarugi2012,Verdin2015}.
+Several recent studies have demonstrated impressive health benefits of dietary supplementation with intermediates of NAD biosynthesis including Nam mononucleotide (NMN) and Nam riboside, (NR) \cite{Yoshino2018}.
+Apparently, the exploitation of NAD biosynthetic routes, in addition to the use of nicotinamide as precursor (Fig.~\ref{fig:pathway_overview}), results in increased NAD concentrations that stimulate NAD-dependent signalling processes, in particular, protein deacetylation by sirtuins \cite{North2004}.
 
@@
 If Nam would not be constantly recycled into NAD, humans would require a much higher daily vitamin B3 intake than the 16 mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}.
-In the first step of the salvage pathway, Nam is converted to the mononucleotide, NMN, by Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate.
+In the first step of the salvage pathway, Nam is converted to the mononucleotide NMN by the Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate.
 The nearly complete recycling of Nam is achieved by an extraordinary high affinity of NamPRT to Nam, the $K_{M}$ being in the low nanomolar range \cite{Burgos2008}.
@@
 The presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012FEBS} is among the most enigmatic and counterintuitive features of NAD metabolism.
-Why is there one enzyme (NamPRT) seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, while at the same time there is another enzyme (NNMT) that seems to have no metabolic function other than to remove Nam from NAD metabolism, This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deamidated Nam to nicotinic acid (NA) using the nicotinamide deamidase (NADA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (Fig.
-1).
+Why is there one enzyme (NamPRT) seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, while at the same time there is another enzyme (NNMT) that seems to have no metabolic function other than to remove Nam from NAD metabolism, This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deamidate Nam to nicotinic acid (NA) using the nicotinamide deamidase (NADA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (Fig.~\ref{fig:pathway_overview}).
 
-We here present a phylogenetic analysis of NAD pathway in eukaryotes that.
-In contrast to our previous analysis, which has been limited by the low number of eukaryotic genomes available at the time, we are now able to provide a comprehensively analysis of the eukaryotic pathway evolution.
-The results show there has seemingly been a selection for the co-existence of NamPRT and NNMT in Deuterostomia, while the pathway dominant in bacteria is lost.
-This was accompanied by a marked increase in the number of NAD-consuming signalling enzymes.
+We here present a phylogenetic analysis of the NAD pathways in eukaryotes.
+In contrast to our previous analysis \cite{Gossmann2012FEBS}, which has been limited by the low number of eukaryotic genomes available at the time, we are now able to provide a comprehensive analysis of the eukaryotic pathway evolution.
+The results suggest that there has been a selection for the co-existence of NamPRT and NNMT in Deuterostomia, while the pathway dominant in bacteria is lost.
+This transition was accompanied by a marked increase in the number of NAD-consuming signalling enzymes.
 To explain these counterintuitive results we built a mathematical model of the pathway and demonstrated that NNMT has a critical role to maintain high NAD-consuming signalling fluxes by preventing accumulation of inhibitory Nam.
-Our model furthermore predicts that NNMT likely exerted an evolutionary pressure on the NamPRT affinity development.
-Simulating the resource competition we furthermore show that the presence of high affinity NamPRT together with NNMT makes the NADA dependent pathway obsolete.
-Based on multiple sequence alignments, we identified a sequence insertion in NamPRT in Deuterostomes that is affecting the affinity of NamPRT, undermining the predictions derived from our mathematical modelling approach
+In addition, our model predicts that NNMT likely exerted an evolutionary pressure on the NamPRT affinity development.
+Simulating the resource competition we furthermore show that the presence of high affinity NamPRT together with NNMT makes the NADA-dependent pathway obsolete.
+Based on multiple sequence alignments, we identified a sequence insertion in NamPRT in Deuterostomes.
+We could show experimentally that this insertion increases the affinity of NamPRT, supporting the predictions derived from our mathematical modelling approach.
 

### 2018-06-26 latex/figurelegends.tex
Wrote first draft of figure legend for fig 8
Fixed error that led to compile error
revised figure legends

@@
   \includegraphics{fig_1_pathway_overview}
-  \caption{\textbf{Schematic overview of NAD biosynthesis and consumption.} NAD can be synthesised from tryptophan (Trp), nicotinamide (Nam), nicotinic acid (NA), and, to a lesser extend, nicotinamide ribose (NR).
-Nam is the main precursor and also the product of NAD-consuming signalling reactions by enzymes such as sirtuins (NAD-dependent histone deacetylases) or PARPs (poly-ADP-ribosylases).
+  \caption{\textbf{Schematic overview of NAD biosynthesis pathways} NAD can be synthesised from tryptophan (Trp), nicotinamide (Nam), nicotinic acid (NA), and, to a lesser extend, nicotinamide ribose (NR).
+Nam is the main precursor in human and also the product of NAD-consuming signalling reactions by enzymes such as sirtuins (NAD-dependent histone deacetylases) or PARPs (poly-ADP-ribosylases).
 For the recycling of Nam, two different pathways exist.
@@
 The pathway found in vertebrates directly converts Nam into the corresponding mononucleotide (NMN) by the Nam phosphoribosyltransferase (NamPRT).
-A third enzyme can consume Nam, the Nam N-methyltransferase (NNMT).
-For more details and abbreviations, see text.}
+The Nam N-methyltransferase (NNMT) degrades Nam to methyl-Nam (MNam), which is in mammals excreted with the urine.}
   \label{fig:pathway_overview}
@@
 NADA is dominant in Bacteria, Fungi, and Plants (Viridiplantae), whereas NamPRT together with NNMT is dominant in Metazoa.
-Numbers at the pie charts show, how many species of the taxon possess the respective enzyme combination indicated by the colour explained in the lower right of the figure.
+Numbers at the pie charts show, how many species of the taxon possess the respective enzyme combination indicated by the colour, which is explained in the lower right of the figure.
 Below the taxon name, the number of species in that taxon is given.
 B) Common tree of selected taxa within the Metazoa, including 334 species.
-The pie charts indicate the distribution of species within the respective taxon that have the enzyme combination indicated by the colour explained in the lower right.
+The pie charts indicates the distribution of species within the respective taxon that encode the enzyme combination indicated by the different colours.
 The size of the pie charts is proportional to the logarithm of the number of species analysed in the particular taxon.
@@
   \includegraphics[width=\textwidth]{fig_3_NNMT_NAD_flux}
-  \caption{\textbf{NNMT enables high NAD consumption flux.} We used a dynamic model of NAD biosynthesis and consumption (details see Materials and Methods) to simulate NAD consumption flux (A) and NAD concentration (B) in the presence of NamPRT and with or without NNMT at different cell division rates.
-NNMT enables higher steady state NAD consumption flux despite reduced NAD concentrations.}
+  \caption{\textbf{NNMT enables high NAD consumption flux.} We used a dynamic model of NAD biosynthesis and consumption (details see Materials and Methods) to simulate NAD consumption flux.
+The amount of NMNAT and NamPRT used in the simulations where adjusted such that the free NAD concentrations were in the range reported in the literature.
+All other enzymes concentration were set equal.
+Details are given in Supplementary table 2.
+In the absence of NNMT (blue lines) steady state NAD-consumption rates (A) are higher despite reduced NAD concentrations (B).
+Increasing the amount of NamPRT in the simulation 4 fold  (blue dotted lines) partially compensated for the decreased NAD steady state concentration caused by Nam degradation through NNMT.
+}
   \label{fig:NNMT_NAD_flux}
@@
   \caption{\textbf{Role of NamPRT affinity for Nam.} Using the dynamic model of NAD biosynthesis and consumption we simulated the affect of different Michaelis-Menten constants ($K_M$) of NamPRT for Nam on steady state NAD consumption flux and NAD concentration at different cell division rates.
+All ther parameter were equal to those used for the simulations in Figure 3.
 In the absence of NNMT (A and B), the $K_M$ of NamPRT has little influence on NAD consumption and concentration, but both are changing with cell division rates.
 In the presence of NNMT (C and D), decreasing $K_M$ of NamPRT enables increasing NAD consumption flux and NAD concentration.
-NNMT furthermore makes both, consumption flux and concentration, relatively independent of cell division rates.
-Comparing the situation with and without NNMT (E and F) at different NamPRT $K_M$ reveals that at low $K_M$ and high cell division rates NNMT no longer enables higher NAD consumption rates compared to NamPRT alone.}
+NNMT furthermore makes both, NAD consumption flux and concentration, relatively independent of cell division rates.
+Comparing the situation with and without NNMT (E and F) at different NamPRT $K_M$ reveals that at low $K_M$ (dashed lines) and high cell division rates NNMT no longer enables higher NAD consumption rates compared to NamPRT alone (green line and dashed black line).}
   \label{fig:NamPRT_affinity_Nam}
@@
   \includegraphics[width=\textwidth]{fig_5_optimal_substr_affinities}
-  \caption{\textbf{The substrate affinities of human NNMT and NamPRT are optimal.} Both NAD consumption flux (A) and NAD concentration (B) are increasing with decreasing $K_M$ of NamPRT but decreasing $K_M$ of NNMT.
+  \caption{\textbf{The substrate affinities of human NNMT and NamPRT are optimal.} Using the same model as in Figure 3 and 4, we simulated the impact of changes in the $K_M$  for both NamPRT and NNMT on NAD consumption rates (A) and NAD concentration (B).
+Both are increasing with decreasing $K_M$ of NamPRT but decreasing $K_M$ of NNMT.
 The affinities reported for human enzymes (indicated by a black asterisk) appear to be close to optimal, as further improvements would have little or no effect on NAD consumption or concentration.}
@@
   \caption{\textbf{The function of the structurally unresolved loop structure of NamPRT.} Most Deuterostomia that encode NNMT show a sequence insertion in the N-terminal region of NamPRT that has been revealed by multiple sequence alignment of NamPRT from different species (A).
-Coloured circles indicate the enzymes present in the species besides NamPRT; blue: NNMT; black: NADA and NNMT; yellow: NADA.
+Coloured circles indicate the enzymes present in the respective species blue: NamPRT and NNMT; black: NamPRT, NADA and NNMT; yellow: NamPRT and NADA.
 For a more comprehensive alignment, please see supplementary figure~S1.
-B) The visualisation of human NamPRT is based on a structure prediction of SWISS-MODEL \cite{Arnold2006,Biasini2014} of the sequence of the human NamPRT using the model 2H3D as template \cite{Wang2006}.
-The inserted region is not resolved in crystal structures of NamPRT and thus appears to be a flexible loop structure at the surface of the NamPRT dimer, coloured in red.
-C) The localisation of the FLAG-tagged mutant protein lacking the unresolved loop is not changed compared to wildtype human NamPRT.
-Both show a heterogeneous nuclear cytosolic localisation in immunofluorescence images in HeLa S3 cells.
-But \textit{in vitro} measurements using recombinant protein show that the mutant NamPRT has lower activity than the wildtype enzyme (D) and is not activated by ATP (E).
-Bars in D) and E) with different letters indicate significant difference as estimated with a T test assuming independent samples and significance at $p < 0.05$.}
+The visualisation of human NamPRT (B) is based on a structure prediction of SWISS-MODEL  \cite{Arnold2006,Biasini2014}  using the model 2H3D of the human NamPRT as template \cite{Wang2006}.
+The inserted region is not resolved in  any of currently available crystal structures of NamPRT and thus appears to be a flexible loop structure at the surface of the NamPRT dimer, coloured in red.
+Immunofluorescence images show that the localisation of the FLAG-tagged mutant protein lacking the unresolved loop (C) is not changed compared to FLAG-tagged human wildtype NamPRT .
+Both show a heterogeneous nuclear cytosolic localisation in HeLa S3 cells.
+\textit{In vitro} measurements using recombinant protein  show that the mutant NamPRT has lower activity than the wildtype enzyme (D) and is not activated by ATP (E).
+Bars in D) and E) with different letters indicate significant difference of measured values as estimated using a T test assuming independent samples and significance at $p < 0.05$.}
   \label{fig:unresolved_loop}
@@
   \caption{\textbf{NNMT provides a competitive advantage and makes NADA obsolete.} To simulate competition for common resources, we created a two compartment model where one compartment contained NADA but no NamPRT and the other compartment contained NamPRT either with or without NNMT but no NADA.
-In the absence of NNMT (A and B) the compartment containing NADA has slightly lower NAD consumption rates (A) but much higher NAD concentrations (B).
-In the presence of NNMT, however, both NAD consumption (C) and NAD concentration (D) are lower in the NADA compartment, but this effect is only observed at low NamPRT $K_M$.}
+NADA and NamPRT were simulated to be present at equal amounts.
+In the absence of NNMT (A and B) the compartment containing NADA has slightly lower NAD consumption rates (A) but much higher steady state NAD concentrations (B).
+In the presence of NNMT, however, both NAD consumption (C) and NAD concentration (D) are lower in the NADA compartment.
+This effect is dependent on a low NamPRT $K_M$.}
   \label{fig:NNMT_comp_advantage}
@@
 \newpage
+
+\subsection{Figure 8}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth]{fig_8_evo_events}
+  \caption{\textbf{Schematic representation of co-evolutionary events in the NAD pathway} Based on the phylogenetic analysis presented here and earlier work  \cite{Lau2010} we summarized and indicated important events in the evolution of NAD metabolism in Metazoa.}
+  \label{fig:evo_events}
+\end{figure}
+
+\todo{What about the positive selection in NNMT?}
+
+\newpage

### 2018-06-26 latex/introduction.tex
Revised intro and results

@@
 Several recent studies have demonstrated impressive health benefits of dietary supplementation with intermediates of NAD biosynthesis including Nam mononucleotide (NMN) and Nam riboside, (NR) \cite{Yoshino2018}.
-Apparently, the exploitation of NAD biosynthetic routes, in addition to the use of nicotinamide as precursor (Fig.~\ref{fig:pathway_overview}), results in increased NAD concentrations that stimulate NAD-dependent signalling processes, in particular, protein deacetylation by sirtuins \cite{North2004}.
+Apparently, the exploitation of NAD biosynthetic routes, in addition to the use of nicotinamide as precursor (fig.~\ref{fig:pathway_overview}), results in increased NAD concentrations that stimulate NAD-dependent signalling processes, in particular, protein deacetylation by sirtuins \cite{North2004}.
 
@@
 The presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012FEBS} is among the most enigmatic and counterintuitive features of NAD metabolism.
-Why is there one enzyme (NamPRT) seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, while at the same time there is another enzyme (NNMT) that seems to have no metabolic function other than to remove Nam from NAD metabolism, This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deamidate Nam to nicotinic acid (NA) using the nicotinamide deamidase (NADA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (Fig.~\ref{fig:pathway_overview}).
+Why is there one enzyme (NamPRT) seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, while at the same time there is another enzyme (NNMT) that seems to have no metabolic function other than to remove Nam from NAD metabolism, This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deamidate Nam to nicotinic acid (NA) using the nicotinamide deamidase (NADA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (fig.~\ref{fig:pathway_overview}).
 

### 2018-06-26 latex/results.tex
Revised intro and results

@@
 
-NAD can be synthesised using several routes from three main precursors: tryptophan, nicotinamide (Nam) and nicotinic acid (NA) (Overview see Figure 1.).
+NAD can be synthesised using several routes from three main precursors: tryptophan, nicotinamide (Nam) and nicotinic acid (NA) (Overview see fig.~\ref{fig:pathway_overview}).
 Nam and NA are together known as vitamin B3 or niacin.
 Alternatively, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead \cite{Yoshino2018}.
-Due to the high turnover of NAD observed and the fact that only 1\% of the tryptophan taken up with our diet is converted into NAD, vitamin B3 (Nam and NA) and to a lower extend NR are the major precursor of human NAD biosynthesis.
+Due to the high turnover of NAD observed and the fact that only 1\% of the tryptophan taken up with our diet is converted into NAD\todo{source}, vitamin B3 (Nam and NA) and to a lower extend NR are the major precursors of human NAD biosynthesis.
 
 Looking across all known species, two different pathways exist that recycle Nam.
-The major pathway found in yeast and plants is using a four-step pathway starting with the deamination of Nam to nicotinic acid by NADA.
+The major pathway found in yeast and plants uses a four-step pathway starting with the deamination of Nam to nicotinic acid by NADA.
 The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates.
@@
 The recycling pathway via NA finally requires an amidation step catalysed by NAD synthase, driven by the conversion of ATP to AMP producing pyrophosphate.
-Even though the latter pathway seems to be very inefficient, it is the pathway preferentially used by most bacteria, fungi and plants (see Figure 2A), whereas most metazoans recycle Nam using NamPRT.
+Even though the latter pathway seems to be very inefficient, it is the pathway preferentially used by most bacteria, fungi and plants (see fig.~\ref{fig:phylo_distribution}A), whereas most metazoans recycle Nam using NamPRT.
 
-Analysing the phylogeny of the NAD recycling enzymes in Metazoa in more detail reveals that not only does NamPRT replace NADA, but in most organisms, especially in Deuterostomia, NamPRT is found together with the Nam methyltransferase NNMT (Figure 2B).
+Analysing the phylogeny of the NAD recycling enzymes in Metazoa in more detail reveals that not only NamPRT replaces NADA, but in most organisms, especially in Deuterostomia, NamPRT is found together with the Nam methyltransferase NNMT (fig.~\ref{fig:phylo_distribution}B).
 NNMT methylates Nam to methyl-Nam that is in mammals excreted with the urine, thus removing Nam from recycling.
 NNMT seems to have arisen \textit{de novo} in the common ancestor of Ecdysozoa and Lophotrochozoa.
-We were not be able to find any gene with considerable similarity in fungi or plants, even though in fungi a genes named NNMT can be found in databases.
-These genes do, however, show very limited sequence similarities to the NNMT of nematodes or deuterostomes and the yeast protein has later on shown to be a lysin-protein-methyltransferase \cite{Wlodarski2011}.
+We were not be able to find any gene with considerable similarity in fungi or plants, even though in fungi genes named NNMT can be found in databases.
+These genes do, however, show very limited sequence similarities to the NNMT of nematodes or deuterostomes.
+The yeast protein has later been shown to be a lysin-protein-methyltransferase \cite{Wlodarski2011}.
 Nematodes are the only organisms that encode NNMT together with NADA without NamPRT being present.
@@
 The reason why a lot of birds do not encode NNMT remains unclear, as the appearance is quite scattered (figure S2\todo[author=Mathias B.]{Has to be made!}).
-It might be related to the excretion system, as the product of NNMT methyl-nicotinamide is in mammals excreted with the urine.
+The lack of NNMT might be related to the excretion system, as the product of NNMT methyl-Nam is in mammals excreted with the urine.
 There are some species where we could not find NamPRT or NADA but NNMT, we assume that this is due to incomplete genomes in the database.
@@
 In addition to the phylogenetic distribution of the two Nam salvage enzymes NADA and NamPRT, we looked at phylogenetic diversity of enzymes catalysing NAD-dependent signalling reactions.
-To do so we used the previously established classification into 10 different enzyme families \cite{Gossmann2012FEBS}.
-The detailed list of templates used for the phylogenetic analysis can be found in Supplementary Table 1 \todo[author=Mathias B.]{Include number and order and name Supplementary Tables accordingly}.
-The numbers in Figure 2B denote the average number of NAD-dependent signalling enzyme families found in each taxa.
-With the exception of Cnidaria and Lophotrochozoa, we find average 3 to 4 families in Protostomia, whereas most Deuterostomia have on average more than 8 families with an increasing diversification of enzymes within some of these families \cite{Gossmann2014DNAR}.
+To do so we used the previously established classification into ten different enzyme families \cite{Gossmann2012FEBS}.
+The detailed list of templates used for the phylogenetic analysis can be found in supplementary table~S1.
+The numbers in figure~\ref{fig:phylo_distribution}B denote the average number of NAD-dependent signalling enzyme families found in each taxa.
+With the exception of Cnidaria and Lophotrochozoa, we find average three to four families in Protostomia, whereas most Deuterostomia have on average more than eight families with an increasing diversification of enzymes within some of these families \cite{Gossmann2014DNAR}.
 
@@
 And why does NADA disappear in Deuterostomia although it is the predominant pathway in bacteria, plants and fungi?
-Given the complexity of the NAD metabolic network, this questions are difficult to be comprehensively addressed experimentally.
-We therefore built a dynamic model of NAD metabolism using existing kinetic data from the literature (for details see materials and methods and supplementary table 2).
+Given the complexity of the NAD metabolic network, these questions are difficult to be comprehensively addressed experimentally.
+We therefore built a dynamic model of NAD metabolism using existing kinetic data from the literature (for details see experimental procedures and supplementary table~S2).
 
 To be able to compare metabolic features of evolutionary quite different systems in our simulations, and as we have limited information about expression levels of enzymes or changes of kinetic constants during evolution, we initially used the kinetic constants found for yeast or human enzymes for all systems analysed and used equal amounts of enzymes for all reactions.
-Wherever possible we did not only include substrate affinities but also known product inhibition or inhibition by downstream metabolites.
+Wherever possible, we did not only include substrate affinities but also known product inhibitions or inhibition by downstream metabolites.
 As we assume that cell growth is, besides NAD-consuming reactions, a major driving force for NAD biosynthesis, we analysed different growth rates (cell division rates) by simulating different dilution rates for all metabolites.
 
-First we addressed the question of the predominant co-existence of NamPRT and NNMT in vertebrates that coincides with an increase in the number of NAD-consuming enzyme families.
+First, we addressed the question of the predominant co-existence of NamPRT and NNMT in vertebrates that coincides with an increase in the number of NAD-consuming enzyme families.
 We thus used our dynamic model to simulate steady state concentrations and NAD consumption rates simulating NAD biosynthesis via NamPRT in the presence or absence of NNMT.
-We see that the presence of NNMT enables higher NAD consumption rates (Figure3A) while, as expected, reducing the steady state concentration of NAD (Figure 3B).
-The decline in NAD concentration can be compensated by a higher expression of NamPRT further increasing NAD consumption flux (Figure 3A and B dashed lines).
+We see that the presence of NNMT enables higher NAD consumption rates (fig.~\ref{fig:NNMT_NAD_flux}A) while, as expected, reducing the steady state concentration of NAD (fig.~\ref{fig:NNMT_NAD_flux}B).
+The decline in NAD concentration can be compensated by a higher expression of NamPRT, further increasing NAD consumption flux (dashed lines in fig.~\ref{fig:NNMT_NAD_flux}A and~B).
 
-These findings can be explained when looking in more detail at the kinetic parameters of NamPRT and NAD-consuming enzymes such as Sirt1p.
+These findings can be explained when looking in more detail at the kinetic parameters of NamPRT and NAD-consuming enzymes such as Sirtuin~1.
 The increase of NAD consumption flux is caused by the fact that most NAD-consuming enzymes are inhibited by their product Nam, explaining why the presence of NNMT enables higher NAD consumption fluxes.
@@
 
-As kinetic parameters of NamPRT are only available for the human enzyme \cite{Burgos2008}, we went on to theoretically analysed the potential effect of the NamPRT affinity ($K_{M}$) on NAD steady state concentration and NAD consumption flux.
-In the absence of NNMT (Figure 4A-B) a change in $K_{M}$ has very little effect on steady state NAD concentration and NAD consumption flux.
-In the presence of NNMT, however, NAD consumption flux and NAD concentration increases with decreasing $K_{M}$ values (Figure 4C-D).
+As kinetic parameters of NamPRT are only available for the human enzyme \cite{Burgos2008}, we went on to theoretically analyse the potential effect of NamPRT affinity ($K_{M}$) on NAD steady state concentration and NAD consumption flux.
+In the absence of NNMT (fig.~\ref{fig:NamPRT_affinity_Nam}A and~B) a change in $K_{M}$ has very little effect on steady state NAD concentration and NAD consumption flux.
+In the presence of NNMT, however, NAD consumption flux and NAD concentration increases with decreasing $K_{M}$ values (fig.~\ref{fig:NamPRT_affinity_Nam}C and~D).
 
-It is interesting to note that, without NNMT, NAD concentration and consumption flux are both considerably affected by cell division rates, at least if the enzyme expression is kept constant.
-This is of course an artificial scenario, as one would assume organisms to regulate enzyme expression to achieve similar levels of metabolite concentrations instead.
+It is interesting to note that without NNMT, NAD concentration and consumption flux are both considerably affected by cell division rates, at least if the enzyme expression is kept constant.
+This is, of course, an artificial scenario, as one would assume organisms to regulate enzyme expression to achieve similar levels of metabolite concentrations instead.
 However, there seems to be an apparent trade off between maintainable NAD concentration and consumption flux in the absence of NNMT.
 In contrast, NAD consumption rates increase with NAD concentration in the presence of NNMT and both are relatively independent of cell division rates.
+\todo{The last sentence is unclear to me.
+Consumption increases with concentration?
+That is not in the figure.
+I needed to read often to understand what "NNMT and both" means.}
 
 When we compare NAD consumption and NAD concentration with and without NNMT with two different substrate affinities of NamPRT, we see that at low affinity ($K_{M}$ of 1 $\mu$M, which is in the range of the $K_{M}$ of NADA for Nam), NAD consumption flux is only higher with NNMT at low cell division rates.
-This effect becomes stronger with decreasing affinity (Figure 4E-F).
+This effect becomes stronger with decreasing affinity (fig.~\ref{fig:NamPRT_affinity_Nam}E and~F).
 
-The pathway dynamics are of course not solely dependent on one enzyme.
+The pathway dynamics are not solely dependent on one enzyme.
 Thus, what is the impact of the substrate affinity of NNMT that is competing with NamPRT for the same substrate?
-In Figure 5 we see that the substrate affinity values found in the human enzymes (indicated by black asterisks) are actually optimal with respect to both achievable steady state NAD concentration and consumption fluxes.
+In figure~\ref{fig:optimal_substr_affinities} we see that the substrate affinity values found in the human enzymes (indicated by black asterisks) are actually optimal with respect to both achievable steady state NAD concentration and consumption fluxes.
 Thus, a further increase of the affinity of NamPRT for Nam would not provide any advantage.
@@
 As our simulations suggest that NNMT might have exerted an evolutionary pressure on the development of NamPRT, we created a multiple sequence alignment to see if we can find sequence variations in the protein sequence of NamPRT that indicate evolutionary changes of NamPRT upon the appearance of NNMT.
-The multiple sequence alignment of selected sequences is shown in Figure 5A and a more comprehensive multiple sequence alignment containing a larger number of species can be found in Supplementary Figure 1.
-We recognised that most Deuterostomia that have only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 43 to 52 of the human enzyme.
+The multiple sequence alignment of selected sequences is shown in figure~\ref{fig:unresolved_loop}A and a more comprehensive multiple sequence alignment containing a larger number of species can be found in supplementary figure~S1.
+We recognised that most Deuterostomia that have only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 42 to 51 of the human enzyme.
 This insert overlaps with a predicted weak nuclear localisation signal, that is lost when removing this sequence stretch.
 The inserted sequence corresponds to a structurally unresolved loop in all available crystal structure of human NamPRT (e.g.
-\cite{Wang2006} structure visualisation Figure 5B).
-The loop is connected to one of the beta-sheets involved in substrate binding.
+\cite{Wang2006} structure visualisation fig.~\ref{fig:unresolved_loop}B).
+The loop is connected to one of the $\beta$-sheets involved in substrate binding.
 
-From these observations we derived two possible hypotheses regarding the role of the loop for NamPRT function.
+From these observations, we derived two possible hypotheses regarding the role of the loop for NamPRT function.
 Firstly, we analysed whether the deletion of the amino acids 42 to 51 has an effect on the localisation of the human enzyme.
-We thus performed immunofluorescence imaging with FLAG-tagged proteins, but both wildtype and mutant protein showed a mixed cytosolic nuclear localisation (Figure 6C).
-We therefore conclude that the partial nuclear localisation of NamPRT is not affected by the sequence deletion.
+We thus performed immunofluorescence imaging with FLAG-tagged proteins, but both wildtype and mutant protein showed a mixed cytosolic nuclear localisation (fig.~\ref{fig:unresolved_loop}C).
+We therefore conclude that the partial nuclear localisation of NamPRT is not affected by the sequence deletion\todo{As there are other NLS in the protein, this one could just be non-functional also with loop, couldn’t it?}.
+
 The second hypothesis was that the sequence insertion influences NamPRT affinity, which is what our mathematical simulations would predict.
 We thus recombinantly expressed both wildtype and mutant NamPRT in \textit{E.~coli}.
-Upon purification we measured the activity using NMR detection of NMN produced in the presence and absence of ATP.
-It appears that the enzymatic activity of the mutant enzyme is much lower than that of the wildtype enzyme (Figure 6D) and that an increase of the substrate concentration has a much stronger effect on the activity of the mutant compared to the wildtype enzyme, indicating a lower affinity.
-We furthermore showed that in contrast to NamPRT wildtype, the mutant does not exhibit an increased activity in the presence of ATP (Figure 6E).
+Upon purification, we measured the activity using NMR detection of NMN produced in the presence and absence of ATP.
+It appears that the enzymatic activity of the mutant enzyme is much lower than that of the wildtype enzyme (fig.~\ref{fig:unresolved_loop}D) and that an increase of the substrate concentration has a much stronger effect on the activity of the mutant compared to the wildtype enzyme, indicating a lower affinity.
+We furthermore showed that in contrast to NamPRT wildtype, the mutant does not exhibit an increased activity in the presence of ATP (fig.~\ref{fig:unresolved_loop}E).
 
@@
 
-The remaining question is now, why NADA is lost in vertebrates, while NADA and NamPRT are co-existing in bacteria?
-To analyse this question we built a two compartment model that has a shared Nam source (details see Materials an Methods and Supplementary Material).
+The remaining question is now, why NADA is lost in vertebrates, while NADA and NamPRT are co-existing in bacteria.
+To analyse this question we built a two compartment model that has a shared Nam source (details see experimental procedures and supplementary table~S2).
 One compartment contains NADA while the other contains either NamPRT alone at equal expression levels, or together with NNMT.
-Without NNMT the compartment containing NADA has slightly lower NAD consumption rates (Figure 7A), but is able to maintain much higher steady state NAD concentrations at low cell division rates (Figure 7B).
+Without NNMT the compartment containing NADA has slightly lower NAD consumption rates (fig.~\ref{fig:NNMT_comp_advantage}A), but is able to maintain much higher steady state NAD concentrations at low cell division rates (fig.~\ref{fig:NNMT_comp_advantage}B).
 At high cell division rates, steady state concentrations in both compartments are similar.
@@
 
-In the presence of NNMT the NamPRT compartment out-competes the NADA containing compartment, both with respect to NAD consumption rates and steady state NAD concentrations (Figure 7 C and D).
-This effect is dependent on a high affinity of NamPRT, as at low affinity (high $K_{M}$) the NADA containing system is again able to maintain higher NAD concentrations.
+In the presence of NNMT, the NamPRT compartment out-competes the NADA-containing compartment, both with respect to NAD consumption rates and steady state NAD concentrations (fig.~\ref{fig:NNMT_comp_advantage}C and~D).
+This effect is dependent on a high affinity of NamPRT, as at low affinity (high $K_{M}$), the NADA containing system is again able to maintain higher NAD concentrations.
 The NAD consumption flux is still higher in the NamPRT and NNMT containing compartment.

### 2018-06-26 supplements/KineticsTable/kinetics_table.tex
Renamed supplementary tables to fit the order in manuscripts
Fixed kinetics table

--- 2018-06-26 supplements/KineticsTable/Table_S1.tex
@@
-\documentclass[a4paper,10pt]{article} \usepackage[utf8x]{inputenc}
+\documentclass[a4paper,10pt]{article}
+\usepackage[utf8]{inputenc}
 \usepackage{amsmath}
@@
 
-\section*{Supplementary Table 1}
-\addcontentsline{toc}{section}{Supplementary Table 1}
+\section*{Supplementary Table 2}
+\addcontentsline{toc}{section}{Supplementary Table 2}
 \begin{longtable}{p{3cm}p{2cm}p{3cm}cc}
@@
     \href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC3/1/3/5.html}{3.1.3.5}
-    &$K_M$:100$\mu$M  & \cite{} & HMM \\
-    & & $k_{cat}$:1$s^{-1}$ 
+    &$K_{M_{NaMN}}$:3.5mM  & \cite{PMID26385918} & HMM \\
+    & & $k_{{cat}_{NaMN}}$:2.8$s^{-1}$
+    & & \\
+    & &$K_{M_{NMN}}$: 5mM   &  \\
+     & & $k_{{cat}_{NMN}}$:0.5$s^{-1}$
+     & & \\  \hline
+     PNP &
+    \href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC3/2/4/2.html}{2.4.2.1}
+    &$K_M$:1.48mM  & \cite{PMID9030766} & HMM \\
+    & & $k_{cat}$:40$s^{-1}$
+    & & \\  \hline
+    NRK &
+    \href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC2/7/1/173.html}{2.7.1.173}
+    &$K_M$:3.4$\mu$M & \cite{PMID19027704} & HMM \\
+    & & $k_{cat}$:0.23$s^{-1}$
     & & \\  \hline
@@
 
-The total enzyme concentration was set to 10 for all enzymes if not mentioned
-otherwise. Concentration of potentiantial cosubstrate was assumed to be constant
-and considered to be not limitng the reaction and thus represented by the
-maximal velocities given.
+The total enzyme concentration was set to 10 for all enzymes  except NamPRT and NMNAT, for which the concentration was set to 100. Concentration of potential co-substrate was assumed to be constant
+and not-limiting for the reaction. Thus they were  implicitly represented by the
+maximal velocities given in the table.
 
@@
 \bibliographystyle{plos2009}
-\bibliography{TableS1}
+\bibliography{tablelibrary}
 \end{document}

### 2018-06-26 supplements/KineticsTable/kinetics_table.tex
Added missing references to kinetics table

@@
 \section*{Supplementary Table 2}
-\addcontentsline{toc}{section}{Supplementary Table 2}
+
 \begin{longtable}{p{3cm}p{2cm}p{3cm}cc}
@@
     \href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC3/1/3/5.html}{3.1.3.5}
-    &$K_{M_{NaMN}}$:3.5mM  & \cite{PMID26385918} & HMM \\
+    &$K_{M_{NaMN}}$:3.5mM  & \cite{Kulikova2015} & HMM \\
     & & $k_{{cat}_{NaMN}}$:2.8$s^{-1}$
@@
     \href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC3/2/4/2.html}{2.4.2.1}
-    &$K_M$:1.48mM  & \cite{PMID9030766} & HMM \\
+    &$K_M$:1.48mM  & \cite{Wielgus-Kutrowska1997} & HMM \\
     & & $k_{cat}$:40$s^{-1}$
@@
     \href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC2/7/1/173.html}{2.7.1.173}
-    &$K_M$:3.4$\mu$M & \cite{PMID19027704} & HMM \\
+    &$K_M$:3.4$\mu$M & \cite{Dolle2009} & HMM \\
     & & $k_{cat}$:0.23$s^{-1}$

### 2018-06-26 latex/results.tex

@@
 However, there seems to be an apparent trade off between maintainable NAD concentration and consumption flux in the absence of NNMT.
-In contrast, NAD consumption rates increase with NAD concentration in the presence of NNMT and both are relatively independent of cell division rates.
+In contrast,  in the presence of NNMT NAD consumption rates increase with NAD concentration and both are relatively independent of cell division rates.
 \todo{The last sentence is unclear to me.
@@
 We thus performed immunofluorescence imaging with FLAG-tagged proteins, but both wildtype and mutant protein showed a mixed cytosolic nuclear localisation (fig.~\ref{fig:unresolved_loop}C).
-We therefore conclude that the partial nuclear localisation of NamPRT is not affected by the sequence deletion\todo{As there are other NLS in the protein, this one could just be non-functional also with loop, couldn’t it?}.
+We therefore conclude that the partial nuclear localisation of NamPRT is not affected by the sequence deletion.
 

### 2018-06-26 latex/figurelegends.tex
Marked sources of events

@@
 In the absence of NNMT (blue lines) steady state NAD-consumption rates (A) are higher despite reduced NAD concentrations (B).
-Increasing the amount of NamPRT in the simulation 4 fold  (blue dotted lines) partially compensated for the decreased NAD steady state concentration caused by Nam degradation through NNMT.
-}
+Increasing the amount of NamPRT in the simulation 4 fold (blue dotted lines) partially compensated for the decreased NAD steady state concentration caused by Nam degradation through NNMT.}
   \label{fig:NNMT_NAD_flux}
@@
   \includegraphics[width=\textwidth]{fig_5_optimal_substr_affinities}
-  \caption{\textbf{The substrate affinities of human NNMT and NamPRT are optimal.} Using the same model as in Figure 3 and 4, we simulated the impact of changes in the $K_M$  for both NamPRT and NNMT on NAD consumption rates (A) and NAD concentration (B).
-Both are increasing with decreasing $K_M$ of NamPRT but decreasing $K_M$ of NNMT.
+  \caption{\textbf{The substrate affinities of human NNMT and NamPRT are optimal.} Using the same model as in Figure 3 and 4, we simulated the impact of changes in the $K_{M}$ for both NamPRT and NNMT on NAD consumption rates (A) and NAD concentration (B).
+Both are increasing with decreasing $K_{M}$ of NamPRT but decreasing $K_{M}$ of NNMT.
 The affinities reported for human enzymes (indicated by a black asterisk) appear to be close to optimal, as further improvements would have little or no effect on NAD consumption or concentration.}
@@
 The inserted region is not resolved in any of currently available crystal structures of NamPRT and thus appears to be a flexible loop structure at the surface of the NamPRT dimer, coloured in red.
-Immunofluorescence images show that the localisation of the FLAG-tagged mutant protein lacking the unresolved loop (C) is not changed compared to FLAG-tagged human wildtype NamPRT .
+Immunofluorescence images show that the localisation of the FLAG-tagged mutant protein lacking the unresolved loop (C) is not changed compared to FLAG-tagged human wildtype NamPRT.
 Both show a heterogeneous nuclear cytosolic localisation in HeLa S3 cells.
@@
   \includegraphics[width=\textwidth]{fig_8_evo_events}
-  \caption{\textbf{Schematic representation of co-evolutionary events in the NAD pathway} Based on the phylogenetic analysis presented here and earlier work  \cite{Lau2010} we summarized and indicated important events in the evolution of NAD metabolism in Metazoa.}
+  \caption{\textbf{Schematic representation of co-evolutionary events in the NAD pathway} Based on the phylogenetic analysis presented here and earlier work \cite{Lau2010} we summarized and indicated important events in the evolution of NAD metabolism in Metazoa.
+Events in italics are from \cite{Lau2010}.
+All other events were found in this work.}
   \label{fig:evo_events}

### 2018-06-26 latex/discussion.tex
Minor changes and library update

@@
 % !TeX spellcheck=en_GB
+
 \section{Discussion}
 
-The dynamic interactions among metabolism, signal transduction, and gene regulation is still not very well understood.
+The dynamic interactions among metabolism, signal transduction, and gene regulation are still not very well understood.
 New approaches are required to disentangle the underlying network, helping us to understand how alterations affect human physiology.
 We here combined phylogenetic analysis and mathematical modelling of the NAD biosynthesis and consumption pathway to better understand the evolution of this pathway and especially the role of the methyltransferase NNMT.
-Our analysis reveals that NNMT plays a vital role for the diversification of NAD-dependent signalling reactions and potentially also for NAD homeostasis.
+Our analyses reveal that NNMT plays a vital role for the diversification of NAD-dependent signalling reactions and potentially also for NAD homeostasis.
 We show that the presence of NNMT enables higher NAD consumption rates in comparison to NamPRT or NADA alone.
-It furthermore appears that NNMT makes both NAD concentration and NAD consumption relatively independent of other NAD requiring processes such as cell growth.
+It furthermore appears that NNMT makes both NAD concentration and NAD consumption relatively independent of other NAD-requiring processes such as cell growth.
 We furthermore suggest that NNMT exerted an evolutionary pressure on NamPRT, driving the development of the high affinity of NamPRT for its substrates found in the human enzyme \cite{Burgos2008}.
@@
 \cite{Okamura1998}).
-Increased NNMT expression in cancer might thus serve to remove  NAM produced by increased NAD-dependent signalling.
+Increased NNMT expression in cancer might thus serve to remove Nam produced by increased NAD-dependent signalling.
 Further analysis is required to better understand which NAD-dependent signalling reactions are affected.
 
-In our analysis we furthermore noted that NNMT only provides an advantage as long as the NamPRT affinity for Nam is sufficiently high, this implies that treatment with competitive inhibitors that affect the apparent $K_M$ would have stronger effects on cells that express NNMT compared to cells that don´t express NNMT (see Figure 4E-F).
+In our analysis, we furthermore noted that NNMT only provides an advantage as long as the NamPRT affinity for Nam is sufficiently high, this implies that treatment with competitive inhibitors that affect the apparent $K_{M}$ would have stronger effects on cells that express NNMT compared to cells that do not express NNMT (see fig.~\ref{fig:NamPRT_affinity_Nam}E and~F).
 This could be important for cancer treatments as the expression of NNMT is relatively low in normal tissues except liver, whereas several types of cancer overexpress NNMT.
@@
 One of the challenges in the analysis of this aspect is the availability of \textit{in vivo} concentration measurements for SAM and the large amount of reactions using it as substrate.
-A simplified mathematical model of this pathway has  been published earlier \cite{PMID14637052}  providing the basis for future studies.
+A simplified mathematical model of this pathway has been published earlier \cite{Reed2004} providing the basis for future studies.
 
-Beside gaining new insights into the role of NNMT for NAD-metabolism, our analyses  provided a suggestion why NADA disappears in vertebrates, while it co-exists with NamPRT in bacteria.
-It seems that the combined enzymatic activity of NNMT and NamPRT has been optimised in vertebrates to maintain stable NAD-concentration and high NAD-consumption flux, making NADA obsolete.
+Beside gaining new insights into the role of NNMT for NAD metabolism, our analyses provided a suggestion why NADA disappears in vertebrates while it co-exists with NamPRT in bacteria.
+It seems that the combined enzymatic activity of NNMT and NamPRT has been optimised in vertebrates to maintain stable NAD concentration and high NAD consumption flux, making NADA obsolete.
 
@@
 
-We here highlight the co-evolution of several enzymes of the NAD-pathway with the appearance of NNMT seemingly initiating and driving complex alterations of the pathway such as an increase and diversification of NAD dependent signalling, followed by an increase in NamPRT substrate affinity (schematic overview see Figure 8).
+We here highlight the co-evolution of several enzymes of the NAD pathway with the appearance of NNMT seemingly initiating and driving complex alterations of the pathway such as an increase and diversification of NAD-dependent signalling, followed by an increase in NamPRT substrate affinity (schematic overview see fig.~\ref{fig:evo_events}).
 This again appears to be accompanied by the loss of NADA in vertebrates and the first gene duplication of NMNATs \cite{Lau2010}.
-We also noted that the second gene duplication of NMNATs and thus the further compartmentalisation of NAD metabolism, is co-occurring with a site specific positive selection event in NNMT (unpublished results).
-To our knowledge this is the first study that comprehensively demonstrates a functional co-evolution of several enzymes of a pathway.
+We also noted that the second gene duplication of NMNATs and thus the further compartmentalisation of NAD metabolism is co-occurring with a site-specific positive selection event in NNMT (unpublished results).
+To our knowledge, this is the first study that comprehensively demonstrates a functional co-evolution of several enzymes of a pathway.

### 2018-06-26 latex/results.tex
Minor changes and library update

@@
 However, there seems to be an apparent trade off between maintainable NAD concentration and consumption flux in the absence of NNMT.
-In contrast,  in the presence of NNMT NAD consumption rates increase with NAD concentration and both are relatively independent of cell division rates.
-\todo{The last sentence is unclear to me.
-Consumption increases with concentration?
-That is not in the figure.
-I needed to read often to understand what "NNMT and both" means.}
+In contrast, in the presence of NNMT, NAD consumption rates increase as do NAD concentrations and both are relatively independent of cell division rates.
 

### 2018-06-26 latex/figurelegends.tex
Some more minor fixes

@@
   \includegraphics[width=0.75\textwidth]{fig_2_phylo_distribution}
-  \caption{\textbf{Evolutionary distribution of NADA, NNMT and NamPRT and their relation to the number of NAD consumers.} A) Distribution of NADA, NNMT and NamPRT in selected major taxa.
+  \caption{\textbf{Evolutionary distribution of NADA, NNMT, and NamPRT and their relation to the number of NAD consumers.} A) Distribution of NADA, NNMT, and NamPRT in selected major taxa.
 NADA is dominant in Bacteria, Fungi, and Plants (Viridiplantae), whereas NamPRT together with NNMT is dominant in Metazoa.
@@
 B) Common tree of selected taxa within the Metazoa, including 334 species.
-The pie charts indicates the distribution of species within the respective taxon that encode the enzyme combination indicated by the different colours.
+The pie charts indicate the distribution of species within the respective taxon that encode the enzyme combination indicated by the different colours.
 The size of the pie charts is proportional to the logarithm of the number of species analysed in the particular taxon.
@@
   \includegraphics[width=\textwidth]{fig_3_NNMT_NAD_flux}
-  \caption{\textbf{NNMT enables high NAD consumption flux.} We used a dynamic model of NAD biosynthesis and consumption (details see Materials and Methods) to simulate NAD consumption flux.
+  \caption{\textbf{NNMT enables high NAD consumption flux.} We used a dynamic model of NAD biosynthesis and consumption (for details, see experimental procedures) to simulate NAD consumption flux.
 The amount of NMNAT and NamPRT used in the simulations where adjusted such that the free NAD concentrations were in the range reported in the literature.
-All other enzymes concentration were set equal.
-Details are given in Supplementary table 2.
-In the absence of NNMT (blue lines) steady state NAD-consumption rates (A) are higher despite reduced NAD concentrations (B).
-Increasing the amount of NamPRT in the simulation 4 fold (blue dotted lines) partially compensated for the decreased NAD steady state concentration caused by Nam degradation through NNMT.}
+All other enzyme concentrations were set equal.
+Details are given in supplementary table~S2.
+In the absence of NNMT (blue lines), steady state NAD consumption rates (A) are higher despite reduced NAD concentrations (B).
+Increasing the amount of NamPRT in the simulation fourfold (blue dotted lines) partially compensates for the decreased NAD steady state concentration caused by Nam degradation through NNMT.}
   \label{fig:NNMT_NAD_flux}
@@
   \caption{\textbf{Role of NamPRT affinity for Nam.} Using the dynamic model of NAD biosynthesis and consumption we simulated the affect of different Michaelis-Menten constants ($K_M$) of NamPRT for Nam on steady state NAD consumption flux and NAD concentration at different cell division rates.
-All ther parameter were equal to those used for the simulations in Figure 3.
+All the parameters were equal to those used for the simulations in figure~\ref{fig:NNMT_NAD_flux}.
 In the absence of NNMT (A and B), the $K_M$ of NamPRT has little influence on NAD consumption and concentration, but both are changing with cell division rates.
 In the presence of NNMT (C and D), decreasing $K_M$ of NamPRT enables increasing NAD consumption flux and NAD concentration.
-NNMT furthermore makes both, NAD consumption flux and concentration, relatively independent of cell division rates.
+NNMT furthermore makes both, NAD consumption flux and concentration, almost independent of cell division rates.
 Comparing the situation with and without NNMT (E and F) at different NamPRT $K_M$ reveals that at low $K_M$ (dashed lines) and high cell division rates NNMT no longer enables higher NAD consumption rates compared to NamPRT alone (green line and dashed black line).}
@@
   \includegraphics[width=\textwidth]{fig_5_optimal_substr_affinities}
-  \caption{\textbf{The substrate affinities of human NNMT and NamPRT are optimal.} Using the same model as in Figure 3 and 4, we simulated the impact of changes in the $K_{M}$ for both NamPRT and NNMT on NAD consumption rates (A) and NAD concentration (B).
-Both are increasing with decreasing $K_{M}$ of NamPRT but decreasing $K_{M}$ of NNMT.
+  \caption{\textbf{The substrate affinities of human NNMT and NamPRT are optimal.} Using the same model as in figures~\ref{fig:NNMT_NAD_flux} and~\ref{fig:NamPRT_affinity_Nam}, we simulated the impact of changes in the $K_{M}$ for both NamPRT and NNMT on NAD consumption rates (A) and NAD concentration (B).
+Both are increasing with decreasing $K_{M}$ of NamPRT but increasing $K_{M}$ of NNMT.
 The affinities reported for human enzymes (indicated by a black asterisk) appear to be close to optimal, as further improvements would have little or no effect on NAD consumption or concentration.}
@@
 Both show a heterogeneous nuclear cytosolic localisation in HeLa S3 cells.
-\textit{In vitro} measurements using recombinant protein show that the mutant NamPRT has lower activity than the wildtype enzyme (D) and is not activated by ATP (E).
+\textit{In vitro} measurements using recombinant protein show that the mutant NamPRT has a lower activity than the wildtype enzyme (D) and is not activated by ATP (E).
 Bars in D) and E) with different letters indicate significant difference of measured values as estimated using a T test assuming independent samples and significance at $p < 0.05$.}
@@
   \includegraphics[width=\textwidth]{fig_8_evo_events}
-  \caption{\textbf{Schematic representation of co-evolutionary events in the NAD pathway} Based on the phylogenetic analysis presented here and earlier work \cite{Lau2010} we summarized and indicated important events in the evolution of NAD metabolism in Metazoa.
+  \caption{\textbf{Schematic representation of evolutionary events in the NAD pathway} Based on the phylogenetic analysis presented here and earlier work \cite{Lau2010} we summarized and indicated important events in the evolution of NAD metabolism in Metazoa.
 Events in italics are from \cite{Lau2010}.

### 2018-06-26 latex/introduction.tex
Some more minor fixes

@@
 Due to the constant release of Nam through NAD-consuming signalling reactions, the NAD salvage pathway using Nam as precursor is the most important NAD synthesis pathway.
-If Nam would not be constantly recycled into NAD, humans would require a much higher daily vitamin B3 intake than the 16 mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}.
+If Nam would not be constantly recycled into NAD, humans would require a much higher daily vitamin B3 intake than the 16\,mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}.
 In the first step of the salvage pathway, Nam is converted to the mononucleotide NMN by the Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate.

### 2018-06-26 latex/results.tex
Some more minor fixes

@@
 In Deuterostomia the only large class that does only have NamPRT and seems to have lost NNMT again are Sauropsida and here especially birds.
-The reason why a lot of birds do not encode NNMT remains unclear, as the appearance is quite scattered (figure S2\todo[author=Mathias B.]{Has to be made!}).
+The reason why a lot of birds do not encode NNMT remains unclear, as the appearance is quite scattered (fig.~S2).
 The lack of NNMT might be related to the excretion system, as the product of NNMT methyl-Nam is in mammals excreted with the urine.
@@
 Given the complexity of the NAD metabolic network, these questions are difficult to be comprehensively addressed experimentally.
-We therefore built a dynamic model of NAD metabolism using existing kinetic data from the literature (for details see experimental procedures and supplementary table~S2).
+We therefore built a dynamic model of NAD metabolism using existing kinetic data from the literature (for details, see experimental procedures and supplementary table~S2).
 
@@
 
-When we compare NAD consumption and NAD concentration with and without NNMT with two different substrate affinities of NamPRT, we see that at low affinity ($K_{M}$ of 1 $\mu$M, which is in the range of the $K_{M}$ of NADA for Nam), NAD consumption flux is only higher with NNMT at low cell division rates.
+When we compare NAD consumption and NAD concentration with and without NNMT with two different substrate affinities of NamPRT, we see that at low affinity ($K_{M}$ of 1\,$\mu$M, which is in the range of the $K_{M}$ of NADA for Nam), NAD consumption flux is only higher with NNMT at low cell division rates.
 This effect becomes stronger with decreasing affinity (fig.~\ref{fig:NamPRT_affinity_Nam}E and~F).
@@
 The remaining question is now, why NADA is lost in vertebrates, while NADA and NamPRT are co-existing in bacteria.
-To analyse this question we built a two compartment model that has a shared Nam source (details see experimental procedures and supplementary table~S2).
+To analyse this question we built a two compartment model that has a shared Nam source (for details, see experimental procedures and supplementary table~S2).
 One compartment contains NADA while the other contains either NamPRT alone at equal expression levels, or together with NNMT.

### 2018-06-26 latex/discussion.tex
Added supplementary figure 3 and corresponding figure legend

@@
 We here neglected the potential effects of the methyl donor S-adenosyl-methionin (SAM) and its precursor methionine.
-Changing SAM availibility is most likely contributing yet another regulatory level for Nam availability and thus an additional interaction point between gene regulation and metabolism.
+Changing SAM availability is most likely contributing yet another regulatory level for Nam availability and thus an additional interaction point between gene regulation and metabolism.
 It has been shown earlier that NNMT expression influences protein methylation dependent on methionine availability \cite{Ulanovskaya2013}.

### 2018-06-26 latex/figurelegends.tex
Added supplementary figure 3 and corresponding figure legend

@@
   \includegraphics[width=0.9\textwidth]{fig_4_NamPRT_affinity_Nam}
-  \caption{\textbf{Role of NamPRT affinity for Nam.} Using the dynamic model of NAD biosynthesis and consumption we simulated the affect of different Michaelis-Menten constants ($K_M$) of NamPRT for Nam on steady state NAD consumption flux and NAD concentration at different cell division rates.
+  \caption{\textbf{Role of NamPRT affinity for Nam.} Using the dynamic model of NAD biosynthesis and consumption we simulated the affect of different Michaelis-Menten constants ($K_{M}$) of NamPRT for Nam on steady state NAD consumption flux and NAD concentration at different cell division rates.
 All the parameters were equal to those used for the simulations in figure~\ref{fig:NNMT_NAD_flux}.
-In the absence of NNMT (A and B), the $K_M$ of NamPRT has little influence on NAD consumption and concentration, but both are changing with cell division rates.
-In the presence of NNMT (C and D), decreasing $K_M$ of NamPRT enables increasing NAD consumption flux and NAD concentration.
+In the absence of NNMT (A and B), the $K_{M}$ of NamPRT has little influence on NAD consumption and concentration, but both are changing with cell division rates.
+In the presence of NNMT (C and D), decreasing $K_{M}$ of NamPRT enables increasing NAD consumption flux and NAD concentration.
 NNMT furthermore makes both, NAD consumption flux and concentration, almost independent of cell division rates.
-Comparing the situation with and without NNMT (E and F) at different NamPRT $K_M$ reveals that at low $K_M$ (dashed lines) and high cell division rates NNMT no longer enables higher NAD consumption rates compared to NamPRT alone (green line and dashed black line).}
+Comparing the situation with and without NNMT (E and F) at different NamPRT $K_{M}$ reveals that at low $K_{M}$ (dashed lines) and high cell division rates NNMT no longer enables higher NAD consumption rates compared to NamPRT alone (green line and dashed black line).}
   \label{fig:NamPRT_affinity_Nam}
@@
 In the presence of NNMT, however, both NAD consumption (C) and NAD concentration (D) are lower in the NADA compartment.
-This effect is dependent on a low NamPRT $K_M$.}
+This effect is dependent on a low NamPRT $K_{M}$.}
   \label{fig:NNMT_comp_advantage}

### 2018-06-26 latex/supplementaries.tex
First round of spell checking done
Minor changes and library update
Added supplementary figure 3 and corresponding figure legend

@@
 
+\subsection{Supplementary figure S3}
+
+\textbf{Purification of wt NamPRT and $\Delta$42-51 NamPRT.} A) Elution profile of wildtype NamPRT and mutant $\Delta$42-51 on size-exclusion chromatography using a Superdex 200 16/60 column.
+B) Coomassie stained denaturating SDS-PAGE analysis of $\Delta$42-51 NamPRT (lane 1) and wt NamPRT (lane 2).
+3\,$\mu$g of pooled enzyme eluted from SEC loaded onto the gel.
+C) The column was calibrated with apronitin 6.5\,kDa, ovalbumine 42.7\,kDa, coalbumine 75\,kDa and blue dextran 2000\,kDa.
+The partition coefficient (Kav) was determined for each standard (light grey squares) and plotted versus log$_{10}$ molecular weight.
+The Kav was determined for wt NamPRT and $\Delta$42-51 NamPRT and the apparent molecular weight calculated to be 135\,kDa and 110\,kDa, respectively.
+
 \subsection{Supplementary table S1}
 
-\textbf{Overview of kinetic constants used for the construction of the model.}
+\textbf{Query proteins used for Blast searches.}
 
@@
 
-\textbf{Query proteins used for Blast searches.}
+\textbf{Overview of kinetic constants used for the construction of the model.}

### 2018-06-26 latex/latex_to_word.py
Fixed two little bugs

@@ -147,10 +147,8 @@ for line in lines:
 
 	if r'_' in line:
 		line = re.sub(r'_\{([^\}]+)\}', r'<sub>\1</sub>', line)
-		line = re.sub(r'_(.)', r'<sub>\1</sub>', line)
 	if r'^' in line:
 		line = re.sub(r'\^\{([^\}]+)\}', r'<sup>\1</sup>', line)
-		line = re.sub(r'^(.)', r'<sup>\1</sup>', line)
 	if '$' in line:
 		line = re.sub(r'\$([^\$]+)\$', r'<i>\1</i>', line)
 	if r'\%' in line:
@@ -197,9 +195,9 @@ for i in range(len(out)):
 		out[i] = '<p>{}</p>'.format(out[i])
 	if 'LABEL' in out[i]:
 		for label in figure_labels:
-			out[i] = out[i].replace(label, 'fig. ' + figure_labels[label])
+			out[i] = out[i].replace(label, figure_labels[label])
 		for label in table_labels:
-			out[i] = out[i].replace(label, 'tab. ' + table_labels[label])
+			out[i] = out[i].replace(label, table_labels[label])
 
 open('/run/shm/manuscript.html', 'w').write('\n'.join(out))
 

### 2018-06-27 latex/manuscript.tex
Minor changes and library update
Added supplementary figure 3 and corresponding figure legend
Included changes by Ines

@@
 \noindent
+\textit{Alternative title:} Functional interdependencies as driving force for the evolution of NAD-dependent signalling and biosynthesis
+
+\vspace{5mm}
+
+\noindent
 Mathias Bockwoldt\textsuperscript{1}, Dorothée Houry\textsuperscript{2}, Marc Niere\textsuperscript{2}, Toni I.
@@
 
-\section{Summary}
+\section*{Summary}
 
@@
 Why is there an enzyme that removes Nam from recycling, seemingly not having any other physiological role?
-Despite the occurance of this Nam degrading enzyme NNMT,  we observe  a diversification of NAD dependent signalling enzymes in vertebrates.
+Despite the occurrence of this Nam degrading enzyme NNMT, we observe a diversification of NAD dependent signalling enzymes in vertebrates.
 This seems counterintuitive.
@@
 
-\section{Keywords}
+\section*{Keywords}
 
@@
 \input{methods}
+
+\bibliographystyle{plain}
+\bibliography{library}
 
@@
 
-\bibliographystyle{plain}
-\bibliography{library}
-
 \end{document}

### 2018-06-27 latex/methods.tex
Minor changes and library update
Included changes by Ines

@@
 For enzymes not present in mammals, values from yeast were used.
-The full list of kinetic parameters including reference to original literature can be found in Supplementary table 1.
+The full list of kinetic parameters including reference to original literature can be found in supplementary table~S2.
 For NMNAT, the previously developed rate law for substrate competition was used \cite{Schauble2013}.
@@
 All simulations were performed using the steady state calculation and parameter scan options provided by COPASI 4.22 \cite{Hoops2006}.
-The model will be available at the Biomodels database accession number xxx\todo[author=Ines]{Insert number}.
-Related figures were generated using gnuplot\todo[author=Ines]{Version?}.
+The model will be available at the Biomodels database upon publication.
+Related figures were generated using Gnuplot~5.0.
 
@@
 NADA, NamPRT, and NNMT enzymes or enzyme candidates were identified using Blastp with known enzymes against the NCBI non-redundant protein sequence database (nr).
-A list of functionally verified enzymes used as templates is given in supplementary table 2.
+A list of functionally verified enzymes used as templates is given in supplementary table~S1.
 This table also includes the length cut-offs for identified enzymes.

### 2018-06-27 latex/results.tex
Added supplementary figure 3 and corresponding figure legend
Included changes by Ines

@@
 Alternatively, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead \cite{Yoshino2018}.
-Due to the high turnover of NAD observed and the fact that only 1\% of the tryptophan taken up with our diet is converted into NAD\todo{source}, vitamin B3 (Nam and NA) and to a lower extend NR are the major precursors of human NAD biosynthesis.
+Due to the high turnover of NAD observed and the fact that only 1\% of the tryptophan taken up with our diet is converted into NAD\todo{Needs a reference}, vitamin B3 (Nam and NA) and to a lower extend NR are the major precursors of human NAD biosynthesis.
 
@@
 In Deuterostomia the only large class that does only have NamPRT and seems to have lost NNMT again are Sauropsida and here especially birds.
-The reason why a lot of birds do not encode NNMT remains unclear, as the appearance is quite scattered (fig.~S2).
+The reason why a lot of birds do not encode NNMT remains unclear, as the appearance is quite scattered (suppl.
+fig.~S2).
 The lack of NNMT might be related to the excretion system, as the product of NNMT methyl-Nam is in mammals excreted with the urine.
@@
 Given the complexity of the NAD metabolic network, these questions are difficult to be comprehensively addressed experimentally.
-We therefore built a dynamic model of NAD metabolism using existing kinetic data from the literature (for details, see experimental procedures and supplementary table~S2).
+We therefore built a dynamic model of NAD metabolism using existing kinetic data from the literature (for details, see experimental procedures and suppl.
+tab.~S2).
 
@@
 The remaining question is now, why NADA is lost in vertebrates, while NADA and NamPRT are co-existing in bacteria.
-To analyse this question we built a two compartment model that has a shared Nam source (for details, see experimental procedures and supplementary table~S2).
+To analyse this question we built a two compartment model that has a shared Nam source (for details, see experimental procedures and suppl.
+tab.~S2).
 One compartment contains NADA while the other contains either NamPRT alone at equal expression levels, or together with NNMT.

### 2018-06-27 latex/latex_to_word.py

@@ -83,7 +83,7 @@ table_labels = {}
 figure_counter = 1
 table_counter = 1
 
-lines_to_skip = [r'\noindent', r'\vspace', r'\centering']
+lines_to_skip = [r'\noindent', r'\vspace', r'\centering', r'\bibliography']
 
 lines = iter(tex.split('\n'))
 
@@ -94,7 +94,7 @@ for line in lines:
 			active = True
 		continue
 
-	if line.startswith(r'\bibliography'):
+	if line.startswith(r'\end{document}'):
 		break
 
 	has_comment = re_comment.search(line)

### 2018-06-28 latex/discussion.tex
Expressions and typos

@@
 We show that the presence of NNMT enables higher NAD consumption rates in comparison to NamPRT or NADA alone.
-It furthermore appears that NNMT makes both NAD concentration and NAD consumption relatively independent of other NAD-requiring processes such as cell growth.
+It furthermore appears that NNMT makes both NAD concentration and NAD consumption relatively independent of other processes requiring NAD such as cell growth.
 We furthermore suggest that NNMT exerted an evolutionary pressure on NamPRT, driving the development of the high affinity of NamPRT for its substrates found in the human enzyme \cite{Burgos2008}.
@@
 
-In our analysis, we furthermore noted that NNMT only provides an advantage as long as the NamPRT affinity for Nam is sufficiently high, this implies that treatment with competitive inhibitors that affect the apparent $K_{M}$ would have stronger effects on cells that express NNMT compared to cells that do not express NNMT (see fig.~\ref{fig:NamPRT_affinity_Nam}E and~F).
-This could be important for cancer treatments as the expression of NNMT is relatively low in normal tissues except liver, whereas several types of cancer overexpress NNMT.
+In our analysis, we furthermore noted that NNMT only provides an advantage as long as the NamPRT affinity for Nam is sufficiently high.
+This implies that treatment with competitive inhibitors that affect the apparent $K_{M}$ would have stronger effects on cells that express NNMT compared to cells that do not express NNMT (see fig.~\ref{fig:NamPRT_affinity_Nam}E and~F).
+This could be important for cancer treatments as the expression of NNMT is relatively low in normal tissues except liver, whereas several types of cancer overexpress NNMT\todo{Needs a reference}.
 
-We here neglected the potential effects of the methyl donor S-adenosyl-methionin (SAM) and its precursor methionine.
+We here neglected the potential effects of the methyl donor \textit{S}-adenosyl methionine (SAM) and its precursor methionine.
 Changing SAM availability is most likely contributing yet another regulatory level for Nam availability and thus an additional interaction point between gene regulation and metabolism.

### 2018-06-30 latex/results.tex

@@
 The recycling pathway via NA finally requires an amidation step catalysed by NAD synthase, driven by the conversion of ATP to AMP producing pyrophosphate.
-Even though the latter pathway seems to be very inefficient, it is the pathway preferentially used by most bacteria, fungi and plants (see fig.~\ref{fig:phylo_distribution}A), whereas most metazoans recycle Nam using NamPRT.
+Even though the latter pathway seems to be very inefficient, it is the pathway preferentially used by most bacteria, fungi and plants, whereas most metazoans recycle Nam using NamPRT (fig.~\ref{fig:phylo_distribution}A).
 
-Analysing the phylogeny of the NAD recycling enzymes in Metazoa in more detail reveals that not only NamPRT replaces NADA, but in most organisms, especially in Deuterostomia, NamPRT is found together with the Nam methyltransferase NNMT (fig.~\ref{fig:phylo_distribution}B).
+Analysing the phylogeny of the NAD-recycling enzymes in Metazoa in more detail reveals that not only NamPRT replaces NADA, but in most organisms, especially in Deuterostomia, NamPRT is found together with the Nam methyltransferase NNMT (fig.~\ref{fig:phylo_distribution}B).
 NNMT methylates Nam to methyl-Nam that is in mammals excreted with the urine, thus removing Nam from recycling.
@@
 The multiple sequence alignment of selected sequences is shown in figure~\ref{fig:unresolved_loop}A and a more comprehensive multiple sequence alignment containing a larger number of species can be found in supplementary figure~S1.
-We recognised that most Deuterostomia that have only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 42 to 51 of the human enzyme.
+We found that most Deuterostomia that have only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 42 to 51 in the human enzyme.
 This insert overlaps with a predicted weak nuclear localisation signal, that is lost when removing this sequence stretch.

### 2018-07-06 latex/introduction.tex
First round of including Mathias Z.s comments. Not finished, yet.

@@
 The nearly complete recycling of Nam is achieved by an extraordinary high affinity of NamPRT to Nam, the $K_{M}$ being in the low nanomolar range \cite{Burgos2008}.
+This extraordinary affinity appears to be mediated by an ATP-dependent phosphorylation of a histidine residue in the catalytic core \todo{Needs reference}.
 Despite the importance of its salvage, Nam can also be marked for excretion by methylation.
@@
 
-We here present a phylogenetic analysis of the NAD pathways in eukaryotes.
-In contrast to our previous analysis \cite{Gossmann2012FEBS}, which has been limited by the low number of eukaryotic genomes available at the time, we are now able to provide a comprehensive analysis of the eukaryotic pathway evolution.
-The results suggest that there has been a selection for the co-existence of NamPRT and NNMT in Deuterostomia, while the pathway dominant in bacteria is lost.
+We here present a phylogenetic analysis of the NAD pathways.
+The results suggest that there has been a selection for the co-existence of NamPRT and NNMT in Deuterostomia, while the deamidation pathway, which is dominant in bacteria, is lost.
 This transition was accompanied by a marked increase in the number of NAD-consuming signalling enzymes.
-To explain these counterintuitive results we built a mathematical model of the pathway and demonstrated that NNMT has a critical role to maintain high NAD-consuming signalling fluxes by preventing accumulation of inhibitory Nam.
-In addition, our model predicts that NNMT likely exerted an evolutionary pressure on the NamPRT affinity development.
+Mathematical modelling of the pathway revealed an unexpected positive kinetic role of NNMT in the maintenance of high NAD-consuming signalling fluxes by preventing accumulation of inhibitory Nam.
+In addition, the model predicts that NNMT likely exerted an evolutionary pressure on NamPRT to develop a high affinity towards its substrate Nam.
+Indeed, we identified a short sequence insertion in NamPRT, which first occurs in Deuterostomes that appears to modulate the affinity of NamPRT.
 Simulating the resource competition we furthermore show that the presence of high affinity NamPRT together with NNMT makes the NADA-dependent pathway obsolete.
-Based on multiple sequence alignments, we identified a sequence insertion in NamPRT in Deuterostomes.
-We could show experimentally that this insertion increases the affinity of NamPRT, supporting the predictions derived from our mathematical modelling approach.
 

### 2018-07-06 latex/results.tex
First round of including Mathias Z.s comments. Not finished, yet.

@@
 
-\subsection{Phylogenetic analysis of NAD biosynthesis and consumption}
+\subsection{Paradoxical evolutionary correlation between NAD-dependent signaling and precursor metabolism}
 
@@
 
-Looking across all known species, two different pathways exist that recycle Nam.
-The major pathway found in yeast and plants uses a four-step pathway starting with the deamination of Nam to nicotinic acid by NADA.
+Two principal pathways exist that recycle Nam.
+As shown in figure \ref{fig:pathway_overview}, the major pathway found in most prokaryotes, yeast, and plants includes four steps pathway starting with the deamidation of Nam to nicotinic acid by NADA.
 The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates.
 The recycling pathway found in mammals directly converts Nam into the corresponding mononucleotide (NMN) a reaction catalysed by NamPRT and driven by a non-stoichiometric ATP-hydrolysis.
-A similar reaction catalysed by an evolutionary related enzyme NAPRT, converts NA into the NA mononucleotide in the Preiss-Handler pathway.
-NMN and NAMN are converted into dinucleotides by the Nam/NA adenylytransferases (NMNATs).
-The recycling pathway via NA finally requires an amidation step catalysed by NAD synthase, driven by the conversion of ATP to AMP producing pyrophosphate.
-Even though the latter pathway seems to be very inefficient, it is the pathway preferentially used by most bacteria, fungi and plants, whereas most metazoans recycle Nam using NamPRT (fig.~\ref{fig:phylo_distribution}A).
+A similar reaction catalysed by an evolutionary related enzyme, NAPRT, converts NA into the NA mononucleotide (NAMN) in the Preiss-Handler pathway.
+NMN and NAMN are converted into dinucleotides by Nam/NA adenylytransferases (NMNATs).
+The recycling pathway via NA finally requires re-amidation of the dinucleotide, which is catalysed by NAD synthase.
+This reaction includes an enzyme adenylation step that consumes ATP.
+Therefore, even though Nam recycling by NADA is energetically less efficient, this pathway is preferentially used by most bacteria, fungi and plants, whereas most metazoans recycle Nam using NamPRT (fig.~\ref{fig:phylo_distribution}A).
 
-Analysing the phylogeny of the NAD-recycling enzymes in Metazoa in more detail reveals that not only NamPRT replaces NADA, but in most organisms, especially in Deuterostomia, NamPRT is found together with the Nam methyltransferase NNMT (fig.~\ref{fig:phylo_distribution}B).
+Analysing the phylogeny of the NAD-recycling enzymes in Metazoa in more detail revealed that not only NamPRT replaced NADA, but in most organisms, especially in Deuterostomes, NamPRT is found together with the Nam methyltransferase NNMT (fig.~\ref{fig:phylo_distribution}B).
 NNMT methylates Nam to methyl-Nam that is in mammals excreted with the urine, thus removing Nam from recycling.
 NNMT seems to have arisen \textit{de novo} in the common ancestor of Ecdysozoa and Lophotrochozoa.
-We were not be able to find any gene with considerable similarity in fungi or plants, even though in fungi genes named NNMT can be found in databases.
-These genes do, however, show very limited sequence similarities to the NNMT of nematodes or deuterostomes.
-The yeast protein has later been shown to be a lysin-protein-methyltransferase \cite{Wlodarski2011}.
-Nematodes are the only organisms that encode NNMT together with NADA without NamPRT being present.
-In Deuterostomia the only large class that does only have NamPRT and seems to have lost NNMT again are Sauropsida and here especially birds.
-The reason why a lot of birds do not encode NNMT remains unclear, as the appearance is quite scattered (suppl.
+We were unable to find any \textit{NNMT} gene with considerable similarity in fungi or plants.
+Even though fungal genes named \textit{NNMT} can be found in databases, these genes show very limited sequence similarities to the NNMT of nematodes or deuterostomes.
+The yeast NNMT protein has meanwhile been shown to be a lysin-protein-methyltransferase \cite{Wlodarski2011}.
+Nematodes are the only organisms that can have any combination of NADA, NamPRT, and NNMT.
+In Deuterostomia, the only large clade that does only have NamPRT and seems to have lost NNMT again are Sauropsida and among them especially birds.
+The reason why a lot of birds do not seem to encode NNMT remains unclear, as the appearance is quite scattered (suppl.
 fig.~S2).
 The lack of NNMT might be related to the excretion system, as the product of NNMT methyl-Nam is in mammals excreted with the urine.
-There are some species where we could not find NamPRT or NADA but NNMT, we assume that this is due to incomplete genomes in the database.
+There are some species for which we could not find NamPRT or NADA, while NNMT was present.
+We assume that this is due to incomplete genomes in the database.
 
-In addition to the phylogenetic distribution of the two Nam salvage enzymes NADA and NamPRT, we looked at phylogenetic diversity of enzymes catalysing NAD-dependent signalling reactions.
-To do so we used the previously established classification into ten different enzyme families \cite{Gossmann2012FEBS}.
-The detailed list of templates used for the phylogenetic analysis can be found in supplementary table~S1.
-The numbers in figure~\ref{fig:phylo_distribution}B denote the average number of NAD-dependent signalling enzyme families found in each taxa.
-With the exception of Cnidaria and Lophotrochozoa, we find average three to four families in Protostomia, whereas most Deuterostomia have on average more than eight families with an increasing diversification of enzymes within some of these families \cite{Gossmann2014DNAR}.
+In addition to the phylogenetic distribution of the two Nam salvage enzymes NADA and NamPRT, we analysed the phylogenetic diversity of enzymes catalysing NAD-dependent signalling reactions.
+To do so, we used the previously established classification into ten different families of NAD-consuming signalling enzymes \cite{Gossmann2012FEBS}.
+The detailed list of templates used for the phylogenetic analyses can be found in supplementary table~S1.
+The numbers shown in figure~\ref{fig:phylo_distribution}B denote the average number of NAD-dependent signalling enzyme families found in each taxon.
+With the exception of Cnidaria and Lophotrochozoa, we find an average of three to four families in Protostomia, whereas most Deuterostomia have on average more than eight families with an increasing diversification of enzymes within some of these families \cite{Gossmann2014DNAR}.
 
-Taken together, we found that the presence of NamPRT and NNMT coincides with an increased diversification of NAD-dependent signalling, whilst NADA is lost in vertebrates.
-This seems counterintuitive, as one would expect that a decrease in precursor concentration caused by the precursor removal through NNMT, should cause a decrease of NAD availability and consequently less active NAD dependent signalling.
+Taken together, we found that NADA is lost in vertebrates, but strongly preserved in most other organisms, despite its higher energetic requirement.
+Moreover, the selection for having both NamPRT and NNMT coincides with an increased diversification of NAD-dependent signalling.
+This observation seems counterintuitive.
+One would expect that increased NAD-dependent signalling should be accompanied by an increase of substrate availability for NAD biosynthesis.
+However, NNMT irreversibly removes Nam from salvage back to NAD.
 
-\subsection{Dynamics of NAD biosynthesis and consumption}
+\subsection{Functional properties of NamPRT and NNMT have evolved to maximize NAD-dependent signaling}
 

### 2018-07-07 latex/acknowledgements.tex
Completely separated content and master file

--- 2018-06-26 latex/manuscript.tex
@@
+%%% TeX-master: "manuscript"
 % !TeX spellcheck=en_GB
-
-\RequirePackage[l2tabu, orthodox]{nag}
-\documentclass[paper=a4, 12pt]{scrartcl}
-
-\usepackage{graphicx}
-\usepackage[font=footnotesize]{caption}
-\usepackage[utf8]{inputenc}
-\usepackage[colorlinks=true,
-	linkcolor=black,
-	citecolor=black,
-	filecolor=black,
-	urlcolor=black]{hyperref}
-\usepackage{textcomp}
-\usepackage[left=1cm,right=5cm,top=2cm,bottom=2cm,nohead,nofoot]{geometry}
-\usepackage[textsize=scriptsize]{todonotes}
-
-\setlength{\marginparwidth}{4cm}
-
-\frenchspacing
-\graphicspath{ {../} }
-
-\begin{document}
-
-\noindent
-{\huge\sffamily\bfseries NamPRT and NNMT – key drivers of NAD-dependent signalling \par}
-
-\vspace{15mm}
-
-\noindent
-Mathias Bockwoldt\textsuperscript{1}, Dorothée Houry\textsuperscript{2}, Marc Niere\textsuperscript{2}, Toni I.
-Gossmann\textsuperscript{3}, Mathias Ziegler\textsuperscript{2}, and Ines Heiland\textsuperscript{1,§}
-
-\vspace{1cm}
-
-\noindent
-\textsuperscript{1}Department of Arctic and Marine Biology, UiT The Arctic University of Norway, Biologibygget, Framstredet 39, 9017 Tromsø, Norway
-
-\noindent
-\textsuperscript{2}Department of Molecular Biology, University of Bergen, Thormøhlensgt. 55, 5020 Bergen, Norway
-
-\noindent
-\textsuperscript{3}Department of Animal and Plant Sciences, Western Bank, University of Sheffield, Sheffield, S10 2TN, United Kingdom
-
-\noindent
-§ Corresponding author: ines.heiland@uit.no
-
-\section{Summary}
-
-NAD is best known as cofactor in redox reactions, but it is also substrate of NAD-dependent signalling reactions that consume NAD and release nicotinamide (Nam).
-In eukaryotes, two different Nam salvage pathways exist.
-Phylogenetic analysis of the pathway show that in lower organisms the initial deamidation of Nam is prevalent, whilst the direct conversion of Nam to the mononucleotide by Nam phosphoribosyltransferase (NamPRT) dominates in animals and eventually remains as the single Nam recycling route in vertebrates.
-Strikingly, loss of the deamidation pathway in early vertebrates is preceded by the occurrence of a new enzyme that marks Nam for excretion by methylation – nicotinamide N-methyltransferase (NNMT).
-The physiological relevance of NNMT is still enigmatic.
-Why is there an enzyme that removes Nam from recycling, seemingly not having any other physiological role?
-Despite the occurance of this Nam degrading enzyme NNMT,  we observe  a diversification of NAD dependent signalling enzymes in vertebrates.
-This seems counterintuitive.
-
-To better understand these evolutionary changes and the role of NNMT for NAD metabolism, we built a mathematical model of the pathway using published enzyme kinetics data.
-Our results indicate that NNMT is required to enable high NAD consumption fluxes necessitated by the increasing diversification of NAD-dependent signalling pathways.
-This kinetic regulation requires a high substrate affinity of the key enzyme for Nam salvage, NamPRT.
-Indeed, the affinity of NamPRT to Nam has previously been measured to be in the nanomolar range.
-Based on our mathematical modelling approach we furthermore suggest that NNMT exerted an evolutionary pressure on NamPRT, enforcing the development of its unusually high substrate affinity.
-Using multiple sequence alignments, we identified a sequence insertion, first occurring in vertebrates, that parallels an experimentally verified increase in the substrate affinity of the enzyme.
-Additional simulations show that the deamidation pathway became obsolete owing to the high substrate affinity of NamPRT.
-Collectively, our results illustrate a close evolutionary relationship between NAD biosynthesis and the diversification NAD-dependent signalling pathways, potentially driven by the concomitant occurrence of a regulator of Nam salvage, NNMT.
-
-\section{Keywords}
-
-\input{introduction}
-
-\input{results}
-
-\input{discussion}
-
-\input{methods}
-
-\input{figurelegends}
-
-\input{supplementaries}
 
@@
 We furthermore thank for the computation time provided through UNINETT Sigma2 – the National Infrastructure for High Performance Computing and Data Storage in Norway.
-
-\bibliographystyle{plain}
-\bibliography{library}
-
-\end{document}

### 2018-07-07 latex/header.tex
Completely separated content and master file

--- 2018-06-27 latex/manuscript.tex
@@
+%%% TeX-master: "manuscript"
 % !TeX spellcheck=en_GB
-
-\RequirePackage[l2tabu, orthodox]{nag}
-\documentclass[paper=a4, 12pt]{scrartcl}
-
-\usepackage{graphicx}
-\usepackage[font=footnotesize]{caption}
-\usepackage[utf8]{inputenc}
-\usepackage[colorlinks=true,
-	linkcolor=black,
-	citecolor=black,
-	filecolor=black,
-	urlcolor=black]{hyperref}
-\usepackage{textcomp}
-\usepackage[left=1cm,right=5cm,top=2cm,bottom=2cm,nohead,nofoot]{geometry}
-\usepackage[textsize=scriptsize]{todonotes}
-
-\setlength{\marginparwidth}{4cm}
-
-\frenchspacing
-\graphicspath{ {../} }
-
-\begin{document}
 
@@
 \section*{Keywords}
-
-\input{introduction}
-
-\input{results}
-
-\input{discussion}
-
-\input{methods}
-
-\bibliographystyle{plain}
-\bibliography{library}
-
-\input{figurelegends}
-
-\input{supplementaries}
-
-\section*{Acknowledgements}
-
-We thank the Norwegian Research Council for funding (grant no.
-250395/F20).
-We furthermore thank for the computation time provided through UNINETT Sigma2 – the National Infrastructure for High Performance Computing and Data Storage in Norway.
-
-\end{document}

### 2018-07-07 latex/manuscript.tex
Completely separated content and master file

--- 2018-06-26 latex/manuscript.tex
@@
 
-\noindent
-{\huge\sffamily\bfseries NamPRT and NNMT – key drivers of NAD-dependent signalling \par}
-
-\vspace{15mm}
-
-\noindent
-Mathias Bockwoldt\textsuperscript{1}, Dorothée Houry\textsuperscript{2}, Marc Niere\textsuperscript{2}, Toni I.
-Gossmann\textsuperscript{3}, Mathias Ziegler\textsuperscript{2}, and Ines Heiland\textsuperscript{1,§}
-
-\vspace{1cm}
-
-\noindent
-\textsuperscript{1}Department of Arctic and Marine Biology, UiT The Arctic University of Norway, Biologibygget, Framstredet 39, 9017 Tromsø, Norway
-
-\noindent
-\textsuperscript{2}Department of Molecular Biology, University of Bergen, Thormøhlensgt. 55, 5020 Bergen, Norway
-
-\noindent
-\textsuperscript{3}Department of Animal and Plant Sciences, Western Bank, University of Sheffield, Sheffield, S10 2TN, United Kingdom
-
-\noindent
-§ Corresponding author: ines.heiland@uit.no
-
-\section{Summary}
-
-NAD is best known as cofactor in redox reactions, but it is also substrate of NAD-dependent signalling reactions that consume NAD and release nicotinamide (Nam).
-In eukaryotes, two different Nam salvage pathways exist.
-While in lower organisms the initial deamidation of Nam is prevalent, the direct conversion of Nam to the mononucleotide by Nam phosphoribosyltransferase (NamPRT) dominates in animals and eventually remains as the single Nam recycling route in vertebrates.
-
-Strikingly, loss of the deamidation pathway in early vertebrates is preceded by the occurrence of a new enzyme that marks Nam for excretion by methylation – nicotinamide N-methyltransferase (NNMT).
-The physiological relevance of NNMT is still enigmatic.
-Why is there an enzyme that removes Nam from recycling, seemingly not having any other physiological role?
-And why is the occurrence of NNMT accompanied by a diversification of NAD-consuming enzymes?
-
-We have used mathematical modelling approaches to resolve these counterintuitive observations.
-Our results indicate that NNMT is required to enable high NAD consumption fluxes necessitated by the increasing diversification of NAD-dependent signalling pathways.
-This kinetic regulation requires a high substrate affinity of the key enzyme for Nam salvage, NamPRT.
-Indeed, the affinity of NamPRT to Nam has previously been measured to be in the nanomolar range.
-Mathematical modelling supports the hypothesis that NNMT exerted an evolutionary pressure on NamPRT enforcing the development of its unusually high substrate affinity.
-Using multiple sequence alignments, we identified a sequence insertion, first occurring in vertebrates, that parallels an experimentally verified increase in the substrate affinity of the enzyme.
-Further simulations show that the deamidation pathway became obsolete owing to the high substrate affinity of NamPRT.
-Collectively, our results illustrate a close evolutionary relationship between NAD biosynthesis and the diversification NAD-dependent signalling pathways, potentially driven by the concomitant occurrence of a regulator of Nam salvage, NNMT.
-
-\section{Keywords}
+\input{header}
 
@@
 
+\bibliographystyle{plain}
+\bibliography{library}
+
+\input{acknowledgements}
+
 \input{figurelegends}
@@
 
-\section{Acknowledgements}
-
-\bibliographystyle{plain}
-\bibliography{library}
-
 \end{document}

### 2018-07-07 latex/introduction.tex
Rewrote introduction and results to improve language and precision

@@
 Imbalances in NAD homeostasis have been linked to various diseases, in particular related to ageing, such as diabetes, neurodegenerative disorders, and cancer \cite{Chiarugi2012,Verdin2015}.
-Several recent studies have demonstrated impressive health benefits of dietary supplementation with intermediates of NAD biosynthesis including Nam mononucleotide (NMN) and Nam riboside, (NR) \cite{Yoshino2018}.
+Several recent studies have demonstrated impressive health benefits of dietary supplementation with intermediates of NAD biosynthesis including Nam mononucleotide (NMN) and Nam riboside (NR) \cite{Yoshino2018}.
 Apparently, the exploitation of NAD biosynthetic routes, in addition to the use of nicotinamide as precursor (fig.~\ref{fig:pathway_overview}), results in increased NAD concentrations that stimulate NAD-dependent signalling processes, in particular, protein deacetylation by sirtuins \cite{North2004}.
@@
 The presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012FEBS} is among the most enigmatic and counterintuitive features of NAD metabolism.
-Why is there one enzyme (NamPRT) seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, while at the same time there is another enzyme (NNMT) that seems to have no metabolic function other than to remove Nam from NAD metabolism, This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deamidate Nam to nicotinic acid (NA) using the nicotinamide deamidase (NADA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (fig.~\ref{fig:pathway_overview}).
+While NamPRT is seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, NNMT seems to have no metabolic function other than to remove Nam from NAD metabolism.
+This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deamidate Nam to nicotinic acid (NA) using the nicotinamide deamidase (NADA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (fig.~\ref{fig:pathway_overview}).
 
@@
 In addition, the model predicts that NNMT likely exerted an evolutionary pressure on NamPRT to develop a high affinity towards its substrate Nam.
-Indeed, we identified a short sequence insertion in NamPRT, which first occurs in Deuterostomes that appears to modulate the affinity of NamPRT.
+Indeed, we identified a short sequence insertion in NamPRT, which first occurs in Deuterostomes and that appears to modulate the affinity of NamPRT.
 Simulating the resource competition we furthermore show that the presence of high affinity NamPRT together with NNMT makes the NADA-dependent pathway obsolete.

### 2018-07-07 latex/results.tex
Rewrote introduction and results to improve language and precision

@@
 
-NAD can be synthesised using several routes from three main precursors: tryptophan, nicotinamide (Nam) and nicotinic acid (NA) (Overview see fig.~\ref{fig:pathway_overview}).
-Nam and NA are together known as vitamin B3 or niacin.
-Alternatively, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead \cite{Yoshino2018}.
-Due to the high turnover of NAD observed and the fact that only 1\% of the tryptophan taken up with our diet is converted into NAD\todo{Needs a reference}, vitamin B3 (Nam and NA) and to a lower extend NR are the major precursors of human NAD biosynthesis.
-
 Two principal pathways exist that recycle Nam.
-As shown in figure \ref{fig:pathway_overview}, the major pathway found in most prokaryotes, yeast, and plants includes four steps pathway starting with the deamidation of Nam to nicotinic acid by NADA.
+As shown in figure \ref{fig:pathway_overview}, the major pathway found in most prokaryotes, yeast, and plants includes four steps starting with the deamidation of Nam to nicotinic acid by NADA.
 The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates.
-The recycling pathway found in mammals directly converts Nam into the corresponding mononucleotide (NMN) a reaction catalysed by NamPRT and driven by a non-stoichiometric ATP-hydrolysis.
-A similar reaction catalysed by an evolutionary related enzyme, NAPRT, converts NA into the NA mononucleotide (NAMN) in the Preiss-Handler pathway.
+The recycling pathway found in mammals directly converts Nam into Nam mononucleotide (NMN) a reaction catalysed by NamPRT and driven by a non-stoichiometric ATP-hydrolysis.
+A similar reaction catalysed by an evolutionary related enzyme, NAPRT, converts NA into the corresponding mononucleotide (NAMN) in the Preiss-Handler pathway.
 NMN and NAMN are converted into dinucleotides by Nam/NA adenylytransferases (NMNATs).
@@
 This reaction includes an enzyme adenylation step that consumes ATP.
-Therefore, even though Nam recycling by NADA is energetically less efficient, this pathway is preferentially used by most bacteria, fungi and plants, whereas most metazoans recycle Nam using NamPRT (fig.~\ref{fig:phylo_distribution}A).
+Even though Nam recycling by NADA is energetically less efficient, this pathway is preferentially used by most bacteria, fungi and plants, whereas most metazoans recycle Nam using NamPRT (fig.~\ref{fig:phylo_distribution}A).
 
-Analysing the phylogeny of the NAD-recycling enzymes in Metazoa in more detail revealed that not only NamPRT replaced NADA, but in most organisms, especially in Deuterostomes, NamPRT is found together with the Nam methyltransferase NNMT (fig.~\ref{fig:phylo_distribution}B).
+Analysing the phylogeny of the NAD-recycling enzymes in Metazoa in more detail revealed that not only NamPRT replaced NADA, but in most organisms, especially in Deuterostomes, NamPRT is found together with NNMT (fig.~\ref{fig:phylo_distribution}B).
 NNMT methylates Nam to methyl-Nam that is in mammals excreted with the urine, thus removing Nam from recycling.
-NNMT seems to have arisen \textit{de novo} in the common ancestor of Ecdysozoa and Lophotrochozoa.
-We were unable to find any \textit{NNMT} gene with considerable similarity in fungi or plants.
-Even though fungal genes named \textit{NNMT} can be found in databases, these genes show very limited sequence similarities to the NNMT of nematodes or deuterostomes.
+NNMT seems to have arisen \textit{de~novo} in the common ancestor of Ecdysozoa and Lophotrochozoa.
+We were unable to find any \textit{NNMT} gene with considerable\todo{define considerable} similarity in fungi or plants.
+Even though fungal genes named \textit{NNMT} can be found in databases, these genes show very limited sequence similarities to the \textit{NNMT} of nematodes or deuterostomes.
 The yeast NNMT protein has meanwhile been shown to be a lysin-protein-methyltransferase \cite{Wlodarski2011}.
-Nematodes are the only organisms that can have any combination of NADA, NamPRT, and NNMT.
-In Deuterostomia, the only large clade that does only have NamPRT and seems to have lost NNMT again are Sauropsida and among them especially birds.
-The reason why a lot of birds do not seem to encode NNMT remains unclear, as the appearance is quite scattered (suppl.
+
+Nematodes are the only organisms, where we observed any combination of NADA, NamPRT, and NNMT.
+In Deuterostomes, the only large clade that possesses only NamPRT and seems to have lost NNMT again are Sauropsida and among them especially birds.
+The reason why a lot of\todo{be more precise!} bird genomes do not seem to encode NNMT remains unclear, as the distribution is quite scattered (suppl.
 fig.~S2).
-The lack of NNMT might be related to the excretion system, as the product of NNMT methyl-Nam is in mammals excreted with the urine.
-There are some species for which we could not find NamPRT or NADA, while NNMT was present.
-We assume that this is due to incomplete genomes in the database.
+The lack of NNMT might be related to the excretion system, as the product of NNMT, methyl-Nam, is in mammals excreted with the urine.
+There are some Metazoan species for which we could not find NamPRT or NADA, while NNMT was present.
+We assume that this is due to incomplete genomes in the database, as the distribution of such species is scarce and widely scattered.
 
@@
 The detailed list of templates used for the phylogenetic analyses can be found in supplementary table~S1.
-The numbers shown in figure~\ref{fig:phylo_distribution}B denote the average number of NAD-dependent signalling enzyme families found in each taxon.
-With the exception of Cnidaria and Lophotrochozoa, we find an average of three to four families in Protostomia, whereas most Deuterostomia have on average more than eight families with an increasing diversification of enzymes within some of these families \cite{Gossmann2014DNAR}.
+The numbers shown in figure~\ref{fig:phylo_distribution}B denote the average number of NAD-dependent signalling enzyme families found in each clade.
+With the exception of Cnidaria and Lophotrochozoa, we find an average of three to four families in Protostomes, whereas most Deuterostome species have, on average, more than eight families with an increasing diversification of enzymes within some of these families \cite{Gossmann2014DNAR}.
 
-Taken together, we found that NADA is lost in vertebrates, but strongly preserved in most other organisms, despite its higher energetic requirement.
+Taken together, we found that NADA is lost in vertebrates, but strongly preserved in most other organisms, despite the higher energetic requirement of that pathway.
 Moreover, the selection for having both NamPRT and NNMT coincides with an increased diversification of NAD-dependent signalling.
@@
 
-So why does the diversity of NAD-dependent signalling increase?
-And why does NADA disappear in Deuterostomia although it is the predominant pathway in bacteria, plants and fungi?
-Given the complexity of the NAD metabolic network, these questions are difficult to be comprehensively addressed experimentally.
-We therefore built a dynamic model of NAD metabolism using existing kinetic data from the literature (for details, see experimental procedures and suppl.
+To resolve this apparent contradiction, we wished to scrutinize the NAD metabolic network.
+Given the complexity of this network, we turned to modelling approaches and built a dynamic model of NAD metabolism based on previously reported kinetic data (for details, see experimental procedures and suppl.
 tab.~S2).
 
-To be able to compare metabolic features of evolutionary quite different systems in our simulations, and as we have limited information about expression levels of enzymes or changes of kinetic constants during evolution, we initially used the kinetic constants found for yeast or human enzymes for all systems analysed and used equal amounts of enzymes for all reactions.
+To be able to compare metabolic features of evolutionary quite different systems in our simulations and as we have limited information about expression levels of enzymes or changes of kinetic constants during evolution, we initially used the kinetic constants found for human or yeast enzymes for all systems analysed and used equal amounts of enzymes for all reactions\todo{Shorten or split sentence and correct "equal amounts"}.
 Wherever possible, we did not only include substrate affinities but also known product inhibitions or inhibition by downstream metabolites.
@@
 
-First, we addressed the question of the predominant co-existence of NamPRT and NNMT in vertebrates that coincides with an increase in the number of NAD-consuming enzyme families.
-We thus used our dynamic model to simulate steady state concentrations and NAD consumption rates simulating NAD biosynthesis via NamPRT in the presence or absence of NNMT.
-We see that the presence of NNMT enables higher NAD consumption rates (fig.~\ref{fig:NNMT_NAD_flux}A) while, as expected, reducing the steady state concentration of NAD (fig.~\ref{fig:NNMT_NAD_flux}B).
+First, we addressed the unexpected correlation between the selection for co-occurrence of NamPRT and NNMT and an increase in the number of NAD-consuming enzyme families.
+We calculated steady state NAD concentrations and NAD consumption rates by simulating NAD biosynthesis proceeding via NamPRT in the presence or absence of NNMT.
+As shown in figure~\ref{fig:NNMT_NAD_flux}, the presence of NNMT enables higher NAD consumption rates (fig.~\ref{fig:NNMT_NAD_flux}A) while reducing the steady state concentration of NAD (fig.~\ref{fig:NNMT_NAD_flux}B).
 The decline in NAD concentration can be compensated by a higher expression of NamPRT, further increasing NAD consumption flux (dashed lines in fig.~\ref{fig:NNMT_NAD_flux}A and~B).
 
-These findings can be explained when looking in more detail at the kinetic parameters of NamPRT and NAD-consuming enzymes such as Sirtuin~1.
-The increase of NAD consumption flux is caused by the fact that most NAD-consuming enzymes are inhibited by their product Nam, explaining why the presence of NNMT enables higher NAD consumption fluxes.
-At the same time, the high substrate affinity of NamPRT maintains a sufficiently high NAD concentration.
+These results can be explained when looking in more detail at the kinetic parameters of NamPRT and NAD-consuming enzymes such as Sirtuin~1.
+Most NAD-consuming enzymes are inhibited by their product Nam, explaining why the presence of NNMT enables higher NAD consumption fluxes.
+At the same time, the high substrate affinity of NamPRT maintains a sufficiently high NAD concentration, although the concentration is lower than in the system without NNMT.
 
-As kinetic parameters of NamPRT are only available for the human enzyme \cite{Burgos2008}, we went on to theoretically analyse the potential effect of NamPRT affinity ($K_{M}$) on NAD steady state concentration and NAD consumption flux.
-In the absence of NNMT (fig.~\ref{fig:NamPRT_affinity_Nam}A and~B) a change in $K_{M}$ has very little effect on steady state NAD concentration and NAD consumption flux.
+As measured kinetic parameters of NamPRT are only available for the human enzyme \cite{Burgos2008}, we analysed the potential effect of NamPRT affinity ($K_{M}$) on NAD steady state concentration and NAD consumption flux.
+In the absence of NNMT, a change in $K_{M}$ has very little effect on steady state NAD concentration and NAD consumption flux (fig.~\ref{fig:NamPRT_affinity_Nam}A and~B).
 In the presence of NNMT, however, NAD consumption flux and NAD concentration increases with decreasing $K_{M}$ values (fig.~\ref{fig:NamPRT_affinity_Nam}C and~D).
 
-It is interesting to note that without NNMT, NAD concentration and consumption flux are both considerably affected by cell division rates, at least if the enzyme expression is kept constant.
-This is, of course, an artificial scenario, as one would assume organisms to regulate enzyme expression to achieve similar levels of metabolite concentrations instead.
-However, there seems to be an apparent trade off between maintainable NAD concentration and consumption flux in the absence of NNMT.
-In contrast, in the presence of NNMT, NAD consumption rates increase as do NAD concentrations and both are relatively independent of cell division rates.
+Remarkably, NAD concentration and consumption flux are both considerably affected by cell division rates in a system without NNMT, at least if the enzyme expression is kept constant at different cell division rates.
+Of course, this is an artificial scenario, as one would assume organisms to regulate enzyme expression to achieve similar levels of metabolite concentrations instead.
+However, there seems to be a tradeoff between maintainable NAD concentration and consumption flux in the absence of NNMT.
+In contrast, in the presence of NNMT, NAD consumption rates and concentrations are almost independent of cell division rates.
 
-When we compare NAD consumption and NAD concentration with and without NNMT with two different substrate affinities of NamPRT, we see that at low affinity ($K_{M}$ of 1\,$\mu$M, which is in the range of the $K_{M}$ of NADA for Nam), NAD consumption flux is only higher with NNMT at low cell division rates.
-This effect becomes stronger with decreasing affinity (fig.~\ref{fig:NamPRT_affinity_Nam}E and~F).
+Figure~\ref{fig:NamPRT_affinity_Nam}E and~F shows a direct comparison of different affinities of NamPRT for Nam with and without NNMT.
+At an affinity of $K_{M}$ = 1\,$\mu$M, which is in the range of the $K_{M}$ of NADA for Nam, NAD consumption flux is higher with NNMT than without, except for low cell division rates (fig.~\ref{fig:NamPRT_affinity_Nam}E).
+At a lower affinity of 5\,nM, increase in consumption rate even more pronounced.
+The NAD concentration is lower with NNMT than without for both tested $K_{M}$ values (fig.~\ref{fig:NamPRT_affinity_Nam}F).
 
-The pathway dynamics are not solely dependent on one enzyme.
-Thus, what is the impact of the substrate affinity of NNMT that is competing with NamPRT for the same substrate?
-In figure~\ref{fig:optimal_substr_affinities} we see that the substrate affinity values found in the human enzymes (indicated by black asterisks) are actually optimal with respect to both achievable steady state NAD concentration and consumption fluxes.
-Thus, a further increase of the affinity of NamPRT for Nam would not provide any advantage.
+As pathway dynamics are not solely dependent on one enzyme, we wanted to estimate the impact of the substrate affinity of NNMT that is competing with NamPRT for the same substrate.
+In figure~\ref{fig:optimal_substr_affinities} is shown that the substrate affinity values found in the human enzymes (indicated by black asterisks) are actually optimal with respect to both achievable steady state NAD concentration and consumption fluxes.
+Thus, a further increase of the affinity of NamPRT for Nam or a change in the affinity of NNMT for Nam would not provide any advantage.
 
@@
 
-As our simulations suggest that NNMT might have exerted an evolutionary pressure on the development of NamPRT, we created a multiple sequence alignment to see if we can find sequence variations in the protein sequence of NamPRT that indicate evolutionary changes of NamPRT upon the appearance of NNMT.
+As our simulations suggested that NNMT might have exerted an evolutionary pressure on the development of NamPRT, we created a multiple sequence alignment to see if we can find sequence variations in NamPRT upon the appearance of NNMT.
 The multiple sequence alignment of selected sequences is shown in figure~\ref{fig:unresolved_loop}A and a more comprehensive multiple sequence alignment containing a larger number of species can be found in supplementary figure~S1.
-We found that most Deuterostomia that have only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 42 to 51 in the human enzyme.
-This insert overlaps with a predicted weak nuclear localisation signal, that is lost when removing this sequence stretch.
-The inserted sequence corresponds to a structurally unresolved loop in all available crystal structure of human NamPRT (e.g.
-\cite{Wang2006} structure visualisation fig.~\ref{fig:unresolved_loop}B).
+We found that most Deuterostomes that possess only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 42 to 51 in the human enzyme.
+This insert overlaps with a predicted weak nuclear localisation signal (NLS), that is lost when the insert is removed.
+The sequence corresponds to a loop at the protein surface that is unresolved in all available crystal structures of human NamPRT (e.g.
+structure visualisation in fig.~\ref{fig:unresolved_loop}B from \cite{Wang2006}).
 The loop is connected to one of the $\beta$-sheets involved in substrate binding.
 
-From these observations, we derived two possible hypotheses regarding the role of the loop for NamPRT function.
-Firstly, we analysed whether the deletion of the amino acids 42 to 51 has an effect on the localisation of the human enzyme.
-We thus performed immunofluorescence imaging with FLAG-tagged proteins, but both wildtype and mutant protein showed a mixed cytosolic nuclear localisation (fig.~\ref{fig:unresolved_loop}C).
-We therefore conclude that the partial nuclear localisation of NamPRT is not affected by the sequence deletion.
+From these observations, we derived two possible hypotheses regarding the role of the loop in NamPRT function.
+The first hypothesis was that the presence of the loop changes the subcellular localisation of NamPRT, since the loop is part of an NLS.
+To test this hypothesis, we recombinatly expressed wildtype NamPRT and a mutated version lacking amino acids 42 to 51.
+Both, wildtype and mutant, were modified to carry a FLAG-tag for easier visualisation.
+Immunofluorescence imaging showed a mixed cytosolic nuclear localisation for the wildtype and the mutant NamPRT (fig.~\ref{fig:unresolved_loop}C).
+The presence of the loop does not seem to have any effect on the subcellular localisation.
 
-The second hypothesis was that the sequence insertion influences NamPRT affinity, which is what our mathematical simulations would predict.
-We thus recombinantly expressed both wildtype and mutant NamPRT in \textit{E.~coli}.
-Upon purification, we measured the activity using NMR detection of NMN produced in the presence and absence of ATP.
-It appears that the enzymatic activity of the mutant enzyme is much lower than that of the wildtype enzyme (fig.~\ref{fig:unresolved_loop}D) and that an increase of the substrate concentration has a much stronger effect on the activity of the mutant compared to the wildtype enzyme, indicating a lower affinity.
-We furthermore showed that in contrast to NamPRT wildtype, the mutant does not exhibit an increased activity in the presence of ATP (fig.~\ref{fig:unresolved_loop}E).
+The second hypothesis was that the sequence insertion influences the affinity of NamPRT, which would be in line with the predictions from our simulations.
+We recombinantly expressed both wildtype and mutant NamPRT in \textit{E.~coli}.
+After purification (suppl.
+fig.~S3), we measured the activity using NMR detection of NMN in the presence and absence of ATP.
+The enzymatic activity of the mutant enzyme lacking the loop was much lower than the activity of the wildtype enzyme (fig.~\ref{fig:unresolved_loop}D).
+An increase of the substrate concentration had a much stronger effect on the activity of the mutant compared to the wildtype enzyme, indicating a lower affinity.
+In contrast to wildtype NamPRT, the mutant does not exhibit an increased activity in the presence of ATP (fig.~\ref{fig:unresolved_loop}E).
 
-\subsection{Loss of NADA in vertebrates}
+\subsection{NamPRT and NNMT make NADA obsolete in vertebrates}
 
-The remaining question is now, why NADA is lost in vertebrates, while NADA and NamPRT are co-existing in bacteria.
-To analyse this question we built a two compartment model that has a shared Nam source (for details, see experimental procedures and suppl.
+A remaining puzzle was, why NADA is lost in vertebrates, while NADA and NamPRT both occur in bacteria.
+We tried to approach the puzzle with a competitive model with two compartments that share a Nam source.
+One compartment contains NADA while the other contains either NamPRT alone, or together with NNMT (for details, see experimental procedures and suppl.
 tab.~S2).
-One compartment contains NADA while the other contains either NamPRT alone at equal expression levels, or together with NNMT.
-Without NNMT the compartment containing NADA has slightly lower NAD consumption rates (fig.~\ref{fig:NNMT_comp_advantage}A), but is able to maintain much higher steady state NAD concentrations at low cell division rates (fig.~\ref{fig:NNMT_comp_advantage}B).
+Without NNMT, the compartment containing NADA shows slightly lower NAD consumption rates (fig.~\ref{fig:NNMT_comp_advantage}A), but is able to maintain much higher NAD concentrations at low cell division rates (fig.~\ref{fig:NNMT_comp_advantage}B).
 At high cell division rates, steady state concentrations in both compartments are similar.
-The latter might explain why in bacteria that have relatively high growth rates both systems co-exist.
+These comparable concentrations might explain why in bacteria that often have relatively high growth rates, both systems co-exist.
 
-In the presence of NNMT, the NamPRT compartment out-competes the NADA-containing compartment, both with respect to NAD consumption rates and steady state NAD concentrations (fig.~\ref{fig:NNMT_comp_advantage}C and~D).
-This effect is dependent on a high affinity of NamPRT, as at low affinity (high $K_{M}$), the NADA containing system is again able to maintain higher NAD concentrations.
-The NAD consumption flux is still higher in the NamPRT and NNMT containing compartment.
-Taken together this might explain why upon the development of a sufficiently high affinity of NamPRT for its substrate, which seems to have been induced by the appearance of NNMT, NADA is lost.
+In the presence of NNMT, the NamPRT compartment has both higher NAD consumption rates and higher steady state NAD concentrations than the compartment containing NADA (fig.~\ref{fig:NNMT_comp_advantage}C and~D).
+The high NAD concentrations are dependent on a high affinity of NamPRT.
+When simulating a low affinity (high $K_{M}$) of NamPRT, the NADA compartment is able to maintain higher NAD concentrations although it still has a lower the NAD consumption flux.
+Taken together, the results suggest that the NADA pathway became less worthwhile when a higher affinity of NamPRT developed, possibly induced by NNMT.

### 2018-07-07 latex/figurelegends.tex
Included changes by Ines
Added Fig S2

@@
 
-\section{Figure Legends}
+\section*{Figure Legends}
 
-\subsection{Figure 1}
+\subsection*{Figure 1}
 
@@
 
-\subsection{Figure 2}
+\subsection*{Figure 2}
 
@@
   \includegraphics[width=0.75\textwidth]{fig_2_phylo_distribution}
-  \caption{\textbf{Evolutionary distribution of NADA, NNMT, and NamPRT and their relation to the number of NAD consumers.} A) Distribution of NADA, NNMT, and NamPRT in selected major taxa.
+  \caption{\textbf{Phylogenetic distribution of NADA, NNMT, and NamPRT and their relation to the number of NAD consumers.} A) Distribution of NADA, NNMT, and NamPRT in selected major taxa.
 NADA is dominant in Bacteria, Fungi, and Plants (Viridiplantae), whereas NamPRT together with NNMT is dominant in Metazoa.
@@
 
-\subsection{Figure 3}
+\subsection*{Figure 3}
 
@@
 
-\subsection{Figure 4}
+\subsection*{Figure 4}
 
@@
 
-\subsection{Figure 5}
+\subsection*{Figure 5}
 
@@
 
-\subsection{Figure 6}
+\subsection*{Figure 6}
 
@@
 
-\subsection{Figure 7}
+\subsection*{Figure 7}
 
@@
 
-\subsection{Figure 8}
+\subsection*{Figure 8}
 

### 2018-07-07 latex/supplementaries.tex
Included changes by Ines
Added Fig S2

@@
 
-\section{Supplementary Legends}
+\section*{Supplementary Legends}
 
-\subsection{Supplementary figure S1}
+\subsection*{Supplementary figure S1}
 
@@
 
-\subsection{Supplementary figure S2}
+\subsection*{Supplementary figure S2}
 
-\todo{Figure must be made!}\textbf{Distribution of NNMT in birds.} ...
+\textbf{The phylogenetic distribution of NamPRT and NNMT in birds and reptiles is scattered.}
 
-\subsection{Supplementary figure S3}
+\subsection*{Supplementary figure S3}
 
@@
 
-\subsection{Supplementary table S1}
+\subsection*{Supplementary table S1}
 
@@
 
-\subsection{Supplementary table S2}
+\subsection*{Supplementary table S2}
 
-\textbf{Overview of kinetic constants used for the construction of the model.}
+\textbf{Overview of kinetic constants and rate laws used for the construction of the mathematical model.}

### 2018-07-08 latex/header.tex
edited header

@@
 Strikingly, loss of the deamidation pathway in early vertebrates is preceded by the occurrence of a new enzyme that marks Nam for excretion by methylation – nicotinamide N-methyltransferase (NNMT).
-The physiological relevance of NNMT is still enigmatic.
-Why is there an enzyme that removes Nam from recycling, seemingly not having any other physiological role?
-Despite the occurrence of this Nam degrading enzyme NNMT, we observe a diversification of NAD dependent signalling enzymes in vertebrates.
+Despite the occurrence of this Nam degrading enzyme, we observe a diversification of NAD dependent signalling enzymes in vertebrates.
 This seems counterintuitive.
@@
 To better understand these evolutionary changes and the role of NNMT for NAD metabolism, we built a mathematical model of the pathway using published enzyme kinetics data.
-Our results indicate that NNMT is required to enable high NAD consumption fluxes necessitated by the increasing diversification of NAD-dependent signalling pathways.
+Our results indicate that NNMT is required to enable high NAD consumption fluxes enabling diversification of the NAD-dependent signalling pathways.
 This kinetic regulation requires a high substrate affinity of the key enzyme for Nam salvage, NamPRT.

### 2018-07-08 latex/methods.tex
First round of including Mathias Z.s comments. Not finished, yet.
Improved phylogenetic methods

@@
 \section{Experimental Procedure}
+
+\subsection{Phylogenetic Analysis}
+
+NADA, NamPRT, and NNMT enzymes or enzyme candidates were identified using Blastp with known protein sequences against the NCBI non-redundant protein sequence database.
+A list of functionally verified enzymes used as templates is given in supplementary table~S1.
+Blastp parameters were set to yield maximum 20\,000 target sequences, using the BLOSUM62 matrix with a word size of 6 and gap opening and extension costs of 11 and 1, respectively.
+Low-complexity filtering was disabled.
+To prevent cross-hits, a matrix was created in which the lowest e-values were given at which Blast yielded the same result for each query protein pair.
+With help of the matrix, the e-value cut-off was set to 1e-30 for all enzymes.
+To further prevent false positives, a minimal length limit was set based on a histogram of the hit lengths found for each query protein, excluding peaks much lower than the total protein length.
+Length limits are given in supplementary table~S1.
+In addition, obvious sequence contaminations were removed by manual inspection of the results.
+The taxonomy IDs of the species for each enzyme was derived from the accession2taxonomy database provided by NCBI.
+Scripts for creating, analysing, and visualising the phylogenetic tree were written in Python~3.5, using the ETE3 toolkit \cite{Huerta-Cepas2016}.
 
@@
 Related figures were generated using Gnuplot~5.0.
-
-\subsection{Phylogenetic Analysis}
-
-NADA, NamPRT, and NNMT enzymes or enzyme candidates were identified using Blastp with known enzymes against the NCBI non-redundant protein sequence database (nr).
-A list of functionally verified enzymes used as templates is given in supplementary table~S1.
-This table also includes the length cut-offs for identified enzymes.
-The e-value cut-off was 1e-30 for all enzymes.
-Blastp parameters were set to yield maximum 20\,000 target sequences, using the BLOSUM62 matrix with a word size of 6 and gap opening and extension costs of 11 and 1, respectively.
-Low-complexity filtering was disabled.
-Obvious sequence contaminations were removed by manual inspection of the results.
-The taxonomy IDs of the species for each enzyme was derived from the accession2taxonomy database provided by NCBI.
-Scripts for creating, analysing, and visualising the phylogenetic tree were written in Python, using the ETE3 toolkit (Huerta-Cepas, 2010).
 
@@
 
-2\,µM of enzyme were incubated with 5-phospho-D-ribose 1-diphosphate (PRPP, 0.1\,mM or 1\,mM) and nicotinamide (Nam, 0.1\,mM or 1\,mM) in reaction buffer (20\,mM Tris-HCl pH 8.0, 500\,mM NaCl, 2\,mM MgCl2, and 0.03\% BSA), in absence or presence of 1\,mM of adenosine triphosphate (ATP).
+2\,µM of enzyme were incubated with 5-phospho-\textsc{d}-ribose 1-diphosphate (PRPP, 0.1\,mM or 1\,mM) and nicotinamide (Nam, 0.1\,mM or 1\,mM) in reaction buffer (20\,mM Tris-HCl pH 8.0, 500\,mM NaCl, 2\,mM MgCl$_{2}$, and 0.03\% BSA), in absence or presence of 1\,mM of adenosine triphosphate (ATP).
 The 1.2\,mL reaction was incubated for 10 minutes at 30\,°C and the enzymatic activity stopped with 0.1\,mM of FK866, the samples were frozen in liquid nitrogen.
@@
 The experiments were acquired with zgesgppe pulse sequence, allowing water suppression using excitation sculpting with gradients and perfect echo.
-The temperature was kept constant at 300\,K and the acquisition parameters were the following: number of scans = 2000, relaxation delay = 1\,s, acquisition time = 1.6\,s, data points = 65K, and spectral width = 14\,ppm.
+The temperature was kept constant at 300\,K and the acquisition was started with 2000 scans, 1\,s relaxation delay, 1.6\,s acquisition time, 65\,000 data points, and a spectral width of 14\,ppm.
 

### 2018-07-08 latex/latex_to_word.py

@@ -162,6 +162,8 @@ for line in lines:
 			line = re.sub(r'\\textbf\{([^\}\{\\]+)\}', r'<b>\1</b>', line)
 		if r'\textit' in line:
 			line = re.sub(r'\\textit\{([^\}\{\\]+)\}', r'<i>\1</i>', line)
+		if r'\textsc' in line:
+			line = re.sub(r'\\textsc\{([^\}\{\\]+)\}', lambda match: match.group(1).upper(), line)
 		if r'\todo' in line:
 			line = re.sub(r'\\todo\{([^\}\{\\]+)\}', r'<span class="todo">\1</span>', line)
 			line = re.sub(r'\\todo\[author=([^\]\{\\]+)\]\{([^\}\{\\]+)\}', r'<span class="todo"><i>\1</i>: \2</span>', line)

### 2018-07-08 latex/discussion.tex
Started work on discussion.
As suggested by Mathias Z., I started with a list of points to discuss (mostly his comments until now).

@@
 
+\subsection{Points to discuss}
+
++ NNMT evolution.
+
+--> Why was the yeast MT defined as NNMT?
+
+--> Are there any ideas where the NNMT came from?
+
+--> Any information on the methylation activity (also in comparison to others)? Does it fit to the found kinetic functions?
+
++ NADA disappears but rest of pathway stays. Why?
+
+--> NADS needed for \textit{de~novo} pathway
+
+--> NMNAT also needed in NamPRT and \textit{de~novo} pathway
+
+-----> Is the \textit{de~novo} pathway still necessary? If we take up enough Vit B3 with the diet, why \textit{de~novo}?
+
+--> NAPRT needed for NA that is taken up with the diet (and if NA is processed, the other two enzymes are also needed to continue processing)
+
++
+
+\subsection{Old discussion}
+
 The dynamic interactions among metabolism, signal transduction, and gene regulation are still not very well understood.
 New approaches are required to disentangle the underlying network, helping us to understand how alterations affect human physiology.
-We here combined phylogenetic analysis and mathematical modelling of the NAD biosynthesis and consumption pathway to better understand the evolution of this pathway and especially the role of the methyltransferase NNMT.
+We here combined phylogenetic analysis, mathematical modelling, and experimental evidence to better understand the evolution of the NAD biosynthesis and consumption pathways and especially the role of the methyltransferase NNMT.
 Our analyses reveal that NNMT plays a vital role for the diversification of NAD-dependent signalling reactions and potentially also for NAD homeostasis.
-We show that the presence of NNMT enables higher NAD consumption rates in comparison to NamPRT or NADA alone.
+We show that the combination of NNMT and NamPRT enables higher NAD consumption rates in comparison to NamPRT or NADA alone.
 It furthermore appears that NNMT makes both NAD concentration and NAD consumption relatively independent of other processes requiring NAD such as cell growth.
-We furthermore suggest that NNMT exerted an evolutionary pressure on NamPRT, driving the development of the high affinity of NamPRT for its substrates found in the human enzyme \cite{Burgos2008}.
+In addition, we suggest that NNMT exerted an evolutionary pressure on NamPRT, driving the development of the high affinity of NamPRT for its substrates found in the human enzyme \cite{Burgos2008}.
 These findings shed new light on the role of NNMT, which has earlier been recognized as potential marker for some types of cancer (e.g.
 \cite{Okamura1998}).
-Increased NNMT expression in cancer might thus serve to remove Nam produced by increased NAD-dependent signalling.
+Increased NNMT expression in cancer might serve to remove Nam produced by increased NAD-dependent signalling.
 Further analysis is required to better understand which NAD-dependent signalling reactions are affected.
@@
 
-We here neglected the potential effects of the methyl donor \textit{S}-adenosyl methionine (SAM) and its precursor methionine.
+We neglected the potential effects of the methyl donor \textit{S}-adenosyl methionine (SAM) and its precursor methionine in our models.
 Changing SAM availability is most likely contributing yet another regulatory level for Nam availability and thus an additional interaction point between gene regulation and metabolism.
 It has been shown earlier that NNMT expression influences protein methylation dependent on methionine availability \cite{Ulanovskaya2013}.
-One of the challenges in the analysis of this aspect is the availability of \textit{in vivo} concentration measurements for SAM and the large amount of reactions using it as substrate.
+One of the challenges in the analysis of this aspect is the availability of \textit{in~vivo} concentration measurements for SAM and the large amount of reactions using it as substrate.
 A simplified mathematical model of this pathway has been published earlier \cite{Reed2004} providing the basis for future studies.

### 2018-07-08 latex/results.tex
Defined how many birds dont have NNMT

@@
 NNMT seems to have arisen \textit{de~novo} in the common ancestor of Ecdysozoa and Lophotrochozoa.
-We were unable to find any \textit{NNMT} gene with considerable\todo{define considerable} similarity in fungi or plants.
+We were unable to find any \textit{NNMT} gene with considerable \todo{define considerable} similarity in fungi or plants.
 Even though fungal genes named \textit{NNMT} can be found in databases, these genes show very limited sequence similarities to the \textit{NNMT} of nematodes or deuterostomes.
@@
 In Deuterostomes, the only large clade that possesses only NamPRT and seems to have lost NNMT again are Sauropsida and among them especially birds.
-The reason why a lot of\todo{be more precise!} bird genomes do not seem to encode NNMT remains unclear, as the distribution is quite scattered (suppl.
+The reason why about half of the sequenced bird genomes do not seem to encode NNMT remains unclear, as the distribution is quite scattered (suppl.
 fig.~S2).

### 2018-07-08 latex/discussion.tex
Minor formatting changes before sending manuscript for internal review

@@
 
---> Why was the yeast MT defined as NNMT?
+-- Why was the yeast MT defined as NNMT?
 
---> Are there any ideas where the NNMT came from?
+-- Are there any ideas where the NNMT came from?
 
---> Any information on the methylation activity (also in comparison to others)? Does it fit to the found kinetic functions?
+-- Any information on the methylation activity (also in comparison to others)? Does it fit to the found kinetic functions?
 
@@
 
---> NADS needed for \textit{de~novo} pathway
+-- NADS needed for \textit{de~novo} pathway
 
---> NMNAT also needed in NamPRT and \textit{de~novo} pathway
+-- NMNAT also needed in NamPRT and \textit{de~novo} pathway
 
------> Is the \textit{de~novo} pathway still necessary? If we take up enough Vit B3 with the diet, why \textit{de~novo}?
+---- Is the \textit{de~novo} pathway still necessary? If we take up enough Vit B3 with the diet, why \textit{de~novo}?
 
---> NAPRT needed for NA that is taken up with the diet (and if NA is processed, the other two enzymes are also needed to continue processing)
-
-+
+-- NAPRT needed for NA that is taken up with the diet (and if NA is processed, the other two enzymes are also needed to continue processing)
 

### 2018-07-08 latex/introduction.tex
Minor formatting changes before sending manuscript for internal review

@@
 The nearly complete recycling of Nam is achieved by an extraordinary high affinity of NamPRT to Nam, the $K_{M}$ being in the low nanomolar range \cite{Burgos2008}.
-This extraordinary affinity appears to be mediated by an ATP-dependent phosphorylation of a histidine residue in the catalytic core \todo{Needs reference}.
+This extraordinary affinity appears to be mediated by an ATP-dependent phosphorylation of a histidine residue in the catalytic core\todo{Needs a reference}.
 Despite the importance of its salvage, Nam can also be marked for excretion by methylation.

### 2018-07-09 latex/figurelegends.tex
Minor formatting changes before sending manuscript for internal review
revised introduction and figure legends, still working on results

@@
 For the recycling of Nam, two different pathways exist.
-The pathway found in yeast and many bacteria starts with the deamination of Nam by Nam deamidase (NADA).
+The pathway found in yeast and many bacteria starts with the deamidation of Nam by Nam deamidase (NADA).
 The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates.
@@
 NADA is dominant in Bacteria, Fungi, and Plants (Viridiplantae), whereas NamPRT together with NNMT is dominant in Metazoa.
-Numbers at the pie charts show, how many species of the taxon possess the respective enzyme combination indicated by the colour, which is explained in the lower right of the figure.
+Numbers at the pie charts show, how many species of the taxon possess the respective enzyme combination indicated by the colour explained in the lower right of the figure.
 Below the taxon name, the number of species in that taxon is given.
@@
   \includegraphics[width=\textwidth]{fig_3_NNMT_NAD_flux}
-  \caption{\textbf{NNMT enables high NAD consumption flux.} We used a dynamic model of NAD biosynthesis and consumption (for details, see experimental procedures) to simulate NAD consumption flux.
+  \caption{\textbf{NNMT enables high NAD consumption flux.} We used a dynamic model of NAD biosynthesis and consumption (for details, see experimental procedures) to simulate steady state NAD  consumption flux  (A) and concentration (B).
 The amount of NMNAT and NamPRT used in the simulations where adjusted such that the free NAD concentrations were in the range reported in the literature.
@@
 Details are given in supplementary table~S2.
-In the absence of NNMT (blue lines), steady state NAD consumption rates (A) are higher despite reduced NAD concentrations (B).
+In the absence of NNMT (blue lines), steady state NAD consumption rates are higher despite reduced NAD concentrations.
 Increasing the amount of NamPRT in the simulation fourfold (blue dotted lines) partially compensates for the decreased NAD steady state concentration caused by Nam degradation through NNMT.}
@@
   \centering
-  \includegraphics[width=0.9\textwidth]{fig_4_NamPRT_affinity_Nam}
-  \caption{\textbf{Role of NamPRT affinity for Nam.} Using the dynamic model of NAD biosynthesis and consumption we simulated the affect of different Michaelis-Menten constants ($K_{M}$) of NamPRT for Nam on steady state NAD consumption flux and NAD concentration at different cell division rates.
-All the parameters were equal to those used for the simulations in figure~\ref{fig:NNMT_NAD_flux}.
-In the absence of NNMT (A and B), the $K_{M}$ of NamPRT has little influence on NAD consumption and concentration, but both are changing with cell division rates.
-In the presence of NNMT (C and D), decreasing $K_{M}$ of NamPRT enables increasing NAD consumption flux and NAD concentration.
+  \includegraphics[width=0.72\textwidth]{fig_4_NamPRT_affinity_Nam}
+  \caption{\textbf{Role of NamPRT substrate affinity} Using the dynamic model of NAD biosynthesis and consumption we simulated the affect of different Michaelis-Menten constants ($K_{M}$) of NamPRT for Nam on the steady state NAD consumption flux and NAD concentration at different cell division rates.
+All parameters were equal to those used for the simulations in figure~\ref{fig:NNMT_NAD_flux}.
+In the absence of NNMT, the $K_{M}$ of NamPRT has little influence on NAD consumption (A) and concentration (B), but both are changing with cell division rates.
+In the presence of NNMT, decreasing $K_{M}$ of NamPRT enables increasing NAD consumption flux (C) and NAD concentration (D).
 NNMT furthermore makes both, NAD consumption flux and concentration, almost independent of cell division rates.
@@
 The inserted region is not resolved in any of currently available crystal structures of NamPRT and thus appears to be a flexible loop structure at the surface of the NamPRT dimer, coloured in red.
-Immunofluorescence images show that the localisation of the FLAG-tagged mutant protein lacking the unresolved loop (C) is not changed compared to FLAG-tagged human wildtype NamPRT.
-Both show a heterogeneous nuclear cytosolic localisation in HeLa S3 cells.
+Immunofluorescence images (C) show that the localisation of the FLAG-tagged mutant protein lacking the unresolved loop  is not changed compared to FLAG-tagged human wildtype NamPRT.
+Both show a heterogeneous nuclear-cytosolic localisation in HeLa S3 cells.
 \textit{In vitro} measurements using recombinant protein show that the mutant NamPRT has a lower activity than the wildtype enzyme (D) and is not activated by ATP (E).
-Bars in D) and E) with different letters indicate significant difference of measured values as estimated using a T test assuming independent samples and significance at $p < 0.05$.}
+Bars in panels D and E that have different letters indicate significant difference of measured values as estimated using a T test assuming independent samples and significance at $p < 0.05$.}
   \label{fig:unresolved_loop}

### 2018-07-09 latex/introduction.tex
revised introduction and figure legends, still working on results

--- 2018-07-07 latex/introduction.tex
@@
 The significance of NAD-dependent signalling for NAD homeostasis has long been underestimated.
-However, it has now been established that substances affecting NAD biosynthesis lead to a rapid decline of the NAD concentration \cite{Buonvicino2018} suggesting that NAD-dependent signalling consumes substantial amounts of NAD, which is why we later refer to them also as NAD-consuming reactions.
+It has now been established, however, that substances affecting NAD biosynthesis lead to a rapid decline of the NAD concentration \cite{Buonvicino2018}.
+This  suggests that NAD-dependent signalling reactions consume substantial amounts of NAD.
+Therefore, we later refer to them also as NAD-consuming reactions.
 The resulting NAD turnover differs in a cell type specific manner and can lead to an NAD half-life as short as two hours \cite{Liu2018}.
 To maintain the NAD concentration at physiological levels, NAD biosynthesis needs to act at an equally rapid rate.
-Imbalances in NAD homeostasis have been linked to various diseases, in particular related to ageing, such as diabetes, neurodegenerative disorders, and cancer \cite{Chiarugi2012,Verdin2015}.
+Imbalances in NAD homeostasis have been linked to various mainly age related diseases, such as diabetes, neurodegenerative disorders, and cancer \cite{Chiarugi2012,Verdin2015}.
 Several recent studies have demonstrated impressive health benefits of dietary supplementation with intermediates of NAD biosynthesis including Nam mononucleotide (NMN) and Nam riboside (NR) \cite{Yoshino2018}.
@@
 If Nam would not be constantly recycled into NAD, humans would require a much higher daily vitamin B3 intake than the 16\,mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}.
-In the first step of the salvage pathway, Nam is converted to the mononucleotide NMN by the Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate.
-The nearly complete recycling of Nam is achieved by an extraordinary high affinity of NamPRT to Nam, the $K_{M}$ being in the low nanomolar range \cite{Burgos2008}.
-This extraordinary affinity appears to be mediated by an ATP-dependent phosphorylation of a histidine residue in the catalytic core \todo{Needs reference}.
+Two principal pathways exist that recycle Nam.
+Firstly, vertebrates use a direct two step pathway staring with the convertion of Nam into the mononucleotide NMN by the Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate.
+The nearly complete recycling of Nam by NamPRT is achieved by an extraordinary high subtrate affinity to Nam, the $K_{M}$ being in the low nanomolar range \cite{Burgos2008}.
+This appears to be mediated by an ATP-dependent phosphorylation of a histidine residue in the catalytic core\todo{PMID:19666527}.
 Despite the importance of its salvage, Nam can also be marked for excretion by methylation.
@@
 While NamPRT is seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, NNMT seems to have no metabolic function other than to remove Nam from NAD metabolism.
-This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deamidate Nam to nicotinic acid (NA) using the nicotinamide deamidase (NADA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (fig.~\ref{fig:pathway_overview}).
+Although it has been suggested that it potentially acts as methylation sink \cite{NNMTreview2018} .
+Methyl-Nam produced by NNMT is in mammals excreted with the urine, thus removing Nam from recycling.
+
+Secondly, in many prokayotes as well as in plants and fungi a four steps starting with the deamidation of Nam to nicotinic acid (NA) by NADA is used.
+(\ref{fig:pathway_overview}).
+The three enzymes that act after NADA belong to the Preiss-Handler pathway that also exists in vertebrates.
+The Preiss-Handler pathway first converts  NA into the corresponding mononucleotide (NAMN), in a reactions performed by the NAD specific phosphoribosyl transferase NAPRT.
+The convertion of the mononucleotides  NMN and NAMN into their dinucleotides NAD and NaAD is performed by the same enzyme, the  Nam/NA adenylytransferases (NMNATs) that is essential in all Organisms\cite{Gossmann2011}.
+The recycling pathway via NA finally requires re-amidation of the dinucleotide, which is catalysed by NAD synthase.
+This reaction includes an enzyme adenylation step that consumes ATP.
+Therefore the Nam recycling by NADA is energetically less efficient under normal conditions.
 
 We here present a phylogenetic analysis of the NAD pathways.
-The results suggest that there has been a selection for the co-existence of NamPRT and NNMT in Deuterostomia, while the deamidation pathway, which is dominant in bacteria, is lost.
+Our results suggest that there has been a selection for the co-existence of NamPRT and NNMT in Deuterostomia, while the deamidation pathway, which is dominant in bacteria, is lost.
 This transition was accompanied by a marked increase in the number of NAD-consuming signalling enzymes.
@@
 Indeed, we identified a short sequence insertion in NamPRT, which first occurs in Deuterostomes and that appears to modulate the affinity of NamPRT.
-Simulating the resource competition we furthermore show that the presence of high affinity NamPRT together with NNMT makes the NADA-dependent pathway obsolete.
+Simulating the resource competition, we furthermore show that the presence of high affinity NamPRT together with NNMT makes the NADA-dependent pathway obsolete.
 

### 2018-07-09 latex/results.tex
revised introduction and figure legends, still working on results

@@
 
-\subsection{Paradoxical evolutionary correlation between NAD-dependent signaling and precursor metabolism}
-
-Two principal pathways exist that recycle Nam.
-As shown in figure \ref{fig:pathway_overview}, the major pathway found in most prokaryotes, yeast, and plants includes four steps starting with the deamidation of Nam to nicotinic acid by NADA.
-The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates.
-The recycling pathway found in mammals directly converts Nam into Nam mononucleotide (NMN) a reaction catalysed by NamPRT and driven by a non-stoichiometric ATP-hydrolysis.
-A similar reaction catalysed by an evolutionary related enzyme, NAPRT, converts NA into the corresponding mononucleotide (NAMN) in the Preiss-Handler pathway.
-NMN and NAMN are converted into dinucleotides by Nam/NA adenylytransferases (NMNATs).
-The recycling pathway via NA finally requires re-amidation of the dinucleotide, which is catalysed by NAD synthase.
-This reaction includes an enzyme adenylation step that consumes ATP.
-Even though Nam recycling by NADA is energetically less efficient, this pathway is preferentially used by most bacteria, fungi and plants, whereas most metazoans recycle Nam using NamPRT (fig.~\ref{fig:phylo_distribution}A).
-
-Analysing the phylogeny of the NAD-recycling enzymes in Metazoa in more detail revealed that not only NamPRT replaced NADA, but in most organisms, especially in Deuterostomes, NamPRT is found together with NNMT (fig.~\ref{fig:phylo_distribution}B).
-NNMT methylates Nam to methyl-Nam that is in mammals excreted with the urine, thus removing Nam from recycling.
+\subsection{Paradoxical evolutionary correlation between NAD-dependent signalling and precursor metabolism}
+As the Preiss-Handler pathway  is  present in virtually all organisms analysed previously\cite{Gossmann2012FEBS} and as based on metabolic pathway analysis we have previously shown tha NMNAT is essential for NAD dependent biosynthesis independent of the biosynthetic route or precursor use \cite{Gossmann2011} we here focused on the phylogenetic analysis of the two enzymes that initiate the two different NAD-salvage pathways NADA and NamPRT as well as on the NAM methyltransferase NNMT as the distribution of these enzymes seems to differ substantially between different organisms.
+As can be seen from Fig.
+(fig.~\ref{fig:phylo_distribution}A, bacteria, fungi and plants predominantly encode NADA and only a few of them harbour NamPRT.
+In contrast, Metazoa predominantly have NamPRT together with NNMT.
+Analysing the phylogeny distribution in Metazoa in more detail revealed that not only is NamPRT replaced by NADA, but in most organisms, especially in Deuterostomes, NamPRT is found together with NNMT (fig.~\ref{fig:phylo_distribution}B).
 NNMT seems to have arisen \textit{de~novo} in the common ancestor of Ecdysozoa and Lophotrochozoa.
 We were unable to find any \textit{NNMT} gene with considerable \todo{define considerable} similarity in fungi or plants.
-Even though fungal genes named \textit{NNMT} can be found in databases, these genes show very limited sequence similarities to the \textit{NNMT} of nematodes or deuterostomes.
-The yeast NNMT protein has meanwhile been shown to be a lysin-protein-methyltransferase \cite{Wlodarski2011}.
 

### 2018-07-09 latex/introduction.tex
further editing of introduction and results first paragraph completed

@@
 While NamPRT is seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, NNMT seems to have no metabolic function other than to remove Nam from NAD metabolism.
-Although it has been suggested that it potentially acts as methylation sink \cite{NNMTreview2018} .
-Methyl-Nam produced by NNMT is in mammals excreted with the urine, thus removing Nam from recycling.
+Although it has been suggested that the process potentially acts as a metabolic methylation sink \cite{PMID:28291578} .
 
@@
 The Preiss-Handler pathway first converts  NA into the corresponding mononucleotide (NAMN), in a reactions performed by the NAD specific phosphoribosyl transferase NAPRT.
-The convertion of the mononucleotides  NMN and NAMN into their dinucleotides NAD and NaAD is performed by the same enzyme, the  Nam/NA adenylytransferases (NMNATs) that is essential in all Organisms\cite{Gossmann2011}.
+The convertion of the mononucleotides  NMN and NAMN into their dinucleotides NAD and NaAD is performed by the same enzyme, the  Nam/NA adenylytransferases (NMNATs) that is essential in all Organisms\cite{PMID:21729004}.
 The recycling pathway via NA finally requires re-amidation of the dinucleotide, which is catalysed by NAD synthase.

### 2018-07-09 latex/results.tex
further editing of introduction and results first paragraph completed

@@
 \subsection{Paradoxical evolutionary correlation between NAD-dependent signalling and precursor metabolism}
-As the Preiss-Handler pathway  is  present in virtually all organisms analysed previously\cite{Gossmann2012FEBS} and as based on metabolic pathway analysis we have previously shown tha NMNAT is essential for NAD dependent biosynthesis independent of the biosynthetic route or precursor use \cite{Gossmann2011} we here focused on the phylogenetic analysis of the two enzymes that initiate the two different NAD-salvage pathways NADA and NamPRT as well as on the NAM methyltransferase NNMT as the distribution of these enzymes seems to differ substantially between different organisms.
+We have  previously shown  that NMNAT is essential for NAD dependent biosynthesis independent of the biosynthetic route or precursor \cite{PMID:21729004}.
+In addition, the Preiss-Handler pathway was found  present in virtually all organisms included in an earlier phylogenetic analysis\cite{Gossmann2012FEBS}.
+Thus, we here focused on the phylogenetic analysis of the two enzymes that initiate the two different NAD-salvage pathways, NADA and NamPRT.
+These show a scattered distribution in bacteria \todo{MathiasB: please find and insert citation} and have been shown to occur concomitantly in some marine invertebrates \cite{Gossmann2012FEBS}.
+In addition, we included  NAM methyltransferase NNMT into our analysis, as its function and origin is unclear.
 As can be seen from Fig.
@@
 In contrast, Metazoa predominantly have NamPRT together with NNMT.
-Analysing the phylogeny distribution in Metazoa in more detail revealed that not only is NamPRT replaced by NADA, but in most organisms, especially in Deuterostomes, NamPRT is found together with NNMT (fig.~\ref{fig:phylo_distribution}B).
+Analysing the phylogeny distribution in Metazoa in more detail revealed that not only is NamPRT dominating and NADA appears to be lost in vertebrates, in most organisms NamPRT is found together with NNMT (fig.~\ref{fig:phylo_distribution}B).
 NNMT seems to have arisen \textit{de~novo} in the common ancestor of Ecdysozoa and Lophotrochozoa.
-We were unable to find any \textit{NNMT} gene with considerable \todo{define considerable} similarity in fungi or plants.
+As we were unable to find any \textit{NNMT} gene with an e-value below ....
+\todo{Mathias B.
+include e-value}  in fungi or plants.
+The gene with the highest sequence similarity in S.
+cerevisiae is  .....
+with proven function .....\todo{Mathias B: include e-value}
 
-Nematodes are the only organisms, where we observed any combination of NADA, NamPRT, and NNMT.
-In Deuterostomes, the only large clade that possesses only NamPRT and seems to have lost NNMT again are Sauropsida and among them especially birds.
-The reason why about half of the sequenced bird genomes do not seem to encode NNMT remains unclear, as the distribution is quite scattered (suppl.
+Nematodes are the only organisms, where we observed a concomitant presence of NADA and NNMT.
+In Deuterostomes, the only large clade that possesses only NamPRT and seems to have lost NNMT are Sauropsida, and among them especially birds.
+The reason why about half of the sequenced bird genomes do not seem to encode for  NNMT,  remains unclear.
+The distribution of NNMT in birds is quite scattered (suppl.
 fig.~S2).
+not corresponding to current species trees \cite{http://dx.doi.org/10.1038/nature15697} .
 The lack of NNMT might be related to the excretion system, as the product of NNMT, methyl-Nam, is in mammals excreted with the urine.
-There are some Metazoan species for which we could not find NamPRT or NADA, while NNMT was present.
+There few Metazoan species for which we could not find NamPRT or NADA, while NNMT was present.
 We assume that this is due to incomplete genomes in the database, as the distribution of such species is scarce and widely scattered.
@@
 Moreover, the selection for having both NamPRT and NNMT coincides with an increased diversification of NAD-dependent signalling.
-This observation seems counterintuitive.
-One would expect that increased NAD-dependent signalling should be accompanied by an increase of substrate availability for NAD biosynthesis.
-However, NNMT irreversibly removes Nam from salvage back to NAD.
+This observation seems counterintuitive, as one would expect that increased NAD-dependent signalling should be accompanied by an increase of substrate availability for NAD biosynthesis.
+However, NNMT irreversibly removes Nam from salvage and thus is assumed to potentially decrease the cellular NAD-concentration.
 

### 2018-07-09 latex/results.tex
results edited except last subsection

@@
 
-\subsection{Functional properties of NamPRT and NNMT have evolved to maximize NAD-dependent signaling}
+\subsection{Functional properties of NamPRT and NNMT have evolved to maximize NAD-dependent signalling}
 
@@
 
-To be able to compare metabolic features of evolutionary quite different systems in our simulations and as we have limited information about expression levels of enzymes or changes of kinetic constants during evolution, we initially used the kinetic constants found for human or yeast enzymes for all systems analysed and used equal amounts of enzymes for all reactions\todo{Shorten or split sentence and correct "equal amounts"}.
+To be able to compare metabolic features of evolutionary quite different systems in our simulations and as we have limited information about expression levels of enzymes, we initially used equal amounts for all enzymes.
+As we have very few cross species kinetic data, we were furthermore mainly relying on kinetic constants found for human or yeast enzymes.
 Wherever possible, we did not only include substrate affinities but also known product inhibitions or inhibition by downstream metabolites.
-As we assume that cell growth is, besides NAD-consuming reactions, a major driving force for NAD biosynthesis, we analysed different growth rates (cell division rates) by simulating different dilution rates for all metabolites.
+As we in addition assumed that cell growth is, besides NAD-consuming reactions, a major driving force for NAD biosynthesis, we analysed different growth rates (cell division rates) by simulating different dilution rates for all metabolites.
 
@@
 We calculated steady state NAD concentrations and NAD consumption rates by simulating NAD biosynthesis proceeding via NamPRT in the presence or absence of NNMT.
+To achieve free NAD concentration in the range reported in the literature and due to the very low turnover of NamPRT, we had to use 10 fold higher NamPRT levels compared to other enzymes and adjusted the amount of NMNAT to the same level to avoid that the NAD synthesis rates are limited by this enzyme.
 As shown in figure~\ref{fig:NNMT_NAD_flux}, the presence of NNMT enables higher NAD consumption rates (fig.~\ref{fig:NNMT_NAD_flux}A) while reducing the steady state concentration of NAD (fig.~\ref{fig:NNMT_NAD_flux}B).
@@
 
-These results can be explained when looking in more detail at the kinetic parameters of NamPRT and NAD-consuming enzymes such as Sirtuin~1.
-Most NAD-consuming enzymes are inhibited by their product Nam, explaining why the presence of NNMT enables higher NAD consumption fluxes.
-At the same time, the high substrate affinity of NamPRT maintains a sufficiently high NAD concentration, although the concentration is lower than in the system without NNMT.
+These results can be explained by looking in more detail at the kinetic parameters of NamPRT and NAD-consuming enzymes such as Sirtuin~1.
+Most NAD-consuming enzymes are inhibited by their product Nam.
+Thus the presence of NNMT enables higher NAD consumption fluxes, by removing excess Nam from the cells.
+At the same time, the high substrate affinity of NamPRT maintains a sufficiently high NAD concentration, although the concentration is of course lower than in the system without NNMT.
 
 As measured kinetic parameters of NamPRT are only available for the human enzyme \cite{Burgos2008}, we analysed the potential effect of NamPRT affinity ($K_{M}$) on NAD steady state concentration and NAD consumption flux.
-In the absence of NNMT, a change in $K_{M}$ has very little effect on steady state NAD concentration and NAD consumption flux (fig.~\ref{fig:NamPRT_affinity_Nam}A and~B).
+In the absence of NNMT, a variation of the substrate affinity of NamPRT for Nam, has very little effect on steady state NAD concentration and NAD consumption flux (fig.~\ref{fig:NamPRT_affinity_Nam}A and~B).
 In the presence of NNMT, however, NAD consumption flux and NAD concentration increases with decreasing $K_{M}$ values (fig.~\ref{fig:NamPRT_affinity_Nam}C and~D).
@@
 Of course, this is an artificial scenario, as one would assume organisms to regulate enzyme expression to achieve similar levels of metabolite concentrations instead.
-However, there seems to be a tradeoff between maintainable NAD concentration and consumption flux in the absence of NNMT.
+Nevertheless, in the absence of NNMT there seems to be a tradeoff between maintainable NAD concentration and consumption flux.
 In contrast, in the presence of NNMT, NAD consumption rates and concentrations are almost independent of cell division rates.
 
-Figure~\ref{fig:NamPRT_affinity_Nam}E and~F shows a direct comparison of different affinities of NamPRT for Nam with and without NNMT.
-At an affinity of $K_{M}$ = 1\,$\mu$M, which is in the range of the $K_{M}$ of NADA for Nam, NAD consumption flux is higher with NNMT than without, except for low cell division rates (fig.~\ref{fig:NamPRT_affinity_Nam}E).
-At a lower affinity of 5\,nM, increase in consumption rate even more pronounced.
-The NAD concentration is lower with NNMT than without for both tested $K_{M}$ values (fig.~\ref{fig:NamPRT_affinity_Nam}F).
+Figure~\ref{fig:NamPRT_affinity_Nam}E and~F show a direct comparison of different affinities of NamPRT for Nam with and without NNMT.
+At an affinity of $K_{M}$ = 1\,$\mu$M, which is in the range of the $K_{M}$ of NADA for Nam, NAD consumption flux is only higher with NNMT than without at low cell division rates (fig.~\ref{fig:NamPRT_affinity_Nam}E).
+If the affinity of NamPRT is high enough, consumption rates are always higher with NNMT than without it.
+The NAD concentration is always lower with NNMT (fig.~\ref{fig:NamPRT_affinity_Nam}F).
 
@@
 In figure~\ref{fig:optimal_substr_affinities} is shown that the substrate affinity values found in the human enzymes (indicated by black asterisks) are actually optimal with respect to both achievable steady state NAD concentration and consumption fluxes.
-Thus, a further increase of the affinity of NamPRT for Nam or a change in the affinity of NNMT for Nam would not provide any advantage.
+Thus, a further increase of the affinity of NamPRT for Nam or a change in the affinity of NNMT for Nam would not provide any further  advantage.
 
@@
 structure visualisation in fig.~\ref{fig:unresolved_loop}B from \cite{Wang2006}).
-The loop is connected to one of the $\beta$-sheets involved in substrate binding.
+The loop is connected to one of the $\beta$-sheets involved in substrate binding \cite{PMID:19666527}.
 
 From these observations, we derived two possible hypotheses regarding the role of the loop in NamPRT function.
-The first hypothesis was that the presence of the loop changes the subcellular localisation of NamPRT, since the loop is part of an NLS.
-To test this hypothesis, we recombinatly expressed wildtype NamPRT and a mutated version lacking amino acids 42 to 51.
-Both, wildtype and mutant, were modified to carry a FLAG-tag for easier visualisation.
+The first hypothesis was that the presence of the loop changes the subcellular localisation of NamPRT, as it is overlapping with a predicted NLS.
+To test this hypothesis, we recombinatly expressed FLAG-tagged  wildtype and mutant NamPRT in Hela S3 cells.
 Immunofluorescence imaging showed a mixed cytosolic nuclear localisation for the wildtype and the mutant NamPRT (fig.~\ref{fig:unresolved_loop}C).
-The presence of the loop does not seem to have any effect on the subcellular localisation.
+Thus, the presence of the loop does not compromise the nuclear localisation.
 
@@
 The enzymatic activity of the mutant enzyme lacking the loop was much lower than the activity of the wildtype enzyme (fig.~\ref{fig:unresolved_loop}D).
-An increase of the substrate concentration had a much stronger effect on the activity of the mutant compared to the wildtype enzyme, indicating a lower affinity.
+Furthermore, an increase of the substrate concentration had a much stronger effect on the activity of the mutant compared to the wildtype enzyme, indicating a lower affinity.
 In contrast to wildtype NamPRT, the mutant does not exhibit an increased activity in the presence of ATP (fig.~\ref{fig:unresolved_loop}E).

### 2018-07-09 latex/results.tex
Resulst revision finished

@@
 
-A remaining puzzle was, why NADA is lost in vertebrates, while NADA and NamPRT both occur in bacteria.
-We tried to approach the puzzle with a competitive model with two compartments that share a Nam source.
-One compartment contains NADA while the other contains either NamPRT alone, or together with NNMT (for details, see experimental procedures and suppl.
+So far we have not discussed the effect of NADA on Nam-salvage and especially the question, why it is lost in vertebrates.
+As selection during evolution results usually from competition for resources, we built a two compartment model, based on the pathway model described above.
+One compartment contains NADA, while the other one contains either NamPRT alone, or together with NNMT.
+Both compartments share a limited  Nam source (for details, see experimental procedures and suppl.
 tab.~S2).
-Without NNMT, the compartment containing NADA shows slightly lower NAD consumption rates (fig.~\ref{fig:NNMT_comp_advantage}A), but is able to maintain much higher NAD concentrations at low cell division rates (fig.~\ref{fig:NNMT_comp_advantage}B).
+Without NNMT, the compartment containing NADA shows slightly lower NAD consumption rates (fig.~\ref{fig:NNMT_comp_advantage}A), but is able to maintain much higher NAD concentrations especially at low cell division rates (fig.~\ref{fig:NNMT_comp_advantage}B).
 At high cell division rates, steady state concentrations in both compartments are similar.
-These comparable concentrations might explain why in bacteria that often have relatively high growth rates, both systems co-exist.
+This might explain why in bacteria, that often have relatively high growth rates, both systems can co-exist.
 
 In the presence of NNMT, the NamPRT compartment has both higher NAD consumption rates and higher steady state NAD concentrations than the compartment containing NADA (fig.~\ref{fig:NNMT_comp_advantage}C and~D).
-The high NAD concentrations are dependent on a high affinity of NamPRT.
-When simulating a low affinity (high $K_{M}$) of NamPRT, the NADA compartment is able to maintain higher NAD concentrations although it still has a lower the NAD consumption flux.
-Taken together, the results suggest that the NADA pathway became less worthwhile when a higher affinity of NamPRT developed, possibly induced by NNMT.
+The higher NAD concentrations in the compartment containing NamPRT and NNMT can, however, only be maintained  if the affinity of NamPRT for Nam is high enough, corresponding to a low $K_{M}$.
+If the affinity of NamPRT is too low (high $K_{M}$), the NADA compartment is able to maintain higher NAD concentrations, but it still has a lower the NAD consumption flux.
+Taken together, the results suggest that the NADA pathway might have become obsolete upon emergence of the  high affinity NamPRT This in turn might have been induced by the appearance of NNMT.

### 2018-07-09 latex/introduction.tex
Inserted correct references and did some fixing

@@
 NAD metabolism represents one of the most critical links that connect cellular signal transduction and energy metabolism.
-Even though best known as cofactor for various redox-reactions, NAD is involved in a number of signalling processes that consume NAD$^{+}$ by cleaving the molecule to nicotinamide (Nam) and ADP-ribose \cite{Verdin2015}.
+Even though it is best known as cofactor for various redox-reactions, NAD is involved in a number of signalling processes that consume NAD$^{+}$ by cleaving the molecule to nicotinamide (Nam) and ADP-ribose \cite{Verdin2015}.
 These NAD-dependent signalling reactions include but are not limited to poly- and mono-ADP-ribosylation \cite{Butepage2015,DeVos2012}, NAD-dependent protein deacylation by sirtuins \cite{Osborne2016}, and the synthesis of calcium-mobilizing molecules such as cyclic ADP-ribose \cite{Lee2012}.
@@
 Therefore, we later refer to them also as NAD-consuming reactions.
-The resulting NAD turnover differs in a cell type specific manner and can lead to an NAD half-life as short as two hours \cite{Liu2018}.
+The resulting NAD turnover differs in a cell-type-specific manner and can lead to an NAD half-life as short as two hours \cite{Liu2018}.
 To maintain the NAD concentration at physiological levels, NAD biosynthesis needs to act at an equally rapid rate.
@@
 Two principal pathways exist that recycle Nam.
-Firstly, vertebrates use a direct two step pathway staring with the convertion of Nam into the mononucleotide NMN by the Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate.
+Firstly, vertebrates use a direct two step pathway starting with the conversion of Nam into the mononucleotide NMN by the Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate.
 The nearly complete recycling of Nam by NamPRT is achieved by an extraordinary high subtrate affinity to Nam, the $K_{M}$ being in the low nanomolar range \cite{Burgos2008}.
-This appears to be mediated by an ATP-dependent phosphorylation of a histidine residue in the catalytic core\todo{PMID:19666527}.
+This appears to be mediated by an ATP-dependent phosphorylation of a histidine residue in the catalytic core \cite{Burgos2009}.
 Despite the importance of its salvage, Nam can also be marked for excretion by methylation.
@@
 While NamPRT is seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, NNMT seems to have no metabolic function other than to remove Nam from NAD metabolism.
-Although it has been suggested that the process potentially acts as a metabolic methylation sink \cite{PMID:28291578} .
+It has been suggested that the process potentially acts as a metabolic methylation sink \cite{Pissios2017}.
 
-Secondly, in many prokayotes as well as in plants and fungi a four steps starting with the deamidation of Nam to nicotinic acid (NA) by NADA is used.
+Secondly, in many prokayotes as well as in plants and fungi, a pathway consisting of four steps starting with the deamidation of Nam to nicotinic acid (NA) by the Nam deamidase (NADA) is used.
 (\ref{fig:pathway_overview}).
 The three enzymes that act after NADA belong to the Preiss-Handler pathway that also exists in vertebrates.
-The Preiss-Handler pathway first converts  NA into the corresponding mononucleotide (NAMN), in a reactions performed by the NAD specific phosphoribosyl transferase NAPRT.
-The convertion of the mononucleotides  NMN and NAMN into their dinucleotides NAD and NaAD is performed by the same enzyme, the  Nam/NA adenylytransferases (NMNATs) that is essential in all Organisms\cite{PMID:21729004}.
+NA is converted into the corresponding mononucleotide (NAMN), in a reactions performed by the NA-specific phosphoribosyltransferase NAPRT.
+The convertion of the mononucleotides NMN and NAMN into their dinucleotides NAD and NAAD is performed by the same enzyme, the Nam/NA adenylytransferases (NMNATs) that is essential in all organisms \cite{DeFigueiredo2011}.
 The recycling pathway via NA finally requires re-amidation of the dinucleotide, which is catalysed by NAD synthase.
-This reaction includes an enzyme adenylation step that consumes ATP.
-Therefore the Nam recycling by NADA is energetically less efficient under normal conditions.
+This final reaction includes an enzyme adenylation step that consumes ATP.
+Therefore the Nam recycling by NADA is energetically less efficient under normal conditions than the recycling pathway starting with NamPRT.
 
 We here present a phylogenetic analysis of the NAD pathways.
-Our results suggest that there has been a selection for the co-existence of NamPRT and NNMT in Deuterostomia, while the deamidation pathway, which is dominant in bacteria, is lost.
+Our results suggest that there has been a selection for the co-existence of NamPRT and NNMT in deuterostomes, while the deamidation pathway, which is dominant in bacteria, is lost.
 This transition was accompanied by a marked increase in the number of NAD-consuming signalling enzymes.

### 2018-07-09 latex/results.tex
Inserted correct references and did some fixing

--- 2018-07-09 latex/results.tex
@@
 \subsection{Paradoxical evolutionary correlation between NAD-dependent signalling and precursor metabolism}
-We have  previously shown  that NMNAT is essential for NAD dependent biosynthesis independent of the biosynthetic route or precursor \cite{PMID:21729004}.
-In addition, the Preiss-Handler pathway was found  present in virtually all organisms included in an earlier phylogenetic analysis\cite{Gossmann2012FEBS}.
-Thus, we here focused on the phylogenetic analysis of the two enzymes that initiate the two different NAD-salvage pathways, NADA and NamPRT.
-These show a scattered distribution in bacteria \todo{MathiasB: please find and insert citation} and have been shown to occur concomitantly in some marine invertebrates \cite{Gossmann2012FEBS}.
-In addition, we included  NAM methyltransferase NNMT into our analysis, as its function and origin is unclear.
-As can be seen from Fig.
-(fig.~\ref{fig:phylo_distribution}A, bacteria, fungi and plants predominantly encode NADA and only a few of them harbour NamPRT.
-In contrast, Metazoa predominantly have NamPRT together with NNMT.
-Analysing the phylogeny distribution in Metazoa in more detail revealed that not only is NamPRT dominating and NADA appears to be lost in vertebrates, in most organisms NamPRT is found together with NNMT (fig.~\ref{fig:phylo_distribution}B).
-NNMT seems to have arisen \textit{de~novo} in the common ancestor of Ecdysozoa and Lophotrochozoa.
-As we were unable to find any \textit{NNMT} gene with an e-value below ....
-\todo{Mathias B.
+
+We have previously shown that NMNAT is essential for NAD biosynthesis independent of the biosynthetic precursor or pathway \cite{DeFigueiredo2011}.
+In addition, the Preiss-Handler pathway was found  present in virtually all organisms included in an earlier phylogenetic analysis \cite{Gossmann2012FEBS}.
+Thus, we here focused on the phylogenetic analysis of the two enzymes that initiate the two different NAD salvage pathways, NADA and NamPRT.
+The phylogenetic distribution of the two enzymes is scattered in bacteria \todo{MathiasB: please find and insert citation} and have been shown to occur together in some marine invertebrates \cite{Gossmann2012FEBS}.
+In addition, we included the Nam methyltransferase NNMT into our analysis, as its function and origin is unclear.
+
+As shown in Figure~\ref{fig:phylo_distribution}A, bacteria, fungi, and plants predominantly possess NADA and only a few of them harbour NamPRT.
+In contrast, Metazoa predominantly lost NADA amd have NamPRT together with NNMT.
+NNMT seems to have arisen \textit{de~novo} in the most recent common ancestor of Ecdysozoa and Lophotrochozoa (fig.~\ref{fig:phylo_distribution}B).
+As we were unable to find any \textit{NNMT} gene with an e-value below xxx \todo{Mathias B.
 include e-value} in fungi or plants.
-The gene with the highest sequence similarity in S.
-cerevisiae is  .....
-with proven function .....\todo{Mathias B: include e-value}
+The gene with the highest sequence similarity in \textit{S.~cerevisiae} is ...
+with proven function ...
+\todo{Mathias B: include values}.
 
 Nematodes are the only organisms, where we observed a concomitant presence of NADA and NNMT.
-In Deuterostomes, the only large clade that possesses only NamPRT and seems to have lost NNMT are Sauropsida, and among them especially birds.
-The reason why about half of the sequenced bird genomes do not seem to encode for  NNMT,  remains unclear.
+In deuterostomes, the only large clade that possesses only NamPRT and seems to have lost NNMT are Sauropsida, and among them especially birds.
+The reason why about half of the sequenced bird genomes do not seem to encode for NNMT remains unclear.
 The distribution of NNMT in birds is quite scattered (suppl.
 fig.~S2).
-not corresponding to current species trees \cite{http://dx.doi.org/10.1038/nature15697} .
-The lack of NNMT might be related to the excretion system, as the product of NNMT, methyl-Nam, is in mammals excreted with the urine.
-There few Metazoan species for which we could not find NamPRT or NADA, while NNMT was present.
+\todo{figure not corresponding to current species trees \cite{Prum2015}} The lack of NNMT might be related to the excretion system, as the product of NNMT, methyl-Nam, is in mammals excreted with the urine.
+There are few metazoan species for which we could not find NamPRT or NADA, while NNMT was present.
 We assume that this is due to incomplete genomes in the database, as the distribution of such species is scarce and widely scattered.
@@
 The numbers shown in figure~\ref{fig:phylo_distribution}B denote the average number of NAD-dependent signalling enzyme families found in each clade.
-With the exception of Cnidaria and Lophotrochozoa, we find an average of three to four families in Protostomes, whereas most Deuterostome species have, on average, more than eight families with an increasing diversification of enzymes within some of these families \cite{Gossmann2014DNAR}.
+With the exception of Cnidaria and Lophotrochozoa, we find an average of three to four families in protostomes, whereas most deuterostome species have, on average, more than eight families with an increasing diversification of enzymes within some of these families \cite{Gossmann2014DNAR}.
 
@@
 This observation seems counterintuitive, as one would expect that increased NAD-dependent signalling should be accompanied by an increase of substrate availability for NAD biosynthesis.
-However, NNMT irreversibly removes Nam from salvage and thus is assumed to potentially decrease the cellular NAD-concentration.
 
@@
 
-To be able to compare metabolic features of evolutionary quite different systems in our simulations and as we have limited information about expression levels of enzymes, we initially used equal amounts for all enzymes.
+To be able to compare metabolic features of evolutionary quite different systems in our simulations and as we had limited information about expression levels of enzymes, we initially used equal amounts for all enzymes.
 As we have very few cross species kinetic data, we were furthermore mainly relying on kinetic constants found for human or yeast enzymes.
@@
 We calculated steady state NAD concentrations and NAD consumption rates by simulating NAD biosynthesis proceeding via NamPRT in the presence or absence of NNMT.
-To achieve free NAD concentration in the range reported in the literature and due to the very low turnover of NamPRT, we had to use 10 fold higher NamPRT levels compared to other enzymes and adjusted the amount of NMNAT to the same level to avoid that the NAD synthesis rates are limited by this enzyme.
+To achieve free NAD concentration in the range reported in the literature and due to the very low turnover of NamPRT, we had to use tenfold higher NamPRT levels compared to other enzymes and adjusted the amount of NMNAT to the same level to avoid that the NAD synthesis rates are limited by this enzyme.
 As shown in figure~\ref{fig:NNMT_NAD_flux}, the presence of NNMT enables higher NAD consumption rates (fig.~\ref{fig:NNMT_NAD_flux}A) while reducing the steady state concentration of NAD (fig.~\ref{fig:NNMT_NAD_flux}B).
@@
 Thus the presence of NNMT enables higher NAD consumption fluxes, by removing excess Nam from the cells.
-At the same time, the high substrate affinity of NamPRT maintains a sufficiently high NAD concentration, although the concentration is of course lower than in the system without NNMT.
+At the same time, the high substrate affinity of NamPRT maintains a sufficiently high NAD concentration, although the concentration is, as expected, lower than in the system without NNMT.
 
@@
 
-Figure~\ref{fig:NamPRT_affinity_Nam}E and~F show a direct comparison of different affinities of NamPRT for Nam with and without NNMT.
-At an affinity of $K_{M}$ = 1\,$\mu$M, which is in the range of the $K_{M}$ of NADA for Nam, NAD consumption flux is only higher with NNMT than without at low cell division rates (fig.~\ref{fig:NamPRT_affinity_Nam}E).
+Figures~\ref{fig:NamPRT_affinity_Nam}E and~F show a direct comparison of different affinities of NamPRT for Nam with and without NNMT.
+At an affinity of $K_{M}$ =~1\,$\mu$M, which is in the range of the $K_{M}$ of NADA for Nam, NAD consumption flux is only higher with NNMT than without at low cell division rates (fig.~\ref{fig:NamPRT_affinity_Nam}E).
 If the affinity of NamPRT is high enough, consumption rates are always higher with NNMT than without it.
@@
 As pathway dynamics are not solely dependent on one enzyme, we wanted to estimate the impact of the substrate affinity of NNMT that is competing with NamPRT for the same substrate.
-In figure~\ref{fig:optimal_substr_affinities} is shown that the substrate affinity values found in the human enzymes (indicated by black asterisks) are actually optimal with respect to both achievable steady state NAD concentration and consumption fluxes.
+The substrate affinity values found in the human enzymes (indicated by black asterisks in fig.~\ref{fig:optimal_substr_affinities}) are actually optimal with respect to both achievable steady state NAD concentration and consumption fluxes.
 Thus, a further increase of the affinity of NamPRT for Nam or a change in the affinity of NNMT for Nam would not provide any further advantage.
@@
 structure visualisation in fig.~\ref{fig:unresolved_loop}B from \cite{Wang2006}).
-The loop is connected to one of the $\beta$-sheets involved in substrate binding \cite{PMID:19666527}.
+The loop is connected to one of the $\beta$-sheets involved in substrate binding \cite{Burgos2009}.
 
@@
 The first hypothesis was that the presence of the loop changes the subcellular localisation of NamPRT, as it is overlapping with a predicted NLS.
-To test this hypothesis, we recombinatly expressed FLAG-tagged  wildtype and mutant NamPRT in Hela S3 cells.
+To test this hypothesis, we created a mutant NamPRT lacking the loop and recombinatly expressed FLAG-tagged wildtype and mutant NamPRT in Hela S3 cells.
 Immunofluorescence imaging showed a mixed cytosolic nuclear localisation for the wildtype and the mutant NamPRT (fig.~\ref{fig:unresolved_loop}C).
@@
 Furthermore, an increase of the substrate concentration had a much stronger effect on the activity of the mutant compared to the wildtype enzyme, indicating a lower affinity.
-In contrast to wildtype NamPRT, the mutant does not exhibit an increased activity in the presence of ATP (fig.~\ref{fig:unresolved_loop}E).
+In contrast to wildtype NamPRT, the mutant did not exhibit an increased activity in the presence of ATP (fig.~\ref{fig:unresolved_loop}E).
 

### 2018-07-09 latex/results.tex
Fixed some things in results

@@
 NNMT seems to have arisen \textit{de~novo} in the most recent common ancestor of Ecdysozoa and Lophotrochozoa (fig.~\ref{fig:phylo_distribution}B).
-As we were unable to find any \textit{NNMT} gene with an e-value below xxx \todo{Mathias B.
-include e-value} in fungi or plants.
+As we were unable to find any \textit{NNMT} gene with an e-value below 0.1 in fungi or plants.
 The gene with the highest sequence similarity in \textit{S.~cerevisiae} is ...
 with proven function ...
-\todo{Mathias B: include values}.
+\todo{Mathias B: include values} \todo{Answer from MB: There was not a single hit outside Metazoa}.
 
@@
 
-So far we have not discussed the effect of NADA on Nam-salvage and especially the question, why it is lost in vertebrates.
+So far, we did not discuss the effect of NADA on Nam salvage and especially the question, why it is lost in vertebrates.
 As selection during evolution results usually from competition for resources, we built a two compartment model, based on the pathway model described above.
-One compartment contains NADA, while the other one contains either NamPRT alone, or together with NNMT.
+One compartment contains NADA, while the other one contains either NamPRT alone or together with NNMT.
 Both compartments share a limited Nam source (for details, see experimental procedures and suppl.
@@
 In the presence of NNMT, the NamPRT compartment has both higher NAD consumption rates and higher steady state NAD concentrations than the compartment containing NADA (fig.~\ref{fig:NNMT_comp_advantage}C and~D).
-The higher NAD concentrations in the compartment containing NamPRT and NNMT can, however, only be maintained  if the affinity of NamPRT for Nam is high enough, corresponding to a low $K_{M}$.
-If the affinity of NamPRT is too low (high $K_{M}$), the NADA compartment is able to maintain higher NAD concentrations, but it still has a lower the NAD consumption flux.
-Taken together, the results suggest that the NADA pathway might have become obsolete upon emergence of the  high affinity NamPRT This in turn might have been induced by the appearance of NNMT.
+The higher NAD concentrations in the compartment containing NamPRT and NNMT can, however, only be maintained if the affinity of NamPRT for Nam is high enough.
+If the substrate affinity of NamPRT is too low (high $K_{M}$), the NADA compartment is able to maintain higher NAD concentrations, but still has a lower NAD consumption flux.
+Taken together, the results suggest that the NADA pathway might have become obsolete upon emergence of the high affinity NamPRT.
+This in turn might have been induced by the appearance of NNMT.

### 2018-07-10 latex/discussion.tex
revised introduction and figure legends, still working on results
new discussion

@@
 
+We here analysed the phylogenetic distribution of the three enzymes using Nam as a substrate, the two NAd-salvage pathway enzymes NADA and NamPRT as well as the Nam degrading enzyme NNMT.
+We found that after the first appearance of NNMT in Protostomia, a diversification of NAD-consuming reactions in Deuterostomia can be observed.
+Using mathematical modelling, we can explain these finding, as NNMT removes excess of Nam in cells and by that reduces product inhibition of NAD-signalling enzymes, enabeling higher fluxes through these reactions.
+This in turn seem to have enabled a diversification and NAD-consuming enzymes in mammals are involved in a wide variety of signalling and gene regulatory mechanisms that due to there sensitivity to NAD+, have the ability to translate differences in metabolic  states into changes in signalling and gene regulation.
+As NAD concentrations are consequently lowered, a high affinity enzyme for recycling of Nam is required to ensure maintenance for cellular NAD-levels for other metabolic processes.
+Whether it was by chance or whether an this is due to mechanistic limitations that NamPRT developed a high affinity and not NADA, or whether other environmental conditions have been responsible is impossible to deduce.
+But our simulations clearly show that a high affinity of NamPRT is favourable and that it therefore appears that NNMT might have been driving NamPRT evolution.
+Looking at the enzyme affinities of the human enzymes it furthermore appears  that both NNMT and NamPRT reached an near optimal state, as further increase of the affinity of NamPRT and decrease of the affinity of NNMT would not result in much higher steady state NAD-concentrations or NAD-consumption fluxes, according to our simulations.
+It furthermore appears that NNMT makes both NAD concentration and NAD consumption relatively independent of other processes requiring NAD, such as cell growth.
+In addition.
+
+Our findings shed also new light on the potential physiological role of NNMT, which has earlier been recognized as potential marker for some types of cancer (e.g.
+\cite{Okamura1998}).
+It might well be that increased NNMT expression in cancer might serve to remove Nam produced by increased NAD-dependent signalling.
+To maintain high NAD-concentration a simultaneous overexpression of NamPRT is required, which is what has been found in some types of cancer (ref).
+As NNMT does, however, provide an advantage as long as NamPRT affinity is sufficiently high, this might provide an useful avenue for treatment, especially for types of cancer that express NNMT at a high lever.
+These would be more susceptible to competitive inhibitors of NamPRT, and several of those inhibitors are currently tested in clinical studies.
+The major healthy tissue expressing NNMT is the liver, while most other tissues do not express NNMT or only at vary low levels.
+Based on our analysis we would furthermore suggest, that it might be reasonable to screen patients before treatment, as non NNMT expressing tumours might respond less to competitive NamPRT inhibitors.
+This is furthermore relevant, as the missing degradation in those cells would potentially lead to an accumulating of Nam that could outcompete the inhibitor in those cells.
+But these aspects require further investigation.
+
+Our combine phylogenetic-modelling analysis furthermore provides a potential explanation both for the co-occurance of NADA and NamPRT in bacteria, but even more strongly for the loss of NADA in vertebrates.
+We here show that when  simulating compartments that share the same limited Nam-source, the compartment  that contains NamPRT and NNMT has a higher steady state NAD concentration and NAD consumption rate as the compartment containing NADA.
+The combination of a high affinity NamPRT with NNMT which is  the dominant enzyme combination found in vertebrates seems to provide a competitive advantages.
+As this may also hold for mammalian associated bacteria, especially pathogens, we wanted to see whether pathogenic bacteria solely express NamPRT.
+Unfortunately, bacterial habitat information is currently not complete and often difficult to access especially at the NCBI database that was used for the analysis provided here, as it is the most comprehensive sequence database to date.
+We therefore manually looked through the bacteria found to encode for NamPRT and indeed found that many but not all of them have been characterised to be pathogenic.
+It should be noted that the distribution of NADA and NamPRT does not follow the bacterial species tree\todo{insert ref}.
+
+A detailed analysis of sequence variances in NamPRT revealed that only Deuterostomia that have NNMT but not NADA, have a sequence insertion in the N-terminal part of NamPRT that according to our experimental analysis seems to enable the high affinity of the enzyme.
+This in turn would suggest that also the bacterial enzymes doe not have a high substrate affinity, a finding that could potentially be exploited for antibiotics development.
+
+We completely neglected the potential effects of cosubstrates of the pathway investigated such as the presence of targets of the NAD-consuming enzymes, such as acetylated proteins for sirtuins for example, or the requirement of PRPP\todo{insert full name and but PRPP in brackets} and ATP that are rrequired for NMN synthesis by NamPRT  or the presence of methyl donor \textit{S}-adenosyl methionine (SAM) and its precursor methionine that have been shown to potentially limit the effect of NNMT\cite{PMID:23455543 }.
+These should be included in future analysis.
+The challenge is, however, that currently information about the in vivo concentration of those co-substrates are very limited.
+We assume that this will improve in the future as new, more sensitive, methods for metabolite measurements are currently developed.
+
+During our analysis we came across several problems related to current sequence database.
+One is sequence contamination, which is a well known problem\todo{insert ref} and we therefore tried to remove all sequences of obvious bacterial origin from the analysis, based on sequence homology analysis.
+Another problem are incomplete genomes ....
+\todo{Mathias B.
+please insert here the paragraph you have written for your thesis--- you should probably write it in other words}
+
+The third problem are wrong annotations.
+We tried to avoid these, by, wherever possible, only relying on template sequences with confirmed function.
+This problem becomes apparent by the fact that in yeast an enzyme named NNMT can be found, which is based on an initial analysis on life span extension in Saccharomyces cerevisae\todo{insert ref} a sequence similarity that is restricted to the SAM-dependent methyl transferase domain.
+The protein has, however, later been shown not function as Nam or small molecule methyl transferase but as ...\todo{Please complete incl.
+refs.}
+
+Nevertheless, we have been able to comprehensively analyse co-evolution of several enzymes of the NAD pathway with the appearance of NNMT seemingly initiating and driving complex alterations of the pathway such as an increase and diversification of NAD-dependent signalling, followed by an increase in NamPRT substrate affinity (schematic overview see fig.~\ref{fig:evo_events}).
+This again appears to be accompanied by the loss of NADA in vertebrates and the first gene duplication of NMNATs \cite{Lau2010}.
+We also noted that the second gene duplication of NMNATs and thus the further compartmentalisation of NAD metabolism is co-occurring with a site-specific positive selection event in NNMT (unpublished results).
+To our knowledge, this is the first study that comprehensively demonstrates a interdependency of enzyme evolution based on functional interaction of several enzymes of a pathway.
+While it is of course well know and described for directly interaction proteins \todo{Not sure that is true, one that might come close is the following work https://academic.oup.com/mbe/article/33/1/268/2579532.....please check and potentially revise statement}.
+
 \subsection{Points to discuss}
-
-+ NNMT evolution.
-
--- Why was the yeast MT defined as NNMT?
-
--- Are there any ideas where the NNMT came from?
-
--- Any information on the methylation activity (also in comparison to others)? Does it fit to the found kinetic functions?
 
@@
 -- NAPRT needed for NA that is taken up with the diet (and if NA is processed, the other two enzymes are also needed to continue processing)
-
-\subsection{Old discussion}
-
-The dynamic interactions among metabolism, signal transduction, and gene regulation are still not very well understood.
-New approaches are required to disentangle the underlying network, helping us to understand how alterations affect human physiology.
-We here combined phylogenetic analysis, mathematical modelling, and experimental evidence to better understand the evolution of the NAD biosynthesis and consumption pathways and especially the role of the methyltransferase NNMT.
-Our analyses reveal that NNMT plays a vital role for the diversification of NAD-dependent signalling reactions and potentially also for NAD homeostasis.
-We show that the combination of NNMT and NamPRT enables higher NAD consumption rates in comparison to NamPRT or NADA alone.
-It furthermore appears that NNMT makes both NAD concentration and NAD consumption relatively independent of other processes requiring NAD such as cell growth.
-In addition, we suggest that NNMT exerted an evolutionary pressure on NamPRT, driving the development of the high affinity of NamPRT for its substrates found in the human enzyme \cite{Burgos2008}.
-These findings shed new light on the role of NNMT, which has earlier been recognized as potential marker for some types of cancer (e.g.
-\cite{Okamura1998}).
-Increased NNMT expression in cancer might serve to remove Nam produced by increased NAD-dependent signalling.
-Further analysis is required to better understand which NAD-dependent signalling reactions are affected.
-
-In our analysis, we furthermore noted that NNMT only provides an advantage as long as the NamPRT affinity for Nam is sufficiently high.
-This implies that treatment with competitive inhibitors that affect the apparent $K_{M}$ would have stronger effects on cells that express NNMT compared to cells that do not express NNMT (see fig.~\ref{fig:NamPRT_affinity_Nam}E and~F).
-This could be important for cancer treatments as the expression of NNMT is relatively low in normal tissues except liver, whereas several types of cancer overexpress NNMT\todo{Needs a reference}.
-
-We neglected the potential effects of the methyl donor \textit{S}-adenosyl methionine (SAM) and its precursor methionine in our models.
-Changing SAM availability is most likely contributing yet another regulatory level for Nam availability and thus an additional interaction point between gene regulation and metabolism.
-It has been shown earlier that NNMT expression influences protein methylation dependent on methionine availability \cite{Ulanovskaya2013}.
-One of the challenges in the analysis of this aspect is the availability of \textit{in~vivo} concentration measurements for SAM and the large amount of reactions using it as substrate.
-A simplified mathematical model of this pathway has been published earlier \cite{Reed2004} providing the basis for future studies.
-
-Beside gaining new insights into the role of NNMT for NAD metabolism, our analyses provided a suggestion why NADA disappears in vertebrates while it co-exists with NamPRT in bacteria.
-It seems that the combined enzymatic activity of NNMT and NamPRT has been optimised in vertebrates to maintain stable NAD concentration and high NAD consumption flux, making NADA obsolete.
-
-\section{Conclusions}
-
-We here highlight the co-evolution of several enzymes of the NAD pathway with the appearance of NNMT seemingly initiating and driving complex alterations of the pathway such as an increase and diversification of NAD-dependent signalling, followed by an increase in NamPRT substrate affinity (schematic overview see fig.~\ref{fig:evo_events}).
-This again appears to be accompanied by the loss of NADA in vertebrates and the first gene duplication of NMNATs \cite{Lau2010}.
-We also noted that the second gene duplication of NMNATs and thus the further compartmentalisation of NAD metabolism is co-occurring with a site-specific positive selection event in NNMT (unpublished results).
-To our knowledge, this is the first study that comprehensively demonstrates a functional co-evolution of several enzymes of a pathway.

### 2018-07-10 latex/discussion.tex
edited new discussion

@@
 A detailed analysis of sequence variances in NamPRT revealed that only Deuterostomia that have NNMT but not NADA, have a sequence insertion in the N-terminal part of NamPRT that according to our experimental analysis seems to enable the high affinity of the enzyme.
-This in turn would suggest that also the bacterial enzymes doe not have a high substrate affinity, a finding that could potentially be exploited for antibiotics development.
+This in turn would suggest that also the bacterial enzymes doe not have a high substrate affinity and although the one bacterial NamPRT has been proven to be enzymatically active, the affinity of a bacterial enzyme has not been measured so far.
+The differences in activity and affinity, implying differences in substrate binding could potentially be exploited for the development of  antibiotics.
+But further analysis as well as potentially the crystallisation of a bacterial enzyme would be required, to see whether this could be a  approach.
 
-We completely neglected the potential effects of cosubstrates of the pathway investigated such as the presence of targets of the NAD-consuming enzymes, such as acetylated proteins for sirtuins for example, or the requirement of PRPP\todo{insert full name and but PRPP in brackets} and ATP that are rrequired for NMN synthesis by NamPRT  or the presence of methyl donor \textit{S}-adenosyl methionine (SAM) and its precursor methionine that have been shown to potentially limit the effect of NNMT\cite{PMID:23455543 }.
+In our analysis, we so far completely neglected the potential effects of cosubstrates of the pathway investigated such as the presence of targets of the NAD-consuming enzymes, such as acetylated proteins for sirtuins for example, or the requirement of PRPP\todo{insert full name and but PRPP in brackets} and ATP that are rrequired for NMN synthesis by NamPRT  or the presence of methyl donor \textit{S}-adenosyl methionine (SAM) and its precursor methionine that have been shown to potentially limit the effect of NNMT\cite{PMID:23455543 }.
 These should be included in future analysis.
@@
 While it is of course well know and described for directly interaction proteins \todo{Not sure that is true, one that might come close is the following work https://academic.oup.com/mbe/article/33/1/268/2579532.....please check and potentially revise statement}.
-
-\subsection{Points to discuss}
-
-+ NADA disappears but rest of pathway stays. Why?
-
--- NADS needed for \textit{de~novo} pathway
-
--- NMNAT also needed in NamPRT and \textit{de~novo} pathway
-
----- Is the \textit{de~novo} pathway still necessary? If we take up enough Vit B3 with the diet, why \textit{de~novo}?
-
--- NAPRT needed for NA that is taken up with the diet (and if NA is processed, the other two enzymes are also needed to continue processing)

### 2018-07-10 latex/discussion.tex
edited new discussion

@@
 
-We here analysed the phylogenetic distribution of the three enzymes using Nam as a substrate, the two NAd-salvage pathway enzymes NADA and NamPRT as well as the Nam degrading enzyme NNMT.
+We here omprehensively analysed the phylogenetic distribution of the three enzymes using Nam as a substrate, the two NAD-salvage pathway enzymes NADA and NamPRT as well as the Nam degrading enzyme NNMT.
 We found that after the first appearance of NNMT in Protostomia, a diversification of NAD-consuming reactions in Deuterostomia can be observed.
-Using mathematical modelling, we can explain these finding, as NNMT removes excess of Nam in cells and by that reduces product inhibition of NAD-signalling enzymes, enabeling higher fluxes through these reactions.
-This in turn seem to have enabled a diversification and NAD-consuming enzymes in mammals are involved in a wide variety of signalling and gene regulatory mechanisms that due to there sensitivity to NAD+, have the ability to translate differences in metabolic  states into changes in signalling and gene regulation.
-As NAD concentrations are consequently lowered, a high affinity enzyme for recycling of Nam is required to ensure maintenance for cellular NAD-levels for other metabolic processes.
-Whether it was by chance or whether an this is due to mechanistic limitations that NamPRT developed a high affinity and not NADA, or whether other environmental conditions have been responsible is impossible to deduce.
-But our simulations clearly show that a high affinity of NamPRT is favourable and that it therefore appears that NNMT might have been driving NamPRT evolution.
+By using mathematical modelling, we can explain these finding, as NNMT removes excess of Nam from cells and by that reduces product inhibition of NAD-signalling enzymes.
+This in turn enables higher fluxes through these reactions.
+Thus the diversification of  NAD-consuming enzymes in mammals semms to have been enabled by the presence of NNMT.
+NAD-consuming enzymes are involved in a wide variety of signalling and gene regulatory mechanisms that due to their sensitivity to NAD+, have the ability to translate differences in metabolic  states into changes in signalling and gene regulation.
+As through the removal of NAD pre-cursor NNMT, NAD concentrations are lowered, a high affinity enzyme for recycling of Nam is required to ensure maintenance for cellular NAD-levels for other metabolic processes.
+Whether it was by chance or whether this is due to thermodynamic or mechanistic limitations that NamPRT developed a high affinity and not NADA, or whether other environmental conditions have been responsible for the evolutionary trajectory observed, is impossible to deduce based on the chosen approach.
+But our simulations clearly show that a high affinity of NamPRT is required for high NAD-consumption fluxes and NAD-concentrations and that it therefore appears that NNMT might have been driving NamPRT evolution.
 Looking at the enzyme affinities of the human enzymes it furthermore appears  that both NNMT and NamPRT reached an near optimal state, as further increase of the affinity of NamPRT and decrease of the affinity of NNMT would not result in much higher steady state NAD-concentrations or NAD-consumption fluxes, according to our simulations.

### 2018-07-10 latex/discussion.tex
Resolved many todos and typos in discussion.

@@
 
-We here omprehensively analysed the phylogenetic distribution of the three enzymes using Nam as a substrate, the two NAD-salvage pathway enzymes NADA and NamPRT as well as the Nam degrading enzyme NNMT.
+We here comprehensively analysed the phylogenetic distribution of the three enzymes using Nam as a substrate that are the two NAD salvage pathway enzymes NADA and NamPRT as well as the Nam-degrading enzyme NNMT.
 We found that after the first appearance of NNMT in Protostomia, a diversification of NAD-consuming reactions in Deuterostomia can be observed.
-By using mathematical modelling, we can explain these finding, as NNMT removes excess of Nam from cells and by that reduces product inhibition of NAD-signalling enzymes.
+By using mathematical modelling, we could explain these finding, as NNMT removes excess Nam from cells and thereby reduces product inhibition of NAD signalling enzymes.
 This in turn enables higher fluxes through these reactions.
-Thus the diversification of  NAD-consuming enzymes in mammals semms to have been enabled by the presence of NNMT.
-NAD-consuming enzymes are involved in a wide variety of signalling and gene regulatory mechanisms that due to their sensitivity to NAD+, have the ability to translate differences in metabolic  states into changes in signalling and gene regulation.
-As through the removal of NAD pre-cursor NNMT, NAD concentrations are lowered, a high affinity enzyme for recycling of Nam is required to ensure maintenance for cellular NAD-levels for other metabolic processes.
+Thus, the diversification of NAD-consuming enzymes in mammals seems to have been enabled by the presence of NNMT.
+NAD-consuming enzymes are involved in a wide variety of signalling and gene regulatory mechanisms that due to their sensitivity to NAD$^{+}$, have the ability to translate differences in metabolic states into changes in signalling and gene regulation.
+As NAD concentrations are lowered by the removal of NAD precursor by NNMT, a high-affinity enzyme for recycling of Nam is required to ensure maintenance for cellular NAD levels for other metabolic processes.
 Whether it was by chance or whether this is due to thermodynamic or mechanistic limitations that NamPRT developed a high affinity and not NADA, or whether other environmental conditions have been responsible for the evolutionary trajectory observed, is impossible to deduce based on the chosen approach.
-But our simulations clearly show that a high affinity of NamPRT is required for high NAD-consumption fluxes and NAD-concentrations and that it therefore appears that NNMT might have been driving NamPRT evolution.
-Looking at the enzyme affinities of the human enzymes it furthermore appears  that both NNMT and NamPRT reached an near optimal state, as further increase of the affinity of NamPRT and decrease of the affinity of NNMT would not result in much higher steady state NAD-concentrations or NAD-consumption fluxes, according to our simulations.
+But our simulations clearly show that a high affinity of NamPRT is required for high NAD consumption fluxes and NAD concentrations and that it therefore appears that NNMT might have been driving NamPRT evolution.
+Looking at the enzyme affinities of the human enzymes it furthermore appears that both NNMT and NamPRT reached an near optimal state, as further increase of the affinity of NamPRT and decrease of the affinity of NNMT would not result in much higher steady state NAD concentrations or NAD consumption fluxes, according to our simulations.
 It furthermore appears that NNMT makes both NAD concentration and NAD consumption relatively independent of other processes requiring NAD, such as cell growth.
-In addition.
 
@@
 It might well be that increased NNMT expression in cancer might serve to remove Nam produced by increased NAD-dependent signalling.
-To maintain high NAD-concentration a simultaneous overexpression of NamPRT is required, which is what has been found in some types of cancer (ref).
-As NNMT does, however, provide an advantage as long as NamPRT affinity is sufficiently high, this might provide an useful avenue for treatment, especially for types of cancer that express NNMT at a high lever.
-These would be more susceptible to competitive inhibitors of NamPRT, and several of those inhibitors are currently tested in clinical studies.
-The major healthy tissue expressing NNMT is the liver, while most other tissues do not express NNMT or only at vary low levels.
-Based on our analysis we would furthermore suggest, that it might be reasonable to screen patients before treatment, as non NNMT expressing tumours might respond less to competitive NamPRT inhibitors.
+To maintain high NAD concentrations a simultaneous overexpression of NamPRT is required, which is what has been found in some types of cancer \cite{Bi2011,Wang2011}.
+As NNMT does, however, provide an advantage as long as NamPRT affinity is sufficiently high, this might provide an useful avenue for treatment, especially for types of cancer that express NNMT at a high level.
+These would be more susceptible to competitive inhibitors of NamPRT, and several of those inhibitors are currently tested in clinical studies \todo{ref.
+Maybe PMID:28733523 \cite{Espindola-Netto2017}?}.
+The major healthy tissue expressing NNMT is the liver, while most other tissues do not express NNMT or only at very low levels.
+Based on our analysis, we would furthermore suggest that it might be reasonable to screen patients before treatment, as non-NNMT expressing tumours might respond less to competitive NamPRT inhibitors.
 This is furthermore relevant, as the missing degradation in those cells would potentially lead to an accumulating of Nam that could outcompete the inhibitor in those cells.
@@
 
-Our combine phylogenetic-modelling analysis furthermore provides a potential explanation both for the co-occurance of NADA and NamPRT in bacteria, but even more strongly for the loss of NADA in vertebrates.
-We here show that when  simulating compartments that share the same limited Nam-source, the compartment  that contains NamPRT and NNMT has a higher steady state NAD concentration and NAD consumption rate as the compartment containing NADA.
-The combination of a high affinity NamPRT with NNMT which is  the dominant enzyme combination found in vertebrates seems to provide a competitive advantages.
-As this may also hold for mammalian associated bacteria, especially pathogens, we wanted to see whether pathogenic bacteria solely express NamPRT.
-Unfortunately, bacterial habitat information is currently not complete and often difficult to access especially at the NCBI database that was used for the analysis provided here, as it is the most comprehensive sequence database to date.
+Our combined phylogenetic/modelling analysis furthermore provides a potential explanation both for the co-occurrence of NADA and NamPRT in bacteria, but even more strongly for the loss of NADA in vertebrates.
+We here show that when simulating compartments that share the same limited source of Nam, the compartment that contains NamPRT and NNMT has a higher steady state NAD concentration and NAD consumption rate as the compartment containing NADA.
+The combination of a high affinity NamPRT with NNMT, which is the dominant enzyme combination found in vertebrates, seems to provide a competitive advantage.
+As this may also hold for mammalian-associated bacteria, particularly pathogens, we wanted to see whether pathogenic bacteria solely express NamPRT.
+Unfortunately, bacterial habitat information is currently far from complete and often difficult to access especially at the NCBI database that was used for the analysis provided here, as it is the most comprehensive sequence database to date.
 We therefore manually looked through the bacteria found to encode for NamPRT and indeed found that many but not all of them have been characterised to be pathogenic.
-It should be noted that the distribution of NADA and NamPRT does not follow the bacterial species tree\todo{insert ref}.
+It should be noted that the distribution of NADA and NamPRT does not follow the bacterial species tree\todo{ref.
+Maybe PMID:19721089 \cite{Gazzaniga2009}?}.
 
-A detailed analysis of sequence variances in NamPRT revealed that only Deuterostomia that have NNMT but not NADA, have a sequence insertion in the N-terminal part of NamPRT that according to our experimental analysis seems to enable the high affinity of the enzyme.
-This in turn would suggest that also the bacterial enzymes doe not have a high substrate affinity and although the one bacterial NamPRT has been proven to be enzymatically active, the affinity of a bacterial enzyme has not been measured so far.
+A detailed analysis of sequence variances in NamPRT revealed that only Deuterostomia that have NNMT but not NADA, have a sequence insertion in the N-terminal part of NamPRT that, according to our experimental analysis, seems to enable the high affinity of the enzyme.
+This in turn would suggest that also the bacterial enzymes do not have a high substrate affinity and although the one bacterial NamPRT has been proven to be enzymatically active, the affinity of a bacterial enzyme has not been measured so far \todo{There were more NamPRTs found and (some of) their kinetic constants were measured.
+Please have a look at e.g.
+PMIDs:11157928, 20926389, 16585762 \cite{Martin2001,Sorci2010,Gerdes2006}}.
 The differences in activity and affinity, implying differences in substrate binding could potentially be exploited for the development of antibiotics.
-But further analysis as well as potentially the crystallisation of a bacterial enzyme would be required, to see whether this could be a  approach.
+But further analysis as well as potentially the crystallisation of a bacterial enzyme would be required, to see whether this could be a promising approach.
 
-In our analysis, we so far completely neglected the potential effects of cosubstrates of the pathway investigated such as the presence of targets of the NAD-consuming enzymes, such as acetylated proteins for sirtuins for example, or the requirement of PRPP\todo{insert full name and but PRPP in brackets} and ATP that are rrequired for NMN synthesis by NamPRT  or the presence of methyl donor \textit{S}-adenosyl methionine (SAM) and its precursor methionine that have been shown to potentially limit the effect of NNMT\cite{PMID:23455543 }.
+In our analysis, we so far completely neglected the potential effects of co-substrates of the pathway investigated such as the presence of targets of the NAD-consuming enzymes, such as acetylated proteins for sirtuins for example, or the requirement of phosphoribosyl pyrophosphate (PRPP) and ATP that are required for NMN synthesis by NamPRT or the presence of the methyl donor \textit{S}-adenosyl methionine (SAM) and its precursor methionine that have been shown to potentially limit the effect of NNMT \cite{Ulanovskaya2013}.
 These should be included in future analysis.
-The challenge is, however, that currently information about the in vivo concentration of those co-substrates are very limited.
-We assume that this will improve in the future as new, more sensitive, methods for metabolite measurements are currently developed.
+The challenge is, however, that currently information about the \textit{in~vivo} concentration of those co-substrates is very limited.
+We assume that this will improve in the future as new, more sensitive, methods for metabolite measurements are currently developed \todo{ref?}.
 
-During our analysis we came across several problems related to current sequence database.
-One is sequence contamination, which is a well known problem\todo{insert ref} and we therefore tried to remove all sequences of obvious bacterial origin from the analysis, based on sequence homology analysis.
-Another problem are incomplete genomes ....
-\todo{Mathias B.
-please insert here the paragraph you have written for your thesis--- you should probably write it in other words}
+During our analysis we came across several problems related to the NCBI sequence database.
+One is sequence contamination, which is a well known problem \todo{ref?} and we therefore tried to remove all sequences of obvious bacterial origin from the analysis, based on sequence homology analysis.
+Another problem is incomplete genomes.
+Although there are tools to assess the completeness of a genome (e.g.~\cite{Simao2015}), none of them could convincingly claim to be reliable.
+The genomes of the common model organisms can probably be assumed to be close to complete, but there are many draft genomes in the databases whose completeness is uncertain.
+Even if the completeness would be known, with the high number of genomes used in this analysis, it is likely that some genes of interest were not sequenced in every genome.
+For our analysis, this means that scattered patterns of few missing genes may be real or stem from an incomplete genome.
 
 The third problem are wrong annotations.
-We tried to avoid these, by, wherever possible, only relying on template sequences with confirmed function.
-This problem becomes apparent by the fact that in yeast an enzyme named NNMT can be found, which is based on an initial analysis on life span extension in Saccharomyces cerevisae\todo{insert ref} a sequence similarity that is restricted to the SAM-dependent methyl transferase domain.
-The protein has, however, later been shown not function as Nam or small molecule methyl transferase but as ...\todo{Please complete incl.
-refs.}
+We tried to avoid these, by only relying on template sequences with confirmed function, wherever possible.
+This problem becomes apparent by the fact that in yeast an enzyme named NNMT can be found, which is based on an initial analysis on life span extension in \textit{Saccharomyces cerevisae} \cite{Anderson2003}.
+The protein has, however, later been shown not function as methyltransferase for Nam but for the eukaryotic elongation factor~1A (eEF1A) giving it its new name elongation factor methyltransferase~7 (Efm7) \cite{Hamey2016}.
+The old name is still present in many databases, though.
 
-Nevertheless, we have been able to comprehensively analyse co-evolution of several enzymes of the NAD pathway with the appearance of NNMT seemingly initiating and driving complex alterations of the pathway such as an increase and diversification of NAD-dependent signalling, followed by an increase in NamPRT substrate affinity (schematic overview see fig.~\ref{fig:evo_events}).
+Nevertheless, we have been able to comprehensively analyse the functional co-evolution of several enzymes of the NAD pathway with the appearance of NNMT seemingly initiating and driving complex alterations of the pathway such as an increase and diversification of NAD-dependent signalling, followed by an increase in NamPRT substrate affinity (schematic overview see fig.~\ref{fig:evo_events}).
 This again appears to be accompanied by the loss of NADA in vertebrates and the first gene duplication of NMNATs \cite{Lau2010}.
@@
 To our knowledge, this is the first study that comprehensively demonstrates a interdependency of enzyme evolution based on functional interaction of several enzymes of a pathway.
-While it is of course well know and described for directly interaction proteins \todo{Not sure that is true, one that might come close is the following work https://academic.oup.com/mbe/article/33/1/268/2579532.....please check and potentially revise statement}.
+While it is of course well know and described for directly interaction proteins \todo{Not sure that is true, one that might come close is the following work PMID:26446903 \cite{Figliuzzi2016}.
+Please check and potentially revise statement}.

### 2018-07-10 latex/latex_to_word.py

@@ -1,206 +1,0 @@
-#!/usr/bin/env python3
-
-import sys
-import re
-import subprocess
-
-
-try:
-	fn = sys.argv[1]
-	if fn.startswith('-h'):
-		raise IndexError
-except IndexError:
-	print('Call me with the main tex file as argument.', file=sys.stderr)
-	print('python3 latex_to_word.py myfile.tex', file=sys.stderr)
-	print('\nNB!', file=sys.stderr)
-	print('Libre Office must be installed and NOT be open!', file=sys.stderr)
-	print('Shared memory under /run/shm must be available!', file=sys.stderr)
-	sys.exit()
-
-if fn.endswith('.'):
-	fn += 'tex'
-if not fn.endswith('.tex'):
-	fn += '.tex'
-
-tex = open(fn).read()
-
-while True:
-	try:
-		idx = tex.index(r'\input{')
-	except ValueError:
-		break
-
-	endidx = tex.index('}', idx)
-	newfn = tex[idx+7:endidx]
-
-	if not newfn.endswith('.tex'):
-		newfn += '.tex'
-
-	newtex = open(newfn).read()
-
-	tex = tex[:idx] + newtex + tex[endidx+1:]
-
-out = ['''<!DOCTYPE html>
-<html>
-<head>
-<meta charset="utf-8"/>
-<title>Thesis</title>
-<style type="text/css">
-.todo {color: red; font-weight: bold;}
-p {text-align: justify;}
-</style>
-</head>
-<body>''']
-active = False
-
-first_replacements = {
-	'~': '&nbsp;',
-	'<': '&lt;',
-	'>': '&gt;'}
-
-hard_replacements = {
-	r'\newpage': '',
-	r'\S': '§',
-	r'$\beta$': 'β',
-	r'$\mu$': 'µ',
-	r'$\Delta$': 'Δ',
-	r'\,': '&thinsp;'}
-
-for key in first_replacements:
-	tex = tex.replace(key, first_replacements[key])
-
-for key in hard_replacements:
-	tex = tex.replace(key, hard_replacements[key])
-
-re_comment = re.compile(r'[^\\]%')
-
-tex = re.sub(r'\\cite\{([^\}]+)\}', r'(\1)', tex)
-
-known_environments = {'figure'}
-environment = None
-figure_labels = {}
-table_labels = {}
-figure_counter = 1
-table_counter = 1
-
[…]

### 2018-07-10 latex/discussion.tex
Incorporated changes suggested by Dorothée

@@
 As NNMT does, however, provide an advantage as long as NamPRT affinity is sufficiently high, this might provide an useful avenue for treatment, especially for types of cancer that express NNMT at a high level.
-These would be more susceptible to competitive inhibitors of NamPRT, and several of those inhibitors are currently tested in clinical studies \todo{ref.
-Maybe PMID:28733523 \cite{Espindola-Netto2017}?}.
+These would be more susceptible to competitive inhibitors of NamPRT, and several of those inhibitors are currently tested in clinical studies \cite{Espindola-Netto2017} \todo{Are there other refs?}.
 The major healthy tissue expressing NNMT is the liver, while most other tissues do not express NNMT or only at very low levels.
@@
 We therefore manually looked through the bacteria found to encode for NamPRT and indeed found that many but not all of them have been characterised to be pathogenic.
-It should be noted that the distribution of NADA and NamPRT does not follow the bacterial species tree\todo{ref.
-Maybe PMID:19721089 \cite{Gazzaniga2009}?}.
+It should be noted that the distribution of NADA and NamPRT does not follow the bacterial species tree \cite{Gazzaniga2009}.
 

### 2018-07-10 latex/figurelegends.tex
Resolved many todos and typos in discussion.
Incorporated changes suggested by Dorothée

@@
   \centering
-  \includegraphics[width=0.75\textwidth]{fig_6_unresolved_loop}
-  \caption{\textbf{The function of the structurally unresolved loop structure of NamPRT.} Most Deuterostomia that encode NNMT show a sequence insertion in the N-terminal region of NamPRT that has been revealed by multiple sequence alignment of NamPRT from different species (A).
+  \includegraphics[width=0.7\textwidth,draft]{fig_6_unresolved_loop}
+  \caption{\textbf{The function of the structurally unresolved loop of NamPRT.} Most Deuterostomia that encode NNMT show a sequence insertion in the N-terminal region of NamPRT that has been revealed by multiple sequence alignment of NamPRT from different species (A).
 Coloured circles indicate the enzymes present in the respective species blue: NamPRT and NNMT; black: NamPRT, NADA and NNMT; yellow: NamPRT and NADA.
@@
 \end{figure}
-
-\todo{What about the positive selection in NNMT?}
-
-\newpage

### 2018-07-10 latex/header.tex
Incorporated changes suggested by Dorothée

@@
 \noindent
-\textsuperscript{2}Department of Molecular Biology, University of Bergen, Thormøhlensgt. 55, 5020 Bergen, Norway
+\textsuperscript{2}Department of Biological Sciences, University of Bergen, Thormøhlensgt. 55, 5020 Bergen, Norway
 
@@
 In eukaryotes, two different Nam salvage pathways exist.
-Phylogenetic analysis of the pathway show that in lower organisms the initial deamidation of Nam is prevalent, whilst the direct conversion of Nam to the mononucleotide by Nam phosphoribosyltransferase (NamPRT) dominates in animals and eventually remains as the single Nam recycling route in vertebrates.
+Phylogenetic analysis of the pathway shows that in lower organisms the initial deamidation of Nam is prevalent, whilst the direct conversion of Nam to the mononucleotide by Nam phosphoribosyltransferase (NamPRT) dominates in animals and eventually remains as the single Nam recycling route in vertebrates.
 Strikingly, loss of the deamidation pathway in early vertebrates is preceded by the occurrence of a new enzyme that marks Nam for excretion by methylation – nicotinamide N-methyltransferase (NNMT).

### 2018-07-10 latex/methods.tex
Incorporated changes suggested by Dorothée

@@
 
-\section{Experimental Procedure}
+\section{Experimental Procedures}
 
@@
 For eukaryotic expression with a C-terminal FLAG-epitope, the open reading frame (ORF) encoding human NamPRT was inserted into pFLAG-CMV-5a (Merck - Sigma Aldrich) via EcoRI/BamHI sites.
-Using a PCR-based approach, this vector provided the basis for the generation of a plasmid encoding a NamPRT deletion mutant lacking amino acid residues 42-51 (NamPRT-$\Delta$42-51).
+Using a PCR-based approach, this vector provided the basis for the generation of a plasmid encoding a NamPRT deletion mutant lacking amino acid residues 42-51 ($\Delta$42-51 NamPRT).
 For prokaryotic expression with an N-terminal 6xHis-tag, the wild-type and mutant ORFs were inserted into pQE-30 (Qiagen) via BamHI and PstI-sites.
@@
 The spectra phase and baseline were automatically and manually corrected using TopSpin 3.5 software (Bruker Biospin).
-Quantification of nicotinamide mononucleotide (NMN) was done by the integration of the peak at 9.57\,ppm and DSS used as an internal standard.
+Quantification of nicotinamide mononucleotide (NMN) was done by the integration of the peak at 9.52\,ppm and DSS used as an internal standard.

### 2018-07-10 latex/results.tex
Incorporated changes suggested by Dorothée

@@
 Thus, we here focused on the phylogenetic analysis of the two enzymes that initiate the two different NAD salvage pathways, NADA and NamPRT.
-The phylogenetic distribution of the two enzymes is scattered in bacteria \todo{MathiasB: please find and insert citation} and have been shown to occur together in some marine invertebrates \cite{Gossmann2012FEBS}.
+The phylogenetic distribution of the two enzymes is scattered in bacteria \cite{Gazzaniga2009} and have been shown to occur together in some marine invertebrates \cite{Gossmann2012FEBS}.
 In addition, we included the Nam methyltransferase NNMT into our analysis, as its function and origin is unclear.
@@
 with proven function ...
-\todo{Mathias B: include values} \todo{Answer from MB: There was not a single hit outside Metazoa}.
+\todo{Mathias B: include values} \todo{Answer by MB: There was not a single hit outside Metazoa}.
 
@@
 
-The second hypothesis was that the sequence insertion influences the affinity of NamPRT, which would be in line with the predictions from our simulations.
-We recombinantly expressed both wildtype and mutant NamPRT in \textit{E.~coli}.
-After purification (suppl.
-fig.~S3), we measured the activity using NMR detection of NMN in the presence and absence of ATP.
-The enzymatic activity of the mutant enzyme lacking the loop was much lower than the activity of the wildtype enzyme (fig.~\ref{fig:unresolved_loop}D).
-Furthermore, an increase of the substrate concentration had a much stronger effect on the activity of the mutant compared to the wildtype enzyme, indicating a lower affinity.
-In contrast to wildtype NamPRT, the mutant did not exhibit an increased activity in the presence of ATP (fig.~\ref{fig:unresolved_loop}E).
+The second hypothesis was that the sequence insertion might influence NamPRT affinity.
+We thus expressed both wildtype and mutant proteins, N-terminally fused to a 6xHis-tag in \textit{E.~coli}, and purified them.
+The size exclusion chromatography profile showed that both wildtype and mutant protein were expressed as dimers (see suppl.
+fig.~S3), indicating that the missing residues in the mutant did not dramatically affect the protein structure.
+The enzymatic activity was measured by NMR spectroscopy using the detection of NMN produced in the presence and absence of ATP.
+Upon incubation with the NamPRT inhibitor FK866 \cite{Hasmann2003} for 30 minutes, both wildtype and mutant NamPRT did not produce any NMN, suggesting that binding of FK866 is not affected by the mutation (see suppl.
+fig.~S4).
+Nevertheless, the enzymatic activity of the mutant enzyme was approximately twice as low than that of the wildtype enzyme (fig.~\ref{fig:unresolved_loop}D).
+In contrast to wildtype NamPRT, the activity of the mutant enzyme was not stimulated in the presence of ATP (fig.~\ref{fig:unresolved_loop}E).
+Our observations suggest that the mutant enzyme is catalytically active, retains its dimeric state and sensitivity to FK866.
+However, the lower activity and the absence of catalytic activation by ATP indicate that deletion of amino acids 42 to 51 might have affected substrate binding.
 

### 2018-07-10 latex/supplementaries.tex
Incorporated changes suggested by Dorothée

@@
 
+\subsection*{Supplementary figure S4}
+
+\textbf{NMR measurement of NamPRT activity.} NamPRT activity was measured as described in Experimental Procedures and product (NMN) formation was detected using 1D 1H NMR spectroscopy.
+Inset on the right: molecular structure of NMN with the atom detected by NMR indicated by an arrow.
+The range used for NMN detection in typical 1D 1H NMR spectra of the enzymatic reactions is shown.
+NMN quantification was done with the singlet detected at 9.52 ppm.
+From the top to the bottom, peak detection of NMN standard (200\,µM), wildtype NamPRT (1\,mM Nam and 1\,mM PRPP), mutant $\Delta$42-51 NamPRT (1\,mM Nam and 1\,mM PRPP), wildtype NamPRT with FK866, and mutant $\Delta$42-51 NamPRT with FK866.
+Incubation with inhibitor FK866 was done for 30\,min at 30\,°C.
+
 \subsection*{Supplementary table S1}

### 2018-07-10 latex/discussion.tex
Proofread documents, esp. discussion

@@
 
-We here comprehensively analysed the phylogenetic distribution of the three enzymes using Nam as a substrate that are the two NAD salvage pathway enzymes NADA and NamPRT as well as the Nam-degrading enzyme NNMT.
+We here comprehensively analysed the phylogenetic distribution of the three enzymes using Nam as a substrate.
+These are the two NAD salvage pathway enzymes NADA and NamPRT as well as the Nam-degrading enzyme NNMT.
 We found that after the first appearance of NNMT in Protostomia, a diversification of NAD-consuming reactions in Deuterostomia can be observed.
-By using mathematical modelling, we could explain these finding, as NNMT removes excess Nam from cells and thereby reduces product inhibition of NAD signalling enzymes.
+We could explain these finding using mathematical modelling, as NNMT removes excess Nam from cells and thereby reduces product inhibition of NAD signalling enzymes.
 This in turn enables higher fluxes through these reactions.
 Thus, the diversification of NAD-consuming enzymes in mammals seems to have been enabled by the presence of NNMT.
-NAD-consuming enzymes are involved in a wide variety of signalling and gene regulatory mechanisms that due to their sensitivity to NAD$^{+}$, have the ability to translate differences in metabolic states into changes in signalling and gene regulation.
+
+NAD-consuming enzymes are involved in a wide variety of signalling and gene regulatory mechanisms that, due to their sensitivity to NAD$^{+}$, have the ability to translate differences in metabolic states into changes in signalling and gene regulation.
 As NAD concentrations are lowered by the removal of NAD precursor by NNMT, a high-affinity enzyme for recycling of Nam is required to ensure maintenance for cellular NAD levels for other metabolic processes.
-Whether it was by chance or whether this is due to thermodynamic or mechanistic limitations that NamPRT developed a high affinity and not NADA, or whether other environmental conditions have been responsible for the evolutionary trajectory observed, is impossible to deduce based on the chosen approach.
-But our simulations clearly show that a high affinity of NamPRT is required for high NAD consumption fluxes and NAD concentrations and that it therefore appears that NNMT might have been driving NamPRT evolution.
-Looking at the enzyme affinities of the human enzymes it furthermore appears that both NNMT and NamPRT reached an near optimal state, as further increase of the affinity of NamPRT and decrease of the affinity of NNMT would not result in much higher steady state NAD concentrations or NAD consumption fluxes, according to our simulations.
-It furthermore appears that NNMT makes both NAD concentration and NAD consumption relatively independent of other processes requiring NAD, such as cell growth.
+It is unclear, why NamPRT developed a higher affinity and not NADA.
+NADA may have thermodynamic or mechanistic limitations or other environmental conditions may have been responsible for the observed evolutionary trajectory.
+Still, our simulations clearly show that a high affinity of NamPRT is required for high NAD consumption fluxes and NAD concentrations.
+It therefore seems plausible that NNMT might have been driving NamPRT evolution.
+Looking at the enzyme affinities of the human enzymes it furthermore appears that both NNMT and NamPRT reached an almost optimal state, as further changes in the affinity of NamPRT or NNMT would not result in much higher steady state NAD concentrations or NAD consumption fluxes, according to our simulations.
+In addition, the data suggest that NNMT makes both NAD concentration and NAD consumption relatively independent of other processes requiring NAD, such as cell growth.
 
@@
 \cite{Okamura1998}).
-It might well be that increased NNMT expression in cancer might serve to remove Nam produced by increased NAD-dependent signalling.
-To maintain high NAD concentrations a simultaneous overexpression of NamPRT is required, which is what has been found in some types of cancer \cite{Bi2011,Wang2011}.
-As NNMT does, however, provide an advantage as long as NamPRT affinity is sufficiently high, this might provide an useful avenue for treatment, especially for types of cancer that express NNMT at a high level.
-These would be more susceptible to competitive inhibitors of NamPRT, and several of those inhibitors are currently tested in clinical studies \cite{Espindola-Netto2017} \todo{Are there other refs?}.
-The major healthy tissue expressing NNMT is the liver, while most other tissues do not express NNMT or only at very low levels.
-Based on our analysis, we would furthermore suggest that it might be reasonable to screen patients before treatment, as non-NNMT expressing tumours might respond less to competitive NamPRT inhibitors.
-This is furthermore relevant, as the missing degradation in those cells would potentially lead to an accumulating of Nam that could outcompete the inhibitor in those cells.
-But these aspects require further investigation.
+The major healthy tissue expressing NNMT is the liver, while no or only little expression of NNMT is observed in most other tissues \todo{ref}.
+Increased NNMT expression in cancer might serve to remove Nam produced by increased NAD-dependent signalling.
+To maintain high NAD concentrations, a simultaneous higher expression of NamPRT is required, which is what has been found in some types of cancer \cite{Bi2011,Wang2011}.
+NNMT is only advantageous as long as NamPRT affinity is sufficiently high.
+This suggests that certain types of cancer that express NNMT at a high level would be more susceptible to competitive inhibitors of NamPRT.
+Several of such inhibitors are currently tested in clinical studies \cite{Espindola-Netto2017} \todo{Are there more refs?}.
+Based on our analysis, we would suggest that it might be reasonable to screen patients before treatment, as non-NNMT expressing tumours might respond less to competitive NamPRT inhibitors and missing Nam degradation in those cancer cells would potentially lead to an accumulation of Nam that could outcompete the inhibitor.
+The latter aspect is unclear and requires further investigation.
 
-Our combined phylogenetic/modelling analysis furthermore provides a potential explanation both for the co-occurrence of NADA and NamPRT in bacteria, but even more strongly for the loss of NADA in vertebrates.
-We here show that when simulating compartments that share the same limited source of Nam, the compartment that contains NamPRT and NNMT has a higher steady state NAD concentration and NAD consumption rate as the compartment containing NADA.
-The combination of a high affinity NamPRT with NNMT, which is the dominant enzyme combination found in vertebrates, seems to provide a competitive advantage.
+Our combined phylogenetic/modelling analysis provides a potential explanation both for the co-occurrence of NADA and NamPRT in bacteria and for the loss of NADA in vertebrates.
+We here show that when two compartments compete for the same limited source of Nam, the compartment that contains NamPRT and NNMT has a higher steady state NAD concentration and NAD consumption rate than the compartment containing NADA.
+The dominant enzyme combination found in vertebrates, a high-affinity NamPRT with NNMT, seems to provide a competitive advantage.
 As this may also hold for mammalian-associated bacteria, particularly pathogens, we wanted to see whether pathogenic bacteria solely express NamPRT.
-Unfortunately, bacterial habitat information is currently far from complete and often difficult to access especially at the NCBI database that was used for the analysis provided here, as it is the most comprehensive sequence database to date.
-We therefore manually looked through the bacteria found to encode for NamPRT and indeed found that many but not all of them have been characterised to be pathogenic.
-It should be noted that the distribution of NADA and NamPRT does not follow the bacterial species tree \cite{Gazzaniga2009}.
+Unfortunately, bacterial habitat information is currently far from complete and often difficult to access.
+We therefore manually checked bacteria that possess NamPRT and indeed found that most of them have been characterised to be pathogenic.
+It should be noted that the distribution of NADA and NamPRT does not follow the bacterial species tree but is rather scattered \cite{Gazzaniga2009}.
+This distribution hints to a functional importance of posssessing NADA or NamPRT that might be due to different environments.
 
-A detailed analysis of sequence variances in NamPRT revealed that only Deuterostomia that have NNMT but not NADA, have a sequence insertion in the N-terminal part of NamPRT that, according to our experimental analysis, seems to enable the high affinity of the enzyme.
-This in turn would suggest that also the bacterial enzymes do not have a high substrate affinity and although the one bacterial NamPRT has been proven to be enzymatically active, the affinity of a bacterial enzyme has not been measured so far \todo{There were more NamPRTs found and (some of) their kinetic constants were measured.
-Please have a look at e.g.
-PMIDs:11157928, 20926389, 16585762 \cite{Martin2001,Sorci2010,Gerdes2006}}.
+A detailed analysis of sequence variances in NamPRT revealed that only deuterostomes that have NNMT but not NADA, have a sequence insertion in the N-terminal part of NamPRT that seems to enable the high affinity of the enzyme.
+This in turn would suggest that also the bacterial enzymes do not have a high substrate affinity.
+The only kinetically characterised NamPRT is from \textit{Acinetobacter baylyi} \cite{Sorci2010} and has a $K_{M}$ for Nam that is about 10\,000 times higher than the $K_{M}$ of human NamPRT, supporting our hypothesis.
+Other bacterial NamPRTs were shown to be functional \cite{Martin2001,Gerdes2006}, but no kinetic parameters were measured.
 The differences in activity and affinity, implying differences in substrate binding could potentially be exploited for the development of antibiotics.
-But further analysis as well as potentially the crystallisation of a bacterial enzyme would be required, to see whether this could be a promising approach.
+Further analysis possibly including the crystallisation of a bacterial NamPRT would be required, to see whether the bacterial NAD metabolism could be a promising target.
 
-In our analysis, we so far completely neglected the potential effects of co-substrates of the pathway investigated such as the presence of targets of the NAD-consuming enzymes, such as acetylated proteins for sirtuins for example, or the requirement of phosphoribosyl pyrophosphate (PRPP) and ATP that are required for NMN synthesis by NamPRT or the presence of the methyl donor \textit{S}-adenosyl methionine (SAM) and its precursor methionine that have been shown to potentially limit the effect of NNMT \cite{Ulanovskaya2013}.
-These should be included in future analysis.
-The challenge is, however, that currently information about the \textit{in~vivo} concentration of those co-substrates is very limited.
-We assume that this will improve in the future as new, more sensitive, methods for metabolite measurements are currently developed \todo{ref?}.
+% Martin2001: Haemophilus ducreyi: no kinetic measurements
+% Sorci2010: Acinetobacter sp.: NamPRT (Nam) Km: 0.04 mM (= 40 000 nM); kcat: 0.12/s (apparent values determined at constant 5 mM PRPP and 2 mM ATP)
+% Gerdes2006: Synechocystis sp.: only activity 0.5 U/mg NamPRT for Nam
+% Human, according to table S2: Km: 5 nM; kcat: 0.0077/s
+
+In our analysis, we completely neglected the potential effects of co-substrates of the investigated pathway.
+Neglected co-substrates include targets of the NAD-consuming enzymes, such as acylated proteins for sirtuins, for example, or phosphoribosyl pyrophosphate (PRPP) and ATP that are required for NMN synthesis by NamPRT.
+Also the presence of the methyl donor \textit{S}-adenosyl methionine (SAM) and its precursor methionine that have been shown to potentially limit the effect of NNMT \cite{Ulanovskaya2013} was not considered.
+All of these co-substrates should be included in future analyses, but information about the \textit{in~vivo} concentration of those co-substrates is currently very limited.
+We anticipate improvements as more sensitive methods for metabolite measurements are currently developed \todo{ref?
+Or how do you know?}.
+
+% todo: The following three paragraphs were not proofread, yet.
 

### 2018-07-10 latex/figurelegends.tex
Proofread documents, esp. discussion

@@
   \includegraphics{fig_1_pathway_overview}
-  \caption{\textbf{Schematic overview of NAD biosynthesis pathways} NAD can be synthesised from tryptophan (Trp), nicotinamide (Nam), nicotinic acid (NA), and, to a lesser extend, nicotinamide ribose (NR).
-Nam is the main precursor in human and also the product of NAD-consuming signalling reactions by enzymes such as sirtuins (NAD-dependent histone deacetylases) or PARPs (poly-ADP-ribosylases).
+  \caption{\textbf{Schematic overview of NAD biosynthesis pathways.} NAD can be synthesised from tryptophan (Trp), nicotinamide (Nam), nicotinic acid (NA), and to a lesser extend nicotinamide ribose (NR).
+Nam is the main precursor in human and also the product of NAD-consuming signalling reactions by enzymes such as sirtuins (NAD-dependent deacylases) or PARPs (poly-ADP-ribosylases).
 For the recycling of Nam, two different pathways exist.
-The pathway found in yeast and many bacteria starts with the deamidation of Nam by Nam deamidase (NADA).
+The pathway found in yeast, plants, and many bacteria starts with the deamidation of Nam by Nam deamidase (NADA).
 The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates.
@@
   \includegraphics[width=0.75\textwidth]{fig_2_phylo_distribution}
-  \caption{\textbf{Phylogenetic distribution of NADA, NNMT, and NamPRT and their relation to the number of NAD consumers.} A) Distribution of NADA, NNMT, and NamPRT in selected major taxa.
-NADA is dominant in Bacteria, Fungi, and Plants (Viridiplantae), whereas NamPRT together with NNMT is dominant in Metazoa.
-Numbers at the pie charts show, how many species of the taxon possess the respective enzyme combination indicated by the colour explained in the lower right of the figure.
-Below the taxon name, the number of species in that taxon is given.
-B) Common tree of selected taxa within the Metazoa, including 334 species.
-The pie charts indicate the distribution of species within the respective taxon that encode the enzyme combination indicated by the different colours.
-The size of the pie charts is proportional to the logarithm of the number of species analysed in the particular taxon.
-The numbers below the taxon names indicate the average number of NAD-consuming enzyme families found in all sub-taxa.
+  \caption{\textbf{Phylogenetic distribution of NADA, NNMT, and NamPRT and their relation to the number of NAD consumers.} A) Distribution of NADA, NNMT, and NamPRT in selected clades.
+NADA is dominant in bacteria, fungi, and plants (Viridiplantae), whereas NamPRT together with NNMT is dominant in Metazoa.
+Numbers at the pie charts show, how many species of the clades possess the respective enzyme combination indicated by the colour explained in the lower right of the figure.
+The number of species in a clade is given below its name.
+B) Common tree of selected clades within the Metazoa, including 334 species.
+The pie charts indicate the distribution of species within the respective clade that encode the enzyme combination indicated by the different colours.
+The size of the pie charts is proportional to the logarithm of the number of species analysed in the particular clade.
+The numbers below the clade names indicate the average number of NAD-consuming enzyme families found in all species of that clade.
 The branch length is arbitrary.}
@@
   \includegraphics[width=\textwidth]{fig_3_NNMT_NAD_flux}
-  \caption{\textbf{NNMT enables high NAD consumption flux.} We used a dynamic model of NAD biosynthesis and consumption (for details, see experimental procedures) to simulate steady state NAD  consumption flux  (A) and concentration (B).
+  \caption{\textbf{NNMT enables high NAD consumption flux.} We used a dynamic model of NAD biosynthesis and consumption (for details, see Experimental Procedures) to simulate steady state NAD consumption flux (A) and concentration (B).
 The amount of NMNAT and NamPRT used in the simulations where adjusted such that the free NAD concentrations were in the range reported in the literature.
@@
 Details are given in supplementary table~S2.
-In the absence of NNMT (blue lines), steady state NAD consumption rates are higher despite reduced NAD concentrations.
-Increasing the amount of NamPRT in the simulation fourfold (blue dotted lines) partially compensates for the decreased NAD steady state concentration caused by Nam degradation through NNMT.}
+In the presence of NNMT (blue lines), steady state NAD consumption rates are higher despite reduced NAD concentrations.
+Increasing the amount of NamPRT in the simulation fourfold (blue dotted lines) partially compensates for the decreased NAD concentration caused by Nam degradation through NNMT.}
   \label{fig:NNMT_NAD_flux}
@@
   \includegraphics[width=0.72\textwidth]{fig_4_NamPRT_affinity_Nam}
-  \caption{\textbf{Role of NamPRT substrate affinity} Using the dynamic model of NAD biosynthesis and consumption we simulated the affect of different Michaelis-Menten constants ($K_{M}$) of NamPRT for Nam on the steady state NAD consumption flux and NAD concentration at different cell division rates.
+  \caption{\textbf{Role of NamPRT substrate affinity.} We simulated the effect of different Michaelis-Menten constants ($K_{M}$) of NamPRT for Nam on the steady state NAD consumption flux and NAD concentration at different cell division rates.
 All parameters were equal to those used for the simulations in figure~\ref{fig:NNMT_NAD_flux}.
-In the absence of NNMT, the $K_{M}$ of NamPRT has little influence on NAD consumption (A) and concentration (B), but both are changing with cell division rates.
+In the absence of NNMT, the $K_{M}$ of NamPRT has little influence on NAD consumption (A) and concentration (B), but both are strongly influenced by cell division rates.
 In the presence of NNMT, decreasing $K_{M}$ of NamPRT enables increasing NAD consumption flux (C) and NAD concentration (D).
 NNMT furthermore makes both, NAD consumption flux and concentration, almost independent of cell division rates.
-Comparing the situation with and without NNMT (E and F) at different NamPRT $K_{M}$ reveals that at low $K_{M}$ (dashed lines) and high cell division rates NNMT no longer enables higher NAD consumption rates compared to NamPRT alone (green line and dashed black line).}
+Comparing the situation with and without NNMT (E and F) at two different NamPRT $K_{M}$ values reveals that at high $K_{M}$ (dashed lines) and high cell division rates NNMT no longer enables higher NAD consumption rates compared to NamPRT alone (green line and dashed grey line).}
   \label{fig:NamPRT_affinity_Nam}
@@
   \includegraphics[width=\textwidth]{fig_5_optimal_substr_affinities}
-  \caption{\textbf{The substrate affinities of human NNMT and NamPRT are optimal.} Using the same model as in figures~\ref{fig:NNMT_NAD_flux} and~\ref{fig:NamPRT_affinity_Nam}, we simulated the impact of changes in the $K_{M}$ for both NamPRT and NNMT on NAD consumption rates (A) and NAD concentration (B).
+  \caption{\textbf{The substrate affinities of human NNMT and NamPRT are optimal.} We simulated the impact of changes in the $K_{M}$ for both NamPRT and NNMT on NAD consumption rates (A) and NAD concentration (B).
 Both are increasing with decreasing $K_{M}$ of NamPRT but increasing $K_{M}$ of NNMT.
@@
   \centering
-  \includegraphics[width=0.7\textwidth,draft]{fig_6_unresolved_loop}
-  \caption{\textbf{The function of the structurally unresolved loop of NamPRT.} Most Deuterostomia that encode NNMT show a sequence insertion in the N-terminal region of NamPRT that has been revealed by multiple sequence alignment of NamPRT from different species (A).
+  \includegraphics[width=0.7\textwidth]{fig_6_unresolved_loop}
+  \caption{\textbf{The function of the structurally unresolved loop of NamPRT.} Most deuterostomes that possess NamPRT and NNMT show a sequence insertion in the N-terminal region of NamPRT that has been revealed by multiple sequence alignment of NamPRT from different species (A).
 Coloured circles indicate the enzymes present in the respective species blue: NamPRT and NNMT; black: NamPRT, NADA and NNMT; yellow: NamPRT and NADA.
 For a more comprehensive alignment, please see supplementary figure~S1.
-The visualisation of human NamPRT (B) is based on a structure prediction of SWISS-MODEL \cite{Arnold2006,Biasini2014} using the model 2H3D of the human NamPRT as template \cite{Wang2006}.
+The structure visualisation of human NamPRT (B) is based on a structure prediction by SWISS-MODEL \cite{Arnold2006,Biasini2014} using the model 2H3D of the human NamPRT as template \cite{Wang2006}.
 The inserted region is not resolved in any of currently available crystal structures of NamPRT and thus appears to be a flexible loop structure at the surface of the NamPRT dimer, coloured in red.
@@
 \textit{In vitro} measurements using recombinant protein show that the mutant NamPRT has a lower activity than the wildtype enzyme (D) and is not activated by ATP (E).
-Bars in panels D and E that have different letters indicate significant difference of measured values as estimated using a T test assuming independent samples and significance at $p < 0.05$.}
+Bars in panels D and E that have different letters indicate significant difference of measured values as estimated using a T~test assuming independent samples and significance at $p < 0.05$.}
   \label{fig:unresolved_loop}
@@
 NADA and NamPRT were simulated to be present at equal amounts.
-In the absence of NNMT (A and B) the compartment containing NADA has slightly lower NAD consumption rates (A) but much higher steady state NAD concentrations (B).
+In the absence of NNMT the compartment containing NADA has slightly lower NAD consumption rates (A) but much higher steady state NAD concentrations (B).
 In the presence of NNMT, however, both NAD consumption (C) and NAD concentration (D) are lower in the NADA compartment.
@@
   \includegraphics[width=\textwidth]{fig_8_evo_events}
-  \caption{\textbf{Schematic representation of evolutionary events in the NAD pathway} Based on the phylogenetic analysis presented here and earlier work \cite{Lau2010} we summarized and indicated important events in the evolution of NAD metabolism in Metazoa.
-Events in italics are from \cite{Lau2010}.
-All other events were found in this work.}
+  \caption{\textbf{Schematic representation of evolutionary events in the NAD pathway.} Based on the phylogenetic analysis presented here (roman font) and earlier work \cite{Lau2010} (italic font) we summarized and indicated important events in the evolution of NAD metabolism in Metazoa.}
   \label{fig:evo_events}

### 2018-07-11 latex/discussion.tex
small edits and comments discussion

@@
 This suggests that certain types of cancer that express NNMT at a high level would be more susceptible to competitive inhibitors of NamPRT.
-Several of such inhibitors are currently tested in clinical studies \cite{Espindola-Netto2017} \todo{Are there more refs?}.
+Several of such inhibitors are currently tested in clinical studies \cite{Espindola-Netto2017} \todo{Yes there are more refs: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4603696/ The  one included does not comp}.
 Based on our analysis, we would suggest that it might be reasonable to screen patients before treatment, as non-NNMT expressing tumours might respond less to competitive NamPRT inhibitors and missing Nam degradation in those cancer cells would potentially lead to an accumulation of Nam that could outcompete the inhibitor.
@@
 This in turn would suggest that also the bacterial enzymes do not have a high substrate affinity.
-The only kinetically characterised NamPRT is from \textit{Acinetobacter baylyi} \cite{Sorci2010} and has a $K_{M}$ for Nam that is about 10\,000 times higher than the $K_{M}$ of human NamPRT, supporting our hypothesis.
-Other bacterial NamPRTs were shown to be functional \cite{Martin2001,Gerdes2006}, but no kinetic parameters were measured.
+The onlyfully  kinetically characterised bacterial NamPRT is from \textit{Acinetobacter baylyi} \cite{Sorci2010} and has a $K_{M}$ for Nam that is about 10\,000 times higher than the $K_{M}$ of human NamPRT, supporting our hypothesis.
+Other bacterial NamPRTs were shown to be functional \cite{Martin2001,Gerdes2006}, but the substrate affinities have not been determined.
 The differences in activity and affinity, implying differences in substrate binding could potentially be exploited for the development of antibiotics.
@@
 We anticipate improvements as more sensitive methods for metabolite measurements are currently developed \todo{ref?
-Or how do you know?}.
+Or how do you know?
+Mathias B.: Include ref for stable isotope labelling of NAD-precursor for example, but there are certainly others}.
 
@@
 We also noted that the second gene duplication of NMNATs and thus the further compartmentalisation of NAD metabolism is co-occurring with a site-specific positive selection event in NNMT (unpublished results).
-To our knowledge, this is the first study that comprehensively demonstrates a interdependency of enzyme evolution based on functional interaction of several enzymes of a pathway.
-While it is of course well know and described for directly interaction proteins \todo{Not sure that is true, one that might come close is the following work PMID:26446903 \cite{Figliuzzi2016}.
-Please check and potentially revise statement}.
+
+\todo{Insert concluding paragraph}

### 2018-07-11 latex/discussion.tex
revised discussion

@@
 It is unclear, why NamPRT developed a higher affinity and not NADA.
-NADA may have thermodynamic or mechanistic limitations or other environmental conditions may have been responsible for the observed evolutionary trajectory.
-Still, our simulations clearly show that a high affinity of NamPRT is required for high NAD consumption fluxes and NAD concentrations.
+NADA may have thermodynamic or mechanistic limitations, other environmental conditions may have been responsible for the observed evolutionary trajectory, or it might just have happened by chance.
+Whatever is the case, we show in our simulations that a high affinity of NamPRT is required for high NAD consumption fluxes and NAD concentrations in the presence of NNMT.
 It therefore seems plausible that NNMT might have been driving NamPRT evolution.
-Looking at the enzyme affinities of the human enzymes it furthermore appears that both NNMT and NamPRT reached an almost optimal state, as further changes in the affinity of NamPRT or NNMT would not result in much higher steady state NAD concentrations or NAD consumption fluxes, according to our simulations.
-In addition, the data suggest that NNMT makes both NAD concentration and NAD consumption relatively independent of other processes requiring NAD, such as cell growth.
+Looking at the enzyme affinities of the human enzymes, it furthermore appears that both NNMT and NamPRT reached an almost optimal state, as further changes in the affinity of either  NamPRT or NNMT would not result in much higher steady state NAD concentrations or NAD consumption fluxes.
+In addition, our simulations suggest that NNMT makes both NAD concentration and NAD consumption relatively independent of other processes requiring NAD, such as cell growth.
 
-Our findings shed also new light on the potential physiological role of NNMT, which has earlier been recognized as potential marker for some types of cancer (e.g.
+Our findings shed new light on the potential physiological role of NNMT, which has earlier been recognized as potential marker for some types of cancer (e.g.
 \cite{Okamura1998}).
-The major healthy tissue expressing NNMT is the liver, while no or only little expression of NNMT is observed in most other tissues \todo{ref}.
-Increased NNMT expression in cancer might serve to remove Nam produced by increased NAD-dependent signalling.
+The main healthy tissue expressing NNMT is the liver, while no or only little expression of NNMT is observed in most other healthy tissues \todo{ref}.
+The increased NNMT expression observed in some types of cancer, might serve to remove Nam derived by increased NAD-dependent signalling.
 To maintain high NAD concentrations, a simultaneous higher expression of NamPRT is required, which is what has been found in some types of cancer \cite{Bi2011,Wang2011}.
 NNMT is only advantageous as long as NamPRT affinity is sufficiently high.
-This suggests that certain types of cancer that express NNMT at a high level would be more susceptible to competitive inhibitors of NamPRT.
-Several of such inhibitors are currently tested in clinical studies \cite{Espindola-Netto2017} \todo{Yes there are more refs: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4603696/ The  one included does not comp}.
-Based on our analysis, we would suggest that it might be reasonable to screen patients before treatment, as non-NNMT expressing tumours might respond less to competitive NamPRT inhibitors and missing Nam degradation in those cancer cells would potentially lead to an accumulation of Nam that could outcompete the inhibitor.
-The latter aspect is unclear and requires further investigation.
+This suggests that certain types of cancer expressing NNMT at a high level, would potentially be more susceptible to competitive inhibitors of NamPRT.
+Several of such inhibitors are currently tested in clinical studies \cite{Espindola-Netto2017} \todo{Yes there are more refs: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4603696/ The  one included does not compile}.
+Based on our analysis, we would suggest that it might be reasonable to screen patients before treatment, as non-NNMT expressing tumours might respond less to competitive NamPRT inhibitors and missing Nam degradation in those cancer cells would potentially lead to an accumulation of Nam that could out-compete the inhibitor.
+The latter aspect is not well investigated and requires further analysis.
 
-Our combined phylogenetic/modelling analysis provides a potential explanation both for the co-occurrence of NADA and NamPRT in bacteria and for the loss of NADA in vertebrates.
-We here show that when two compartments compete for the same limited source of Nam, the compartment that contains NamPRT and NNMT has a higher steady state NAD concentration and NAD consumption rate than the compartment containing NADA.
-The dominant enzyme combination found in vertebrates, a high-affinity NamPRT with NNMT, seems to provide a competitive advantage.
+The scattered distribution of NamPRT and NADA that is especially pronounced in bacteria \cite{Gazzaniga2009} but has also been observed in eukaryotes as well as the disappearance of NADA in vertebrates has not been understood earlier.
+Our combined phylogenetic-modelling analysis now provides a potential explanation for both observations.
+Using simulated competition between two compartments that share for the same limited source of Nam, we show that the compartment that contains NamPRT and NNMT has a higher steady state NAD concentration and NAD consumption rate than the compartment containing NADA.
+This is, however, only  the case if NamPRT substrate affinity is sufficiently high.
+The dominant enzyme combination found in vertebrates, a high-affinity NamPRT with NNMT, thus seems to provide a competitive advantage.
 As this may also hold for mammalian-associated bacteria, particularly pathogens, we wanted to see whether pathogenic bacteria solely express NamPRT.
@@
 We therefore manually checked bacteria that possess NamPRT and indeed found that most of them have been characterised to be pathogenic.
-It should be noted that the distribution of NADA and NamPRT does not follow the bacterial species tree but is rather scattered \cite{Gazzaniga2009}.
-This distribution hints to a functional importance of posssessing NADA or NamPRT that might be due to different environments.
+It should be noted that the distribution of NADA and NamPRT does not follow the bacterial species tree \cite{Gazzaniga2009}.
+Besides the suggestion made here, there might might well be other environmental aspects that influence the phylogenetic distribution in bacteria.
 
@@
 This in turn would suggest that also the bacterial enzymes do not have a high substrate affinity.
-The onlyfully  kinetically characterised bacterial NamPRT is from \textit{Acinetobacter baylyi} \cite{Sorci2010} and has a $K_{M}$ for Nam that is about 10\,000 times higher than the $K_{M}$ of human NamPRT, supporting our hypothesis.
+We have only been able to find susbtrate affinities for one bacterial NamPRT from \textit{Acinetobacter baylyi} \cite{Sorci2010}.
+This enzyme has a $K_{M}$ of 0.04 mM for Nam, which  is about 10\,000 times higher than the $K_{M}$ of human NamPRT, supporting our hypothesis.
 Other bacterial NamPRTs were shown to be functional \cite{Martin2001,Gerdes2006}, but the substrate affinities have not been determined.
@@
 Neglected co-substrates include targets of the NAD-consuming enzymes, such as acylated proteins for sirtuins, for example, or phosphoribosyl pyrophosphate (PRPP) and ATP that are required for NMN synthesis by NamPRT.
-Also the presence of the methyl donor \textit{S}-adenosyl methionine (SAM) and its precursor methionine that have been shown to potentially limit the effect of NNMT \cite{Ulanovskaya2013} was not considered.
-All of these co-substrates should be included in future analyses, but information about the \textit{in~vivo} concentration of those co-substrates is currently very limited.
+Furtermore, the presence of the methyl donor \textit{S}-adenosyl methionine (SAM) and its precursor methionine that have been shown to potentially limit the effect of NNMT \cite{Ulanovskaya2013} was not considered here.
+As co-substrate availability might alter the behaviour of the system, these should be included in future analyses.
+Information about the \textit{in~vivo} concentration of those co-substrates is, however, currently very limited.
 We anticipate improvements as more sensitive methods for metabolite measurements are currently developed \todo{ref?
@@
 
-During our analysis we came across several problems related to the NCBI sequence database.
+During our analysis we came across several problems related to the use of NCBI sequence database for phylogenetic analysis.
 One is sequence contamination, which is a well known problem \todo{ref?} and we therefore tried to remove all sequences of obvious bacterial origin from the analysis, based on sequence homology analysis.
-Another problem is incomplete genomes.
+Another problem is posed by incomplete genomes.
 Although there are tools to assess the completeness of a genome (e.g.~\cite{Simao2015}), none of them could convincingly claim to be reliable.
@@
 We tried to avoid these, by only relying on template sequences with confirmed function, wherever possible.
-This problem becomes apparent by the fact that in yeast an enzyme named NNMT can be found, which is based on an initial analysis on life span extension in \textit{Saccharomyces cerevisae} \cite{Anderson2003}.
+This problem becomes apparent by the fact that in yeast an enzyme named NNMT can be found,.
+The initial naming was based on a very weak homology to human NNMT and an analysis of life span extension of the mutant in \textit{Saccharomyces cerevisae} \cite{Anderson2003}, which showed simmilar effects as other mutants of the NAD pathway.
 The protein has, however, later been shown not function as methyltransferase for Nam but for the eukaryotic elongation factor~1A (eEF1A) giving it its new name elongation factor methyltransferase~7 (Efm7) \cite{Hamey2016}.
@@
 
-Nevertheless, we have been able to comprehensively analyse the functional co-evolution of several enzymes of the NAD pathway with the appearance of NNMT seemingly initiating and driving complex alterations of the pathway such as an increase and diversification of NAD-dependent signalling, followed by an increase in NamPRT substrate affinity (schematic overview see fig.~\ref{fig:evo_events}).
+Taken together, we have been able to comprehensively analyse the functional co-evolution of several enzymes of the NAD pathway with the appearance of NNMT seemingly initiating and driving complex alterations of the pathway such as an increase and diversification of NAD-dependent signalling, followed by an increase in NamPRT substrate affinity (schematic overview see fig.~\ref{fig:evo_events}).
 This again appears to be accompanied by the loss of NADA in vertebrates and the first gene duplication of NMNATs \cite{Lau2010}.
@@
 
-\todo{Insert concluding paragraph}
+We here developed a new approach that combines detailed phylogenetic analysis with dynamic metabolic modelling of the respective pathway and have been able to explain observed evolutionary changes in the NAD biosynthesis and consumption pathway.
+Based on the simulated pathway dynamics, we have furthermore derived predictions for potentially disease relevant physiological interdependencies between several enzymes of the pathway, that could upon verification provide the basis for new disease treatments.
+Our results and the point that we could verify part of our predictions, demonstrates the potential of  the approach for the analysis of dynamic networks and how it can be used to unravel functional interdependencies within pathways of interest.

### 2018-07-11 latex/discussion.tex
Did some general rewriting and todo solving

@@
 It is unclear, why NamPRT developed a higher affinity and not NADA.
-NADA may have thermodynamic or mechanistic limitations, other environmental conditions may have been responsible for the observed evolutionary trajectory, or it might just have happened by chance.
-Whatever is the case, we show in our simulations that a high affinity of NamPRT is required for high NAD consumption fluxes and NAD concentrations in the presence of NNMT.
+NADA may have thermodynamic or mechanistic limitations, other environmental conditions may apply, or it might just have happened by chance.
+Whatever the reason, we show in our simulations that a high affinity of NamPRT is required for high NAD consumption fluxes and NAD concentrations in the presence of NNMT.
 It therefore seems plausible that NNMT might have been driving NamPRT evolution.
@@
 \cite{Okamura1998}).
-The main healthy tissue expressing NNMT is the liver, while no or only little expression of NNMT is observed in most other healthy tissues \todo{ref}.
+The main healthy tissue expressing NNMT is the liver, while no or only little expression of NNMT is observed in most other healthy tissues \cite{Aksoy1994}.
 The increased NNMT expression observed in some types of cancer, might serve to remove Nam derived by increased NAD-dependent signalling.
@@
 This suggests that certain types of cancer expressing NNMT at a high level, would potentially be more susceptible to competitive inhibitors of NamPRT.
-Several of such inhibitors are currently tested in clinical studies \cite{Espindola-Netto2017} \todo{Yes there are more refs: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4603696/ The  one included does not compile}.
-Based on our analysis, we would suggest that it might be reasonable to screen patients before treatment, as non-NNMT expressing tumours might respond less to competitive NamPRT inhibitors and missing Nam degradation in those cancer cells would potentially lead to an accumulation of Nam that could out-compete the inhibitor.
+Several of such inhibitors are currently tested in clinical studies \cite{EspindolaNetto2017,Xu2015SR}.
+Based on our analysis, we would suggest that it might be reasonable to screen patients before treatment, as non-NNMT expressing tumours might respond less to competitive NamPRT inhibitors and missing Nam degradation in those cancer cells would potentially lead to an accumulation of Nam that could outcompete the inhibitor.
 The latter aspect is not well investigated and requires further analysis.
 
-The scattered distribution of NamPRT and NADA that is especially pronounced in bacteria \cite{Gazzaniga2009} but has also been observed in eukaryotes as well as the disappearance of NADA in vertebrates has not been understood earlier.
+Neither the scattered distribution of NamPRT and NADA that is especially pronounced in bacteria \cite{Gazzaniga2009} but that has also been observed in eukaryotes, nor the disappearance of NADA in vertebrates has not been understood earlier.
 Our combined phylogenetic-modelling analysis now provides a potential explanation for both observations.
@@
 This in turn would suggest that also the bacterial enzymes do not have a high substrate affinity.
-We have only been able to find susbtrate affinities for one bacterial NamPRT from \textit{Acinetobacter baylyi} \cite{Sorci2010}.
-This enzyme has a $K_{M}$ of 0.04 mM for Nam, which  is about 10\,000 times higher than the $K_{M}$ of human NamPRT, supporting our hypothesis.
+We have only been able to find substrate affinities for one bacterial NamPRT from \textit{Acinetobacter baylyi} \cite{Sorci2010}.
+This enzyme has a $K_{M}$ of 0.04\,mM for Nam, which is about 10\,000 times higher than the $K_{M}$ of human NamPRT, supporting our hypothesis.
 Other bacterial NamPRTs were shown to be functional \cite{Martin2001,Gerdes2006}, but the substrate affinities have not been determined.
@@
 Neglected co-substrates include targets of the NAD-consuming enzymes, such as acylated proteins for sirtuins, for example, or phosphoribosyl pyrophosphate (PRPP) and ATP that are required for NMN synthesis by NamPRT.
-Furtermore, the presence of the methyl donor \textit{S}-adenosyl methionine (SAM) and its precursor methionine that have been shown to potentially limit the effect of NNMT \cite{Ulanovskaya2013} was not considered here.
+Furthermore, the presence of the methyl donor \textit{S}-adenosyl methionine (SAM) and its precursor methionine that have been shown to potentially limit the effect of NNMT \cite{Ulanovskaya2013} was not considered here.
 As co-substrate availability might alter the behaviour of the system, these should be included in future analyses.
 Information about the \textit{in~vivo} concentration of those co-substrates is, however, currently very limited.
-We anticipate improvements as more sensitive methods for metabolite measurements are currently developed \todo{ref?
-Or how do you know?
-Mathias B.: Include ref for stable isotope labelling of NAD-precursor for example, but there are certainly others}.
+We anticipate improvements as more sensitive methods for metabolite measurements are currently developed \cite{Frederick2017} \todo{Mathias Z., do you know others?}.
 
-% todo: The following three paragraphs were not proofread, yet.
-
-During our analysis we came across several problems related to the use of NCBI sequence database for phylogenetic analysis.
-One is sequence contamination, which is a well known problem \todo{ref?} and we therefore tried to remove all sequences of obvious bacterial origin from the analysis, based on sequence homology analysis.
-Another problem is posed by incomplete genomes.
+During our analysis we came across several problems related to the use of NCBI sequence databases for phylogenetic analyses.
+One is sequence contamination, which is a well known problem \cite{Ballenghien2017,Longo2011}.
+To avoid contamination, we used sequence homology analysis to remove all sequences of obvious bacterial origin from the results in eukaryotic species.
+Another problem is incomplete genomes.
 Although there are tools to assess the completeness of a genome (e.g.~\cite{Simao2015}), none of them could convincingly claim to be reliable.
 The genomes of the common model organisms can probably be assumed to be close to complete, but there are many draft genomes in the databases whose completeness is uncertain.
-Even if the completeness would be known, with the high number of genomes used in this analysis, it is likely that some genes of interest were not sequenced in every genome.
+Even if the completeness would be known, due to the high number of genomes used in this analysis, it is likely that some genes of interest were not sequenced in every genome.
 For our analysis, this means that scattered patterns of few missing genes may be real or stem from an incomplete genome.
@@
 We tried to avoid these, by only relying on template sequences with confirmed function, wherever possible.
-This problem becomes apparent by the fact that in yeast an enzyme named NNMT can be found,.
-The initial naming was based on a very weak homology to human NNMT and an analysis of life span extension of the mutant in \textit{Saccharomyces cerevisae} \cite{Anderson2003}, which showed simmilar effects as other mutants of the NAD pathway.
-The protein has, however, later been shown not function as methyltransferase for Nam but for the eukaryotic elongation factor~1A (eEF1A) giving it its new name elongation factor methyltransferase~7 (Efm7) \cite{Hamey2016}.
+This problem becomes apparent by the fact that in yeast an enzyme named NNMT can be found.
+The initial naming was based on a very weak homology to human NNMT and an analysis of life span extension of the mutant in \textit{Saccharomyces cerevisae} \cite{Anderson2003}, which showed similar effects as other mutants of the NAD pathway.
+The protein has later been shown not to function as methyltransferase for Nam but for the eukaryotic elongation factor~1A (eEF1A) giving it its new name elongation factor methyltransferase~7 (Efm7) \cite{Hamey2016}.
 The old name is still present in many databases, though.
 
-Taken together, we have been able to comprehensively analyse the functional co-evolution of several enzymes of the NAD pathway with the appearance of NNMT seemingly initiating and driving complex alterations of the pathway such as an increase and diversification of NAD-dependent signalling, followed by an increase in NamPRT substrate affinity (schematic overview see fig.~\ref{fig:evo_events}).
-This again appears to be accompanied by the loss of NADA in vertebrates and the first gene duplication of NMNATs \cite{Lau2010}.
+Taken together, we have been able to comprehensively analyse the functional co-evolution of several enzymes of the NAD pathway.
+The appearance of NNMT seemingly initiated and drove complex alterations of the pathway such as an increase and diversification of NAD-dependent signalling, followed by an increase in NamPRT substrate affinity.
+A schematic overview is given in figure~\ref{fig:evo_events}.
+This transition appears to be accompanied by the loss of NADA in vertebrates and the first gene duplication of NMNATs \cite{Lau2010}.
 We also noted that the second gene duplication of NMNATs and thus the further compartmentalisation of NAD metabolism is co-occurring with a site-specific positive selection event in NNMT (unpublished results).
 
-We here developed a new approach that combines detailed phylogenetic analysis with dynamic metabolic modelling of the respective pathway and have been able to explain observed evolutionary changes in the NAD biosynthesis and consumption pathway.
-Based on the simulated pathway dynamics, we have furthermore derived predictions for potentially disease relevant physiological interdependencies between several enzymes of the pathway, that could upon verification provide the basis for new disease treatments.
-Our results and the point that we could verify part of our predictions, demonstrates the potential of  the approach for the analysis of dynamic networks and how it can be used to unravel functional interdependencies within pathways of interest.
+We here developed a new approach that combines detailed phylogenetic analysis with dynamic metabolic modelling and have been able to explain observed evolutionary changes in the NAD biosynthesis and consumption pathway.
+Based on the simulated pathway dynamics, we have furthermore derived predictions for physiological interdependencies between several enzymes of the pathway that are potentially relevant for new disease treatments.
+Our results, including the experimental verification of parts of our predictions, demonstrate the potential of our approach for the analysis of dynamic networks and how the approach can be used to unravel functional interdependencies within pathways of interest.

### 2018-07-11 latex/introduction.tex
Did some general rewriting and todo solving

@@
 Firstly, vertebrates use a direct two step pathway starting with the conversion of Nam into the mononucleotide NMN by the Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate.
-The nearly complete recycling of Nam by NamPRT is achieved by an extraordinary high subtrate affinity to Nam, the $K_{M}$ being in the low nanomolar range \cite{Burgos2008}.
+The nearly complete recycling of Nam by NamPRT is achieved by an extraordinary high substrate affinity to Nam, the $K_{M}$ being in the low nanomolar range \cite{Burgos2008}.
 This appears to be mediated by an ATP-dependent phosphorylation of a histidine residue in the catalytic core \cite{Burgos2009}.
@@
 NA is converted into the corresponding mononucleotide (NAMN), in a reactions performed by the NA-specific phosphoribosyltransferase NAPRT.
-The convertion of the mononucleotides NMN and NAMN into their dinucleotides NAD and NAAD is performed by the same enzyme, the Nam/NA adenylytransferases (NMNATs) that is essential in all organisms \cite{DeFigueiredo2011}.
+The conversion of the mononucleotides NMN and NAMN into their dinucleotides NAD and NAAD is performed by the same enzyme, the Nam/NA adenylytransferases (NMNATs) that is essential in all organisms \cite{DeFigueiredo2011}.
 The recycling pathway via NA finally requires re-amidation of the dinucleotide, which is catalysed by NAD synthase.

### 2018-07-11 latex/methods.tex
Proofread documents, esp. discussion
Did some general rewriting and todo solving

@@
 
-\subsection{Transient transfection, immunocytochemstry, and confocal laser scanning microscopy}
+\subsection{Transient transfection, immunocytochemistry, and confocal laser scanning microscopy}
 
@@
 1D 1H NMR spectra were acquired on a 850\,MHz Ascend Bruker spectrometer equipped with 5\,mm TCI triple-resonance CryoProbe and a pulse field gradients along the z-axis.
-The experiments were acquired with zgesgppe pulse sequence, allowing water suppression using excitation sculpting with gradients and perfect echo.
+The experiments were acquired with the zgesgppe pulse sequence, allowing water suppression using excitation sculpting with gradients and perfect echo.
 The temperature was kept constant at 300\,K and the acquisition was started with 2000 scans, 1\,s relaxation delay, 1.6\,s acquisition time, 65\,000 data points, and a spectral width of 14\,ppm.

### 2018-07-11 latex/results.tex
Did some general rewriting and todo solving

@@
 As shown in Figure~\ref{fig:phylo_distribution}A, bacteria, fungi, and plants predominantly possess NADA and only a few of them harbour NamPRT.
-In contrast, Metazoa predominantly lost NADA amd have NamPRT together with NNMT.
+In contrast, Metazoa predominantly lost NADA and have NamPRT together with NNMT.
 NNMT seems to have arisen \textit{de~novo} in the most recent common ancestor of Ecdysozoa and Lophotrochozoa (fig.~\ref{fig:phylo_distribution}B).
-As we were unable to find any \textit{NNMT} gene with an e-value below 0.1 in fungi or plants.
-The gene with the highest sequence similarity in \textit{S.~cerevisiae} is ...
-with proven function ...
-\todo{Mathias B: include values} \todo{Answer by MB: There was not a single hit outside Metazoa}.
+We were unable to find any \textit{NNMT} gene with an e-value below 0.1 in fungi or plants.
 
@@
 fig.~S2).
-\todo{figure not corresponding to current species trees \cite{Prum2015}} The lack of NNMT might be related to the excretion system, as the product of NNMT, methyl-Nam, is in mammals excreted with the urine.
+\todo{figure not corresponding to current species tree \cite{Prum2015}.
+MB: Will recreate this figure with the tree from the paper tonight.} The lack of NNMT might be related to the excretion system, as the product of NNMT, methyl-Nam, is in mammals excreted with the urine.
 There are few metazoan species for which we could not find NamPRT or NADA, while NNMT was present.
@@
 Of course, this is an artificial scenario, as one would assume organisms to regulate enzyme expression to achieve similar levels of metabolite concentrations instead.
-Nevertheless, in the absence of NNMT there seems to be a tradeoff between maintainable NAD concentration and consumption flux.
+Nevertheless, in the absence of NNMT there seems to be a trade-off between maintainable NAD concentration and consumption flux.
 In contrast, in the presence of NNMT, NAD consumption rates and concentrations are almost independent of cell division rates.
@@
 The first hypothesis was that the presence of the loop changes the subcellular localisation of NamPRT, as it is overlapping with a predicted NLS.
-To test this hypothesis, we created a mutant NamPRT lacking the loop and recombinatly expressed FLAG-tagged wildtype and mutant NamPRT in Hela S3 cells.
+To test this hypothesis, we created a mutant NamPRT lacking the loop and recombinantly expressed FLAG-tagged wildtype and mutant NamPRT in HeLa S3 cells.
 Immunofluorescence imaging showed a mixed cytosolic nuclear localisation for the wildtype and the mutant NamPRT (fig.~\ref{fig:unresolved_loop}C).

### 2018-07-13 latex/discussion.tex
Incorporated all changes and comments by MZ. Comments not directly addressed are included as todos.

@@
 
+\todo[inline]{MZ: Ist Geschmackssache. Ich bin ja mehr so der Typ: "Three major observations have emerged from the present study. 1., 2., 3..." Aber es geht natürlich auch so wie es steht.}
+
 NAD-consuming enzymes are involved in a wide variety of signalling and gene regulatory mechanisms that, due to their sensitivity to NAD$^{+}$, have the ability to translate differences in metabolic states into changes in signalling and gene regulation.
@@
 It is unclear, why NamPRT developed a higher affinity and not NADA.
-NADA may have thermodynamic or mechanistic limitations, other environmental conditions may apply, or it might just have happened by chance.
+NADA may have thermodynamic or mechanistic limitations, other environmental conditions may apply, or it might just have happened by chance \todo{MZ: Irgendwo vorher steht, dass dieser Weg energetisch ungünstiger ist.
+Wäre das nicht zumindest eine überlegenswerte Erklärung?}.
 Whatever the reason, we show in our simulations that a high affinity of NamPRT is required for high NAD consumption fluxes and NAD concentrations in the presence of NNMT.
@@
 It should be noted that the distribution of NADA and NamPRT does not follow the bacterial species tree \cite{Gazzaniga2009}.
-Besides the suggestion made here, there might might well be other environmental aspects that influence the phylogenetic distribution in bacteria.
+Besides the suggestion made here, there might well be other environmental aspects that influence the phylogenetic distribution in bacteria.
 
@@
 
-In our analysis, we completely neglected the potential effects of co-substrates of the investigated pathway.
-Neglected co-substrates include targets of the NAD-consuming enzymes, such as acylated proteins for sirtuins, for example, or phosphoribosyl pyrophosphate (PRPP) and ATP that are required for NMN synthesis by NamPRT.
+In our analyses, we did not consider the potential effects of co-substrates of the investigated pathway.
+Such co-substrates include targets of the NAD-consuming enzymes, such as acylated proteins for sirtuins, for example, or phosphoribosyl pyrophosphate (PRPP) and ATP that are required for NMN synthesis by NamPRT.
 Furthermore, the presence of the methyl donor \textit{S}-adenosyl methionine (SAM) and its precursor methionine that have been shown to potentially limit the effect of NNMT \cite{Ulanovskaya2013} was not considered here.
 As co-substrate availability might alter the behaviour of the system, these should be included in future analyses.
-Information about the \textit{in~vivo} concentration of those co-substrates is, however, currently very limited.
-We anticipate improvements as more sensitive methods for metabolite measurements are currently developed \cite{Frederick2017} \todo{Mathias Z., do you know others?}.
+Unfortunately, information about the \textit{in~vivo} concentrations of these co-substrates is currently very limited.
 

### 2018-07-13 latex/figurelegends.tex
Incorporated all changes and comments by MZ. Comments not directly addressed are included as todos.

@@
 Both are increasing with decreasing $K_{M}$ of NamPRT but increasing $K_{M}$ of NNMT.
-The affinities reported for human enzymes (indicated by a black asterisk) appear to be close to optimal, as further improvements would have little or no effect on NAD consumption or concentration.}
+The affinities reported for human enzymes (indicated by a black asterisk) appear to be close to the theoretical optimum, as further improvements would have little or no effect on NAD consumption or concentration.}
   \label{fig:optimal_substr_affinities}

### 2018-07-13 latex/header.tex
Incorporated all changes and comments by MZ. Comments not directly addressed are included as todos.

--- 2018-07-08 latex/header.tex
@@
 \noindent
-\textsuperscript{2}Department of Molecular Biology, University of Bergen, Thormøhlensgt. 55, 5020 Bergen, Norway
+\textsuperscript{2}Department of Biomedicine, University of Bergen, Jonas Lies Vei 91, 5020 Bergen, Norway
 
@@
 NAD is best known as cofactor in redox reactions, but it is also substrate of NAD-dependent signalling reactions that consume NAD and release nicotinamide (Nam).
-In eukaryotes, two different Nam salvage pathways exist.
-Phylogenetic analysis of the pathway show that in lower organisms the initial deamidation of Nam is prevalent, whilst the direct conversion of Nam to the mononucleotide by Nam phosphoribosyltransferase (NamPRT) dominates in animals and eventually remains as the single Nam recycling route in vertebrates.
-Strikingly, loss of the deamidation pathway in early vertebrates is preceded by the occurrence of a new enzyme that marks Nam for excretion by methylation – nicotinamide N-methyltransferase (NNMT).
-Despite the occurrence of this Nam degrading enzyme, we observe a diversification of NAD dependent signalling enzymes in vertebrates.
-This seems counterintuitive.
+Two different Nam salvage pathways exist.
+We conducted extensive phylogenetic analyses of these pathways and show that in lower organisms the initial deamidation of Nam is prevalent, whilst the direct conversion of Nam to the mononucleotide by Nam phosphoribosyltransferase (NamPRT) dominates in animals and eventually remains as the single Nam recycling route in vertebrates.
+Strikingly, loss of the deamidation pathway in early vertebrates is preceded \todo{Statt "preceded", schlägt MZ "accompanied" for.
+Ich (MB) denke, dass preceded richtig ist.} by the emergence of a new enzyme that marks Nam for excretion by methylation – nicotinamide N-methyltransferase (NNMT).
+Paradoxically, the occurrence of this Nam degrading enzyme is paralleled by a diversification of NAD dependent signalling enzymes in vertebrates.
 
-To better understand these evolutionary changes and the role of NNMT for NAD metabolism, we built a mathematical model of the pathway using published enzyme kinetics data.
-Our results indicate that NNMT is required to enable high NAD consumption fluxes enabling diversification of the NAD-dependent signalling pathways.
-This kinetic regulation requires a high substrate affinity of the key enzyme for Nam salvage, NamPRT.
-Indeed, the affinity of NamPRT to Nam has previously been measured to be in the nanomolar range.
-Based on our mathematical modelling approach we furthermore suggest that NNMT exerted an evolutionary pressure on NamPRT, enforcing the development of its unusually high substrate affinity.
+To better understand these evolutionary changes and the role of NNMT for NAD metabolism, we built a mathematical model of the pathway using available enzyme kinetics data.
+Our simulations indicate that NNMT is required to maintain high NAD consumption fluxes, thereby enabling diversification of the NAD-dependent signalling pathways.
+This kinetic regulation requires and explains the unusually high substrate affinity of the key enzyme for Nam salvage, NamPRT.
+Moreover, we suggest that NNMT exerted evolutionary pressure on NamPRT, enforcing the development of its high substrate affinity.
 Using multiple sequence alignments, we identified a sequence insertion, first occurring in vertebrates, that parallels an experimentally verified increase in the substrate affinity of the enzyme.
 Additional simulations show that the deamidation pathway became obsolete owing to the high substrate affinity of NamPRT.
-Collectively, our results illustrate a close evolutionary relationship between NAD biosynthesis and the diversification NAD-dependent signalling pathways, potentially driven by the concomitant occurrence of a regulator of Nam salvage, NNMT.
+Collectively, our results illustrate a close evolutionary relationship between NAD biosynthesis and the diversification of NAD-dependent signalling pathways, potentially driven by the concomitant occurrence of a regulator of Nam salvage, NNMT.
 

### 2018-07-13 latex/introduction.tex
Incorporated all changes and comments by MZ. Comments not directly addressed are included as todos.

@@
 Even though it is best known as cofactor for various redox-reactions, NAD is involved in a number of signalling processes that consume NAD$^{+}$ by cleaving the molecule to nicotinamide (Nam) and ADP-ribose \cite{Verdin2015}.
-These NAD-dependent signalling reactions include but are not limited to poly- and mono-ADP-ribosylation \cite{Butepage2015,DeVos2012}, NAD-dependent protein deacylation by sirtuins \cite{Osborne2016}, and the synthesis of calcium-mobilizing molecules such as cyclic ADP-ribose \cite{Lee2012}.
+These NAD-dependent signalling reactions include poly- and mono-ADP-ribosylation \cite{Butepage2015,DeVos2012}, NAD-dependent protein deacylation by sirtuins \cite{Osborne2016}, and the synthesis of calcium-mobilizing molecules such as cyclic ADP-ribose \cite{Lee2012}.
 These NAD-dependent signalling processes participate in the regulation of virtually all cellular activities.
@@
 Due to the constant release of Nam through NAD-consuming signalling reactions, the NAD salvage pathway using Nam as precursor is the most important NAD synthesis pathway.
-If Nam would not be constantly recycled into NAD, humans would require a much higher daily vitamin B3 intake than the 16\,mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}.
+If Nam were not continuously recycled into NAD, humans would require a much higher daily vitamin B3 intake than the 16\,mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}.
 Two principal pathways exist that recycle Nam.
@@
 
-Secondly, in many prokayotes as well as in plants and fungi, a pathway consisting of four steps starting with the deamidation of Nam to nicotinic acid (NA) by the Nam deamidase (NADA) is used.
-(\ref{fig:pathway_overview}).
+Secondly, in most prokayotes as well as in plants and fungi, a pathway consisting of four steps starting with the deamidation of Nam to nicotinic acid (NA) by the Nam deamidase (NADA) is used.
+(fig.~\ref{fig:pathway_overview}).
 The three enzymes that act after NADA belong to the Preiss-Handler pathway that also exists in vertebrates.
 NA is converted into the corresponding mononucleotide (NAMN), in a reactions performed by the NA-specific phosphoribosyltransferase NAPRT.
-The conversion of the mononucleotides NMN and NAMN into their dinucleotides NAD and NAAD is performed by the same enzyme, the Nam/NA adenylytransferases (NMNATs) that is essential in all organisms \cite{DeFigueiredo2011}.
-The recycling pathway via NA finally requires re-amidation of the dinucleotide, which is catalysed by NAD synthase.
+The conversion of both mononucleotides, NMN and NAMN, into their corresponding dinucleotides, NAD and NAAD, is catalysed by the Nam/NA adenylytransferases (NMNATs) that are essential in all organisms \cite{DeFigueiredo2011}.
+The recycling pathway via NA finally requires re-amidation of NAAD by NAD synthase.
 This final reaction includes an enzyme adenylation step that consumes ATP.
-Therefore the Nam recycling by NADA is energetically less efficient under normal conditions than the recycling pathway starting with NamPRT.
+Therefore, the Nam recycling by NADA is energetically less efficient under normal conditions than the recycling pathway starting with NamPRT.
+\todo{MZ: Hier müssten Fragen/Probleme aufgeworfen werden, z.B.: Wieso sind die beiden Möglichkeiten offenbar evolutionär getrennt?
+Warum gibt es ein weiteres Enzym, NNMT, das Nam "abbaut"?
+...}
 

### 2018-07-13 latex/results.tex
Incorporated all changes and comments by MZ. Comments not directly addressed are included as todos.

@@
 
-We have previously shown that NMNAT is essential for NAD biosynthesis independent of the biosynthetic precursor or pathway \cite{DeFigueiredo2011}.
-In addition, the Preiss-Handler pathway was found present in virtually all organisms included in an earlier phylogenetic analysis \cite{Gossmann2012FEBS}.
-Thus, we here focused on the phylogenetic analysis of the two enzymes that initiate the two different NAD salvage pathways, NADA and NamPRT.
-The phylogenetic distribution of the two enzymes is scattered in bacteria \cite{Gazzaniga2009} and have been shown to occur together in some marine invertebrates \cite{Gossmann2012FEBS}.
-In addition, we included the Nam methyltransferase NNMT into our analysis, as its function and origin is unclear.
+\todo[inline]{MZ: Ich finde, die NAD-abhängigen Signalwege kommen in den Resultaten zu kurz weg.
+Es genügt vielleicht, einfach die 10 Klassen noch mal aufzuzählen (also z.B.
+x verschiedene PARP-Klassen, y Sirtuin-Klassen usw.).
+Dies würde ich gegenüberstellen an den Stellen, wo große Sprünge passieren.
+Daraus kann man z.B.
+ablesen, an welchen Stellen sich welche Signalwege speziell weiterentwickelt/diversifiziert haben.
+MB: Ich versuch mal, die entsprechenden Infos aus den Ergebnissen zu bekommen.}
 
+To understand the functional roles and potential interplay between the three known enzymes that use Nam as substrate (NamPRT, NADA and NNMT), we first conducted a comprehensive phylogenetic analysis.
+The phylogenetic distribution of the two enzymes that initiate the two different NAD salvage pathways, NADA and NamPRT is scattered in bacteria \cite{Gazzaniga2009}, while their co-occurrence has been detected in some marine invertebrates \cite{Gossmann2012FEBS}.
 As shown in Figure~\ref{fig:phylo_distribution}A, bacteria, fungi, and plants predominantly possess NADA and only a few of them harbour NamPRT.
@@
 \todo{figure not corresponding to current species tree \cite{Prum2015}.
-MB: Will recreate this figure with the tree from the paper tonight.} The lack of NNMT might be related to the excretion system, as the product of NNMT, methyl-Nam, is in mammals excreted with the urine.
+MB: Will recreate this figure with the tree from the paper asap.} The lack of NNMT might be related to the excretion system, as the product of NNMT, methyl-Nam, is in mammals excreted with the urine.
 There are few metazoan species for which we could not find NamPRT or NADA, while NNMT was present.
@@
 As we have very few cross species kinetic data, we were furthermore mainly relying on kinetic constants found for human or yeast enzymes.
-Wherever possible, we did not only include substrate affinities but also known product inhibitions or inhibition by downstream metabolites.
+Wherever possible, we included both substrate affinities and known product inhibitions or inhibition by downstream metabolites.
 As we in addition assumed that cell growth is, besides NAD-consuming reactions, a major driving force for NAD biosynthesis, we analysed different growth rates (cell division rates) by simulating different dilution rates for all metabolites.
@@
 We calculated steady state NAD concentrations and NAD consumption rates by simulating NAD biosynthesis proceeding via NamPRT in the presence or absence of NNMT.
-To achieve free NAD concentration in the range reported in the literature and due to the very low turnover of NamPRT, we had to use tenfold higher NamPRT levels compared to other enzymes and adjusted the amount of NMNAT to the same level to avoid that the NAD synthesis rates are limited by this enzyme.
-As shown in figure~\ref{fig:NNMT_NAD_flux}, the presence of NNMT enables higher NAD consumption rates (fig.~\ref{fig:NNMT_NAD_flux}A) while reducing the steady state concentration of NAD (fig.~\ref{fig:NNMT_NAD_flux}B).
+To achieve free NAD concentrations in the range reported in the literature and due to the very low turnover of NamPRT, we used tenfold higher NamPRT levels compared to other enzymes.
+We also adjusted the amount of NMNAT accordingly to avoid that the NAD synthesis rates are limited by this enzyme.
+Surprisingly, as shown in figure~\ref{fig:NNMT_NAD_flux}, the presence of NNMT enables higher rather than lower NAD consumption rates (fig.~\ref{fig:NNMT_NAD_flux}A).
+However, it diminishes the steady state concentration of NAD (fig.~\ref{fig:NNMT_NAD_flux}B).
 The decline in NAD concentration can be compensated by a higher expression of NamPRT, further increasing NAD consumption flux (dashed lines in fig.~\ref{fig:NNMT_NAD_flux}A and~B).
@@
 In the absence of NNMT, a variation of the substrate affinity of NamPRT for Nam, has very little effect on steady state NAD concentration and NAD consumption flux (fig.~\ref{fig:NamPRT_affinity_Nam}A and~B).
-In the presence of NNMT, however, NAD consumption flux and NAD concentration increases with decreasing $K_{M}$ values (fig.~\ref{fig:NamPRT_affinity_Nam}C and~D).
+In the presence of NNMT, however, NAD consumption flux and NAD concentration increases with decreasing $K_{M}$ values of NamPRT (fig.~\ref{fig:NamPRT_affinity_Nam}C and~D).
 
@@
 
-Figures~\ref{fig:NamPRT_affinity_Nam}E and~F show a direct comparison of different affinities of NamPRT for Nam with and without NNMT.
-At an affinity of $K_{M}$ =~1\,$\mu$M, which is in the range of the $K_{M}$ of NADA for Nam, NAD consumption flux is only higher with NNMT than without at low cell division rates (fig.~\ref{fig:NamPRT_affinity_Nam}E).
+Figures~\ref{fig:NamPRT_affinity_Nam}E and~F visualise a direct comparison of simulations assuming different affinities of NamPRT for Nam, in the presence or absence of NNMT.
+Interestingly, at an affinity of $K_{M}$ =~1\,$\mu$M, which is in the range of the $K_{M}$ of NADA for Nam, NAD consumption flux is only higher with NNMT present when cell division rates are low (fig.~\ref{fig:NamPRT_affinity_Nam}E).
 If the affinity of NamPRT is high enough, consumption rates are always higher with NNMT than without it.
@@
 
+\todo[inline]{MZ: Der obige Absatz ist sicher nicht einfach zu beschreiben, aber so ist es kaum zu verstehen. Mein Versuch ist der folgende Absatz:}
+
+To understand the interplay and competition for Nam between NNMT and NamPRT, we scanned a wide range of possible $K_{M}$ values for both enzymes in our simulations.
+As shown in figure~\ref{fig:optimal_substr_affinities}, the simulations indicate that both NAD consumption flux and NAD concentration would be minimal in case of a high $K_{M}$ for NamPRT and a low $K_{M}$ for NNMT.
+Conversely, lowering the $K_{M}$ of NamPRT to the nanomolar range substantially increases NAD consumption and concentration, which reach a maximum when the $K_{M}$ of NNMT is concomitantly elevated to the submillimolar range.
+The asterisks in figure~\ref{fig:optimal_substr_affinities} denote the $K_{M}$ values actually found for the human enzymes.
+Astonishingly, the naturally occurring $K_{M}$ values are very close to the theoretical optimum.
+
 \subsection{Sequence variance acquired in metazoans enhances substrate affinity}
 
-As our simulations suggested that NNMT might have exerted an evolutionary pressure on the development of NamPRT, we created a multiple sequence alignment to see if we can find sequence variations in NamPRT upon the appearance of NNMT.
-The multiple sequence alignment of selected sequences is shown in figure~\ref{fig:unresolved_loop}A and a more comprehensive multiple sequence alignment containing a larger number of species can be found in supplementary figure~S1.
+Given that NNMT might have exerted an evolutionary pressure on the development of NamPRT, one would expect to observe adaptations that are reflected in the NamPRT protein sequence arising with the occurrence of NNMT.
+To explore this, we created a multiple sequence alignment.
+The alignment of selected sequences is shown in figure~\ref{fig:unresolved_loop}A and a more comprehensive multiple sequence alignment containing a larger number of species can be found in supplementary figure~S1.
 We found that most Deuterostomes that possess only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 42 to 51 in the human enzyme.
 This insert overlaps with a predicted weak nuclear localisation signal (NLS), that is lost when the insert is removed.
-The sequence corresponds to a loop at the protein surface that is unresolved in all available crystal structures of human NamPRT (e.g.
+These ten amino acids correspond to a stretch at the protein surface that is unresolved in all available crystal structures of human NamPRT (e.g.
 structure visualisation in fig.~\ref{fig:unresolved_loop}B from \cite{Wang2006}).
-The loop is connected to one of the $\beta$-sheets involved in substrate binding \cite{Burgos2009}.
+Intriguingly, this presumed loop, depicted in red in figure~\ref{fig:unresolved_loop}B, is connected to one of the $\beta$-sheets involved in substrate binding \cite{Burgos2009} and in the functional homodimer, the two loops are placed side-by-side.
 
 From these observations, we derived two possible hypotheses regarding the role of the loop in NamPRT function.
-The first hypothesis was that the presence of the loop changes the subcellular localisation of NamPRT, as it is overlapping with a predicted NLS.
+The first hypothesis was that the presence of the loop could change the subcellular localisation of NamPRT, as it is overlapping with a predicted NLS.
 To test this hypothesis, we created a mutant NamPRT lacking the loop and recombinantly expressed FLAG-tagged wildtype and mutant NamPRT in HeLa S3 cells.
-Immunofluorescence imaging showed a mixed cytosolic nuclear localisation for the wildtype and the mutant NamPRT (fig.~\ref{fig:unresolved_loop}C).
-Thus, the presence of the loop does not compromise the nuclear localisation.
+Immunofluorescence imaging showed a mixed cytosolic nuclear localisation for both the wildtype and the mutant NamPRT (fig.~\ref{fig:unresolved_loop}C).
+Thus, deletion of the loop did not compromise the nuclear localisation.
 
-The second hypothesis was that the sequence insertion might influence NamPRT affinity.
+The second hypothesis was that the sequence insertion might influence substrate binding of NamPRT.
 We thus expressed both wildtype and mutant proteins, N-terminally fused to a 6xHis-tag in \textit{E.~coli}, and purified them.
@@
 In contrast to wildtype NamPRT, the activity of the mutant enzyme was not stimulated in the presence of ATP (fig.~\ref{fig:unresolved_loop}E).
-Our observations suggest that the mutant enzyme is catalytically active, retains its dimeric state and sensitivity to FK866.
+These observations suggest that the mutant enzyme is catalytically active, retains its dimeric state and sensitivity to FK866.
 However, the lower activity and the absence of catalytic activation by ATP indicate that deletion of amino acids 42 to 51 might have affected substrate binding.
 
-\subsection{NamPRT and NNMT make NADA obsolete in vertebrates}
+\subsection{NamPRT and NNMT made NADA obsolete in vertebrates}
 
-So far, we did not discuss the effect of NADA on Nam salvage and especially the question, why it is lost in vertebrates.
+Next, we wished to understand why NADA was lost in vertebrates.
 As selection during evolution results usually from competition for resources, we built a two compartment model, based on the pathway model described above.

### 2018-07-13 latex/header.tex
agreed and removed todo from summary

@@
 We conducted extensive phylogenetic analyses of these pathways and show that in lower organisms the initial deamidation of Nam is prevalent, whilst the direct conversion of Nam to the mononucleotide by Nam phosphoribosyltransferase (NamPRT) dominates in animals and eventually remains as the single Nam recycling route in vertebrates.
-Strikingly, loss of the deamidation pathway in early vertebrates is preceded \todo{Statt "preceded", schlägt MZ "accompanied" for.
-Ich (MB) denke, dass preceded richtig ist.} by the emergence of a new enzyme that marks Nam for excretion by methylation – nicotinamide N-methyltransferase (NNMT).
+Strikingly, loss of the deamidation pathway in early vertebrates is preceded by the emergence of a new enzyme that marks Nam for excretion by methylation – nicotinamide N-methyltransferase (NNMT).
 Paradoxically, the occurrence of this Nam degrading enzyme is paralleled by a diversification of NAD dependent signalling enzymes in vertebrates.

### 2018-07-13 latex/introduction.tex
revised introduction resolved todo, included new ones

@@
 Therefore, the Nam recycling by NADA is energetically less efficient under normal conditions than the recycling pathway starting with NamPRT.
-\todo{MZ: Hier müssten Fragen/Probleme aufgeworfen werden, z.B.: Wieso sind die beiden Möglichkeiten offenbar evolutionär getrennt?
-Warum gibt es ein weiteres Enzym, NNMT, das Nam "abbaut"?
-...}
 
-We here present a phylogenetic analysis of the NAD pathways.
+We and others have earlier shown that the two pathway co-exists in some eukaryotes\todo{Insert Refs}, as well as in some bacterial species\todo{MB: please check in the respective bacterial artical if they reported the co-eixtence.
+I think they did.}.
+But why we observe such a scattered distribution of the two pathways is still unknown.
+We furthermore have little understanding of the physiological role of NNMT and its impact on NAD-metabolism so far.
+
+As earlier analysis have been limited by the few eukaryotic genomes available at the time, we here performed a comprehensive phylogenetic analysis of the NAD pathways using ...
+eukaryotic and ...
+prokaryotic genomes \todo{include numbers}.
 Our results suggest that there has been a selection for the co-existence of NamPRT and NNMT in deuterostomes, while the deamidation pathway, which is dominant in bacteria, is lost.
 This transition was accompanied by a marked increase in the number of NAD-consuming signalling enzymes.
-Mathematical modelling of the pathway revealed an unexpected positive kinetic role of NNMT in the maintenance of high NAD-consuming signalling fluxes by preventing accumulation of inhibitory Nam.
+Mathematical modelling of the pathway revealed an unexpected positive kinetic role of NNMT in the maintenance of high NAD-consuming signalling fluxes, preventing accumulation of inhibitory Nam.
 In addition, the model predicts that NNMT likely exerted an evolutionary pressure on NamPRT to develop a high affinity towards its substrate Nam.

### 2018-07-13 latex/discussion.tex
revised results and discussion

@@
 
-\todo[inline]{MZ: Ist Geschmackssache. Ich bin ja mehr so der Typ: "Three major observations have emerged from the present study. 1., 2., 3..." Aber es geht natürlich auch so wie es steht.}
-
 NAD-consuming enzymes are involved in a wide variety of signalling and gene regulatory mechanisms that, due to their sensitivity to NAD$^{+}$, have the ability to translate differences in metabolic states into changes in signalling and gene regulation.
-As NAD concentrations are lowered by the removal of NAD precursor by NNMT, a high-affinity enzyme for recycling of Nam is required to ensure maintenance for cellular NAD levels for other metabolic processes.
-It is unclear, why NamPRT developed a higher affinity and not NADA.
-NADA may have thermodynamic or mechanistic limitations, other environmental conditions may apply, or it might just have happened by chance \todo{MZ: Irgendwo vorher steht, dass dieser Weg energetisch ungünstiger ist.
-Wäre das nicht zumindest eine überlegenswerte Erklärung?}.
-Whatever the reason, we show in our simulations that a high affinity of NamPRT is required for high NAD consumption fluxes and NAD concentrations in the presence of NNMT.
+As NAD concentrations are lowered by the removal of NAD precursor by NNMT,  a high affinity of NamPRT is required for high NAD consumption fluxes and NAD concentrations in the presence of NNMT.
 It therefore seems plausible that NNMT might have been driving NamPRT evolution.
@@
 To maintain high NAD concentrations, a simultaneous higher expression of NamPRT is required, which is what has been found in some types of cancer \cite{Bi2011,Wang2011}.
-NNMT is only advantageous as long as NamPRT affinity is sufficiently high.
+It is worth noticing, that NNMT is only advantageous as long as NamPRT affinity is sufficiently high.
 This suggests that certain types of cancer expressing NNMT at a high level, would potentially be more susceptible to competitive inhibitors of NamPRT.
@@
 
-Neither the scattered distribution of NamPRT and NADA that is especially pronounced in bacteria \cite{Gazzaniga2009} but that has also been observed in eukaryotes, nor the disappearance of NADA in vertebrates has not been understood earlier.
+Neither the scattered distribution of NamPRT and NADA that is especially pronounced in bacteria \cite{Gazzaniga2009} but that has also been observed in eukaryotes, nor the disappearance of NADA in vertebrates has been understood earlier.
 Our combined phylogenetic-modelling analysis now provides a potential explanation for both observations.
-Using simulated competition between two compartments that share for the same limited source of Nam, we show that the compartment that contains NamPRT and NNMT has a higher steady state NAD concentration and NAD consumption rate than the compartment containing NADA.
+Using simulated competition between two compartments that share the same limited source of Nam, we show that the compartment that contains NamPRT and NNMT can maintain a higher steady state NAD concentration and NAD consumption rate than the compartment containing NADA.
 This is, however, only the case if NamPRT substrate affinity is sufficiently high.
@@
 This in turn would suggest that also the bacterial enzymes do not have a high substrate affinity.
-We have only been able to find substrate affinities for one bacterial NamPRT from \textit{Acinetobacter baylyi} \cite{Sorci2010}.
-This enzyme has a $K_{M}$ of 0.04\,mM for Nam, which is about 10\,000 times higher than the $K_{M}$ of human NamPRT, supporting our hypothesis.
+The substrate affinity meassured for a bacterial NamPRT from \textit{Acinetobacter baylyi} \cite{Sorci2010} is indeed 10\,000 times higher lower ($K_{M}$ of 0.04mM) than that of the   human NamPRT, supporting our hypothesis.
 Other bacterial NamPRTs were shown to be functional \cite{Martin2001,Gerdes2006}, but the substrate affinities have not been determined.
@@
 Furthermore, the presence of the methyl donor \textit{S}-adenosyl methionine (SAM) and its precursor methionine that have been shown to potentially limit the effect of NNMT \cite{Ulanovskaya2013} was not considered here.
-As co-substrate availability might alter the behaviour of the system, these should be included in future analyses.
+As co-substrate availability might alter the behaviour of the system, these should thus be included in future analyses.
 Unfortunately, information about the \textit{in~vivo} concentrations of these co-substrates is currently very limited.
@@
 Even if the completeness would be known, due to the high number of genomes used in this analysis, it is likely that some genes of interest were not sequenced in every genome.
-For our analysis, this means that scattered patterns of few missing genes may be real or stem from an incomplete genome.
+For our analysis, this means that scattered patterns of few missing genes could be real, but are in general thought more likely to stem from an incomplete genome.
 
 The third problem are wrong annotations.
-We tried to avoid these, by only relying on template sequences with confirmed function, wherever possible.
+We tried to avoid these, by only relying, wherever possible,  on template sequences with confirmed function.
 This problem becomes apparent by the fact that in yeast an enzyme named NNMT can be found.

### 2018-07-13 latex/results.tex
revised results and discussion

@@
 
-As measured kinetic parameters of NamPRT are only available for the human enzyme \cite{Burgos2008}, we analysed the potential effect of NamPRT affinity ($K_{M}$) on NAD steady state concentration and NAD consumption flux.
+Kinetic paramters of NamPRT were previously meassured for the human enzyme \cite{Burgos2008} as well as for some bacterial enzymes\todo{insert ref}, the latter having a much lower substrate affinity for Nam.
+We thus analysed the potential effect of NamPRT affinity ($K_{M}$) on NAD steady state concentration and NAD consumption flux.
 In the absence of NNMT, a variation of the substrate affinity of NamPRT for Nam, has very little effect on steady state NAD concentration and NAD consumption flux (fig.~\ref{fig:NamPRT_affinity_Nam}A and~B).
@@
 The NAD concentration is always lower with NNMT (fig.~\ref{fig:NamPRT_affinity_Nam}F).
-
-As pathway dynamics are not solely dependent on one enzyme, we wanted to estimate the impact of the substrate affinity of NNMT that is competing with NamPRT for the same substrate.
-The substrate affinity values found in the human enzymes (indicated by black asterisks in fig.~\ref{fig:optimal_substr_affinities}) are actually optimal with respect to both achievable steady state NAD concentration and consumption fluxes.
-Thus, a further increase of the affinity of NamPRT for Nam or a change in the affinity of NNMT for Nam would not provide any further advantage.
-
-\todo[inline]{MZ: Der obige Absatz ist sicher nicht einfach zu beschreiben, aber so ist es kaum zu verstehen. Mein Versuch ist der folgende Absatz:}
 
@@
 fig.~S4).
-Nevertheless, the enzymatic activity of the mutant enzyme was approximately twice as low than that of the wildtype enzyme (fig.~\ref{fig:unresolved_loop}D).
+Nevertheless, the enzymatic activity of the mutant enzyme was only approximately ...
+percent \todo{insert number} of that of the wildtype enzyme (fig.~\ref{fig:unresolved_loop}D).
 In contrast to wildtype NamPRT, the activity of the mutant enzyme was not stimulated in the presence of ATP (fig.~\ref{fig:unresolved_loop}E).

### 2018-07-13 latex/header.tex
removed alternative title, changed title

@@
 \noindent
-{\huge\sffamily\bfseries NamPRT and NNMT – key drivers of NAD-dependent signalling \par}
-
-\vspace{15mm}
-
-\noindent
-\textit{Alternative title:} Functional interdependencies as driving force for the evolution of NAD-dependent signalling and biosynthesis
+{\huge\sffamily\bfseries NamPRT and NNMT – evolutionary and kinetic drivers of NAD-dependent signalling \par}
 

### 2018-07-14 latex/figurelegends.tex
revised figure legends and materials and methods

--- 2018-07-10 latex/figurelegends.tex
@@
   \includegraphics[width=\textwidth]{fig_5_optimal_substr_affinities}
-  \caption{\textbf{The substrate affinities of human NNMT and NamPRT are optimal.} We simulated the impact of changes in the $K_{M}$ for both NamPRT and NNMT on NAD consumption rates (A) and NAD concentration (B).
+  \caption{\textbf{The substrate affinities of human NNMT and NamPRT are close to optimum.} We simulated the impact of changes in the $K_{M}$ for both NamPRT and NNMT on NAD consumption rates (A) and NAD concentration (B).
 Both are increasing with decreasing $K_{M}$ of NamPRT but increasing $K_{M}$ of NNMT.
-The affinities reported for human enzymes (indicated by a black asterisk) appear to be close to optimal, as further improvements would have little or no effect on NAD consumption or concentration.}
+The affinities reported for human enzymes (indicated by a black asterisk) appear to be close to the theoretical optimum, as further improvements would have little effect on NAD consumption or concentration.}
   \label{fig:optimal_substr_affinities}
@@
   \caption{\textbf{The function of the structurally unresolved loop of NamPRT.} Most deuterostomes that possess NamPRT and NNMT show a sequence insertion in the N-terminal region of NamPRT that has been revealed by multiple sequence alignment of NamPRT from different species (A).
-Coloured circles indicate the enzymes present in the respective species blue: NamPRT and NNMT; black: NamPRT, NADA and NNMT; yellow: NamPRT and NADA.
+Coloured circles indicate the enzymes present in the respective species; blue: NamPRT and NNMT; black: NamPRT, NADA and NNMT; yellow: NamPRT and NADA.
 For a more comprehensive alignment, please see supplementary figure~S1.

### 2018-07-14 latex/methods.tex
revised figure legends and materials and methods

@@
 \subsection{Phylogenetic Analysis}
-
-NADA, NamPRT, and NNMT enzymes or enzyme candidates were identified using Blastp with known protein sequences against the NCBI non-redundant protein sequence database.
-A list of functionally verified enzymes used as templates is given in supplementary table~S1.
+Functionally verified NNMT, NADA, NamPRT and NAD-consuming enzyme sequences were used (supplementary table~S1) as sequence templates for a Blastp analysis against the NCBI non-redundant protein sequence database.
 Blastp parameters were set to yield maximum 20\,000 target sequences, using the BLOSUM62 matrix with a word size of 6 and gap opening and extension costs of 11 and 1, respectively.
@@
 Parameter values from mammals were used if available.
-For enzymes not present in mammals, values from yeast were used.
+For enzymes not present in mammals, values from yeast were integrated.
 The full list of kinetic parameters including reference to original literature can be found in supplementary table~S2.
 For NMNAT, the previously developed rate law for substrate competition was used \cite{Schauble2013}.
-Despite these modifications, Henri-Michaelis-Menten kinetics were used for all reactions except the import and efflux of Nam, which were simulated using constant flux and mass action kinetics, respectively.
-All simulations were performed using the steady state calculation and parameter scan options provided by COPASI 4.22 \cite{Hoops2006}.
+Otherwise, Henri-Michaelis-Menten kinetics were applied for all reactions except the import and efflux of Nam, which were simulated using constant flux and mass action kinetics, respectively.
+Steady state calculation and parameter scan tasks provided by COPASI 4.24 \cite{Hoops2006} were used for all simulations.
 The model will be available at the Biomodels database upon publication.
@@
 
-\subsection{Generation of expression vectors encoding encoding wild-type and mutant human NamPRT}
+\subsection{Generation of expression vectors encoding wild-type and mutant human NamPRT}
 
 For eukaryotic expression with a C-terminal FLAG-epitope, the open reading frame (ORF) encoding human NamPRT was inserted into pFLAG-CMV-5a (Merck - Sigma Aldrich) via EcoRI/BamHI sites.
-Using a PCR-based approach, this vector provided the basis for the generation of a plasmid encoding a NamPRT deletion mutant lacking amino acid residues 42-51 ($\Delta$42-51 NamPRT).
+Using a PCR-approach, this vector provided the basis for the generation of a plasmid encoding a NamPRT deletion mutant lacking amino acid residues 42-51 ($\Delta$42-51 NamPRT).
 For prokaryotic expression with an N-terminal 6xHis-tag, the wild-type and mutant ORFs were inserted into pQE-30 (Qiagen) via BamHI and PstI-sites.

### 2018-07-14 latex/supplementaries.tex
Added new fig S2 - Birds based on Prum 2015

@@
 
-\textbf{The phylogenetic distribution of NamPRT and NNMT in birds and reptiles is scattered.}
+\textbf{The phylogenetic distribution of NamPRT and NNMT in birds and reptiles is scattered.} The phylogenetic distribution of birds and reptiles was adopted from \cite{Prum2015}.
+Families are marked with a green circle if they possess NamPRT without NNMT or a blue circle if they possess both NamPRT and NNMT.
 

### 2018-07-14 figures/NamPRT_activity/make_figure.py
Updated NamPRT activity in the unresolved loop figure

@@ -91,7 +91,7 @@ ax2.legend((wt_bar2[0], mu_bar2[0]), majorlabel, loc='upper left', frameon=False
 
 ax1.xaxis.grid(False)
 ax1.set_xlim(0, 2)
-ax1.set_ylim(0, 0.03)
+ax1.set_ylim(0, 0.014)
 ax1.set_xticks(ind1 + width + 0.165)
 ax1.set_xticklabels(minorlabel1)
 ax1.set_ylabel('kcat (1/s)')
@@ -100,7 +100,7 @@ ax1.yaxis.set_ticks_position('left')
 
 ax2.xaxis.grid(False)
 ax2.set_xlim(0, 2)
-ax2.set_ylim(0, 3)
+ax2.set_ylim(0, 2.5)
 ax2.set_xticks(ind2 + width + 0.165)
 ax2.set_xticklabels(minorlabel2)
 ax2.set_ylabel('kcat (rel. to mean WT w/o ATP)')

### 2018-07-14 latex/introduction.tex
Solved all todos :)

@@
 
-We and others have earlier shown that the two pathway co-exists in some eukaryotes\todo{Insert Refs}, as well as in some bacterial species\todo{MB: please check in the respective bacterial artical if they reported the co-eixtence.
-I think they did.}.
+We and others have earlier shown that the two pathway co-exists in some eukaryotes \cite{Gossmann2012FEBS,Carneiro2013}, as well as in some bacterial species \cite{Gazzaniga2009}.
 But why we observe such a scattered distribution of the two pathways is still unknown.
@@
 
-As earlier analysis have been limited by the few eukaryotic genomes available at the time, we here performed a comprehensive phylogenetic analysis of the NAD pathways using ...
-eukaryotic and ...
-prokaryotic genomes \todo{include numbers}.
+As earlier analyses have been limited by the few eukaryotic genomes available at the time, we here performed a comprehensive phylogenetic analysis of the NAD pathways using 793 eukaryotic and 7892 prokaryotic genomes.
 Our results suggest that there has been a selection for the co-existence of NamPRT and NNMT in deuterostomes, while the deamidation pathway, which is dominant in bacteria, is lost.

### 2018-07-14 latex/results.tex
Solved all todos :)

@@
 
-\todo[inline]{MZ: Ich finde, die NAD-abhängigen Signalwege kommen in den Resultaten zu kurz weg.
+% \to%do[inline]{MZ: Ich finde, die NAD-abhängigen Signalwege kommen in den Resultaten zu kurz weg.
 Es genügt vielleicht, einfach die 10 Klassen noch mal aufzuzählen (also z.B.
@@
 fig.~S2).
-\todo{figure not corresponding to current species tree \cite{Prum2015}.
-MB: Will recreate this figure with the tree from the paper asap.} The lack of NNMT might be related to the excretion system, as the product of NNMT, methyl-Nam, is in mammals excreted with the urine.
+The lack of NNMT might be related to the excretion system, as the product of NNMT, methyl-Nam, is in mammals excreted with the urine.
 There are few metazoan species for which we could not find NamPRT or NADA, while NNMT was present.
@@
 
-Kinetic paramters of NamPRT were previously meassured for the human enzyme \cite{Burgos2008} as well as for some bacterial enzymes\todo{insert ref}, the latter having a much lower substrate affinity for Nam.
+Kinetic paramters of NamPRT were previously meassured for the human enzyme \cite{Burgos2008} as well as for some bacterial enzymes \cite{Sorci2010}, the latter having a much lower substrate affinity for Nam.
 We thus analysed the potential effect of NamPRT affinity ($K_{M}$) on NAD steady state concentration and NAD consumption flux.
@@
 fig.~S4).
-Nevertheless, the enzymatic activity of the mutant enzyme was only approximately ...
-percent \todo{insert number} of that of the wildtype enzyme (fig.~\ref{fig:unresolved_loop}D).
+Nevertheless, the enzymatic activity of the mutant enzyme was only approximately 30\% of that of the wildtype enzyme (fig.~\ref{fig:unresolved_loop}D).
 In contrast to wildtype NamPRT, the activity of the mutant enzyme was not stimulated in the presence of ATP (fig.~\ref{fig:unresolved_loop}E).

### 2018-07-16 latex/discussion.tex
Solved all todos :)
Incorporated changes by Toni

@@
 
-Our findings shed new light on the potential physiological role of NNMT, which has earlier been recognized as potential marker for some types of cancer (e.g.
-\cite{Okamura1998}).
-The main healthy tissue expressing NNMT is the liver, while no or only little expression of NNMT is observed in most other healthy tissues \cite{Aksoy1994}.
+Our findings shed new light on the potential physiological role of NNMT, which has earlier been recognized as potential marker for some types of cancer \citep[e.g.][]{Okamura1998}.
+The main healthy tissue expressing NNMT is the liver, while no or only little expression of NNMT is observed in most other healthy tissues \citep{Aksoy1994}.
 The increased NNMT expression observed in some types of cancer, might serve to remove Nam derived by increased NAD-dependent signalling.
-To maintain high NAD concentrations, a simultaneous higher expression of NamPRT is required, which is what has been found in some types of cancer \cite{Bi2011,Wang2011}.
+To maintain high NAD concentrations, a simultaneous higher expression of NamPRT is required, which is what has been found in some types of cancer \citep{Bi2011,Wang2011}.
 It is worth noticing, that NNMT is only advantageous as long as NamPRT affinity is sufficiently high.
 This suggests that certain types of cancer expressing NNMT at a high level, would potentially be more susceptible to competitive inhibitors of NamPRT.
-Several of such inhibitors are currently tested in clinical studies \cite{EspindolaNetto2017,Xu2015SR}.
+Several of such inhibitors are currently tested in clinical studies \citep{EspindolaNetto2017,Xu2015SR}.
 Based on our analysis, we would suggest that it might be reasonable to screen patients before treatment, as non-NNMT expressing tumours might respond less to competitive NamPRT inhibitors and missing Nam degradation in those cancer cells would potentially lead to an accumulation of Nam that could outcompete the inhibitor.
@@
 
-Neither the scattered distribution of NamPRT and NADA that is especially pronounced in bacteria \cite{Gazzaniga2009} but that has also been observed in eukaryotes, nor the disappearance of NADA in vertebrates has been understood earlier.
+Neither the scattered distribution of NamPRT and NADA that is especially pronounced in bacteria \citep{Gazzaniga2009}, but that has also been observed in eukaryotes, nor the disappearance of NADA in vertebrates has been understood earlier.
 Our combined phylogenetic-modelling analysis now provides a potential explanation for both observations.
@@
 We therefore manually checked bacteria that possess NamPRT and indeed found that most of them have been characterised to be pathogenic.
-It should be noted that the distribution of NADA and NamPRT does not follow the bacterial species tree \cite{Gazzaniga2009}.
+It should be noted that the distribution of NADA and NamPRT does not follow the bacterial species tree \citep{Gazzaniga2009}.
 Besides the suggestion made here, there might well be other environmental aspects that influence the phylogenetic distribution in bacteria.
 
-A detailed analysis of sequence variances in NamPRT revealed that only deuterostomes that have NNMT but not NADA, have a sequence insertion in the N-terminal part of NamPRT that seems to enable the high affinity of the enzyme.
+A detailed analysis of sequence variances in NamPRT revealed that only deuterostomes that have NNMT, but not NADA, have a sequence insertion in the N-terminal part of NamPRT that seems to enable the high affinity of the enzyme.
 This in turn would suggest that also the bacterial enzymes do not have a high substrate affinity.
-The substrate affinity meassured for a bacterial NamPRT from \textit{Acinetobacter baylyi} \cite{Sorci2010} is indeed 10\,000 times higher lower ($K_{M}$ of 0.04mM) than that of the   human NamPRT, supporting our hypothesis.
-Other bacterial NamPRTs were shown to be functional \cite{Martin2001,Gerdes2006}, but the substrate affinities have not been determined.
+The substrate affinity meassured for a bacterial NamPRT from \textit{Acinetobacter baylyi} \citep{Sorci2010} is indeed 10\,000 times higher lower ($K_{M}$ of 0.04mM) than that of the human NamPRT, supporting our hypothesis.
+Other bacterial NamPRTs were shown to be functional \citep{Martin2001,Gerdes2006}, but the substrate affinities have not been determined.
 The differences in activity and affinity, implying differences in substrate binding could potentially be exploited for the development of antibiotics.
@@
 Such co-substrates include targets of the NAD-consuming enzymes, such as acylated proteins for sirtuins, for example, or phosphoribosyl pyrophosphate (PRPP) and ATP that are required for NMN synthesis by NamPRT.
-Furthermore, the presence of the methyl donor \textit{S}-adenosyl methionine (SAM) and its precursor methionine that have been shown to potentially limit the effect of NNMT \cite{Ulanovskaya2013} was not considered here.
+Furthermore, the presence of the methyl donor \textit{S}-adenosyl methionine (SAM) and its precursor methionine that have been shown to potentially limit the effect of NNMT \citep{Ulanovskaya2013} was not considered here.
 As co-substrate availability might alter the behaviour of the system, these should thus be included in future analyses.
@@
 During our analysis we came across several problems related to the use of NCBI sequence databases for phylogenetic analyses.
-One is sequence contamination, which is a well known problem \cite{Ballenghien2017,Longo2011}.
+One is sequence contamination, which is a well known problem \citep{Ballenghien2017,Longo2011}.
 To avoid contamination, we used sequence homology analysis to remove all sequences of obvious bacterial origin from the results in eukaryotic species.
 Another problem is incomplete genomes.
-Although there are tools to assess the completeness of a genome (e.g.~\cite{Simao2015}), none of them could convincingly claim to be reliable.
+Although there are tools to assess the completeness of a genome \citep[e.g.][]{Simao2015}, none of them could convincingly claim to be reliable.
 The genomes of the common model organisms can probably be assumed to be close to complete, but there are many draft genomes in the databases whose completeness is uncertain.
@@
 This problem becomes apparent by the fact that in yeast an enzyme named NNMT can be found.
-The initial naming was based on a very weak homology to human NNMT and an analysis of life span extension of the mutant in \textit{Saccharomyces cerevisae} \cite{Anderson2003}, which showed similar effects as other mutants of the NAD pathway.
-The protein has later been shown not to function as methyltransferase for Nam but for the eukaryotic elongation factor~1A (eEF1A) giving it its new name elongation factor methyltransferase~7 (Efm7) \cite{Hamey2016}.
+The initial naming was based on a very weak homology to human NNMT and an analysis of life span extension of the mutant in \textit{Saccharomyces cerevisae} \citep{Anderson2003}, which showed similar effects as other mutants of the NAD pathway.
+The protein has later been shown not to function as methyltransferase for Nam, but for the eukaryotic elongation factor~1A (eEF1A) giving it its new name elongation factor methyltransferase~7 (Efm7) \citep{Hamey2016}.
 The old name is still present in many databases, though.
@@
 A schematic overview is given in figure~\ref{fig:evo_events}.
-This transition appears to be accompanied by the loss of NADA in vertebrates and the first gene duplication of NMNATs \cite{Lau2010}.
+This transition appears to be accompanied by the loss of NADA in vertebrates and the first gene duplication of NMNATs \citep{Lau2010}.
 We also noted that the second gene duplication of NMNATs and thus the further compartmentalisation of NAD metabolism is co-occurring with a site-specific positive selection event in NNMT (unpublished results).

### 2018-07-16 latex/figurelegends.tex
Incorporated changes by Toni

@@
   \caption{\textbf{The substrate affinities of human NNMT and NamPRT are close to optimum.} We simulated the impact of changes in the $K_{M}$ for both NamPRT and NNMT on NAD consumption rates (A) and NAD concentration (B).
-Both are increasing with decreasing $K_{M}$ of NamPRT but increasing $K_{M}$ of NNMT.
+Both are increasing with decreasing $K_{M}$ of NamPRT, but increasing $K_{M}$ of NNMT.
 The affinities reported for human enzymes (indicated by a black asterisk) appear to be close to the theoretical optimum, as further improvements would have little effect on NAD consumption or concentration.}
@@
   \includegraphics[width=\textwidth]{fig_7_NNMT_comp_advantage}
-  \caption{\textbf{NNMT provides a competitive advantage and makes NADA obsolete.} To simulate competition for common resources, we created a two compartment model where one compartment contained NADA but no NamPRT and the other compartment contained NamPRT either with or without NNMT but no NADA.
+  \caption{\textbf{NNMT provides a competitive advantage and makes NADA obsolete.} To simulate competition for common resources, we created a two compartment model where one compartment contained NADA, but no NamPRT and the other compartment contained NamPRT either with or without NNMT, but no NADA.
 NADA and NamPRT were simulated to be present at equal amounts.
-In the absence of NNMT the compartment containing NADA has slightly lower NAD consumption rates (A) but much higher steady state NAD concentrations (B).
+In the absence of NNMT the compartment containing NADA has slightly lower NAD consumption rates (A), but much higher steady state NAD concentrations (B).
 In the presence of NNMT, however, both NAD consumption (C) and NAD concentration (D) are lower in the NADA compartment.

### 2018-07-16 latex/introduction.tex
Incorporated changes by Toni

@@
 NAD metabolism represents one of the most critical links that connect cellular signal transduction and energy metabolism.
-Even though it is best known as cofactor for various redox-reactions, NAD is involved in a number of signalling processes that consume NAD$^{+}$ by cleaving the molecule to nicotinamide (Nam) and ADP-ribose \cite{Verdin2015}.
-These NAD-dependent signalling reactions include poly- and mono-ADP-ribosylation \cite{Butepage2015,DeVos2012}, NAD-dependent protein deacylation by sirtuins \cite{Osborne2016}, and the synthesis of calcium-mobilizing molecules such as cyclic ADP-ribose \cite{Lee2012}.
+Even though it is best known as cofactor for various redox-reactions, NAD is involved in a number of signalling processes that consume NAD$^{+}$ by cleaving the molecule to nicotinamide (Nam) and ADP-ribose \citep{Verdin2015}.
+These NAD-dependent signalling reactions include poly- and mono-ADP-ribosylation \citep{Butepage2015,DeVos2012}, NAD-dependent protein deacylation by sirtuins \citep{Osborne2016}, and the synthesis of calcium-mobilizing molecules such as cyclic ADP-ribose \citep{Lee2012}.
 These NAD-dependent signalling processes participate in the regulation of virtually all cellular activities.
-The enzymes involved in these processes are sensitive to the available NAD concentration \cite{Ruggieri2015}, which in turn is dependent on the NAD$^{+}$/NADH redox ratio.
-Therefore, NAD-dependent signalling can act as a transmitter of changes in the cellular energy homeostasis, for example, to regulate gene expression or metabolic activity \cite{Koch-Nolte2009}.
+The enzymes involved in these processes are sensitive to the available NAD concentration \citep{Ruggieri2015}, which in turn is dependent on the NAD$^{+}$/NADH redox ratio.
+Therefore, NAD-dependent signalling can act as a transmitter of changes in the cellular energy homeostasis, for example, to regulate gene expression or metabolic activity \citep{Koch-Nolte2009}.
 
 The significance of NAD-dependent signalling for NAD homeostasis has long been underestimated.
-It has now been established, however, that substances affecting NAD biosynthesis lead to a rapid decline of the NAD concentration \cite{Buonvicino2018}.
+It has now been established, however, that substances affecting NAD biosynthesis lead to a rapid decline of the NAD concentration \citep{Buonvicino2018}.
 This suggests that NAD-dependent signalling reactions consume substantial amounts of NAD.
 Therefore, we later refer to them also as NAD-consuming reactions.
-The resulting NAD turnover differs in a cell-type-specific manner and can lead to an NAD half-life as short as two hours \cite{Liu2018}.
+The resulting NAD turnover differs in a cell-type-specific manner and can lead to an NAD half-life as short as two hours \citep{Liu2018}.
 To maintain the NAD concentration at physiological levels, NAD biosynthesis needs to act at an equally rapid rate.
-Imbalances in NAD homeostasis have been linked to various mainly age related diseases, such as diabetes, neurodegenerative disorders, and cancer \cite{Chiarugi2012,Verdin2015}.
-Several recent studies have demonstrated impressive health benefits of dietary supplementation with intermediates of NAD biosynthesis including Nam mononucleotide (NMN) and Nam riboside (NR) \cite{Yoshino2018}.
-Apparently, the exploitation of NAD biosynthetic routes, in addition to the use of nicotinamide as precursor (fig.~\ref{fig:pathway_overview}), results in increased NAD concentrations that stimulate NAD-dependent signalling processes, in particular, protein deacetylation by sirtuins \cite{North2004}.
+Imbalances in NAD homeostasis have been linked to various mainly age related diseases, such as diabetes, neurodegenerative disorders, and cancer \citep{Chiarugi2012,Verdin2015}.
+Several recent studies have demonstrated impressive health benefits of dietary supplementation with intermediates of NAD biosynthesis including Nam mononucleotide (NMN) and Nam riboside (NR) \citep{Yoshino2018}.
+Apparently, the exploitation of NAD biosynthetic routes, in addition to the use of nicotinamide as precursor (fig.~\ref{fig:pathway_overview}), results in increased NAD concentrations that stimulate NAD-dependent signalling processes, in particular, protein deacetylation by sirtuins \citep{North2004}.
 
 Due to the constant release of Nam through NAD-consuming signalling reactions, the NAD salvage pathway using Nam as precursor is the most important NAD synthesis pathway.
-If Nam were not continuously recycled into NAD, humans would require a much higher daily vitamin B3 intake than the 16\,mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}.
+If Nam were not continuously recycled into NAD, humans would require a much higher daily vitamin B3 intake than the 16\,mg that are the current daily recommendation \citep{CommissionofEuropeanCommunities2008}.
 Two principal pathways exist that recycle Nam.
 Firstly, vertebrates use a direct two step pathway starting with the conversion of Nam into the mononucleotide NMN by the Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate.
-The nearly complete recycling of Nam by NamPRT is achieved by an extraordinary high substrate affinity to Nam, the $K_{M}$ being in the low nanomolar range \cite{Burgos2008}.
-This appears to be mediated by an ATP-dependent phosphorylation of a histidine residue in the catalytic core \cite{Burgos2009}.
+The nearly complete recycling of Nam by NamPRT is achieved by an extraordinary high substrate affinity to Nam, the $K_{M}$ being in the low nanomolar range \citep{Burgos2008}.
+This appears to be mediated by an ATP-dependent phosphorylation of a histidine residue in the catalytic core \citep{Burgos2009}.
 Despite the importance of its salvage, Nam can also be marked for excretion by methylation.
-The presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012FEBS} is among the most enigmatic and counterintuitive features of NAD metabolism.
+The presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \citep{Gossmann2012FEBS} is among the most enigmatic and counterintuitive features of NAD metabolism.
 While NamPRT is seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, NNMT seems to have no metabolic function other than to remove Nam from NAD metabolism.
-It has been suggested that the process potentially acts as a metabolic methylation sink \cite{Pissios2017}.
+It has been suggested that the process potentially acts as a metabolic methylation sink \citep{Pissios2017}.
 
@@
 NA is converted into the corresponding mononucleotide (NAMN), in a reactions performed by the NA-specific phosphoribosyltransferase NAPRT.
-The conversion of both mononucleotides, NMN and NAMN, into their corresponding dinucleotides, NAD and NAAD, is catalysed by the Nam/NA adenylytransferases (NMNATs) that are essential in all organisms \cite{DeFigueiredo2011}.
+The conversion of both mononucleotides, NMN and NAMN, into their corresponding dinucleotides, NAD and NAAD, is catalysed by the Nam/NA adenylytransferases (NMNATs) that are essential in all organisms \citep{DeFigueiredo2011}.
 The recycling pathway via NA finally requires re-amidation of NAAD by NAD synthase.
@@
 
-We and others have earlier shown that the two pathway co-exists in some eukaryotes \cite{Gossmann2012FEBS,Carneiro2013}, as well as in some bacterial species \cite{Gazzaniga2009}.
+We and others have earlier shown that the two pathway co-exists in some eukaryotes \citep{Gossmann2012FEBS,Carneiro2013}, as well as in some bacterial species \citep{Gazzaniga2009}.
 But why we observe such a scattered distribution of the two pathways is still unknown.

### 2018-07-16 latex/methods.tex
Incorporated changes by Toni

@@
 \subsection{Phylogenetic Analysis}
-Functionally verified NNMT, NADA, NamPRT and NAD-consuming enzyme sequences were used (supplementary table~S1) as sequence templates for a Blastp analysis against the NCBI non-redundant protein sequence database.
+Functionally verified sequences of NNMT, NADA, NamPRT, and NAD-consuming enzymes were used (suppl.
+table~S1) as sequence templates for a Blastp analysis against the NCBI non-redundant protein sequence database.
 Blastp parameters were set to yield maximum 20\,000 target sequences, using the BLOSUM62 matrix with a word size of 6 and gap opening and extension costs of 11 and 1, respectively.
@@
 The taxonomy IDs of the species for each enzyme was derived from the accession2taxonomy database provided by NCBI.
-Scripts for creating, analysing, and visualising the phylogenetic tree were written in Python~3.5, using the ETE3 toolkit \cite{Huerta-Cepas2016}.
+Scripts for creating, analysing, and visualising the phylogenetic tree were written in Python~3.5, using the ETE3 toolkit \citep{Huerta-Cepas2016}.
 
@@
 The full list of kinetic parameters including reference to original literature can be found in supplementary table~S2.
-For NMNAT, the previously developed rate law for substrate competition was used \cite{Schauble2013}.
+For NMNAT, the previously developed rate law for substrate competition was used \citep{Schauble2013}.
 Otherwise, Henri-Michaelis-Menten kinetics were applied for all reactions except the import and efflux of Nam, which were simulated using constant flux and mass action kinetics, respectively.
-Steady state calculation and parameter scan tasks provided by COPASI 4.24 \cite{Hoops2006} were used for all simulations.
+Steady state calculation and parameter scan tasks provided by COPASI~4.24 \citep{Hoops2006} were used for all simulations.
 The model will be available at the Biomodels database upon publication.
@@
 For eukaryotic expression with a C-terminal FLAG-epitope, the open reading frame (ORF) encoding human NamPRT was inserted into pFLAG-CMV-5a (Merck - Sigma Aldrich) via EcoRI/BamHI sites.
-Using a PCR-approach, this vector provided the basis for the generation of a plasmid encoding a NamPRT deletion mutant lacking amino acid residues 42-51 ($\Delta$42-51 NamPRT).
+Using a PCR approach, this vector provided the basis for the generation of a plasmid encoding a NamPRT deletion mutant lacking amino acid residues 42-51 ($\Delta$42-51 NamPRT).
 For prokaryotic expression with an N-terminal 6xHis-tag, the wild-type and mutant ORFs were inserted into pQE-30 (Qiagen) via BamHI and PstI-sites.

### 2018-07-16 latex/results.tex
Incorporated changes by Toni

@@
 
-To understand the functional roles and potential interplay between the three known enzymes that use Nam as substrate (NamPRT, NADA and NNMT), we first conducted a comprehensive phylogenetic analysis.
-The phylogenetic distribution of the two enzymes that initiate the two different NAD salvage pathways, NADA and NamPRT is scattered in bacteria \cite{Gazzaniga2009}, while their co-occurrence has been detected in some marine invertebrates \cite{Gossmann2012FEBS}.
+To understand the functional roles and potential interplay between the three known enzymes that use Nam as substrate (NamPRT, NADA and NNMT), we first conducted a comprehensive phylogenetic analysis of these three enzymes.
+The phylogenetic distribution of the two enzymes that initiate the two different NAD salvage pathways, NADA and NamPRT is scattered in bacteria \citep{Gazzaniga2009}, while their co\-/occurrence has been detected in some marine invertebrates \citep{Gossmann2012FEBS}.
 As shown in Figure~\ref{fig:phylo_distribution}A, bacteria, fungi, and plants predominantly possess NADA and only a few of them harbour NamPRT.
 In contrast, Metazoa predominantly lost NADA and have NamPRT together with NNMT.
-NNMT seems to have arisen \textit{de~novo} in the most recent common ancestor of Ecdysozoa and Lophotrochozoa (fig.~\ref{fig:phylo_distribution}B).
+NNMT seems to have arisen \textit{de~novo} or diverged rapidly in the most recent common ancestor of Ecdysozoa and Lophotrochozoa (fig.~\ref{fig:phylo_distribution}B).
 We were unable to find any \textit{NNMT} gene with an e-value below 0.1 in fungi or plants.
@@
 The distribution of NNMT in birds is quite scattered (suppl.
-fig.~S2).
-The lack of NNMT might be related to the excretion system, as the product of NNMT, methyl-Nam, is in mammals excreted with the urine.
+fig.~S2) but could be explained by the fact that numerous bird genes are high in GC content \citep{Hron2015}.
+The lack of NNMT might be related to the differences in the excretion system, as the product of NNMT, methyl-Nam, is in mammals excreted with the urine.
 There are few metazoan species for which we could not find NamPRT or NADA, while NNMT was present.
@@
 In addition to the phylogenetic distribution of the two Nam salvage enzymes NADA and NamPRT, we analysed the phylogenetic diversity of enzymes catalysing NAD-dependent signalling reactions.
-To do so, we used the previously established classification into ten different families of NAD-consuming signalling enzymes \cite{Gossmann2012FEBS}.
+To do so, we used the previously established classification into ten different families of NAD-consuming signalling enzymes \citep{Gossmann2012FEBS}.
 The detailed list of templates used for the phylogenetic analyses can be found in supplementary table~S1.
 The numbers shown in figure~\ref{fig:phylo_distribution}B denote the average number of NAD-dependent signalling enzyme families found in each clade.
-With the exception of Cnidaria and Lophotrochozoa, we find an average of three to four families in protostomes, whereas most deuterostome species have, on average, more than eight families with an increasing diversification of enzymes within some of these families \cite{Gossmann2014DNAR}.
+With the exception of Cnidaria and Lophotrochozoa, we find an average of three to four families in protostomes, whereas most deuterostome species have, on average, more than eight families with an increasing diversification of enzymes within some of these families \citep{Gossmann2014DNAR}.
 
@@
 
-To be able to compare metabolic features of evolutionary quite different systems in our simulations and as we had limited information about expression levels of enzymes, we initially used equal amounts for all enzymes.
+To be able to compare metabolic features of evolutionary quite different systems in our simulations and as we had limited information about species-specific expression levels of enzymes, we initially assumed equal expression rates for all enzymes.
 As we have very few cross species kinetic data, we were furthermore mainly relying on kinetic constants found for human or yeast enzymes.
@@
 
-First, we addressed the unexpected correlation between the selection for co-occurrence of NamPRT and NNMT and an increase in the number of NAD-consuming enzyme families.
+First, we addressed the unexpected correlation between the selection for co\-/occurrence of NamPRT and NNMT and an increase in the number of NAD-consuming enzyme families.
 We calculated steady state NAD concentrations and NAD consumption rates by simulating NAD biosynthesis proceeding via NamPRT in the presence or absence of NNMT.
@@
 
-Kinetic paramters of NamPRT were previously meassured for the human enzyme \cite{Burgos2008} as well as for some bacterial enzymes \cite{Sorci2010}, the latter having a much lower substrate affinity for Nam.
+Kinetic paramters of NamPRT were previously meassured for the human enzyme \citep{Burgos2008} as well as for some bacterial enzymes \citep{Sorci2010}, the latter having a much lower substrate affinity for Nam.
 We thus analysed the potential effect of NamPRT affinity ($K_{M}$) on NAD steady state concentration and NAD consumption flux.
@@
 
-Given that NNMT might have exerted an evolutionary pressure on the development of NamPRT, one would expect to observe adaptations that are reflected in the NamPRT protein sequence arising with the occurrence of NNMT.
+Given that NNMT might have exerted an evolutionary pressure on the development of NamPRT, one would expect to observe adaptations that are reflected in the NamPRT protein sequence arising shortly after the occurrence of NNMT.
 To explore this, we created a multiple sequence alignment.
@@
 These ten amino acids correspond to a stretch at the protein surface that is unresolved in all available crystal structures of human NamPRT (e.g.
-structure visualisation in fig.~\ref{fig:unresolved_loop}B from \cite{Wang2006}).
-Intriguingly, this presumed loop, depicted in red in figure~\ref{fig:unresolved_loop}B, is connected to one of the $\beta$-sheets involved in substrate binding \cite{Burgos2009} and in the functional homodimer, the two loops are placed side-by-side.
+structure visualisation in fig.~\ref{fig:unresolved_loop}B from \citet{Wang2006}).
+Intriguingly, this presumed loop, depicted in red in figure~\ref{fig:unresolved_loop}B, is connected to one of the $\beta$-sheets involved in substrate binding \citep{Burgos2009} and in the functional homodimer, the two loops are placed side-by-side.
 
@@
 The second hypothesis was that the sequence insertion might influence substrate binding of NamPRT.
-We thus expressed both wildtype and mutant proteins, N-terminally fused to a 6xHis-tag in \textit{E.~coli}, and purified them.
+We thus expressed both wildtype and mutant proteins, N\=/terminally fused to a 6xHis-tag in \textit{E.~coli}, and purified them.
 The size exclusion chromatography profile showed that both wildtype and mutant protein were expressed as dimers (see suppl.
@@
 The enzymatic activity was measured by NMR spectroscopy using the detection of NMN produced in the presence and absence of ATP.
-Upon incubation with the NamPRT inhibitor FK866 \cite{Hasmann2003} for 30 minutes, both wildtype and mutant NamPRT did not produce any NMN, suggesting that binding of FK866 is not affected by the mutation (see suppl.
+Upon incubation with the NamPRT inhibitor FK866 \citep{Hasmann2003} for 30 minutes, both wildtype and mutant NamPRT did not produce any NMN, suggesting that binding of FK866 is not affected by the mutation (see suppl.
 fig.~S4).

### 2018-07-16 latex/discussion.tex
Ran Word spell and grammar check...

@@
 This in turn would suggest that also the bacterial enzymes do not have a high substrate affinity.
-The substrate affinity meassured for a bacterial NamPRT from \textit{Acinetobacter baylyi} \citep{Sorci2010} is indeed 10\,000 times higher lower ($K_{M}$ of 0.04mM) than that of the human NamPRT, supporting our hypothesis.
+The substrate affinity measured for a bacterial NamPRT from \textit{Acinetobacter baylyi} \citep{Sorci2010} is indeed 10\,000 times higher lower ($K_{M}$ of 0.04mM) than that of the human NamPRT, supporting our hypothesis.
 Other bacterial NamPRTs were shown to be functional \citep{Martin2001,Gerdes2006}, but the substrate affinities have not been determined.
@@
 
-During our analysis we came across several problems related to the use of NCBI sequence databases for phylogenetic analyses.
-One is sequence contamination, which is a well known problem \citep{Ballenghien2017,Longo2011}.
+During our analysis, we came across several problems related to the use of NCBI sequence databases for phylogenetic analyses.
+One is sequence contamination, which is a well-known problem \citep{Ballenghien2017,Longo2011}.
 To avoid contamination, we used sequence homology analysis to remove all sequences of obvious bacterial origin from the results in eukaryotic species.

### 2018-07-16 latex/figurelegends.tex
Ran Word spell and grammar check...

@@
   \includegraphics[width=\textwidth]{fig_7_NNMT_comp_advantage}
-  \caption{\textbf{NNMT provides a competitive advantage and makes NADA obsolete.} To simulate competition for common resources, we created a two compartment model where one compartment contained NADA, but no NamPRT and the other compartment contained NamPRT either with or without NNMT, but no NADA.
+  \caption{\textbf{NNMT provides a competitive advantage and makes NADA obsolete.} To simulate competition for common resources, we created a two-compartment model where one compartment contained NADA, but no NamPRT and the other compartment contained NamPRT either with or without NNMT, but no NADA.
 NADA and NamPRT were simulated to be present at equal amounts.

### 2018-07-16 latex/introduction.tex
Ran Word spell and grammar check...

@@
 Two principal pathways exist that recycle Nam.
-Firstly, vertebrates use a direct two step pathway starting with the conversion of Nam into the mononucleotide NMN by the Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate.
+Firstly, vertebrates use a direct two-step pathway starting with the conversion of Nam into the mononucleotide NMN by the Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate.
 The nearly complete recycling of Nam by NamPRT is achieved by an extraordinary high substrate affinity to Nam, the $K_{M}$ being in the low nanomolar range \citep{Burgos2008}.
@@
 
-Secondly, in most prokayotes as well as in plants and fungi, a pathway consisting of four steps starting with the deamidation of Nam to nicotinic acid (NA) by the Nam deamidase (NADA) is used.
+Secondly, in most prokaryotes as well as in plants and fungi, a pathway consisting of four steps starting with the deamidation of Nam to nicotinic acid (NA) by the Nam deamidase (NADA) is used.
 (fig.~\ref{fig:pathway_overview}).
@@
 NA is converted into the corresponding mononucleotide (NAMN), in a reactions performed by the NA-specific phosphoribosyltransferase NAPRT.
-The conversion of both mononucleotides, NMN and NAMN, into their corresponding dinucleotides, NAD and NAAD, is catalysed by the Nam/NA adenylytransferases (NMNATs) that are essential in all organisms \citep{DeFigueiredo2011}.
+The conversion of both mononucleotides, NMN and NAMN, into their corresponding dinucleotides, NAD and NAAD, is catalysed by the Nam/NA adenylyltransferases (NMNATs) that are essential in all organisms \citep{DeFigueiredo2011}.
 The recycling pathway via NA finally requires re-amidation of NAAD by NAD synthase.

### 2018-07-16 latex/results.tex
Ran Word spell and grammar check...

@@
 Most NAD-consuming enzymes are inhibited by their product Nam.
-Thus the presence of NNMT enables higher NAD consumption fluxes, by removing excess Nam from the cells.
+Thus, the presence of NNMT enables higher NAD consumption fluxes, by removing excess Nam from the cells.
 At the same time, the high substrate affinity of NamPRT maintains a sufficiently high NAD concentration, although the concentration is, as expected, lower than in the system without NNMT.
 
-Kinetic paramters of NamPRT were previously meassured for the human enzyme \citep{Burgos2008} as well as for some bacterial enzymes \citep{Sorci2010}, the latter having a much lower substrate affinity for Nam.
+Kinetic parameters of NamPRT were previously measured for the human enzyme \citep{Burgos2008} as well as for some bacterial enzymes \citep{Sorci2010}, the latter having a much lower substrate affinity for Nam.
 We thus analysed the potential effect of NamPRT affinity ($K_{M}$) on NAD steady state concentration and NAD consumption flux.
-In the absence of NNMT, a variation of the substrate affinity of NamPRT for Nam, has very little effect on steady state NAD concentration and NAD consumption flux (fig.~\ref{fig:NamPRT_affinity_Nam}A and~B).
+In the absence of NNMT, a variation of the substrate affinity of NamPRT for Nam has very little effect on steady state NAD concentration and NAD consumption flux (fig.~\ref{fig:NamPRT_affinity_Nam}A and~B).
 In the presence of NNMT, however, NAD consumption flux and NAD concentration increases with decreasing $K_{M}$ values of NamPRT (fig.~\ref{fig:NamPRT_affinity_Nam}C and~D).
@@
 We found that most Deuterostomes that possess only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 42 to 51 in the human enzyme.
-This insert overlaps with a predicted weak nuclear localisation signal (NLS), that is lost when the insert is removed.
+This insert overlaps with a predicted weak nuclear localisation signal (NLS) that is lost when the insert is removed.
 These ten amino acids correspond to a stretch at the protein surface that is unresolved in all available crystal structures of human NamPRT (e.g.
@@
 Next, we wished to understand why NADA was lost in vertebrates.
-As selection during evolution results usually from competition for resources, we built a two compartment model, based on the pathway model described above.
+As selection during evolution results usually from competition for resources, we built a two-compartment model, based on the pathway model described above.
 One compartment contains NADA, while the other one contains either NamPRT alone or together with NNMT.

### 2018-07-17 latex/discussion.tex
Some changes from AE to BE

@@
 
-Our findings shed new light on the potential physiological role of NNMT, which has earlier been recognized as potential marker for some types of cancer \citep[e.g.][]{Okamura1998}.
+Our findings shed new light on the potential physiological role of NNMT, which has earlier been recognised as potential marker for some types of cancer \citep[e.g.][]{Okamura1998}.
 The main healthy tissue expressing NNMT is the liver, while no or only little expression of NNMT is observed in most other healthy tissues \citep{Aksoy1994}.

### 2018-07-17 latex/figurelegends.tex
Some changes from AE to BE

@@
   \includegraphics[width=\textwidth]{fig_8_evo_events}
-  \caption{\textbf{Schematic representation of evolutionary events in the NAD pathway.} Based on the phylogenetic analysis presented here (roman font) and earlier work \cite{Lau2010} (italic font) we summarized and indicated important events in the evolution of NAD metabolism in Metazoa.}
+  \caption{\textbf{Schematic representation of evolutionary events in the NAD pathway.} Based on the phylogenetic analysis presented here (roman font) and earlier work \cite{Lau2010} (italic font) we summarised and indicated important events in the evolution of NAD metabolism in Metazoa.}
   \label{fig:evo_events}

### 2018-07-17 latex/results.tex
Some changes from AE to BE

@@
 
-\subsection{Functional properties of NamPRT and NNMT have evolved to maximize NAD-dependent signalling}
+\subsection{Functional properties of NamPRT and NNMT have evolved to maximise NAD-dependent signalling}
 
-To resolve this apparent contradiction, we wished to scrutinize the NAD metabolic network.
+To resolve this apparent contradiction, we wished to scrutinise the NAD metabolic network.
 Given the complexity of this network, we turned to modelling approaches and built a dynamic model of NAD metabolism based on previously reported kinetic data (for details, see experimental procedures and suppl.

### 2018-07-17 latex/header.tex
Solved all todos :)
Incorporated changes by Toni
Some minor changes

@@
 Moreover, we suggest that NNMT exerted evolutionary pressure on NamPRT, enforcing the development of its high substrate affinity.
-Using multiple sequence alignments, we identified a sequence insertion, first occurring in vertebrates, that parallels an experimentally verified increase in the substrate affinity of the enzyme.
+Using multiple sequence alignments, we identified a sequence insertion first occurring in vertebrates that parallels an experimentally verified increase in the substrate affinity of the enzyme.
 Additional simulations show that the deamidation pathway became obsolete owing to the high substrate affinity of NamPRT.
@@
 
-\section*{Keywords}
+%\section*{Keywords}

### 2018-07-20 latex/results.tex
Some minor changes
Final print version for PhD thesis

@@
 To resolve this apparent contradiction, we wished to scrutinise the NAD metabolic network.
-Given the complexity of this network, we turned to modelling approaches and built a dynamic model of NAD metabolism based on previously reported kinetic data (for details, see experimental procedures and suppl.
+Given the complexity of this network, we turned to modelling approaches and built a dynamic model of NAD metabolism based on previously reported kinetic data (for details, see Experimental Procedures and suppl.
 tab.~S2).
@@
 
-To understand the interplay and competition for Nam between NNMT and NamPRT, we scanned a wide range of possible $K_{M}$ values for both enzymes in our simulations.
+To understand the interplay and competition for Nam between NamPRT and NNMT, we scanned a wide range of possible $K_{M}$ values for both enzymes in our simulations.
 As shown in figure~\ref{fig:optimal_substr_affinities}, the simulations indicate that both NAD consumption flux and NAD concentration would be minimal in case of a high $K_{M}$ for NamPRT and a low $K_{M}$ for NNMT.
@@
 One compartment contains NADA, while the other one contains either NamPRT alone or together with NNMT.
-Both compartments share a limited Nam source (for details, see experimental procedures and suppl.
+Both compartments share a limited Nam source (for details, see Experimental Procedures and suppl.
 tab.~S2).
@@
 At high cell division rates, steady state concentrations in both compartments are similar.
-This might explain why in bacteria, that often have relatively high growth rates, both systems can co-exist.
+This might explain why in bacteria that often have relatively high growth rates, both systems can co-exist.
 

### 2018-08-24 latex/discussion.tex
Some minor changes
Reverted changes resulting from xelatex back to pdflatex

@@
 
-Our findings shed new light on the potential physiological role of NNMT, which has earlier been recognised as potential marker for some types of cancer \citep[e.g.][]{Okamura1998}.
-The main healthy tissue expressing NNMT is the liver, while no or only little expression of NNMT is observed in most other healthy tissues \citep{Aksoy1994}.
+Our findings shed new light on the potential physiological role of NNMT, which has earlier been recognised as potential marker for some types of cancer e.g.~\cite{Okamura1998}.
+The main healthy tissue expressing NNMT is the liver, while no or only little expression of NNMT is observed in most other healthy tissues \cite{Aksoy1994}.
 The increased NNMT expression observed in some types of cancer, might serve to remove Nam derived by increased NAD-dependent signalling.
-To maintain high NAD concentrations, a simultaneous higher expression of NamPRT is required, which is what has been found in some types of cancer \citep{Bi2011,Wang2011}.
-It is worth noticing, that NNMT is only advantageous as long as NamPRT affinity is sufficiently high.
+To maintain high NAD concentrations, a simultaneous higher expression of NamPRT is required, which is what has been found in some types of cancer \cite{Bi2011,Wang2011}.
+It is worth noticing that NNMT is only advantageous as long as NamPRT affinity is sufficiently high.
 This suggests that certain types of cancer expressing NNMT at a high level, would potentially be more susceptible to competitive inhibitors of NamPRT.
-Several of such inhibitors are currently tested in clinical studies \citep{EspindolaNetto2017,Xu2015SR}.
+Several of such inhibitors are currently tested in clinical studies \cite{EspindolaNetto2017,Xu2015SR}.
 Based on our analysis, we would suggest that it might be reasonable to screen patients before treatment, as non-NNMT expressing tumours might respond less to competitive NamPRT inhibitors and missing Nam degradation in those cancer cells would potentially lead to an accumulation of Nam that could outcompete the inhibitor.
@@
 
-Neither the scattered distribution of NamPRT and NADA that is especially pronounced in bacteria \citep{Gazzaniga2009}, but that has also been observed in eukaryotes, nor the disappearance of NADA in vertebrates has been understood earlier.
+Neither the scattered distribution of NamPRT and NADA that is especially pronounced in bacteria \cite{Gazzaniga2009}, but that has also been observed in eukaryotes, nor the disappearance of NADA in vertebrates has been understood earlier.
 Our combined phylogenetic-modelling analysis now provides a potential explanation for both observations.
@@
 We therefore manually checked bacteria that possess NamPRT and indeed found that most of them have been characterised to be pathogenic.
-It should be noted that the distribution of NADA and NamPRT does not follow the bacterial species tree \citep{Gazzaniga2009}.
+It should be noted that the distribution of NADA and NamPRT does not follow the bacterial species tree \cite{Gazzaniga2009}.
 Besides the suggestion made here, there might well be other environmental aspects that influence the phylogenetic distribution in bacteria.
@@
 This in turn would suggest that also the bacterial enzymes do not have a high substrate affinity.
-The substrate affinity measured for a bacterial NamPRT from \textit{Acinetobacter baylyi} \citep{Sorci2010} is indeed 10\,000 times higher lower ($K_{M}$ of 0.04mM) than that of the human NamPRT, supporting our hypothesis.
-Other bacterial NamPRTs were shown to be functional \citep{Martin2001,Gerdes2006}, but the substrate affinities have not been determined.
+The substrate affinity measured for a bacterial NamPRT from \textit{Acinetobacter baylyi} \cite{Sorci2010} is indeed 10\,000 times higher lower ($K_{M}$ of 0.04mM) than that of the human NamPRT, supporting our hypothesis.
+Other bacterial NamPRTs were shown to be functional \cite{Martin2001,Gerdes2006}, but the substrate affinities have not been determined.
 The differences in activity and affinity, implying differences in substrate binding could potentially be exploited for the development of antibiotics.
@@
 Such co-substrates include targets of the NAD-consuming enzymes, such as acylated proteins for sirtuins, for example, or phosphoribosyl pyrophosphate (PRPP) and ATP that are required for NMN synthesis by NamPRT.
-Furthermore, the presence of the methyl donor \textit{S}-adenosyl methionine (SAM) and its precursor methionine that have been shown to potentially limit the effect of NNMT \citep{Ulanovskaya2013} was not considered here.
+Furthermore, the presence of the methyl donor \textit{S}-adenosyl methionine (SAM) and its precursor methionine that have been shown to potentially limit the effect of NNMT \cite{Ulanovskaya2013} was not considered here.
 As co-substrate availability might alter the behaviour of the system, these should thus be included in future analyses.
@@
 During our analysis, we came across several problems related to the use of NCBI sequence databases for phylogenetic analyses.
-One is sequence contamination, which is a well-known problem \citep{Ballenghien2017,Longo2011}.
+One is sequence contamination, which is a well-known problem \cite{Ballenghien2017,Longo2011}.
 To avoid contamination, we used sequence homology analysis to remove all sequences of obvious bacterial origin from the results in eukaryotic species.
 Another problem is incomplete genomes.
-Although there are tools to assess the completeness of a genome \citep[e.g.][]{Simao2015}, none of them could convincingly claim to be reliable.
+Although there are tools to assess the completeness of a genome e.g.~\cite{Simao2015}, none of them could convincingly claim to be reliable.
 The genomes of the common model organisms can probably be assumed to be close to complete, but there are many draft genomes in the databases whose completeness is uncertain.
@@
 This problem becomes apparent by the fact that in yeast an enzyme named NNMT can be found.
-The initial naming was based on a very weak homology to human NNMT and an analysis of life span extension of the mutant in \textit{Saccharomyces cerevisae} \citep{Anderson2003}, which showed similar effects as other mutants of the NAD pathway.
-The protein has later been shown not to function as methyltransferase for Nam, but for the eukaryotic elongation factor~1A (eEF1A) giving it its new name elongation factor methyltransferase~7 (Efm7) \citep{Hamey2016}.
+The initial naming was based on a very weak homology to human NNMT and an analysis of life span extension of the mutant in \textit{Saccharomyces cerevisae} \cite{Anderson2003}, which showed similar effects as other mutants of the NAD pathway.
+The protein has later been shown not to function as methyltransferase for Nam, but for the eukaryotic elongation factor~1A (eEF1A) giving it its new name elongation factor methyltransferase~7 (Efm7) \cite{Hamey2016}.
 The old name is still present in many databases, though.
@@
 A schematic overview is given in figure~\ref{fig:evo_events}.
-This transition appears to be accompanied by the loss of NADA in vertebrates and the first gene duplication of NMNATs \citep{Lau2010}.
+This transition appears to be accompanied by the loss of NADA in vertebrates and the first gene duplication of NMNATs \cite{Lau2010}.
 We also noted that the second gene duplication of NMNATs and thus the further compartmentalisation of NAD metabolism is co-occurring with a site-specific positive selection event in NNMT (unpublished results).

### 2018-08-24 latex/introduction.tex
Reverted changes resulting from xelatex back to pdflatex

--- 2018-07-14 latex/introduction.tex
@@
 Two principal pathways exist that recycle Nam.
-Firstly, vertebrates use a direct two step pathway starting with the conversion of Nam into the mononucleotide NMN by the Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate.
+Firstly, vertebrates use a direct two-step pathway starting with the conversion of Nam into the mononucleotide NMN by the Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate.
 The nearly complete recycling of Nam by NamPRT is achieved by an extraordinary high substrate affinity to Nam, the $K_{M}$ being in the low nanomolar range \cite{Burgos2008}.
@@
 NA is converted into the corresponding mononucleotide (NAMN), in a reactions performed by the NA-specific phosphoribosyltransferase NAPRT.
-The conversion of both mononucleotides, NMN and NAMN, into their corresponding dinucleotides, NAD and NAAD, is catalysed by the Nam/NA adenylytransferases (NMNATs) that are essential in all organisms \cite{DeFigueiredo2011}.
+The conversion of both mononucleotides, NMN and NAMN, into their corresponding dinucleotides, NAD and NAAD, is catalysed by the Nam/NA adenylyltransferases (NMNATs) that are essential in all organisms \cite{DeFigueiredo2011}.
 The recycling pathway via NA finally requires re-amidation of NAAD by NAD synthase.

### 2018-08-24 latex/methods.tex
Reverted changes resulting from xelatex back to pdflatex

@@
 The taxonomy IDs of the species for each enzyme was derived from the accession2taxonomy database provided by NCBI.
-Scripts for creating, analysing, and visualising the phylogenetic tree were written in Python~3.5, using the ETE3 toolkit \citep{Huerta-Cepas2016}.
+Scripts for creating, analysing, and visualising the phylogenetic tree were written in Python~3.5, using the ETE3 toolkit \cite{Huerta-Cepas2016}.
 
@@
 The full list of kinetic parameters including reference to original literature can be found in supplementary table~S2.
-For NMNAT, the previously developed rate law for substrate competition was used \citep{Schauble2013}.
+For NMNAT, the previously developed rate law for substrate competition was used \cite{Schauble2013}.
 Otherwise, Henri-Michaelis-Menten kinetics were applied for all reactions except the import and efflux of Nam, which were simulated using constant flux and mass action kinetics, respectively.
-Steady state calculation and parameter scan tasks provided by COPASI~4.24 \citep{Hoops2006} were used for all simulations.
+Steady state calculation and parameter scan tasks provided by COPASI~4.24 \cite{Hoops2006} were used for all simulations.
 The model will be available at the Biomodels database upon publication.

### 2018-08-24 latex/results.tex
Reverted changes resulting from xelatex back to pdflatex

@@
 To understand the functional roles and potential interplay between the three known enzymes that use Nam as substrate (NamPRT, NADA and NNMT), we first conducted a comprehensive phylogenetic analysis of these three enzymes.
-The phylogenetic distribution of the two enzymes that initiate the two different NAD salvage pathways, NADA and NamPRT is scattered in bacteria \citep{Gazzaniga2009}, while their co\-/occurrence has been detected in some marine invertebrates \citep{Gossmann2012FEBS}.
+The phylogenetic distribution of the two enzymes that initiate the two different NAD salvage pathways, NADA and NamPRT is scattered in bacteria \cite{Gazzaniga2009}, while their co-occurrence has been detected in some marine invertebrates \cite{Gossmann2012FEBS}.
 As shown in Figure~\ref{fig:phylo_distribution}A, bacteria, fungi, and plants predominantly possess NADA and only a few of them harbour NamPRT.
@@
 The distribution of NNMT in birds is quite scattered (suppl.
-fig.~S2) but could be explained by the fact that numerous bird genes are high in GC content \citep{Hron2015}.
+fig.~S2) but could be explained by the fact that numerous bird genes are high in GC content \cite{Hron2015}.
 The lack of NNMT might be related to the differences in the excretion system, as the product of NNMT, methyl-Nam, is in mammals excreted with the urine.
@@
 In addition to the phylogenetic distribution of the two Nam salvage enzymes NADA and NamPRT, we analysed the phylogenetic diversity of enzymes catalysing NAD-dependent signalling reactions.
-To do so, we used the previously established classification into ten different families of NAD-consuming signalling enzymes \citep{Gossmann2012FEBS}.
+To do so, we used the previously established classification into ten different families of NAD-consuming signalling enzymes \cite{Gossmann2012FEBS}.
 The detailed list of templates used for the phylogenetic analyses can be found in supplementary table~S1.
 The numbers shown in figure~\ref{fig:phylo_distribution}B denote the average number of NAD-dependent signalling enzyme families found in each clade.
-With the exception of Cnidaria and Lophotrochozoa, we find an average of three to four families in protostomes, whereas most deuterostome species have, on average, more than eight families with an increasing diversification of enzymes within some of these families \citep{Gossmann2014DNAR}.
+With the exception of Cnidaria and Lophotrochozoa, we find an average of three to four families in protostomes, whereas most deuterostome species have, on average, more than eight families with an increasing diversification of enzymes within some of these families \cite{Gossmann2014DNAR}.
 
@@
 
-First, we addressed the unexpected correlation between the selection for co\-/occurrence of NamPRT and NNMT and an increase in the number of NAD-consuming enzyme families.
+First, we addressed the unexpected correlation between the selection for co-occurrence of NamPRT and NNMT and an increase in the number of NAD-consuming enzyme families.
 We calculated steady state NAD concentrations and NAD consumption rates by simulating NAD biosynthesis proceeding via NamPRT in the presence or absence of NNMT.
@@
 
-Kinetic parameters of NamPRT were previously measured for the human enzyme \citep{Burgos2008} as well as for some bacterial enzymes \citep{Sorci2010}, the latter having a much lower substrate affinity for Nam.
+Kinetic parameters of NamPRT were previously measured for the human enzyme \cite{Burgos2008} as well as for some bacterial enzymes \cite{Sorci2010}, the latter having a much lower substrate affinity for Nam.
 We thus analysed the potential effect of NamPRT affinity ($K_{M}$) on NAD steady state concentration and NAD consumption flux.
@@
 These ten amino acids correspond to a stretch at the protein surface that is unresolved in all available crystal structures of human NamPRT (e.g.
-structure visualisation in fig.~\ref{fig:unresolved_loop}B from \citet{Wang2006}).
-Intriguingly, this presumed loop, depicted in red in figure~\ref{fig:unresolved_loop}B, is connected to one of the $\beta$-sheets involved in substrate binding \citep{Burgos2009} and in the functional homodimer, the two loops are placed side-by-side.
+structure visualisation in fig.~\ref{fig:unresolved_loop}B from \cite{Wang2006}).
+Intriguingly, this presumed loop, depicted in red in figure~\ref{fig:unresolved_loop}B, is connected to one of the $\beta$-sheets involved in substrate binding \cite{Burgos2009} and in the functional homodimer, the two loops are placed side-by-side.
 
@@
 The second hypothesis was that the sequence insertion might influence substrate binding of NamPRT.
-We thus expressed both wildtype and mutant proteins, N\=/terminally fused to a 6xHis-tag in \textit{E.~coli}, and purified them.
+We thus expressed both wildtype and mutant proteins, N-terminally fused to a 6xHis-tag in \textit{E.~coli}, and purified them.
 The size exclusion chromatography profile showed that both wildtype and mutant protein were expressed as dimers (see suppl.
@@
 The enzymatic activity was measured by NMR spectroscopy using the detection of NMN produced in the presence and absence of ATP.
-Upon incubation with the NamPRT inhibitor FK866 \citep{Hasmann2003} for 30 minutes, both wildtype and mutant NamPRT did not produce any NMN, suggesting that binding of FK866 is not affected by the mutation (see suppl.
+Upon incubation with the NamPRT inhibitor FK866 \cite{Hasmann2003} for 30 minutes, both wildtype and mutant NamPRT did not produce any NMN, suggesting that binding of FK866 is not affected by the mutation (see suppl.
 fig.~S4).

### 2018-11-15 latex/acknowledgements.tex
Incorporated changes suggested by Dorothée
added additional authors and affiliations, revised introduction

@@
 We thank the Norwegian Research Council for funding (grant no.
-250395/F20).
+250395/F20 and grant no.
+226244/F50).
 We furthermore thank for the computation time provided through UNINETT Sigma2 – the National Infrastructure for High Performance Computing and Data Storage in Norway.
+
+\section*{Author contribution}
+MB and TG performed the phylogenetic analysis, IH performed the mathematical modelling, DH and MN performed the experiments, IR performed the MD analyses guided by AS, MZ and IH were supervising and guiding the investigations.
+All authors contributed to the manuscript preparation..

### 2018-11-15 latex/header.tex
added additional authors and affiliations, revised introduction

@@
 \noindent
-{\huge\sffamily\bfseries NamPRT and NNMT – evolutionary and kinetic drivers of NAD-dependent signalling \par}
+{\huge\sffamily\bfseries Identifying evolutionary and kinetic drivers of NAD-dependent signalling \par}
 
@@
 Mathias Bockwoldt\textsuperscript{1}, Dorothée Houry\textsuperscript{2}, Marc Niere\textsuperscript{2}, Toni I.
-Gossmann\textsuperscript{3}, Mathias Ziegler\textsuperscript{2}, and Ines Heiland\textsuperscript{1,§}
+Gossmann\textsuperscript{3}, Ines Reinartz\textsuperscript{4}, Alexander Schug\textsuperscript{4,5}, Mathias Ziegler\textsuperscript{2}, and Ines Heiland\textsuperscript{1,§}
 
@@
 \textsuperscript{3}Department of Animal and Plant Sciences, Western Bank, University of Sheffield, Sheffield, S10 2TN, United Kingdom
+
+\noindent
+\textsuperscript{4}Karlsruhe Institute of Technology (KIT), Steinbuch Centre for Computing (SCC), Helmholtz Young Investigator Group Leader, Multiscale Biomolecular Simulation, Hermann-von-Helmholtz-Platz 1, 76344 Eggenstein-Leopoldshafen, Germany
+
+\noindent
+\textsuperscript{5}Research Center Juelich,52425 Juelich, Germany
 

### 2018-11-15 latex/introduction.tex
added additional authors and affiliations, revised introduction

@@
 
-NAD metabolism represents one of the most critical links that connect cellular signal transduction and energy metabolism.
+NAD metabolism represents one of the most critical links connecting cellular signal transduction and energy metabolism.
 Even though it is best known as cofactor for various redox-reactions, NAD is involved in a number of signalling processes that consume NAD$^{+}$ by cleaving the molecule to nicotinamide (Nam) and ADP-ribose \cite{Verdin2015}.
@@
 This suggests that NAD-dependent signalling reactions consume substantial amounts of NAD.
-Therefore, we later refer to them also as NAD-consuming reactions.
-The resulting NAD turnover differs in a cell-type-specific manner and can lead to an NAD half-life as short as two hours \cite{Liu2018}.
+Therefore, we hereafter refer to them also as NAD-consuming reactions.
+The resulting NAD turnover differs in a cell-type-specific manner and can lead to an NAD half-life as short as 15 minutes \cite{Liu2018}.
 To maintain the NAD concentration at physiological levels, NAD biosynthesis needs to act at an equally rapid rate.
@@
 Due to the constant release of Nam through NAD-consuming signalling reactions, the NAD salvage pathway using Nam as precursor is the most important NAD synthesis pathway.
-If Nam were not continuously recycled into NAD, humans would require a much higher daily vitamin B3 intake than the 16\,mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}.
-Two principal pathways exist that recycle Nam.
-Firstly, vertebrates use a direct two-step pathway starting with the conversion of Nam into the mononucleotide NMN by the Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate.
+In general, two principal pathways exist that recycle Nam.
+Vertebrates use a direct two-step pathway starting with the conversion of Nam into the mononucleotide NMN by the Nam phosphoribosyltransferase (NamPT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate.
 The nearly complete recycling of Nam by NamPRT is achieved by an extraordinary high substrate affinity to Nam, the $K_{M}$ being in the low nanomolar range \cite{Burgos2008}.
@@
 
-Secondly, in most prokaryotes as well as in plants and fungi, a pathway consisting of four steps starting with the deamidation of Nam to nicotinic acid (NA) by the Nam deamidase (NADA) is used.
+In most prokaryotes as well as in plants and fungi, another pathway consisting of four steps starting with the deamidation of Nam to nicotinic acid (NA) by the Nam deamidase (NADA) is used.
 (fig.~\ref{fig:pathway_overview}).
@@
 This final reaction includes an enzyme adenylation step that consumes ATP.
-Therefore, the Nam recycling by NADA is energetically less efficient under normal conditions than the recycling pathway starting with NamPRT.
+Therefore, the Nam recycling by NADA appears to be energetically less efficient than the recycling pathway starting with NamPT.
 
-We and others have earlier shown that the two pathway co-exists in some eukaryotes \cite{Gossmann2012FEBS,Carneiro2013}, as well as in some bacterial species \cite{Gazzaniga2009}.
-But why we observe such a scattered distribution of the two pathways is still unknown.
+We and others have earlier shown that the two NAD biosynthesis pathways starting from Nam co-exists in some eukaryotes \cite{Gossmann2012FEBS,Carneiro2013}, as well as in some bacterial species \cite{Gazzaniga2009}.
+Why these pathways co-exists over in some organisms and over a very long evolutionary time frame and why NADA nevertheless disappears in vertebrates, is not known.
 We furthermore have little understanding of the physiological role of NNMT and its impact on NAD-metabolism so far.
@@
 As earlier analyses have been limited by the few eukaryotic genomes available at the time, we here performed a comprehensive phylogenetic analysis of the NAD pathways using 793 eukaryotic and 7892 prokaryotic genomes.
-Our results suggest that there has been a selection for the co-existence of NamPRT and NNMT in deuterostomes, while the deamidation pathway, which is dominant in bacteria, is lost.
+Our results suggest that there has been a selection for the co-existence of NamPRT and NNMT in deuterostomes, while the deamidation pathway, which is dominant in bacteria, became superfluous.
 This transition was accompanied by a marked increase in the number of NAD-consuming signalling enzymes.
-Mathematical modelling of the pathway revealed an unexpected positive kinetic role of NNMT in the maintenance of high NAD-consuming signalling fluxes, preventing accumulation of inhibitory Nam.
-In addition, the model predicts that NNMT likely exerted an evolutionary pressure on NamPRT to develop a high affinity towards its substrate Nam.
-Indeed, we identified a short sequence insertion in NamPRT, which first occurs in Deuterostomes and that appears to modulate the affinity of NamPRT.
-Simulating the resource competition, we furthermore show that the presence of high affinity NamPRT together with NNMT makes the NADA-dependent pathway obsolete.
+Mathematical modelling of the pathway revealed an unexpected positive kinetic role of NNMT for NAD-consuming signalling fluxes, through prevention of accumulation of Nam.
+In addition, the model predicts that NNMT likely exerted an evolutionary pressure on NamPT to develop a high affinity towards its substrate Nam.
+Indeed, we identified a short sequence insertion in NamPT, which first occurs in Deuterostomes and that appears to modulate the affinity of NamPT.
+Simulating the resource competition, we furthermore show that the presence of high affinity NamPT together with NNMT makes the NADA-dependent pathway obsolete.
 
-Taken together, our analyses suggest that the co-existence of NamPRT and NNMT has been a prerequisite to enable the evolutionary development of versatile NAD-dependent signalling mechanisms present in vertebrates.
+Taken together, our analyses suggest that the co-existence of NamPT and NNMT has been a prerequisite to enable the evolutionary development of versatile NAD-dependent signalling mechanisms present in vertebrates.

### 2018-11-17 latex/results.tex
started to revise results section, inserted md simulation part

@@
 First, we addressed the unexpected correlation between the selection for co-occurrence of NamPRT and NNMT and an increase in the number of NAD-consuming enzyme families.
-We calculated steady state NAD concentrations and NAD consumption rates by simulating NAD biosynthesis proceeding via NamPRT in the presence or absence of NNMT.
+We calculated steady state NAD concentrations and NAD consumption fluxes by simulating NAD biosynthesis proceeding via NamPRT in the presence or absence of NNMT.
 To achieve free NAD concentrations in the range reported in the literature and due to the very low turnover of NamPRT, we used tenfold higher NamPRT levels compared to other enzymes.
 We also adjusted the amount of NMNAT accordingly to avoid that the NAD synthesis rates are limited by this enzyme.
-Surprisingly, as shown in figure~\ref{fig:NNMT_NAD_flux}, the presence of NNMT enables higher rather than lower NAD consumption rates (fig.~\ref{fig:NNMT_NAD_flux}A).
+Surprisingly, as shown in figure~\ref{fig:NNMT_NAD_flux}, the presence of NNMT enables higher rather than lower NAD consumption fluxes (fig.~\ref{fig:NNMT_NAD_flux}A).
 However, it diminishes the steady state concentration of NAD (fig.~\ref{fig:NNMT_NAD_flux}B).
@@
 Thus, the presence of NNMT enables higher NAD consumption fluxes, by removing excess Nam from the cells.
-At the same time, the high substrate affinity of NamPRT maintains a sufficiently high NAD concentration, although the concentration is, as expected, lower than in the system without NNMT.
+At the same time, high substrate affinity of NamPRT maintains a sufficiently high NAD concentration, although the concentration is, as expected, lower than in the system without NNMT.
 
@@
 Of course, this is an artificial scenario, as one would assume organisms to regulate enzyme expression to achieve similar levels of metabolite concentrations instead.
-Nevertheless, in the absence of NNMT there seems to be a trade-off between maintainable NAD concentration and consumption flux.
-In contrast, in the presence of NNMT, NAD consumption rates and concentrations are almost independent of cell division rates.
+In addition, there seems to be ad trade-off  between maintainable NAD concentration and consumption flux, in the absence of NNMT.
+In the presence of NNMT, however, NAD consumption rates and concentrations are almost independent of cell division rates.
 
@@
 As shown in figure~\ref{fig:optimal_substr_affinities}, the simulations indicate that both NAD consumption flux and NAD concentration would be minimal in case of a high $K_{M}$ for NamPRT and a low $K_{M}$ for NNMT.
-Conversely, lowering the $K_{M}$ of NamPRT to the nanomolar range substantially increases NAD consumption and concentration, which reach a maximum when the $K_{M}$ of NNMT is concomitantly elevated to the submillimolar range.
+Conversely, lowering the $K_{M}$ of NamPRT to the nanomolar range substantially increases NAD consumption and concentration, which reach a plateau when the $K_{M}$ of NNMT is concomitantly elevated to the submillimolar range.
 The asterisks in figure~\ref{fig:optimal_substr_affinities} denote the $K_{M}$ values actually found for the human enzymes.
-Astonishingly, the naturally occurring $K_{M}$ values are very close to the theoretical optimum.
+Astonishingly, the naturally occurring $K_{M}$ values are in a range where further adjustment would lead to little or no increase of NAD consumption flux or NAD concentration.
 
@@
 
-Given that NNMT might have exerted an evolutionary pressure on the development of NamPRT, one would expect to observe adaptations that are reflected in the NamPRT protein sequence arising shortly after the occurrence of NNMT.
-To explore this, we created a multiple sequence alignment.
-The alignment of selected sequences is shown in figure~\ref{fig:unresolved_loop}A and a more comprehensive multiple sequence alignment containing a larger number of species can be found in supplementary figure~S1.
+Given the kinetic interdeence of  NNMT and NamPRT shown in figure fig.~\ref{fig:NamPRT_affinity_Nam and \ref{fig:optimal_substr_affinities},it is possible that NNMT has exerted an evolutionary pressure on the development of NamPRT.
+In this case, one would expect to observe adaptations that are reflected in the NamPRT protein sequence arising shortly after the occurrence of NNMT.
+To explore this, we created a multiple sequence alignment of NamPRT protein sequences from metazoa.
+An alignment of selected sequences is shown in figure~\ref{fig:unresolved_loop}A and a more comprehensive multiple sequence alignment containing a larger number of species can be found in supplementary figure~S1.
 We found that most Deuterostomes that possess only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 42 to 51 in the human enzyme.
-This insert overlaps with a predicted weak nuclear localisation signal (NLS) that is lost when the insert is removed.
-These ten amino acids correspond to a stretch at the protein surface that is unresolved in all available crystal structures of human NamPRT (e.g.
+This insert overlaps with a predicted weak nuclear localisation signal (NLS).
+The NLS prediction is lost when the insert is removed.
+These ten amino acids furthermore correspond to a stretch at the protein surface that is unresolved in all available crystal structures of human NamPRT (e.g.
 structure visualisation in fig.~\ref{fig:unresolved_loop}B from \cite{Wang2006}).
@@
 The second hypothesis was that the sequence insertion might influence substrate binding of NamPRT.
+We first compared and analysed different available protein structure of NamPRT and tested whether  the loop insertion leads to dynamic structural rearrangements using homology modelling with and without a loop insertion (see materials and methods to be included).
+In addition, we performed molecular dynamics simulations both with and without loop insertion.
+We were not able to show any obvious structural rearrangements and  the molecular dynamics simulations showed little structural changes upon loop insertion and we observe a structurally stable catalytic core.
+This might be based on the fact that  all available protein structures of NamPRT differ very little even at the catalytic (side between 0.15Å and 0.4Å)  down tot the atomic level  [see suppl.
+tables ?].
+
 We thus expressed both wildtype and mutant proteins, N-terminally fused to a 6xHis-tag in \textit{E.~coli}, and purified them.
@@
 Nevertheless, the enzymatic activity of the mutant enzyme was only approximately 30\% of that of the wildtype enzyme (fig.~\ref{fig:unresolved_loop}D).
-In contrast to wildtype NamPRT, the activity of the mutant enzyme was not stimulated in the presence of ATP (fig.~\ref{fig:unresolved_loop}E).
 These observations suggest that the mutant enzyme is catalytically active, retains its dimeric state and sensitivity to FK866.

### 2018-11-17 latex/discussion.tex
revised results and discussion, changed naming to NamPT

@@
 
-We here comprehensively analysed the phylogenetic distribution of the three enzymes using Nam as a substrate.
+We here comprehensively analysed the phylogenetic distribution of the three enzymes metabolising  Nam.
 These are the two NAD salvage pathway enzymes NADA and NamPRT as well as the Nam-degrading enzyme NNMT.
-We found that after the first appearance of NNMT in Protostomia, a diversification of NAD-consuming reactions in Deuterostomia can be observed.
-We could explain these finding using mathematical modelling, as NNMT removes excess Nam from cells and thereby reduces product inhibition of NAD signalling enzymes.
+We found that after the first appearance of NNMT in Protostomia, a diversification of NAD-consuming reactions  can be observed in Deuterostomia.
+We could explain these finding using mathematical modelling, as NNMT can remove excess Nam from cells and thereby reduce product inhibition of NAD signalling enzymes.
 This in turn enables higher fluxes through these reactions.
-Thus, the diversification of NAD-consuming enzymes in mammals seems to have been enabled by the presence of NNMT.
+Thus, the diversification of NAD-consuming enzymes in mammals might have been enabled by the presence of NNMT.
 
 NAD-consuming enzymes are involved in a wide variety of signalling and gene regulatory mechanisms that, due to their sensitivity to NAD$^{+}$, have the ability to translate differences in metabolic states into changes in signalling and gene regulation.
-As NAD concentrations are lowered by the removal of NAD precursor by NNMT,  a high affinity of NamPRT is required for high NAD consumption fluxes and NAD concentrations in the presence of NNMT.
-It therefore seems plausible that NNMT might have been driving NamPRT evolution.
-Looking at the enzyme affinities of the human enzymes, it furthermore appears that both NNMT and NamPRT reached an almost optimal state, as further changes in the affinity of either NamPRT or NNMT would not result in much higher steady state NAD concentrations or NAD consumption fluxes.
+As NAD concentrations are lowered by the removal of NAD precursor through  NNMT,  a high affinity of NamPT is required for high NAD consumption fluxes and NAD concentrations in the presence of NNMT.
+It therefore seems plausible that NNMT might have been driving NamPT evolution.
+Looking at the enzyme affinities of the human enzymes, it furthermore appears that both NNMT and NamPT reached an almost optimal state, as further changes in the affinity of either NamPT or NNMT would not result in much higher steady state NAD concentrations or NAD consumption fluxes.
 In addition, our simulations suggest that NNMT makes both NAD concentration and NAD consumption relatively independent of other processes requiring NAD, such as cell growth.
@@
 The increased NNMT expression observed in some types of cancer, might serve to remove Nam derived by increased NAD-dependent signalling.
-To maintain high NAD concentrations, a simultaneous higher expression of NamPRT is required, which is what has been found in some types of cancer \cite{Bi2011,Wang2011}.
-It is worth noticing that NNMT is only advantageous as long as NamPRT affinity is sufficiently high.
-This suggests that certain types of cancer expressing NNMT at a high level, would potentially be more susceptible to competitive inhibitors of NamPRT.
+To maintain high NAD concentrations, a simultaneous higher expression of NamPT is required, which is what has been found in some types of cancer \cite{Bi2011,Wang2011}.
+It is worth noticing that NNMT is only advantageous as long as NamPT affinity is sufficiently high.
+This suggests that certain types of cancer expressing NNMT at a high level, would potentially be more susceptible to competitive inhibitors of NamPT.
 Several of such inhibitors are currently tested in clinical studies \cite{EspindolaNetto2017,Xu2015SR}.
-Based on our analysis, we would suggest that it might be reasonable to screen patients before treatment, as non-NNMT expressing tumours might respond less to competitive NamPRT inhibitors and missing Nam degradation in those cancer cells would potentially lead to an accumulation of Nam that could outcompete the inhibitor.
+Based on our analysis, we would suggest that it might be reasonable to screen patients before treatment, as non-NNMT expressing tumours might respond less to competitive NamPT inhibitors and missing Nam degradation in those cancer cells would potentially lead to an accumulation of Nam that could outcompete the inhibitor.
 The latter aspect is not well investigated and requires further analysis.
 
-Neither the scattered distribution of NamPRT and NADA that is especially pronounced in bacteria \cite{Gazzaniga2009}, but that has also been observed in eukaryotes, nor the disappearance of NADA in vertebrates has been understood earlier.
+Neither the scattered distribution of NamPT and NADA that is especially pronounced in bacteria \cite{Gazzaniga2009}, nor the disappearance of NADA in vertebrates has been understood earlier.
 Our combined phylogenetic-modelling analysis now provides a potential explanation for both observations.
-Using simulated competition between two compartments that share the same limited source of Nam, we show that the compartment that contains NamPRT and NNMT can maintain a higher steady state NAD concentration and NAD consumption rate than the compartment containing NADA.
-This is, however, only the case if NamPRT substrate affinity is sufficiently high.
-The dominant enzyme combination found in vertebrates, a high-affinity NamPRT with NNMT, thus seems to provide a competitive advantage.
-As this may also hold for mammalian-associated bacteria, particularly pathogens, we wanted to see whether pathogenic bacteria solely express NamPRT.
+Using simulated competition between two compartments that share the same limited source of Nam, we show that the compartment that contains NamPT and NNMT can maintain a higher steady state NAD concentration and NAD consumption rate than the compartment containing NADA.
+This is, however, only the case if NamPT substrate affinity is sufficiently high.
+The dominant enzyme combination found in vertebrates, a high-affinity NamPT with NNMT, thus seems to provide a competitive advantage.
+
+%As this may also hold for mammalian-associated bacteria, particularly pathogens, we wanted to see whether pathogenic bacteria solely express NamPT.
 Unfortunately, bacterial habitat information is currently far from complete and often difficult to access.
-We therefore manually checked bacteria that possess NamPRT and indeed found that most of them have been characterised to be pathogenic.
-It should be noted that the distribution of NADA and NamPRT does not follow the bacterial species tree \cite{Gazzaniga2009}.
+We therefore manually checked bacteria that possess NamPT and indeed found that most of them have been characterised to be pathogenic.
+It should be noted that the distribution of NADA and NamPT does not follow the bacterial species tree \cite{Gazzaniga2009}.
 Besides the suggestion made here, there might well be other environmental aspects that influence the phylogenetic distribution in bacteria.
 
-A detailed analysis of sequence variances in NamPRT revealed that only deuterostomes that have NNMT, but not NADA, have a sequence insertion in the N-terminal part of NamPRT that seems to enable the high affinity of the enzyme.
+A detailed analysis of sequence variances in NamPT revealed that only deuterostomes that have NNMT, but not NADA, have a sequence insertion in the N-terminal part of NamPT that seems to enable the high affinity of the enzyme.
 This in turn would suggest that also the bacterial enzymes do not have a high substrate affinity.
-The substrate affinity measured for a bacterial NamPRT from \textit{Acinetobacter baylyi} \cite{Sorci2010} is indeed 10\,000 times higher lower ($K_{M}$ of 0.04mM) than that of the human NamPRT, supporting our hypothesis.
-Other bacterial NamPRTs were shown to be functional \cite{Martin2001,Gerdes2006}, but the substrate affinities have not been determined.
-The differences in activity and affinity, implying differences in substrate binding could potentially be exploited for the development of antibiotics.
-Further analysis possibly including the crystallisation of a bacterial NamPRT would be required, to see whether the bacterial NAD metabolism could be a promising target.
+The substrate affinity measured for a bacterial NamPT from \textit{Acinetobacter baylyi} \cite{Sorci2010} is indeed 10\,000 times higher  ($K_{M}$ of 0.04mM) than that of the human NamPT, supporting our hypothesis.
+Other bacterial NamPTs were shown to be functional \cite{Martin2001,Gerdes2006}, but the substrate affinities have not been determined.
+As many pathogeneic bacteria encode NamPT, the difference in enzyme activity and affinity between the human and the bacterial enzyme could potentially be exploited for the development of antibiotics.
+Further analysis including the crystallisation of a bacterial NamPT would be required, to see whether the bacterial NAD metabolism could be a promising target.
 
 % Martin2001: Haemophilus ducreyi: no kinetic measurements
-% Sorci2010: Acinetobacter sp.: NamPRT (Nam) Km: 0.04 mM (= 40 000 nM); kcat: 0.12/s (apparent values determined at constant 5 mM PRPP and 2 mM ATP)
-% Gerdes2006: Synechocystis sp.: only activity 0.5 U/mg NamPRT for Nam
+% Sorci2010: Acinetobacter sp.: NamPT (Nam) Km: 0.04 mM (= 40 000 nM); kcat: 0.12/s (apparent values determined at constant 5 mM PRPP and 2 mM ATP)
+% Gerdes2006: Synechocystis sp.: only activity 0.5 U/mg NamPT for Nam
 % Human, according to table S2: Km: 5 nM; kcat: 0.0077/s
@@
 In our analyses, we did not consider the potential effects of co-substrates of the investigated pathway.
-Such co-substrates include targets of the NAD-consuming enzymes, such as acylated proteins for sirtuins, for example, or phosphoribosyl pyrophosphate (PRPP) and ATP that are required for NMN synthesis by NamPRT.
+Such co-substrates include targets of the NAD-consuming enzymes, such as acylated proteins for sirtuins, for example, or phosphoribosyl pyrophosphate (PRPP) and ATP that are required for NMN synthesis by NamPT.
 Furthermore, the presence of the methyl donor \textit{S}-adenosyl methionine (SAM) and its precursor methionine that have been shown to potentially limit the effect of NNMT \cite{Ulanovskaya2013} was not considered here.
@@
 
-During our analysis, we came across several problems related to the use of NCBI sequence databases for phylogenetic analyses.
+%During our analysis, we came across several problems related to the use of NCBI sequence databases for phylogenetic analyses.
 One is sequence contamination, which is a well-known problem \cite{Ballenghien2017,Longo2011}.
@@
 
-The third problem are wrong annotations.
+%The third problem are wrong annotations.
 We tried to avoid these, by only relying, wherever possible,  on template sequences with confirmed function.
@@
 Taken together, we have been able to comprehensively analyse the functional co-evolution of several enzymes of the NAD pathway.
-The appearance of NNMT seemingly initiated and drove complex alterations of the pathway such as an increase and diversification of NAD-dependent signalling, followed by an increase in NamPRT substrate affinity.
+The appearance of NNMT seemingly initiated and drove complex alterations of the pathway such as an increase and diversification of NAD-dependent signalling, followed by an increase in NamPT substrate affinity.
 A schematic overview is given in figure~\ref{fig:evo_events}.

### 2018-11-17 latex/results.tex
further revised results section
revised results and discussion, changed naming to NamPT

@@
 
-To understand the functional roles and potential interplay between the three known enzymes that use Nam as substrate (NamPRT, NADA and NNMT), we first conducted a comprehensive phylogenetic analysis of these three enzymes.
-The phylogenetic distribution of the two enzymes that initiate the two different NAD salvage pathways, NADA and NamPRT is scattered in bacteria \cite{Gazzaniga2009}, while their co-occurrence has been detected in some marine invertebrates \cite{Gossmann2012FEBS}.
+To understand the functional roles and potential interplay between the three known enzymes that use Nam as substrate (NamPT, NADA and NNMT), we first conducted a comprehensive phylogenetic analysis of these three enzymes.
+The phylogenetic distribution of the two enzymes that initiate the two different NAD salvage pathways, NADA and NamPT is scattered in bacteria \cite{Gazzaniga2009}, while their co-occurrence has been detected in some marine invertebrates \cite{Gossmann2012FEBS}.
 As shown in Figure~\ref{fig:phylo_distribution}A, bacteria, fungi, and plants predominantly possess NADA and only a few of them harbour NamPRT.
-In contrast, Metazoa predominantly lost NADA and have NamPRT together with NNMT.
+In contrast, Metazoa predominantly lost NADA and have NamPT together with NNMT.
 NNMT seems to have arisen \textit{de~novo} or diverged rapidly in the most recent common ancestor of Ecdysozoa and Lophotrochozoa (fig.~\ref{fig:phylo_distribution}B).
@@
 Nematodes are the only organisms, where we observed a concomitant presence of NADA and NNMT.
-In deuterostomes, the only large clade that possesses only NamPRT and seems to have lost NNMT are Sauropsida, and among them especially birds.
+In deuterostomes, the only large clade that possesses only NamPT and seems to have lost NNMT are Sauropsida, and among them especially birds.
 The reason why about half of the sequenced bird genomes do not seem to encode for NNMT remains unclear.
@@
 The lack of NNMT might be related to the differences in the excretion system, as the product of NNMT, methyl-Nam, is in mammals excreted with the urine.
-There are few metazoan species for which we could not find NamPRT or NADA, while NNMT was present.
+There are few metazoan species for which we could not find NamPT or NADA, while NNMT was present.
 We assume that this is due to incomplete genomes in the database, as the distribution of such species is scarce and widely scattered.
 
-In addition to the phylogenetic distribution of the two Nam salvage enzymes NADA and NamPRT, we analysed the phylogenetic diversity of enzymes catalysing NAD-dependent signalling reactions.
+In addition to the phylogenetic distribution of the two Nam salvage enzymes NADA and NamPT, we analysed the phylogenetic diversity of enzymes catalysing NAD-dependent signalling reactions.
 To do so, we used the previously established classification into ten different families of NAD-consuming signalling enzymes \cite{Gossmann2012FEBS}.
@@
 Taken together, we found that NADA is lost in vertebrates, but strongly preserved in most other organisms, despite the higher energetic requirement of that pathway.
-Moreover, the selection for having both NamPRT and NNMT coincides with an increased diversification of NAD-dependent signalling.
+Moreover, the selection for having both NamPT and NNMT coincides with an increased diversification of NAD-dependent signalling.
 This observation seems counterintuitive, as one would expect that increased NAD-dependent signalling should be accompanied by an increase of substrate availability for NAD biosynthesis.
 
-\subsection{Functional properties of NamPRT and NNMT have evolved to maximise NAD-dependent signalling}
+\subsection{Functional properties of NamPT and NNMT have evolved to maximise NAD-dependent signalling}
 
@@
 
-First, we addressed the unexpected correlation between the selection for co-occurrence of NamPRT and NNMT and an increase in the number of NAD-consuming enzyme families.
-We calculated steady state NAD concentrations and NAD consumption fluxes by simulating NAD biosynthesis proceeding via NamPRT in the presence or absence of NNMT.
-To achieve free NAD concentrations in the range reported in the literature and due to the very low turnover of NamPRT, we used tenfold higher NamPRT levels compared to other enzymes.
+First, we addressed the unexpected correlation between the selection for co-occurrence of NamPT and NNMT and an increase in the number of NAD-consuming enzyme families.
+We calculated steady state NAD concentrations and NAD consumption fluxes by simulating NAD biosynthesis proceeding via NamPT in the presence or absence of NNMT.
+To achieve free NAD concentrations in the range reported in the literature and due to the very low turnover of NamPT, we used tenfold higher NamPT levels compared to other enzymes.
 We also adjusted the amount of NMNAT accordingly to avoid that the NAD synthesis rates are limited by this enzyme.
@@
 However, it diminishes the steady state concentration of NAD (fig.~\ref{fig:NNMT_NAD_flux}B).
-The decline in NAD concentration can be compensated by a higher expression of NamPRT, further increasing NAD consumption flux (dashed lines in fig.~\ref{fig:NNMT_NAD_flux}A and~B).
+The decline in NAD concentration can be compensated by a higher expression of NamPT, further increasing NAD consumption flux (dashed lines in fig.~\ref{fig:NNMT_NAD_flux}A and~B).
 
-These results can be explained by looking in more detail at the kinetic parameters of NamPRT and NAD-consuming enzymes such as Sirtuin~1.
+These results can be explained by looking in more detail at the kinetic parameters of NamPT and NAD-consuming enzymes such as Sirtuin~1.
 Most NAD-consuming enzymes are inhibited by their product Nam.
 Thus, the presence of NNMT enables higher NAD consumption fluxes, by removing excess Nam from the cells.
-At the same time, high substrate affinity of NamPRT maintains a sufficiently high NAD concentration, although the concentration is, as expected, lower than in the system without NNMT.
+At the same time, high substrate affinity of NamPT maintains a sufficiently high NAD concentration, although the concentration is, as expected, lower than in the system without NNMT.
 
-Kinetic parameters of NamPRT were previously measured for the human enzyme \cite{Burgos2008} as well as for some bacterial enzymes \cite{Sorci2010}, the latter having a much lower substrate affinity for Nam.
-We thus analysed the potential effect of NamPRT affinity ($K_{M}$) on NAD steady state concentration and NAD consumption flux.
-In the absence of NNMT, a variation of the substrate affinity of NamPRT for Nam has very little effect on steady state NAD concentration and NAD consumption flux (fig.~\ref{fig:NamPRT_affinity_Nam}A and~B).
-In the presence of NNMT, however, NAD consumption flux and NAD concentration increases with decreasing $K_{M}$ values of NamPRT (fig.~\ref{fig:NamPRT_affinity_Nam}C and~D).
+Kinetic parameters of NamPT were previously measured for the human enzyme \cite{Burgos2008} as well as for some bacterial enzymes \cite{Sorci2010}, the latter having a much lower substrate affinity for Nam.
+We thus analysed the potential effect of NamPT affinity ($K_{M}$) on NAD steady state concentration and NAD consumption flux.
+In the absence of NNMT, a variation of the substrate affinity of NamPT for Nam has very little effect on steady state NAD concentration and NAD consumption flux (fig.~\ref{fig:NamPT_affinity_Nam}A and~B).
+In the presence of NNMT, however, NAD consumption flux and NAD concentration increases with decreasing $K_{M}$ values of NamPT (fig.~\ref{fig:NamPT_affinity_Nam}C and~D).
 
@@
 
-Figures~\ref{fig:NamPRT_affinity_Nam}E and~F visualise a direct comparison of simulations assuming different affinities of NamPRT for Nam, in the presence or absence of NNMT.
-Interestingly, at an affinity of $K_{M}$ =~1\,$\mu$M, which is in the range of the $K_{M}$ of NADA for Nam, NAD consumption flux is only higher with NNMT present when cell division rates are low (fig.~\ref{fig:NamPRT_affinity_Nam}E).
-If the affinity of NamPRT is high enough, consumption rates are always higher with NNMT than without it.
-The NAD concentration is always lower with NNMT (fig.~\ref{fig:NamPRT_affinity_Nam}F).
+Figures~\ref{fig:NamPT_affinity_Nam}E and~F visualise a direct comparison of simulations assuming different affinities of NamPT for Nam, in the presence or absence of NNMT.
+Interestingly, at an affinity of $K_{M}$ =~1\,$\mu$M, which is in the range of the $K_{M}$ of NADA for Nam, NAD consumption flux is only higher with NNMT present when cell division rates are low (fig.~\ref{fig:NamPT_affinity_Nam}E).
+If the affinity of NamPT is high enough, consumption rates are always higher with NNMT than without it.
+The NAD concentration is always lower with NNMT (fig.~\ref{fig:NamPT_affinity_Nam}F).
 
-To understand the interplay and competition for Nam between NamPRT and NNMT, we scanned a wide range of possible $K_{M}$ values for both enzymes in our simulations.
-As shown in figure~\ref{fig:optimal_substr_affinities}, the simulations indicate that both NAD consumption flux and NAD concentration would be minimal in case of a high $K_{M}$ for NamPRT and a low $K_{M}$ for NNMT.
-Conversely, lowering the $K_{M}$ of NamPRT to the nanomolar range substantially increases NAD consumption and concentration, which reach a plateau when the $K_{M}$ of NNMT is concomitantly elevated to the submillimolar range.
+To understand the interplay and competition for Nam between NamPT and NNMT, we scanned a wide range of possible $K_{M}$ values for both enzymes in our simulations.
+As shown in figure~\ref{fig:optimal_substr_affinities}, the simulations indicate that both NAD consumption flux and NAD concentration would be minimal in case of a high $K_{M}$ for NamPT and a low $K_{M}$ for NNMT.
+Conversely, lowering the $K_{M}$ of NamPT to the nanomolar range substantially increases NAD consumption and concentration, which reach a plateau when the $K_{M}$ of NNMT is concomitantly elevated to the submillimolar range.
 The asterisks in figure~\ref{fig:optimal_substr_affinities} denote the $K_{M}$ values actually found for the human enzymes.
@@
 
-Given the kinetic interdeence of  NNMT and NamPRT shown in figure fig.~\ref{fig:NamPRT_affinity_Nam and \ref{fig:optimal_substr_affinities},it is possible that NNMT has exerted an evolutionary pressure on the development of NamPRT.
-In this case, one would expect to observe adaptations that are reflected in the NamPRT protein sequence arising shortly after the occurrence of NNMT.
-To explore this, we created a multiple sequence alignment of NamPRT protein sequences from metazoa.
+Given the kinetic interdeence of  NNMT and NamPT shown in figure fig.~\ref{fig:NamPT_affinity_Nam} and \ref{fig:optimal_substr_affinities},it is possible that NNMT has exerted an evolutionary pressure on the development of NamPT.
+In this case, one would expect to observe adaptations that are reflected in the NamPT protein sequence arising shortly after the occurrence of NNMT.
+To explore this, we created a multiple sequence alignment of NamPT protein sequences from metazoa.
 An alignment of selected sequences is shown in figure~\ref{fig:unresolved_loop}A and a more comprehensive multiple sequence alignment containing a larger number of species can be found in supplementary figure~S1.
-We found that most Deuterostomes that possess only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 42 to 51 in the human enzyme.
+We found that most Deuterostomes that possess only NamPT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 42 to 51 in the human enzyme.
 This insert overlaps with a predicted weak nuclear localisation signal (NLS).
 The NLS prediction is lost when the insert is removed.
-These ten amino acids furthermore correspond to a stretch at the protein surface that is unresolved in all available crystal structures of human NamPRT (e.g.
+These ten amino acids furthermore correspond to a stretch at the protein surface that is unresolved in all available crystal structures of human NamPT (e.g.
 structure visualisation in fig.~\ref{fig:unresolved_loop}B from \cite{Wang2006}).
@@
 
-From these observations, we derived two possible hypotheses regarding the role of the loop in NamPRT function.
-The first hypothesis was that the presence of the loop could change the subcellular localisation of NamPRT, as it is overlapping with a predicted NLS.
-To test this hypothesis, we created a mutant NamPRT lacking the loop and recombinantly expressed FLAG-tagged wildtype and mutant NamPRT in HeLa S3 cells.
-Immunofluorescence imaging showed a mixed cytosolic nuclear localisation for both the wildtype and the mutant NamPRT (fig.~\ref{fig:unresolved_loop}C).
+From these observations, we derived two possible hypotheses regarding the role of the loop in NamPT function.
+The first hypothesis was that the presence of the loop could change the subcellular localisation of NamPT, as it is overlapping with a predicted NLS.
+To test this hypothesis, we created a mutant NamPT lacking the loop and recombinantly expressed FLAG-tagged wildtype and mutant NamPT in HeLa S3 cells.
+Immunofluorescence imaging showed a mixed cytosolic nuclear localisation for both the wildtype and the mutant NamPT (fig.~\ref{fig:unresolved_loop}C).
 Thus, deletion of the loop did not compromise the nuclear localisation.
 
-The second hypothesis was that the sequence insertion might influence substrate binding of NamPRT.
-We first compared and analysed different available protein structure of NamPRT and tested whether  the loop insertion leads to dynamic structural rearrangements using homology modelling with and without a loop insertion (see materials and methods to be included).
+The second hypothesis was  based on our model simulations which predict that NNMT might have excerted evolutionary pressure on  NamPT and that therefore the sequence insertion might influence substrate binding of NamPT.
+To analalyze this possibility, we first compared and analysed different available protein structure of NamPT and tested whether  the loop insertion could lead to dynamic structural rearrangements.
+Initial this was done using homology modelling with and without a loop insertion (see Figure ...).
 In addition, we performed molecular dynamics simulations both with and without loop insertion.
-We were not able to show any obvious structural rearrangements and  the molecular dynamics simulations showed little structural changes upon loop insertion and we observe a structurally stable catalytic core.
-This might be based on the fact that  all available protein structures of NamPRT differ very little even at the catalytic (side between 0.15Å and 0.4Å)  down tot the atomic level  [see suppl.
-tables ?].
+Taken together we were not able to show any obvious structural rearrangements.
+The molecular dynamics simulations showed little structural changes upon loop insertion and we observe a structurally stable catalytic core.
+This might be based on the fact that  all available protein structures of NamPT differ very little even at the catalytic side ( between 0.15Å and 0.4Å).
 
-We thus expressed both wildtype and mutant proteins, N-terminally fused to a 6xHis-tag in \textit{E.~coli}, and purified them.
+We nevertheless expressed and purified  both wildtype and mutant proteins, N-terminally fused to a 6xHis-tag in \textit{E.~coli}..
 The size exclusion chromatography profile showed that both wildtype and mutant protein were expressed as dimers (see suppl.
-fig.~S3), indicating that the missing residues in the mutant did not dramatically affect the protein structure.
-The enzymatic activity was measured by NMR spectroscopy using the detection of NMN produced in the presence and absence of ATP.
-Upon incubation with the NamPRT inhibitor FK866 \cite{Hasmann2003} for 30 minutes, both wildtype and mutant NamPRT did not produce any NMN, suggesting that binding of FK866 is not affected by the mutation (see suppl.
+fig.~S3), indicating that the  mutant protein is likely to be folded correctly.
+The enzymatic activity was measured by NMR spectroscopy using the detection of NMN produced in the presence of ATP.
+Upon incubation with the NamPT inhibitor FK866 \cite{Hasmann2003} for 30 minutes, both wildtype and mutant NamPT did not produce any NMN, suggesting that binding of FK866 is not affected by the mutation (see suppl.
 fig.~S4).
-Nevertheless, the enzymatic activity of the mutant enzyme was only approximately 30\% of that of the wildtype enzyme (fig.~\ref{fig:unresolved_loop}D).
-These observations suggest that the mutant enzyme is catalytically active, retains its dimeric state and sensitivity to FK866.
-However, the lower activity and the absence of catalytic activation by ATP indicate that deletion of amino acids 42 to 51 might have affected substrate binding.
+The mutant enzyme showed no detectable activity with 100 $\mu$M  after 10 min of incubation at 30.
+With 1mM Nam ezymatic activity could be detected with the mutant enzyme.
+But the measssured activity is reduced to ...\% compared to the wildtype enzyme (fig.~\ref{fig:unresolved_loop}D).
+These observations suggest that the mutant enzyme is catalytically active, retains its dimeric state and sensitivity to FK866, but, has a  lower affinity to Nam, supporting the predictions derived from the modelling approach.
 
-\subsection{NamPRT and NNMT made NADA obsolete in vertebrates}
+\subsection{NamPT and NNMT made NADA obsolete in vertebrates}
 
@@
 As selection during evolution results usually from competition for resources, we built a two-compartment model, based on the pathway model described above.
-One compartment contains NADA, while the other one contains either NamPRT alone or together with NNMT.
+One compartment contains NADA, while the other one contains either NamPT alone or together with NNMT.
 Both compartments share a limited Nam source (for details, see Experimental Procedures and suppl.
@@
 
-In the presence of NNMT, the NamPRT compartment has both higher NAD consumption rates and higher steady state NAD concentrations than the compartment containing NADA (fig.~\ref{fig:NNMT_comp_advantage}C and~D).
-The higher NAD concentrations in the compartment containing NamPRT and NNMT can, however, only be maintained if the affinity of NamPRT for Nam is high enough.
-If the substrate affinity of NamPRT is too low (high $K_{M}$), the NADA compartment is able to maintain higher NAD concentrations, but still has a lower NAD consumption flux.
-Taken together, the results suggest that the NADA pathway might have become obsolete upon emergence of the high affinity NamPRT.
+In the presence of NNMT, the NamPT compartment has both higher NAD consumption rates and higher steady state NAD concentrations than the compartment containing NADA (fig.~\ref{fig:NNMT_comp_advantage}C and~D).
+The higher NAD concentrations in the compartment containing NamPT and NNMT can, however, only be maintained if the affinity of NamPT for Nam is high enough.
+If the substrate affinity of NamPT is too low (high $K_{M}$), the NADA compartment is able to maintain higher NAD concentrations, but still has a lower NAD consumption flux.
+Taken together, the results suggest that the NADA pathway might have become obsolete upon emergence of the high affinity NamPT.
 This in turn might have been induced by the appearance of NNMT.

### 2018-11-18 latex/figurelegends.tex
revised results again, changed naming into NamPT

@@
 The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates.
-The pathway found in vertebrates directly converts Nam into the corresponding mononucleotide (NMN) by the Nam phosphoribosyltransferase (NamPRT).
+The pathway found in vertebrates directly converts Nam into the corresponding mononucleotide (NMN) by the Nam phosphoribosyltransferase (NamPT).
 The Nam N-methyltransferase (NNMT) degrades Nam to methyl-Nam (MNam), which is in mammals excreted with the urine.}
@@
   \includegraphics[width=0.75\textwidth]{fig_2_phylo_distribution}
-  \caption{\textbf{Phylogenetic distribution of NADA, NNMT, and NamPRT and their relation to the number of NAD consumers.} A) Distribution of NADA, NNMT, and NamPRT in selected clades.
-NADA is dominant in bacteria, fungi, and plants (Viridiplantae), whereas NamPRT together with NNMT is dominant in Metazoa.
+  \caption{\textbf{Phylogenetic distribution of NADA, NNMT, and NamPT and their relation to the number of NAD consumers.} A) Distribution of NADA, NNMT, and NamPT in selected clades.
+NADA is dominant in bacteria, fungi, and plants (Viridiplantae), whereas NamPT together with NNMT is dominant in Metazoa.
 Numbers at the pie charts show, how many species of the clades possess the respective enzyme combination indicated by the colour explained in the lower right of the figure.
@@
   \caption{\textbf{NNMT enables high NAD consumption flux.} We used a dynamic model of NAD biosynthesis and consumption (for details, see Experimental Procedures) to simulate steady state NAD consumption flux (A) and concentration (B).
-The amount of NMNAT and NamPRT used in the simulations where adjusted such that the free NAD concentrations were in the range reported in the literature.
+The amount of NMNAT and NamPT used in the simulations where adjusted such that the free NAD concentrations were in the range reported in the literature.
 All other enzyme concentrations were set equal.
@@
 In the presence of NNMT (blue lines), steady state NAD consumption rates are higher despite reduced NAD concentrations.
-Increasing the amount of NamPRT in the simulation fourfold (blue dotted lines) partially compensates for the decreased NAD concentration caused by Nam degradation through NNMT.}
+Increasing the amount of NamPT in the simulation fourfold (blue dotted lines) partially compensates for the decreased NAD concentration caused by Nam degradation through NNMT.}
   \label{fig:NNMT_NAD_flux}
@@
   \centering
-  \includegraphics[width=0.72\textwidth]{fig_4_NamPRT_affinity_Nam}
-  \caption{\textbf{Role of NamPRT substrate affinity.} We simulated the effect of different Michaelis-Menten constants ($K_{M}$) of NamPRT for Nam on the steady state NAD consumption flux and NAD concentration at different cell division rates.
+  \includegraphics[width=0.72\textwidth]{fig_4_NamPT_affinity_Nam}
+  \caption{\textbf{Role of NamPT substrate affinity.} We simulated the effect of different Michaelis-Menten constants ($K_{M}$) of NamPT for Nam on the steady state NAD consumption flux and NAD concentration at different cell division rates.
 All parameters were equal to those used for the simulations in figure~\ref{fig:NNMT_NAD_flux}.
-In the absence of NNMT, the $K_{M}$ of NamPRT has little influence on NAD consumption (A) and concentration (B), but both are strongly influenced by cell division rates.
-In the presence of NNMT, decreasing $K_{M}$ of NamPRT enables increasing NAD consumption flux (C) and NAD concentration (D).
+In the absence of NNMT, the $K_{M}$ of NamPT has little influence on NAD consumption (A) and concentration (B), but both are strongly influenced by cell division rates.
+In the presence of NNMT, decreasing $K_{M}$ of NamPT enables increasing NAD consumption flux (C) and NAD concentration (D).
 NNMT furthermore makes both, NAD consumption flux and concentration, almost independent of cell division rates.
-Comparing the situation with and without NNMT (E and F) at two different NamPRT $K_{M}$ values reveals that at high $K_{M}$ (dashed lines) and high cell division rates NNMT no longer enables higher NAD consumption rates compared to NamPRT alone (green line and dashed grey line).}
-  \label{fig:NamPRT_affinity_Nam}
+Comparing the situation with and without NNMT (E and F) at two different NamPT $K_{M}$ values reveals that at high $K_{M}$ (dashed lines) and high cell division rates NNMT no longer enables higher NAD consumption rates compared to NamPT alone (green line and dashed grey line).}
+  \label{fig:NamPT_affinity_Nam}
 \end{figure}
@@
   \includegraphics[width=\textwidth]{fig_5_optimal_substr_affinities}
-  \caption{\textbf{The substrate affinities of human NNMT and NamPRT are close to optimum.} We simulated the impact of changes in the $K_{M}$ for both NamPRT and NNMT on NAD consumption rates (A) and NAD concentration (B).
-Both are increasing with decreasing $K_{M}$ of NamPRT, but increasing $K_{M}$ of NNMT.
+  \caption{\textbf{The substrate affinities of human NNMT and NamPT are close to optimum.} We simulated the impact of changes in the $K_{M}$ for both NamPT and NNMT on NAD consumption rates (A) and NAD concentration (B).
+Both are increasing with decreasing $K_{M}$ of NamPT, but increasing $K_{M}$ of NNMT.
 The affinities reported for human enzymes (indicated by a black asterisk) appear to be close to the theoretical optimum, as further improvements would have little effect on NAD consumption or concentration.}
@@
   \includegraphics[width=0.7\textwidth]{fig_6_unresolved_loop}
-  \caption{\textbf{The function of the structurally unresolved loop of NamPRT.} Most deuterostomes that possess NamPRT and NNMT show a sequence insertion in the N-terminal region of NamPRT that has been revealed by multiple sequence alignment of NamPRT from different species (A).
-Coloured circles indicate the enzymes present in the respective species; blue: NamPRT and NNMT; black: NamPRT, NADA and NNMT; yellow: NamPRT and NADA.
+  \caption{\textbf{The function of the structurally unresolved loop of NamPT.} Most deuterostomes that possess NamPT and NNMT show a sequence insertion in the N-terminal region of NamPT that has been revealed by multiple sequence alignment of NamPT from different species (A).
+Coloured circles indicate the enzymes present in the respective species; blue: NamPT and NNMT; black: NamPT, NADA and NNMT; yellow: NamPT and NADA.
 For a more comprehensive alignment, please see supplementary figure~S1.
-The structure visualisation of human NamPRT (B) is based on a structure prediction by SWISS-MODEL \cite{Arnold2006,Biasini2014} using the model 2H3D of the human NamPRT as template \cite{Wang2006}.
-The inserted region is not resolved in any of currently available crystal structures of NamPRT and thus appears to be a flexible loop structure at the surface of the NamPRT dimer, coloured in red.
-Immunofluorescence images (C) show that the localisation of the FLAG-tagged mutant protein lacking the unresolved loop is not changed compared to FLAG-tagged human wildtype NamPRT.
+The structure visualisation of human NamPT (B) is based on a structure prediction by SWISS-MODEL \cite{Arnold2006,Biasini2014} using the model 2H3D of the human NamPT as template \cite{Wang2006}.
+The inserted region is not resolved in any of currently available crystal structures of NamPT and thus appears to be a flexible loop structure at the surface of the NamPT dimer, coloured in red.
+Immunofluorescence images (C) show that the localisation of the FLAG-tagged mutant protein lacking the unresolved loop is not changed compared to FLAG-tagged human wildtype NamPT.
 Both show a heterogeneous nuclear-cytosolic localisation in HeLa S3 cells.
-\textit{In vitro} measurements using recombinant protein show that the mutant NamPRT has a lower activity than the wildtype enzyme (D) and is not activated by ATP (E).
+\textit{In vitro} measurements using recombinant protein show that the mutant NamPT has a lower activity than the wildtype enzyme (D) and is not activated by ATP (E).
 Bars in panels D and E that have different letters indicate significant difference of measured values as estimated using a T~test assuming independent samples and significance at $p < 0.05$.}
@@
   \includegraphics[width=\textwidth]{fig_7_NNMT_comp_advantage}
-  \caption{\textbf{NNMT provides a competitive advantage and makes NADA obsolete.} To simulate competition for common resources, we created a two-compartment model where one compartment contained NADA, but no NamPRT and the other compartment contained NamPRT either with or without NNMT, but no NADA.
-NADA and NamPRT were simulated to be present at equal amounts.
+  \caption{\textbf{NNMT provides a competitive advantage and makes NADA obsolete.} To simulate competition for common resources, we created a two-compartment model where one compartment contained NADA, but no NamPT and the other compartment contained NamPT either with or without NNMT, but no NADA.
+NADA and NamPT were simulated to be present at equal amounts.
 In the absence of NNMT the compartment containing NADA has slightly lower NAD consumption rates (A), but much higher steady state NAD concentrations (B).
 In the presence of NNMT, however, both NAD consumption (C) and NAD concentration (D) are lower in the NADA compartment.
-This effect is dependent on a low NamPRT $K_{M}$.}
+This effect is dependent on a low NamPT $K_{M}$.}
   \label{fig:NNMT_comp_advantage}

### 2018-11-18 latex/results.tex
revised results again, changed naming into NamPT

@@
 
-To understand the functional roles and potential interplay between the three known enzymes that use Nam as substrate (NamPT, NADA and NNMT), we first conducted a comprehensive phylogenetic analysis of these three enzymes.
-The phylogenetic distribution of the two enzymes that initiate the two different NAD salvage pathways, NADA and NamPT is scattered in bacteria \cite{Gazzaniga2009}, while their co-occurrence has been detected in some marine invertebrates \cite{Gossmann2012FEBS}.
+To understand the functional roles and potential interplay between the three known enzymes that use Nam as substrate (NamPT, NADA and NNMT), we first conducted a comprehensive analysis of the phylogenetic distribution of these three enzymes.
 As shown in Figure~\ref{fig:phylo_distribution}A, bacteria, fungi, and plants predominantly possess NADA and only a few of them harbour NamPRT.
-In contrast, Metazoa predominantly lost NADA and have NamPT together with NNMT.
+In contrast, Metazoa predominantly lost NADA and encode for NamPT together with NNMT.
+NADA and NamPT, the  two enzymes that initiate the two different NAD salvage pathways,  show a scattered distribution in bacteria.
+Co-occurrence of these enzymes is rather rare, and has mainly been found in bacteria \cite{Gazzaniga2009} and some marine invertebrates \cite{Gossmann2012FEBS}.
 NNMT seems to have arisen \textit{de~novo} or diverged rapidly in the most recent common ancestor of Ecdysozoa and Lophotrochozoa (fig.~\ref{fig:phylo_distribution}B).
@@
 The distribution of NNMT in birds is quite scattered (suppl.
-fig.~S2) but could be explained by the fact that numerous bird genes are high in GC content \cite{Hron2015}.
-The lack of NNMT might be related to the differences in the excretion system, as the product of NNMT, methyl-Nam, is in mammals excreted with the urine.
-There are few metazoan species for which we could not find NamPT or NADA, while NNMT was present.
-We assume that this is due to incomplete genomes in the database, as the distribution of such species is scarce and widely scattered.
+fig.~S2).
+It is possible that detection of \textit{NNMT} in some bird genomes failed because  of high in GC content \cite{Hron2015} or because of the existing of very small chromosomes in birds that are difficult to assemble.
+The lack of NNMT might alternatively be related to the differences in the excretion system, as the product of NNMT, methyl-Nam, is in mammals excreted with the urine.
+There are few metazoan species for which we could not find NamPT or NADA, while NNMT was detected.
+We assume that this is due to incomplete genomes in the database, as these species are scarce and their distribution is widely scattered.
 
 In addition to the phylogenetic distribution of the two Nam salvage enzymes NADA and NamPT, we analysed the phylogenetic diversity of enzymes catalysing NAD-dependent signalling reactions.
-To do so, we used the previously established classification into ten different families of NAD-consuming signalling enzymes \cite{Gossmann2012FEBS}.
+To do so, we used the previously established classification into ten different families of NAD-consuming signalling enzymes \cite{Gossmann2012FEBS}, including PARP1-3,PARP4, PARP6/8, PARP7/9-15, PARP16, sirtuins, tankyrases, ADPR-cyclases, mono-ADP-ribosyltransferases and t-RNA-phosphotransferases.
 The detailed list of templates used for the phylogenetic analyses can be found in supplementary table~S1.
 The numbers shown in figure~\ref{fig:phylo_distribution}B denote the average number of NAD-dependent signalling enzyme families found in each clade.
-With the exception of Cnidaria and Lophotrochozoa, we find an average of three to four families in protostomes, whereas most deuterostome species have, on average, more than eight families with an increasing diversification of enzymes within some of these families \cite{Gossmann2014DNAR}.
+With the exception of Cnidaria and Lophotrochozoa, we find an average of three to four families in protostomes, whereas most deuterostome species have, on average, more than eight families with an increasing diversification of enzymes within some of these families, especially PARPs \cite{Gossmann2014DNAR}.
 
-Taken together, we found that NADA is lost in vertebrates, but strongly preserved in most other organisms, despite the higher energetic requirement of that pathway.
+Taken together, we found that NADA is lost in vertebrates, but strongly preserved in most other organisms, despite the higher energetic requirement of this pathway.
 Moreover, the selection for having both NamPT and NNMT coincides with an increased diversification of NAD-dependent signalling.
-This observation seems counterintuitive, as one would expect that increased NAD-dependent signalling should be accompanied by an increase of substrate availability for NAD biosynthesis.
+This observation seems counter-intuitive, as one would expect that increased NAD-dependent signalling should be accompanied by an increase of substrate availability for NAD biosynthesis.
 

### 2018-11-18 latex/results.tex
further revision of results

@@
 
-The second hypothesis was  based on our model simulations which predict that NNMT might have excerted evolutionary pressure on  NamPT and that therefore the sequence insertion might influence substrate binding of NamPT.
-To analalyze this possibility, we first compared and analysed different available protein structure of NamPT and tested whether  the loop insertion could lead to dynamic structural rearrangements.
-Initial this was done using homology modelling with and without a loop insertion (see Figure ...).
-In addition, we performed molecular dynamics simulations both with and without loop insertion.
+The second hypothesis was  based on our model simulations that predict that NNMT might have exerted evolutionary pressure on  NamPT and that therefore the sequence insertion might have an effect on substrate binding of NamPT.
+To analyse this possibility, we first compared different available protein structures of NamPT and tested whether  the loop insertion could potentially lead to dynamic structural rearrangements.
+This was done using homology modelling (see Figure ...) and molecular dynamics simulations for structures with and without the loop insertion.
 Taken together we were not able to show any obvious structural rearrangements.
-The molecular dynamics simulations showed little structural changes upon loop insertion and we observe a structurally stable catalytic core.
+The molecular dynamics simulations showed little structural changes upon loop insertion and we observed a structurally stable catalytic core.
 This might be based on the fact that  all available protein structures of NamPT differ very little even at the catalytic side ( between 0.15Å and 0.4Å).
@@
 fig.~S3), indicating that the  mutant protein is likely to be folded correctly.
-The enzymatic activity was measured by NMR spectroscopy using the detection of NMN produced in the presence of ATP.
+The enzymatic activity was measured by NMR spectroscopy using the detection of NMN produced in the absence of ATP.
 Upon incubation with the NamPT inhibitor FK866 \cite{Hasmann2003} for 30 minutes, both wildtype and mutant NamPT did not produce any NMN, suggesting that binding of FK866 is not affected by the mutation (see suppl.
 fig.~S4).
-The mutant enzyme showed no detectable activity with 100 $\mu$M  after 10 min of incubation at 30.
+The mutant enzyme showed no detectable activity with 100 $\mu$M  Nam after 10 min of incubation at 30\textdegree{}C.
 With 1mM Nam ezymatic activity could be detected with the mutant enzyme.
-But the measssured activity is reduced to ...\% compared to the wildtype enzyme (fig.~\ref{fig:unresolved_loop}D).
-These observations suggest that the mutant enzyme is catalytically active, retains its dimeric state and sensitivity to FK866, but, has a  lower affinity to Nam, supporting the predictions derived from the modelling approach.
+But the meassured activity is reduced to 30\% compared to the wildtype enzyme (fig.~\ref{fig:unresolved_loop}D).
+These observations suggest that the mutant enzyme is catalytically active, retains its dimeric state and sensitivity to FK866, but, has a  lower affinity to Nam, supporting the predictions derived from the metabolic modelling approach.
 
@@
 
-Next, we wished to understand why NADA was lost in vertebrates.
+Finally, we wanted to understand why NADA was lost in vertebrates.
 As selection during evolution results usually from competition for resources, we built a two-compartment model, based on the pathway model described above.
@@
 At high cell division rates, steady state concentrations in both compartments are similar.
-This might explain why in bacteria that often have relatively high growth rates, both systems can co-exist.
+This might explain why in bacteria that often have relatively high growth rates, both systems co-exist.
 

### 2018-11-26 supplements/MD.tex
moved md-files into folder supplements

\documentclass[a4paper,10pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[left=3cm]{geometry}
\usepackage{booktabs}
\usepackage{textcomp}

\begin{document}

We performed all-atom molecular dynamics simulations with explicit solvent for NamPT (PDB Code: 2H3D \cite{Wang2006}) with and without loop insertion. We used the AMBER99SB-ILDN force field \cite{Lindorff-Larsen2010} with the TIP3P water model \cite{Jorgensen1983} in GROMACS 5.1.2 \cite{Abraham2015}. The structures were simulated each in a box of water with distance between the solute and the box set to 0.2\,nm at a temperature of 300\,K for a total time of 1\,\textmu s. A time step of 2\,fs and the stochastic dynamics integrator were used. For the evaluation of the root mean square fluctuations (RMSF) the first 100\,ns of the simulations were omitted.

\begin{table}
 \caption{Root mean square deviation (RMSD) values between different structures (in \AA). The alignment and RMSD calculation was done with PyMOL\cite{PyMOL}. The structures are 2H3D (human NAMPT) \cite{Wang2006}, 3DGR (human NAMPT$\cdot$AMPcP complex) \cite{Burgos2009}, 3DHD (human NAMPT$\cdot$NMN$\cdot$Mg$_2$PPi complex) \cite{Burgos2009}, 3DHF (human BeF$_{3^-}$-NAMPT$\cdot$NMN$\cdot$Mg$_2$PPi complex) \cite{Burgos2009}, 3DKJ (human NAMPT$\cdot$PRPP$\cdot$BzAM complex) \cite{Burgos2009}, and 3DKL (human BeF$_{3^-}$-NAMPT$\cdot$Mg$_2$PRPP$\cdot$BzAM complex) \cite{Burgos2009}. The structural resolution of the PDB structures ranges from 1.8\,\AA~to 2.1\,\AA.}
 \begin{tabular}{l|cccccc}
  \toprule
  %\hline
  PDB Code & 2H3D & 3DGR & 3DHD & 3DHF & 3DKJ & 3DKL \\ 
  \midrule
  %\hline
  2H3D & - & 0.95 & 0.85 & 0.86 & 0.88 & 0.88 \\
  %\hline
  3DGR &   &  -   & 0.61 & 0.61 & 0.55 & 0.57 \\
  %\hlineM
  3DHD &   &      &  -   & 0.43 & 0.40 & 0.43 \\
  %\hline
  3DHF &   &      &      &  -   & 0.42 & 0.33 \\
  %\hline
  3DKJ &   &      &      &      &  -   & 0.39 \\
  %\hline
  3DKL &   &      &      &      &      &  -   \\
  %\hline
 \bottomrule
 \end{tabular}
\end{table}
\end{document}

### 2018-11-26 supplements/KineticsTable/Table_S1.tex
revised Kinetics Table

@@
     \\ \hline
-NMNT & \href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC2/1/1/1.html}{2.1.1.1}
+NNMT & \href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC2/1/1/1.html}{2.1.1.1}
 &$K_M$:400$\mu$M  & \cite{Aksoy1994} & Product inhibition\\
@@
     \\  \hline
-NamPRT & \href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC6/3/5/1.html}{6.3.5.1}
+NamPT & \href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC6/3/5/1.html}{6.3.5.1}
 &$K_M$:5nM  & \cite{Burgos2008} & HMM\\
@@
        \href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC2/4/2/11.html}{2.4.2.11}
-    &$K_M$:23$\mu$M  & \cite{Hanna1983} & HMM\\
+    &$K_M$:1.5$\mu$M  & \cite{Burgos2008} & HMM\\
     & & $k_{cat}$:3.3$s^{-1}$  & & 
@@
     \href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC3/1/3/5.html}{3.1.3.5}
-    &$K_M$:100$\mu$M  & \cite{} & HMM \\
-    & & $k_{cat}$:1$s^{-1}$ 
+    &$K_{M_{NaMN}}$:3.5mM  & \cite{PMID26385918} & HMM \\
+    & & $k_{{cat}_{NaMN}}$:2.8$s^{-1}$ 
+    & & \\ 
+    & &$K_{M_{NMN}}$: 5mM   &  \\
+     & & $k_{{cat}_{NMN}}$:0.5$s^{-1}$ 
+     & & \\  \hline  
+     PNP &
+    \href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC3/2/4/2.html}{2.4.2.1}
+    &$K_M$:1.48mM  & \cite{PMID9030766} & HMM \\
+    & & $k_{cat}$:40$s^{-1}$ 
+    & & \\  \hline    
+    NRK &
+    \href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC2/7/1/173.html}{2.7.1.173}
+    &$K_M$:3.4$\mu$M & \cite{PMID19027704} & HMM \\
+    & & $k_{cat}$:0.23$s^{-1}$ 
     & & \\  \hline    
@@
 
-The total enzyme concentration was set to 10 for all enzymes if not mentioned
-otherwise. Concentration of potentiantial cosubstrate was assumed to be constant
-and considered to be not limitng the reaction and thus represented by the
-maximal velocities given.
+The total enzyme concentration was set to 10 for all enzymes  except NamPT and NMNAT, for which the concentration was set to 100. Concentration of potential co-substrate was assumed to be constant
+and not-limiting for the reaction. Thus they were  implicitly represented by the
+maximal velocities given in the table.
 

### 2018-11-27 latex/discussion.tex
revised results again, changed naming into NamPT
revised supplementary table and implemented part of the comments from Alexander and Ines R.

@@
 We here comprehensively analysed the phylogenetic distribution of the three enzymes metabolising  Nam.
-These are the two NAD salvage pathway enzymes NADA and NamPRT as well as the Nam-degrading enzyme NNMT.
+These are the two NAD salvage pathway enzymes NADA and NamPT as well as the Nam-degrading enzyme NNMT.
 We found that after the first appearance of NNMT in Protostomia, a diversification of NAD-consuming reactions  can be observed in Deuterostomia.
@@
 
-We here developed a new approach that combines detailed phylogenetic analysis with dynamic metabolic modelling and have been able to explain observed evolutionary changes in the NAD biosynthesis and consumption pathway.
+We here developed an approach that combines detailed phylogenetic analysis with dynamic metabolic modelling and have been able to explain observed evolutionary changes in the NAD biosynthesis and consumption pathway.
 Based on the simulated pathway dynamics, we have furthermore derived predictions for physiological interdependencies between several enzymes of the pathway that are potentially relevant for new disease treatments.

### 2018-11-27 latex/header.tex
revised supplementary table and implemented part of the comments from Alexander and Ines R.

--- 2018-07-17 latex/header.tex
@@
 Mathias Bockwoldt\textsuperscript{1}, Dorothée Houry\textsuperscript{2}, Marc Niere\textsuperscript{2}, Toni I.
-Gossmann\textsuperscript{3}, Mathias Ziegler\textsuperscript{2}, and Ines Heiland\textsuperscript{1,§}
+Gossmann\textsuperscript{3}, Ines Reinartz\textsuperscript{4,5}, Alexander Schug\textsuperscript{6}, Mathias Ziegler\textsuperscript{2}, and Ines Heiland\textsuperscript{1,§}
 
@@
 \textsuperscript{3}Department of Animal and Plant Sciences, Western Bank, University of Sheffield, Sheffield, S10 2TN, United Kingdom
+
+\noindent
+\textsuperscript{4}Department of Physics, Karlsruhe Institute of Technology, Wolfgang-Gaede-Str. 1, 76131 Karlsruhe, Germany
+\textsuperscript{5}Steinbuch Centre for Computing, Karlsruhe Institute of Technology, Hermann-von-Helmholtz-Platz 1, 76344 Eggenstein-Leopoldshafen, Germany
+
+\noindent
+\textsuperscript{6}John von Neumann Institute for Computing, Jülich Supercomputing Centre, Forschungszentrum Jülich, 52425 Jülich, Germany
 

### 2018-11-27 latex/methods.tex
revised supplementary table and implemented part of the comments from Alexander and Ines R.

@@
 Quantification of nicotinamide mononucleotide (NMN) was done by the integration of the peak at 9.52\,ppm and DSS used as an internal standard.
+
+\subsection{Molecular dynamics simulations}
+We performed all-atom molecular dynamics simulations with explicit solvent for NamPT (PDB Code: 2H3D \cite{Wang2006}) with and without loop insertion.
+We used the AMBER99SB-ILDN force field \cite{Lindorff-Larsen2010} with the TIP3P water model \cite{Jorgensen1983} in GROMACS 5.1.2 \cite{Abraham2015}.
+The structures were simulated each in a box of water with distance between the solute and the box set to 0.2\,nm at a temperature of 300\,K for a total time of 1\,\textmu s.
+A time step of 2\,fs and the stochastic dynamics integrator were used.
+For the evaluation of the root mean square fluctuations (RMSF) the first 100\,ns of the simulations were omitted.

### 2018-11-27 latex/results.tex
revised supplementary table and implemented part of the comments from Alexander and Ines R.

@@
 The second hypothesis was  based on our model simulations that predict that NNMT might have exerted evolutionary pressure on  NamPT and that therefore the sequence insertion might have an effect on substrate binding of NamPT.
-To analyse this possibility, we first compared different available protein structures of NamPT and tested whether  the loop insertion could potentially lead to dynamic structural rearrangements.
-This was done using homology modelling (see Figure ...) and molecular dynamics simulations for structures with and without the loop insertion.
-Taken together we were not able to show any obvious structural rearrangements.
-The molecular dynamics simulations showed little structural changes upon loop insertion and we observed a structurally stable catalytic core.
-This might be based on the fact that  all available protein structures of NamPT differ very little even at the catalytic side ( between 0.15Å and 0.4Å).
-
-We nevertheless expressed and purified  both wildtype and mutant proteins, N-terminally fused to a 6xHis-tag in \textit{E.~coli}..
+To analyse this possibility, we expressed and purified  both the wildtype and the mutant enzyme in \textit{E.~coli}., N-terminally fused to a 6xHis-tag.
 The size exclusion chromatography profile showed that both wildtype and mutant protein were expressed as dimers (see suppl.
@@
 The mutant enzyme showed no detectable activity with 100 $\mu$M  Nam after 10 min of incubation at 30\textdegree{}C.
-With 1mM Nam ezymatic activity could be detected with the mutant enzyme.
-But the meassured activity is reduced to 30\% compared to the wildtype enzyme (fig.~\ref{fig:unresolved_loop}D).
-These observations suggest that the mutant enzyme is catalytically active, retains its dimeric state and sensitivity to FK866, but, has a  lower affinity to Nam, supporting the predictions derived from the metabolic modelling approach.
+With 1mM Nam enzymatic activity could be detected with the mutant enzyme.
+But the measured activity is reduced to 80\% of that observed for the wildtype enzyme under the same condition (fig.~\ref{fig:unresolved_loop}D).
+These observations suggest that the mutant enzyme is catalytically active, retains its dimeric state and sensitivity to FK866, but, has a  lower activity and affinity to Nam, supporting the predictions derived from the metabolic modelling approach.
+
+To see whether we can find a mechanistic explanation for the reduced affinity of the mutant enzyme, we analysed different available protein structures of NamPT and tested whether  the loop insertion could potentially lead to dynamic structural rearrangements.
+To this end we applied homology modelling (fig.~\ref{fig:unresolved_loop}C) and molecular dynamics simulations for structures with and without the loop insertion (fig.~\ref{fig:unresolved_loop}F) Taken together we did not observe substantial structural rearrangements and the molecular dynamics simulations showed only limited structural changes upon loop insertion and we observed a mostly structurally stable catalytic core.
+This might be based on the fact that  all available protein structures of NamPT differ very little even at the catalytic side ( between 0.33Å and 0.95Å see supplementary table S3).
+Some residues close to the catalytic, nevertheless, showed slightly elevated mobilities in the wildtype structure, but, these elevated mobilities were dominated by rare events during the simulation time 1$~\mu s$.
+Thus, they do not appear as a robust change of structural dynamics upon loop insertion.
+They could nevertheless provide a hind towards reduced mobility of the mutant enzyme that explain the observed changes in affinity and activity upon loop deletion.
 

### 2018-11-27 supplements/KineticsTable/kinetics_table.tex
suppl update

@@
 \section*{Supplementary Table 2}
-
-\begin{longtable}{p{3cm}p{2cm}p{3cm}cc}
+\setcounter{table}{1}
+\begin{longtable}{p{2cm}p{2cm}p{2cm}cc}
 \multicolumn{5}{l}{\textbf{Overview of kinetic constants used for the
@@
        \href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC2/4/2/11.html}{2.4.2.11}
-    &$K_M$:23$\mu$M  & \cite{Hanna1983} & HMM\\
+    &$K_M$:1.5$\mu$M  & \cite{Burgos2008} & HMM\\
     & & $k_{cat}$:3.3$s^{-1}$  & &
@@
   \bottomrule
+   \caption{The total enzyme concentration was set to 10 for all enzymes except NamPT and NMNAT, for which the concentration was set to 100.
+Concentration of potential co-substrate was assumed to be constant and not-limiting for the reaction.
+Thus they were implicitly represented by the
+maximal velocities given in the table.}
   \label{tab:kinetic}
+
 \end{longtable}
-
-The total enzyme concentration was set to 10 for all enzymes except NamPRT and NMNAT, for which the concentration was set to 100. Concentration of potential co-substrate was assumed to be constant
-and not-limiting for the reaction. Thus they were implicitly represented by the
-maximal velocities given in the table.
 

### 2018-11-27 supplements/MD.tex
suppl update

@@
 \begin{document}
-
-We performed all-atom molecular dynamics simulations with explicit solvent for NamPT (PDB Code: 2H3D \cite{Wang2006}) with and without loop insertion.
-We used the AMBER99SB-ILDN force field \cite{Lindorff-Larsen2010} with the TIP3P water model \cite{Jorgensen1983} in GROMACS 5.1.2 \cite{Abraham2015}.
-The structures were simulated each in a box of water with distance between the solute and the box set to 0.2\,nm at a temperature of 300\,K for a total time of 1\,\textmu s.
-A time step of 2\,fs and the stochastic dynamics integrator were used.
-For the evaluation of the root mean square fluctuations (RMSF) the first 100\,ns of the simulations were omitted.
-
+\setcounter{table}{2}
 \begin{table}
- \caption{Root mean square deviation (RMSD) values between different structures (in \AA).
-The alignment and RMSD calculation was done with PyMOL\cite{PyMOL}.
-The structures are 2H3D (human NAMPT) \cite{Wang2006}, 3DGR (human NAMPT$\cdot$AMPcP complex) \cite{Burgos2009}, 3DHD (human NAMPT$\cdot$NMN$\cdot$Mg$_2$PPi complex) \cite{Burgos2009}, 3DHF (human BeF$_{3^-}$-NAMPT$\cdot$NMN$\cdot$Mg$_2$PPi complex) \cite{Burgos2009}, 3DKJ (human NAMPT$\cdot$PRPP$\cdot$BzAM complex) \cite{Burgos2009}, and 3DKL (human BeF$_{3^-}$-NAMPT$\cdot$Mg$_2$PRPP$\cdot$BzAM complex) \cite{Burgos2009}.
-The structural resolution of the PDB structures ranges from 1.8\,\AA~to 2.1\,\AA.}
+\section*{Supplementary Table 3}
  \begin{tabular}{l|cccccc}
@@
  \end{tabular}
+  \caption{Root mean square deviation (RMSD) values between different structures (in \AA).
+The alignment and RMSD calculation was done with PyMOL\cite{PyMOL}.
+The structures are 2H3D (human NAMPT) \cite{Wang2006}, 3DGR (human NAMPT$\cdot$AMPcP complex) \cite{Burgos2009}, 3DHD (human NAMPT$\cdot$NMN$\cdot$Mg$_2$PPi complex) \cite{Burgos2009}, 3DHF (human BeF$_{3^-}$-NAMPT$\cdot$NMN$\cdot$Mg$_2$PPi complex) \cite{Burgos2009}, 3DKJ (human NAMPT$\cdot$PRPP$\cdot$BzAM complex) \cite{Burgos2009}, and 3DKL (human BeF$_{3^-}$-NAMPT$\cdot$Mg$_2$PRPP$\cdot$BzAM complex) \cite{Burgos2009}.
+The structural resolution of the PDB structures ranges from 1.8\,\AA~to 2.1\,\AA.}
 \end{table}
+
+\bibliographystyle{plos2009}
+\bibliography{MD}
 \end{document}

### 2018-11-27 latex/introduction.tex
included all changes from Alexander and Ines R.

@@
 The presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012FEBS} is among the most enigmatic and counterintuitive features of NAD metabolism.
-While NamPRT is seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, NNMT seems to have no metabolic function other than to remove Nam from NAD metabolism.
+While NamPRT is seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, NNMT seems to have no function other than to remove Nam from NAD metabolism.
 It has been suggested that the process potentially acts as a metabolic methylation sink \cite{Pissios2017}.
@@
 We and others have earlier shown that the two NAD biosynthesis pathways starting from Nam co-exists in some eukaryotes \cite{Gossmann2012FEBS,Carneiro2013}, as well as in some bacterial species \cite{Gazzaniga2009}.
-Why these pathways co-exists over in some organisms and over a very long evolutionary time frame and why NADA nevertheless disappears in vertebrates, is not known.
+Why these pathways co-exists in some organisms and over a very long evolutionary time frame and why NADA nevertheless disappears in vertebrates, is not known.
 We furthermore have little understanding of the physiological role of NNMT and its impact on NAD-metabolism so far.

### 2018-11-27 latex/results.tex
included all changes from Alexander and Ines R.

@@
 fig.~S2).
-It is possible that detection of \textit{NNMT} in some bird genomes failed because  of high in GC content \cite{Hron2015} or because of the existing of very small chromosomes in birds that are difficult to assemble.
+It is possible that detection of \textit{NNMT} in some bird genomes failed because  of their high GC content \cite{Hron2015} or because of difficulties in assembling very small chromosomes common in birds.
 The lack of NNMT might alternatively be related to the differences in the excretion system, as the product of NNMT, methyl-Nam, is in mammals excreted with the urine.
@@
 
-Remarkably, NAD concentration and consumption flux are both considerably affected by cell division rates in a system without NNMT, at least if the enzyme expression is kept constant at different cell division rates.
-Of course, this is an artificial scenario, as one would assume organisms to regulate enzyme expression to achieve similar levels of metabolite concentrations instead.
-In addition, there seems to be ad trade-off  between maintainable NAD concentration and consumption flux, in the absence of NNMT.
+Remarkably, NAD concentration and consumption flux are both considerably affected by cell division rates in a system without NNMT, if the enzyme expression is kept constant at different cell division rates.
+Although, one assume organisms to regulate enzyme expression under physiological conditions to achieve similar levels of metabolite concentrations instead.
+Our simulations furthermore predict  a trade-off  between maintainable NAD concentration and consumption flux, in the absence of NNMT.
 In the presence of NNMT, however, NAD consumption rates and concentrations are almost independent of cell division rates.

### 2018-11-28 latex/figurelegends.tex
further revision of results
Revised introduction, Suppl. Table and supplements MD simulations

@@
   \centering
-  \includegraphics[width=0.72\textwidth]{fig_4_NamPT_affinity_Nam}
+  \includegraphics[width=0.72\textwidth]{fig_4_NamPRT_affinity_Nam}
   \caption{\textbf{Role of NamPT substrate affinity.} We simulated the effect of different Michaelis-Menten constants ($K_{M}$) of NamPT for Nam on the steady state NAD consumption flux and NAD concentration at different cell division rates.
@@
 \textit{In vitro} measurements using recombinant protein show that the mutant NamPT has a lower activity than the wildtype enzyme (D) and is not activated by ATP (E).
-Bars in panels D and E that have different letters indicate significant difference of measured values as estimated using a T~test assuming independent samples and significance at $p < 0.05$.}
+Bars in panels D and E that have different letters indicate significant difference of measured values as estimated using a T~test assuming independent samples and significance at $p < 0.05$.
+(F) Root mean square deviation (RMSD) with respect to initial structure for simulation of NamPT with loop insertion (red) and without loop insertion (blue), respectively.
+The RMSD values for the entire simulation (in total 1000 ns) show mostly stable structures with small fluctuations related to rare events.}
   \label{fig:unresolved_loop}

### 2018-11-28 latex/introduction.tex
Revised introduction, Suppl. Table and supplements MD simulations

@@
 
-NAD metabolism represents one of the most critical links connecting cellular signal transduction and energy metabolism.
+NAD metabolism has received increasing interest as changes in NAD metabolism are associated with a large number of diseases including but not limited to neurodegeneration, obesity, heart diseases, renal dysfunction and different types of cancer.
+It has been established that the gradual decrease in NAD during ageing is one of the major driving force of age related pathologies \cite{Chini17}.
+In addition,  NAD metabolism has emerged as key regulator for axonal degradation \cite{Araki04}.
+It is therefore not surprising that NAD metabolism and has emerged as promising pharmacological target for disease treatment  \cite{EspindolaNetto2017,Yoshino18, Sinclair18}.
+However, to develop efficient new therapeutic strategies a better understanding of the pathway dynamics and the pathway components determining it is required.
+
+NAD represents one of the most critical links connecting cellular signal transduction and energy metabolism.
 Even though it is best known as cofactor for various redox-reactions, NAD is involved in a number of signalling processes that consume NAD$^{+}$ by cleaving the molecule to nicotinamide (Nam) and ADP-ribose \cite{Verdin2015}.

### 2018-11-28 latex/header.tex
Replaced remaining NamPRT with NamPT in text

@@
 Two different Nam salvage pathways exist.
-We conducted extensive phylogenetic analyses of these pathways and show that in lower organisms the initial deamidation of Nam is prevalent, whilst the direct conversion of Nam to the mononucleotide by Nam phosphoribosyltransferase (NamPRT) dominates in animals and eventually remains as the single Nam recycling route in vertebrates.
+We conducted extensive phylogenetic analyses of these pathways and show that in lower organisms the initial deamidation of Nam is prevalent, whilst the direct conversion of Nam to the mononucleotide by Nam phosphoribosyltransferase (NamPT) dominates in animals and eventually remains as the single Nam recycling route in vertebrates.
 Strikingly, loss of the deamidation pathway in early vertebrates is preceded by the emergence of a new enzyme that marks Nam for excretion by methylation – nicotinamide N-methyltransferase (NNMT).
@@
 Our simulations indicate that NNMT is required to maintain high NAD consumption fluxes, thereby enabling diversification of the NAD-dependent signalling pathways.
-This kinetic regulation requires and explains the unusually high substrate affinity of the key enzyme for Nam salvage, NamPRT.
-Moreover, we suggest that NNMT exerted evolutionary pressure on NamPRT, enforcing the development of its high substrate affinity.
+This kinetic regulation requires and explains the unusually high substrate affinity of the key enzyme for Nam salvage, NamPT.
+Moreover, we suggest that NNMT exerted evolutionary pressure on NamPT, enforcing the development of its high substrate affinity.
 Using multiple sequence alignments, we identified a sequence insertion first occurring in vertebrates that parallels an experimentally verified increase in the substrate affinity of the enzyme.
-Additional simulations show that the deamidation pathway became obsolete owing to the high substrate affinity of NamPRT.
+Additional simulations show that the deamidation pathway became obsolete owing to the high substrate affinity of NamPT.
 Collectively, our results illustrate a close evolutionary relationship between NAD biosynthesis and the diversification of NAD-dependent signalling pathways, potentially driven by the concomitant occurrence of a regulator of Nam salvage, NNMT.

### 2018-11-28 latex/introduction.tex
Replaced remaining NamPRT with NamPT in text

@@
 Vertebrates use a direct two-step pathway starting with the conversion of Nam into the mononucleotide NMN by the Nam phosphoribosyltransferase (NamPT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate.
-The nearly complete recycling of Nam by NamPRT is achieved by an extraordinary high substrate affinity to Nam, the $K_{M}$ being in the low nanomolar range \cite{Burgos2008}.
+The nearly complete recycling of Nam by NamPT is achieved by an extraordinary high substrate affinity to Nam, the $K_{M}$ being in the low nanomolar range \cite{Burgos2008}.
 This appears to be mediated by an ATP-dependent phosphorylation of a histidine residue in the catalytic core \cite{Burgos2009}.
@@
 The presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012FEBS} is among the most enigmatic and counterintuitive features of NAD metabolism.
-While NamPRT is seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, NNMT seems to have no function other than to remove Nam from NAD metabolism.
+While NamPT is seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, NNMT seems to have no function other than to remove Nam from NAD metabolism.
 It has been suggested that the process potentially acts as a metabolic methylation sink \cite{Pissios2017}.
@@
 As earlier analyses have been limited by the few eukaryotic genomes available at the time, we here performed a comprehensive phylogenetic analysis of the NAD pathways using 793 eukaryotic and 7892 prokaryotic genomes.
-Our results suggest that there has been a selection for the co-existence of NamPRT and NNMT in deuterostomes, while the deamidation pathway, which is dominant in bacteria, became superfluous.
+Our results suggest that there has been a selection for the co-existence of NamPT and NNMT in deuterostomes, while the deamidation pathway, which is dominant in bacteria, became superfluous.
 This transition was accompanied by a marked increase in the number of NAD-consuming signalling enzymes.

### 2018-11-28 latex/methods.tex
Replaced remaining NamPRT with NamPT in text

@@
 \subsection{Phylogenetic Analysis}
-Functionally verified sequences of NNMT, NADA, NamPRT, and NAD-consuming enzymes were used (suppl.
+Functionally verified sequences of NNMT, NADA, NamPT, and NAD-consuming enzymes were used (suppl.
 table~S1) as sequence templates for a Blastp analysis against the NCBI non-redundant protein sequence database.
@@
 
-\subsection{Generation of expression vectors encoding wild-type and mutant human NamPRT}
+\subsection{Generation of expression vectors encoding wild-type and mutant human NamPT}
 
-For eukaryotic expression with a C-terminal FLAG-epitope, the open reading frame (ORF) encoding human NamPRT was inserted into pFLAG-CMV-5a (Merck - Sigma Aldrich) via EcoRI/BamHI sites.
-Using a PCR approach, this vector provided the basis for the generation of a plasmid encoding a NamPRT deletion mutant lacking amino acid residues 42-51 ($\Delta$42-51 NamPRT).
+For eukaryotic expression with a C-terminal FLAG-epitope, the open reading frame (ORF) encoding human NamPT was inserted into pFLAG-CMV-5a (Merck - Sigma Aldrich) via EcoRI/BamHI sites.
+Using a PCR approach, this vector provided the basis for the generation of a plasmid encoding a NamPT deletion mutant lacking amino acid residues 42-51 ($\Delta$42-51 NamPT).
 For prokaryotic expression with an N-terminal 6xHis-tag, the wild-type and mutant ORFs were inserted into pQE-30 (Qiagen) via BamHI and PstI-sites.
@@
 
-\subsection{Purification of NamPRT}
+\subsection{Purification of NamPT}
 

### 2018-11-28 latex/supplementaries.tex
Replaced remaining NamPRT with NamPT in text

--- 2018-07-07 latex/supplementaries.tex
@@
 
-\textbf{The structurally unresolved loop structure of NamPRT.} Sequence alignment of NamPRT of different species cropped to the region around the unresolved loop structure.
-Coloured rectangles indicate the enzymes present in the species besides NamPRT; blue: NNMT; black: NADA and NNMT; yellow: NADA; green: NamPRT only.
+\textbf{The structurally unresolved loop structure of NamPT.} Sequence alignment of NamPT of different species cropped to the region around the unresolved loop structure.
+Coloured rectangles indicate the enzymes present in the species besides NamPT; blue: NNMT; black: NADA and NNMT; yellow: NADA; green: NamPT only.
 Major clades are indicated for better orientation.
@@
 
-\textbf{The phylogenetic distribution of NamPRT and NNMT in birds and reptiles is scattered.}
+\textbf{The phylogenetic distribution of NamPT and NNMT in birds and reptiles is scattered.} The phylogenetic distribution of birds and reptiles was adopted from \cite{Prum2015}.
+Families are marked with a green circle if they possess NamPT without NNMT or a blue circle if they possess both NamPT and NNMT.
 
@@
 
-\textbf{Purification of wt NamPRT and $\Delta$42-51 NamPRT.} A) Elution profile of wildtype NamPRT and mutant $\Delta$42-51 on size-exclusion chromatography using a Superdex 200 16/60 column.
-B) Coomassie stained denaturating SDS-PAGE analysis of $\Delta$42-51 NamPRT (lane 1) and wt NamPRT (lane 2).
+\textbf{Purification of wt NamPT and $\Delta$42-51 NamPT.} A) Elution profile of wildtype NamPT and mutant $\Delta$42-51 on size-exclusion chromatography using a Superdex 200 16/60 column.
+B) Coomassie stained denaturating SDS-PAGE analysis of $\Delta$42-51 NamPT (lane 1) and wt NamPT (lane 2).
 3\,$\mu$g of pooled enzyme eluted from SEC loaded onto the gel.
@@
 The partition coefficient (Kav) was determined for each standard (light grey squares) and plotted versus log$_{10}$ molecular weight.
-The Kav was determined for wt NamPRT and $\Delta$42-51 NamPRT and the apparent molecular weight calculated to be 135\,kDa and 110\,kDa, respectively.
+The Kav was determined for wt NamPT and $\Delta$42-51 NamPT and the apparent molecular weight calculated to be 135\,kDa and 110\,kDa, respectively.
+
+\subsection*{Supplementary figure S4}
+
+\textbf{NMR measurement of NamPT activity.} NamPT activity was measured as described in Experimental Procedures and product (NMN) formation was detected using 1D 1H NMR spectroscopy.
+Inset on the right: molecular structure of NMN with the atom detected by NMR indicated by an arrow.
+The range used for NMN detection in typical 1D 1H NMR spectra of the enzymatic reactions is shown.
+NMN quantification was done with the singlet detected at 9.52 ppm.
+From the top to the bottom, peak detection of NMN standard (200\,µM), wildtype NamPT (1\,mM Nam and 1\,mM PRPP), mutant $\Delta$42-51 NamPT (1\,mM Nam and 1\,mM PRPP), wildtype NamPT with FK866, and mutant $\Delta$42-51 NamPT with FK866.
+Incubation with inhibitor FK866 was done for 30\,min at 30\,°C.
 

### 2018-11-28 latex/methods.tex
Minor changes fixing some special chars

@@
 We used the AMBER99SB-ILDN force field \cite{Lindorff-Larsen2010} with the TIP3P water model \cite{Jorgensen1983} in GROMACS 5.1.2 \cite{Abraham2015}.
-The structures were simulated each in a box of water with distance between the solute and the box set to 0.2\,nm at a temperature of 300\,K for a total time of 1\,\textmu s.
+The structures were simulated each in a box of water with distance between the solute and the box set to 0.2\,nm at a temperature of 300\,K for a total time of 1\,$\mu$s.
 A time step of 2\,fs and the stochastic dynamics integrator were used.

### 2018-11-28 latex/results.tex
Replaced remaining NamPRT with NamPT in text
Minor changes fixing some special chars

@@
 To understand the functional roles and potential interplay between the three known enzymes that use Nam as substrate (NamPT, NADA and NNMT), we first conducted a comprehensive analysis of the phylogenetic distribution of these three enzymes.
-As shown in Figure~\ref{fig:phylo_distribution}A, bacteria, fungi, and plants predominantly possess NADA and only a few of them harbour NamPRT.
+As shown in Figure~\ref{fig:phylo_distribution}A, bacteria, fungi, and plants predominantly possess NADA and only a few of them harbour NamPT.
 In contrast, Metazoa predominantly lost NADA and encode for NamPT together with NNMT.
@@
 Figures~\ref{fig:NamPT_affinity_Nam}E and~F visualise a direct comparison of simulations assuming different affinities of NamPT for Nam, in the presence or absence of NNMT.
-Interestingly, at an affinity of $K_{M}$ =~1\,$\mu$M, which is in the range of the $K_{M}$ of NADA for Nam, NAD consumption flux is only higher with NNMT present when cell division rates are low (fig.~\ref{fig:NamPT_affinity_Nam}E).
+Interestingly, at an affinity of $K_{M}$~= 1\,$\mu$M, which is in the range of the $K_{M}$ of NADA for Nam, NAD consumption flux is only higher with NNMT present when cell division rates are low (fig.~\ref{fig:NamPT_affinity_Nam}E).
 If the affinity of NamPT is high enough, consumption rates are always higher with NNMT than without it.
@@
 fig.~S4).
-The mutant enzyme showed no detectable activity with 100 $\mu$M  Nam after 10 min of incubation at 30\textdegree{}C.
+The mutant enzyme showed no detectable activity with 100\,$\mu$M  Nam after 10 min of incubation at 30\,\textdegree{}C.
 With 1mM Nam enzymatic activity could be detected with the mutant enzyme.
@@
 This might be based on the fact that  all available protein structures of NamPT differ very little even at the catalytic side ( between 0.33Å and 0.95Å see supplementary table S3).
-Some residues close to the catalytic, nevertheless, showed slightly elevated mobilities in the wildtype structure, but, these elevated mobilities were dominated by rare events during the simulation time 1$~\mu s$.
+Some residues close to the catalytic, nevertheless, showed slightly elevated mobilities in the wildtype structure, but, these elevated mobilities were dominated by rare events during the simulation time 1\,$\mu$s.
 Thus, they do not appear as a robust change of structural dynamics upon loop insertion.

### 2018-11-30 latex/figurelegends.tex
Added own file for figure legends and updated Description of suppl. files
revised supplementary information, status manuscript send to Bergen group 30th of November

--- 2018-11-18 latex/figurelegends.tex
@@
   \centering
-  \includegraphics[width=0.7\textwidth]{fig_6_unresolved_loop}
+  \includegraphics[width=0.83\textwidth]{fig_6_unresolved_loop}
+\end{figure}
+
+\newpage
+
+\begin{figure}[ht]
   \caption{\textbf{The function of the structurally unresolved loop of NamPT.} Most deuterostomes that possess NamPT and NNMT show a sequence insertion in the N-terminal region of NamPT that has been revealed by multiple sequence alignment of NamPT from different species (A).
@@
 For a more comprehensive alignment, please see supplementary figure~S1.
-The structure visualisation of human NamPT (B) is based on a structure prediction by SWISS-MODEL \cite{Arnold2006,Biasini2014} using the model 2H3D of the human NamPT as template \cite{Wang2006}.
+The structure visualisation of human NamPT (B) is based on a structure prediction by SWISS-MODEL \cite{Arnold2006,Biasini2014} using the model 3DGR of the human NamPT as template \cite{Burgos2009}.
 The inserted region is not resolved in any of currently available crystal structures of NamPT and thus appears to be a flexible loop structure at the surface of the NamPT dimer, coloured in red.
@@
 Both show a heterogeneous nuclear-cytosolic localisation in HeLa S3 cells.
-\textit{In vitro} measurements using recombinant protein show that the mutant NamPT has a lower activity than the wildtype enzyme (D) and is not activated by ATP (E).
-Bars in panels D and E that have different letters indicate significant difference of measured values as estimated using a T~test assuming independent samples and significance at $p < 0.05$.}
+In vitro measurements using recombinant protein show that the mutant NamPT has no detectable activity with 100 $\mu$M Nam and a significantly reduced activity at 1mM Nam (D) The p-value was determined using a non-parametric one-tailed Mann-Whitney test.
+(F) Root mean square fluctuations (RMSF) for every residue of chain A of NamPT with loop insertion (red) and without loop insertion (blue) are shown, respectively (top).
+The difference RMSF for every residue is shown in the bottom (green).
+For better comparison the residue IDs for the structure without loop are aligned to accord with the structure with loop and the average RMSF of residues 43 and 54 displayed in the blue curve between these residues.
+For the RMSF calculation, the first 100 ns of the simulation are omitted to allow equilibration.}
   \label{fig:unresolved_loop}

### 2018-11-30 supplements/KineticsTable/kinetics_table.tex

@@
 \setcounter{table}{1}
-\begin{longtable}{p{2cm}p{2cm}p{2cm}cc}
+\begin{longtable}{p{1.5cm}p{1.5cm}p{2.5cm}cc}
 \multicolumn{5}{l}{\textbf{Overview of kinetic constants used for the
@@
 NMNAT &
-\href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC2/7/7/1.html}{2.7.7.1}/\href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC2/7/7/18.html}{2.7.7.18}
+\href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC2/7/7/1.html}{2.7.7.1}  
 &$K_{M_{NaMN}}$:67.7$\mu$M  & \cite{Sorci2007}\footnote{Values for NMNAT1 used}
 & Substrate Competition\\
-    & & $k_{{cat}_{NaMN}}$:42.9$s^{-1}$  & & \\
+&  \href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC2/7/7/18.html}{2.7.7.18}   & $k_{{cat}_{NaMN}}$:42.9$s^{-1}$  & & \\
     & & $K_{M_{NMN}}$:22.3$\mu$M & & \\
@@
 NamPT & \href{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC6/3/5/1.html}{6.3.5.1}
-&$K_M$:5nM  & \cite{Burgos2008} & HMM\\
+&$K_M$:5nM  & \cite{Burgos2008} & Competitive inhibition\\
     & & $k_{cat}$:0.0077$s^{-1}$  & &
+    \\
+     & & $K_{i_{NAD}}$:  2.1$\mu$M& &
     \\  \hline
@@
   \bottomrule
-   \caption{The total enzyme concentration was set to 10 for all enzymes except NamPT and NMNAT, for which the concentration was set to 100.
-Concentration of potential co-substrate was assumed to be constant and not-limiting for the reaction.
-Thus they were implicitly represented by the
-maximal velocities given in the table.}
   \label{tab:kinetic}
@@
 \end{longtable}
+
+\subsection*{Amount of enzymes and import rates}
+The total enzyme concentration was set to 10 for all enzymes except NamPT and NMNAT, for which the concentration was set to 100.
+Concentration of potential co-substrate was assumed to be constant and not-limiting for the reaction.
+Thus being implicitly represented by turnover rates measured at saturating conditions for the cosubstrate.
+Nam import rates for iport into the system was set to 0.1 $\mu$M/ls for all simulations.
+For the two compartment simulation compartmentsize was equal to ...
+for both compartments.
+The Nam import rates were set to 100/s for both compartments.
+The amount of NADA present was set to 100.
+Thus equal to the amount of NamPT used.
 
@@
 v=\cfrac{E_T\cdot k_{cat}\cdot S}{K_M + S + \cfrac{K_M\cdot P}{K_{iP}} }
+\end{equation}
+
+\subsubsection*{Competitive Inhibition}
+\begin{equation}
+v=\cfrac{E_T\cdot k_{cat}\cdot S}{K_M + S + \cfrac{K_M\cdot I}{K_{iI}} }
 \end{equation}

### 2018-12-09 supplements/KineticsTable/kinetics_table.tex
included all comments from Marc

@@
 \subsection*{Amount of enzymes and import rates}
-The total enzyme concentration was set to 10 for all enzymes except NamPT and NMNAT, for which the concentration was set to 100.
-Concentration of potential co-substrate was assumed to be constant and not-limiting for the reaction.
-Thus being implicitly represented by turnover rates measured at saturating conditions for the cosubstrate.
-Nam import rates for iport into the system was set to 0.1 $\mu$M/ls for all simulations.
+The total enzyme concentration was set to 10 times the scaling factor, for all enzymes except NamPT and NMNAT, for which the concentration was set to 100 times the scaling factor.
+As enzyme concentrations here have an arbitrary unit a scaling factor of 0.1$\mu$M  was applied to all enzymatic reactions to achieve consumption rates that are in the range of reported values  \cite{Liu2018}.
+Concentration of potential co-substrates were assumed to be constant and not-limiting for the reaction.
+Thus being implicitly represented by maximal velocities consiting of total enzyme concentration times tunrover rates.
+Nam import rates for import into the system was set to 0.1 $\mu$M/s for all simulations, being in the range meassured for Nam uptake in mammalian cells \cite{Namuptake}.
 For the two compartment simulation compartmentsize was equal to ...

### 2018-12-13 supplements/KineticsTable/kinetics_table.tex
updated simulations for competitions
Discussion under  revision, preparing models for submission. Both tasks not yet finished

@@
 Concentration of potential co-substrates were assumed to be constant and not-limiting for the reaction.
-Thus being implicitly represented by maximal velocities consiting of total enzyme concentration times tunrover rates.
+Thus being implicitly represented by maximal velocities consisting of total enzyme concentration times tunrover rates.
 Nam import rates for import into the system was set to 0.1 $\mu$M/s for all simulations, being in the range meassured for Nam uptake in mammalian cells \cite{Namuptake}.
-For the two compartment simulation compartmentsize was equal to ...
-for both compartments.
-The Nam import rates were set to 100/s for both compartments.
+In addition to the reactions listed above an additional NAD consumption was simulated using HMM-kinetics with a susbtrate affinity of 0.3 mM and a turnover rate of 1.
+Furthermore, reversible NAD binding to proteins was simulated using reversible mass actions kinetics with an equilibrium constant of 0.1, which is in a range of values reported in the literature, dissociation and association constants where set to 10 and 100$s^{-1}$ respectively.
+s
+For the two compartment simulation, compartmentsize was equal for both compartments and set to 1$\mu$l.
+The actual compartmentsize does not change the outcome of the simulations as long as both compartments have equal volumes.
+The Nam import rates were set to 100$s^{-1}$ for both compartments.
 The amount of NADA present was set to 100.
